### 2023-07-03 notes/main.tex
added main.tex

\documentclass{article}

\let\oldb\b
\let\oldl\l
\usepackage[dvipsnames]{xcolor}
\usepackage{notes,hyperref}
\let\l\lambda
\let\wtilde\widetilde
\let\bar\overline
\def\O{\mathcal O}
\def\sing{{\mathrm{sing}}}
\let\b\beta
\def\F{\mathcal F}

\let\inc\hookrightarrow
\let\xinc\xhookrightarrow

\numberwithin{equation}{section}
\let\d\partial

\theoremstyle{theorem}
\renewtheorem{thm}{Theorem}
\renewtheorem{lem}[thm]{Lemma}
\renewtheorem{cor}[thm]{Corollary}
\renewtheorem{prop}[thm]{Proposition}
\renewtheorem{conj}[thm]{Conjecture}
\numberwithin{thm}{section}
\theoremstyle{definition}
\renewtheorem{dfn}[thm]{Definition}
\renewtheorem{exa}[thm]{Example}
\renewtheorem{note}[thm]{Note}
\renewtheorem{nota}[thm]{Notation}
\renewtheorem{quest}[thm]{Question}
\renewtheorem{rem}[thm]{Remark}

\def\L{\mathcal L}
\def\M{\mathcal M}

\newcommand{\mahrud}[1]{{\color{ForestGreen} \sf $\blacklozenge$ Mahrud: [#1]}}
\newcommand{\devlin}[1]{{\color{red} \sf $\clubsuit$ Devlin: [#1]}}

\title{(Frobenius) splittings in Macaulay2}
\author{Devlin Mallory and Mahrud Sayrafi}
\begin{document}
\maketitle

Throughout, $R$ will be an $\N$-graded ring with $R_0=k$ a field\mahrud{I think most of this applies to any grading as long as the latter condition holds}; we write $m$ for the homogeneous maximal ideal. Note that if $M$ is a finitely generated $R$-module, and $A\in \End_R(M)$, then $A$ acts on the $k$-vector space $M/mM$.

\begin{lem}
  Let $M$ be an $R$-module, and let $A\in\End_R(M)$. If the induced action of $A$ on $M/mM$ is idempotent\mahrud{Maybe you're proving something stronger, but it's possible that action of $A$ on $M$ is either idempotent of nonzero degree or not idempotent, but the induced action on $M/mM$ is}, then $M$ admits a direct sum decomposition $N\oplus M/N$, where $N\cong \im A$.
\end{lem}

\begin{proof}
  By assumption, we can write $A^2 = A + B$, where $B\in\End_R(M)$ with $B(M)\subset mM$.
Note that if $n \in m^kM$, then $A^2(n) - A(n) = B(n)$  lies in $m^{k+1}M$.

Let $N = \im A$. Consider the composition $N\inc M \xra{A} \im(A) = N$. We claim that this composition is surjective. We may complete at the maximal ideal to check this, and thus assume $R$ and $M$ are complete\mahrud{out of curiosity, what can go wrong?}. Let $n_0\in N$. By assumption, $n_0=A(m_1)$ for some $m_1\in M$. Applying $A$ again, we get 
\[ A(n_0) = A^2(m_1) = A(m_1) + n_1 = n_0 + n_1, \]
or equivalently \mahrud{should this say $n_1 = A(n_0) - n_0$ instead?}
\[ n_0 = A(n_0) - n_1 \]
for some $n_1\in mM$. In fact, since $n_0$ and $A(n_0)$ are both in $N$, we have $n_1\in N$ also, so $n_1\in mM\cap N$.

Thus, we can write $n_1 = A(m_2)$ for $m_2\in M$.
Now, apply $A$ to both sides: by the assumption that $A$ is idempotent modulo $m$\mahrud{how is this used here?}, we have
\[ A(n_1)=A^2(m_2) = A(m_2) + n_2 = n_1+n_2, \]
Thus, $n_2=A(n_1)-n_1$, so $n_2\in m^2M$; clearly also $n_2\in N$ as well. 
Combining the previous equations, we can write
\[ n_0 = A(n_0) - n_1 = A(n_0) - A(n_1) + n_2 = A(n_0 - n_1) + n_2, \]
with $n_1\in mM\cap N$ and $n_2\in m^2M\cap N$.

Continuing in this fashion, for any $k$ we can write 
\[ n_0=A(n_0-n_1+\dots \pm n_k) \mp n_{k+1}, \]
with $n_i \in m^i M\cap N$.

By the Artin--Rees lemma, for there's some positive integer $k$ such that for $n\gg0$ we can write
\[ m^n M\cap N = m^{n-k} ( m^kM\cap N)\subset m^{n-k} N. \]
That is, the terms of $n_0-n_1+n_2-\dots$ are going to 0 in the $m$-adic topology on $N$. Thus, we can write
\[ n_0=A(n_0-n_1+n_2-\dots), \]
with $n_0-n_1+n_2-\dots\in N$. Thus, we have that $A$ is surjective as a map $N\to N$.

Since a surjective endomorphism of finitely generated modules is invertible, we have that this composition is an isomorphism on $N$; say $\a$.
Finally, we then have that
\[ 0 \inc N \inc M \]
is split by the map of $R$-modules $M\xra A N\xra{\a\inv} N$. 
\end{proof}

\begin{lem}
Let $k$ be a finite field of characteristic $p$, and let $A$ be an endomorphism of a $k$-vector space such that all eigenvalues of $A$ are contained in $k$. If $\l$ is an eigenvalue of $A$, then some power of $A-\l I$ acts idempotently.
Furthermore, if $\l$ is not the only eigenvalue of $A$, then a nonzero power of $A-\l I$ acts idempotently.
\end{lem}

\begin{proof}
Since all eigenvalues of $A$ are contained in $k$, we can without loss of generality put $A$ in Jordan canonical form, with each Jordan block having the form
$$
r_i\Biggl\{
\underbrace{\begin{pmatrix}
\l_i & 1 & 0 &\dots & 0
\\
0& \l_i & 1  &\dots & 0
\\
\vdots & &&&\vdots
\\
0& 0 & 0  &\dots & \l_i
\end{pmatrix}}_{r_i}
$$
with each $\l_i$ an eigenvalue of $A$.
In this basis, $A-\l I$ will be block-diagonal with blocks of form
$$
\begin{pmatrix}
\l_i-\l & 1 & 0 &\dots & 0
\\
0& \l_i-\l & 1  &\dots & 0
\\
\vdots & &&&\vdots
\\
0& 0 & 0  &\dots & \l_i-\l
\end{pmatrix}
$$
Set $\mu_i=\l_i-\l$. Then $(A-\l I)^l$ is block-diagonal with blocks of the form
$$
\begin{pmatrix}
\mu_i^l & \binom{l}1\mu_i^{l-1} & \binom{l}2\mu_i^{l-2} &\dots & \binom{l}{r_i}\mu_i^{l-r_i}
\\
0& \mu_i^l &
\binom{l}1\mu_i^{l-1}  &\dots & \binom{l}{r_i-1}\mu_i^{l-r_i+1}
\\
\vdots & &&&\vdots
\\
0& 0 & 0  &\dots & \mu_i^l
\end{pmatrix}
$$
If $l$ is $> r_i$ and divisible by $p$, then all nondiagonal terms will vanish, so all blocks will have the form
$$
\begin{pmatrix}
\mu_i^l & 0 & 0 &\dots & 0
\\
0& \mu_i^l & 0  &\dots & 0
\\
\vdots & &&&\vdots
\\
0& 0 & 0  &\dots & \mu_i^l 
\end{pmatrix}.
$$
Finally, choosing $l$ to be divisible also by $p^e-1$ for some $e>0$, we have that $$\mu_i^{l-1} = (\mu_i^{p^{e}-1})^{l/(p^{e}-1)} =1,$$ since $\mu_i$ is contained in a finite field of characteristic $p$.
Thus, $(A-\l I)^l$ is simply a diagonal matrix with 1 or 0 on the diagonal, hence idempotent. Note moreover that if some $\l_i\neq \l$, then $(A-\l I)^l$ is not the zero matrix.

\end{proof}

\end{document}

### 2023-10-27 notes/main.tex
committed Devlin's proof of the idempotent lemma

@@
 
+\begin{lem}
+Let $A\in \End_R M$ be idempotent. Then $M=\im  A\oplus \ker A$.
+\end{lem}
+
+\begin{proof}
+Consider the short exact sequence
+$$
+0\to \ker A \to M \to \im A\to 0.
+$$
+The map $M\to M$ given by $-A+\id_M$ sends $M\to \ker A$, since $-A(A-\id_M)(m) = -(A^2-A)(m)=0$.
+In fact, this map $M\to \ker A$ is a splitting of $\ker A\to M$, since if $m\in \ker A$ we have $-(A-\id_M)(m) = -A(m)+\id_M(m)=m$.
+\end{proof}
+
+\begin{lem}
+Let $A\in \End_R M$ satisfy $A^d=A$. Then $M=\im  A\oplus \ker A$.
+\end{lem}
+
+\begin{proof}
+Consider the short exact sequence
+$$
+0\to \ker A \to M \to \im A\to 0.
+$$
+The map $M\to M$ given by $-A^{d-1}+\id_M$ sends $M\to \ker A$, since $-A(A^{d-1}-\id_M)(m) = -(A^d-A)(m)=0$.
+In fact, this map $M\to \ker A$ is a splitting of $\ker A\to M$, since if $m\in \ker A$ then also $m\in \ker A^{d-1}$ and thus we have $-(A^{d-1}-\id_M)(m) = -A^d(m)+\id_M(m)=m$.
+
+\end{proof}
+
 \end{document}

### 2024-01-29 beamer/beamer.tex
added JMM presentation

\documentclass[beamer]{paper}

\usepackage{quiver}
\usepackage[italic,symbolgreek]{mathastext}

% for highlighting
\usepackage{xcolor, soul}
\makeatletter
\let\HL\hl
\renewcommand\hl{%
  \let\set@color\beamerorig@set@color
  \let\reset@color\beamerorig@reset@color
  \HL}
\makeatother

\usetheme{Warsaw}
\beamertemplatenavigationsymbolsempty
%\setbeamertemplate{footline}[frame number]
%\setbeamercovered{transparent} % gray out hidden stuff
\AtBeginSection{\frame{\sectionpage}}

\newcommand{\lc}{\MakeLowercase}

\newcommand{\PPN}{\PP{\underline{\nn}}}

\newcommand{\reg}{\operatorname{reg}}

\newcommand{\cH}{\mathcal{H}} % Hirzebruch surface

\renewcommand{\aa}{\mathbf a}
\newcommand{\bb}{\mathbf b}
\newcommand{\cc}{\mathbf c}
\newcommand{\dd}{\mathbf d}
\newcommand{\ee}{\mathbf e}
\newcommand{\vv}{\mathbf v}
\newcommand{\ww}{\mathbf w}
\newcommand{\pp}{\mathbf p}
\newcommand{\qq}{\mathbf q}
\newcommand{\nn}{\mathbf n}
\newcommand{\one}{\mathbf 1}
\newcommand{\zero}{\mathbf 0}
\newcommand{\Om}{\Omega}
\newcommand{\om}{\omega}
\newcommand{\cech}{\check{C}}
\newcommand{\Alt}{\bigwedge\nolimits}
\newcommand{\R}{\mathbf R} % right derived

\def\OO{\mathcal{O}}
\def\FF{\mathbb{F}}

\def\a{\alpha}
\def\b{\beta}
\def\g{\gamma}
\def\del{\partial}
\def\cech{\check{C}}
\def\Gs{\Gamma_{\!*}}

\def\ub{\underbrace}
\def\on{\operatorname}
\def\ph{\phantom{-}}

\def\Sym{\operatorname{Sym}}

\def\L{\mathcal{L}}
\def\E{\mathcal{E}}
\def\X{\mathcal{X}}
\def\B{\mathbf{B}}
\def\RR{\mathbf{R}}
\def\cR{\mathcal{R}}
\def\L{\mathbf{L}}
\def\U{\mathbf{U}}
\def\H{\mathbb{F}}
\def\F{\mathcal{F}}
\def\G{\mathcal{G}}
\def\C{\mathcal{C}}
\def\Z{\ZZ}

\def\Db{\operatorname{D}^{\operatorname{b}}} % -> \Db{X}
\def\coh{\operatorname{coh}}
\def\cL{\mathcal{L}}

\def\from{\leftarrow}
\def\into{\hookrightarrow}
\def\operatornameto{\twoheadrightarrow}

\def\id{\operatorname{id}}
\def\th{\operatorname{th}}
\def\colim{\operatorname{colim}}
\def\lenef{\le_{\operatorname{nef}}}
\def\lnef{<_{\operatorname{nef}}}
\def\Dsat{D_{\operatorname{sat}}}

\def\cS{\mathcal{S}}
\def\cT{\mathcal{T}}
\def\cK{\mathcal{K}}
\def\cB{\mathcal{B}}
\def\tS{\widetilde{S}}
\def\cU{\mathcal{U}}
\def\cW{\mathcal{W}}
\def\cC{\mathcal{C}}
\def\tr{\widetilde{\rho}}
\def\ts{\widetilde{\sigma}}
\def\tt{\widetilde{\tau}}

\newcommand{\PG}[1]{{\color{PineGreen}#1} }
\newcommand{\dao}[1]{{\color{DarkOrchid}#1} }

\begin{document}

% Use Computer Modern with Serif Roman Font Family
\rmfamily
\setbeamerfont{frametitle}{family=\rmfamily\selectfont}

\title[Computing summands of the Frobenius over Cox rings]
      {\textsc{Computing Summands of the Frobenius \\ over Multigraded Cox Rings}}

% I acknowledge that the University of Minnesota stands on Miní Sóta Makhóčhe, the homelands of the Dakhóta Oyáte.

% Abstract: It is known that the Frobenius pushforward of the structure sheaf on toric varieties splits as a sum of line bundles, and high enough pushforwards of it generate the derived category. Working over the multigraded Cox ring of a Mori Dream Space, we discuss algorithms for determining the summands of the Frobenius pushforward of the structure sheaf.

\author[Mahrud Sayrafi]{M\lc{ahrud} S\lc{ayrafi} \\
  {\footnotesize U\lc{niversity} of M\lc{innesota}, T\lc{win} C\lc{ities}} \\[0em]
  {\footnotesize Joint work with Devlin Mallory (University of Utah)}}

%\institute[UMN]{U\lc{niversity} of M\lc{innesota}, T\lc{win} C\lc{ities}}

% https://www.jointmathematicsmeetings.org/meetings/national/jmm2024/2300_progfull.html
\date{{\footnotesize AMS Special Session on \\
    Derived Categories, Arithmetic, and Geometry, \\
    Joint Mathematics Meetings} \\ January 4, 2024}

\frame[plain]{\titlepage}

\begin{frame}[t]{Warm-up: Frobenius pushforward on $\PPn$}
  Let $\kk$ be an algebraically closed field of char $p>0$.

  Let $S = \kk[x_0,\dots,x_n]$ be a polynomial ring with $\deg x_i = 1$.

  \vfill
  %\pause
  Consider the Frobenius endomorphism
  \[ F\colon S\to S \quad \text{given by} \quad f \to f^p \]

  %\pause
  - Hartshorne: any $L\in\Pic\PPn$, $F_*L$ splits as a sum of line bundles.
  %\pause
  \begin{block}{Example: $F_*\o_{\PPn}$}
    \vspace*{-0.2in}
    \begin{align*}
      \text{When $p=3$, $n=2$:} \\
      F_*\o_{\PP2} = \o \oplus &\o(-1)^7 \oplus \o(-2). \\
      \text{When $p=2$, $n=5$:} \\
      F_*\o_{\PP5} = \o \oplus &\o(-1)^{15} \oplus \o(-2)^{15} \oplus \o(-3). \\
      F_*^2\o_{\PP5} = \o \oplus &\o(-1)^{120} \oplus \o(-2)^{546} \oplus \o(-3)^{336} \oplus \o(-4)^{21}
    \end{align*}
  \end{block}
\end{frame}

\begin{frame}[t]{Warm-up: Frobenius pushforward on toric varieties}
  Let $X$ be a smooth toric variety and consider the Cox ring
  \[ S = \bigoplus\!{}_{[D]\in\Pic{X}} \; \Gamma(X, \OO(D)) \]

  %\pause
  \begin{theorem}[Thm.~3.2, \cite{[Cox95]}]
    Every coherent sheaf on a simplicial toric variety $X$ corresponds \\
    to a finitely generated $\Pic X$-graded $S$-module.
  \end{theorem}

  \vfill
  %\pause
  Fact: for any toric variety, $S$ is a \textbf{multigraded} polynomial ring.
  %\pause
  Upshot: we can do commutative algebra on a toric variety! \\
  %\pause
  \hfill + easy monomial orderings, Gr\"obner bases, syzygies, etc.
\end{frame}

\begin{frame}[t]{Warm-up: Frobenius pushforward on toric varieties}
  Let $X$ be a smooth toric variety and consider the Cox ring
  \[ S = \bigoplus\!{}_{[D]\in\Pic{X}} \; \Gamma(X, \OO(D)) \]

  %\pause
  - Bogvad '98, Thomsen '00: $F_*L$ totally splits for any $L\in\Pic X$.
  %\pause
  \begin{block}{Example: $X = $ Hirzebruch surface of type 3, $F_*\o_X, p = 2$}
    \vspace*{-0.2in}
    \[ F_*\o_X = \o_X \oplus \o_X(-1,0)^2 \oplus \o_X(0,-1)^2 \oplus \o_X(1,-1)^3 \oplus \o_X(2,-1). \]
  \end{block}

  - Achinger '13: \\ \hfill splitting of $F_*L$ characterizes smooth projective toric varieties.

  \vfill
  %\pause
  \begin{block}{Remark}
    The toric Frobenius is simply multiplication by $p$ on the torus.
  \end{block}
  %\pause
  \begin{alertblock}{Funky Remark}
    The $p$ need not be prime, nor even match the characteristic!
  \end{alertblock}
\end{frame}

\begin{frame}{Frobenius pushforward on Grassmannians}
  We can still consider the Cox ring
  \[ S = \bigoplus\!{}_{[D]\in\Pic{X}} \; \Gamma(X, \OO(D)) \]
  which in this case coincides with the coordinate ring.

  %\pause
  \begin{block}{Example: $X = \mathrm{Gr(2,4)}, F_*\o_X, p = 3$}
    Working over the coordinate ring:
    \[ S = \frac{\kk[p_{0,1},\dots p_{0,2},p_{1,2},p_{0,3},p_{1,3},p_{2,3}]}{p_{1,2}p_{0,3}-p_{0,2}p_{1,3}+p_{0,1}p_{2,3}} \]
    When $p = 3$:
    \[ F_*\o_X = \o \oplus \o(-1)^{44} \oplus \o(-2)^{20} \oplus A^4 \oplus B^4, \]
    where $A$ and $B$ are rank 2 indecomposable bundles. \\
    \hfill (c.f. [Raedschelders--Špenko--Van den Bergh, '17])
  \end{block}
\end{frame}

\begin{frame}{Frobenius pushforward on Mori Dream Spaces}
  This is the natural setting to work over the Cox ring
  \[ S = \bigoplus\!{}_{[D]\in\Pic{X}} \; \Gamma(X, \OO(D)). \]
  Hu--Keel: $X$ is an $MDS$ $\iff$ the Cox ring is finitely generated.

  \vfill
  %\pause
  \begin{exampleblock}{Examples}
  \; - Projective toric varieties \\
  \; - Grassmannians and flag varieties \\
  \; - Determinantal varieties \\
  \; - Many complete intersection rings \\
  \; - Smooth Fano varieties over $\CC$ \\
  \; - $\overline{M}_{0,n}$ for $n\leq 6$ (7,8,9 are unknown)
  \end{exampleblock}
\end{frame}

\begin{frame}{Frobenius pushforward on Mori Dream Spaces}
  \begin{block}{Example: $X = \mathrm{Bl}_4\PP2, F_*\o_X, p = 2$}
    Working over the $\ZZ^5$-graded Cox ring \vspace*{-0.15in}
    \[ S = \kk[x_1,\dots,x_{10}]/\text{(five quadric Pl\"ucker relations)} \]
    %\pause
    With degrees \vspace*{-0.15in}
    \[
    \left(\!\begin{array}{cccccccccc}
      0&0&0&0&1&1&1&1&1&1 \\
      1&0&0&0&-1&-1&-1&0&0&0 \\
      0&1&0&0&-1&0&0&-1&-1&0 \\
      0&0&1&0&0&-1&0&-1&0&-1 \\
      0&0&0&1&0&0&-1&0&-1&-1
    \end{array}\!\right)
    \]
  \end{block}
\end{frame}

\begin{frame}{Frobenius pushforward on Mori Dream Spaces}
  \begin{block}{Example: $X = \mathrm{Bl}_4\PP2, F_*\o_X, p = 2$}
    \vspace*{-0.15in}
    h\begin{align*}
      F_*^2\o_X = {\o_{X}^{1}}
      &\oplus {\o_{X}^{2}\ \left(-2,\,1,\,1,\,1,\,1\right)} \oplus {\o_{X}^{2}\ \left(-1,\,0,\,0,\,0,\,1\right)} \\ 
      &\oplus {\o_{X}^{2}\ \left(-1,\,0,\,0,\,1,\,0\right)} \oplus {\o_{X}^{2}\ \left(-1,\,0,\,1,\,0,\,0\right)} \\
      &\oplus {\o_{X}^{2}\ \left(-1,\,1,\,0,\,0,\,0\right)} \oplus B \oplus G,
    \end{align*}
    where $B, G$ are rank 3 and rank 2 indecomposables. \\
    \hfill (c.f. [Hara, '15])
  \end{block}
\end{frame}

\begin{frame}{The splitting algorithm}
  \begin{alertblock}{The ``Meataxe'' algorithm}
    Input: finitely generated graded module $M$ over an algebra.
    %\pause
    \begin{enumerate}
    \item compute $\Hom(M, M)$; %\pause
    \item find element $h$ corresponding to an idempotent; %\pause
    \item find the image and cokernel of $h$; %\pause
    \item repeat.
    \end{enumerate}
  \end{alertblock}
\end{frame}

\begin{frame}[t]{Frobenius pushforward on determinantal varieties}
  \textbf{Fact:} a determinantal variety $X$ is a Mori dream space. \\
  \hfill (i.e. the Cox ring is a finitely generated algebra.)

  \vfill
  Singh: does $F^e_*\OO_X$ have finitely many classes of summands? \\
  This is known as having \emph{globally finite $F$-representation type}.

  \begin{block}{Problem}
    Classify MDS with globally finite $F$-representation type.
  \end{block}

  \begin{alertblock}{Upshot}
    Determining line bundle summands is easy over the Cox ring.
  \end{alertblock}

  \vfill
\end{frame}

\begin{frame}[t]{References}
  \scriptsize{
    - Hartshorne \hfill\textit{Ample subvarieties of algebraic varieties} \\
    - Bogvad, \hfill\textit{Splitting of the direct image of sheaves under the Frobenius} \\
    - Thomsen, \hfill\textit{Frobenius direct images of line bundles on toric varieties} \\
    - Achinger, \hfill\textit{A characterization of toric varieties in characteristic $p$} \\
    - Hara, \hfill\textit{Looking out for Frobenius summands on a blown-up surface of $\PP2$} \\
    - Raedschelders, Špenko, Van den Bergh, \hfill\textit{The Frobenius morphism in invariant theory}
  }

  \renewcommand{\section}{\subsection}
  \printbibliography
\end{frame}

\end{document}

% Local Variables:
% tab-width: 2
% eval: (visual-line-mode)
% End:

### 2024-02-09 paper/main.tex
added MEGA 2024 draft (main.tex)
added changes as of Friday morning

--- 2023-10-27 notes/main.tex
@@
 
-\title{(Frobenius) splittings in Macaulay2}
+\title{Computing direct sum decompositions}
 \author{Devlin Mallory and Mahrud Sayrafi}
@@
 
-Throughout, $R$ will be an $\N$-graded ring with $R_0=k$ a field\mahrud{I think most of this applies to any grading as long as the latter condition holds}; we write $m$ for the homogeneous maximal ideal.
+% Regular papers, aimed to contributed talks:
+% for these, authors can choose to submit either an extended abstract of at least six pages (not counting bibliography) that should contain full results, an outline of the proofs and all of the main ideas; or else a full paper with complete proofs.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Introduction}
+
+%% The introduction of the paper must explicitly address the following questions in succinct and informal manner:
+%% – What is the problem?
+%% – Why is the problem important?
+%% – What has been done so far on the problem?
+%% – What is the main contribution of the paper on the problem?
+%% – Why is the contribution original? (see below for clarification)
+%% – Why is the contribution non-trivial?
+%% – How does the journal paper differ from the conference paper (if relevant)?
+
+The problem of finding isomorphism classes of indecomposable modules with certain properties is ubiquitous in commutative algebra, group theory, representation theory, and other fields.
+Within commutative algebra, for instance, the classification of Cohen--Macaulay local rings $R$ for which there are only finitely many indecomposable maximal Cohen--Macaulay $R$-modules (the \emph{finite CM-type} property), or determining whether iterated Frobenius pushforwards of a Noetherian ring in positive characteristic have finitely many isomorphism classes of indecomposable summands (the \emph{finite F-representation type} property) are two well established research problems where making and testing conjectures depends on the ability to find all summands of modules or verify their indecomposability.
+
+Currently there are no widely used algorithms for checking indecomposability or finding summands of modules over commutative rings.
+In contrast, variants of the ``Meat-Axe'' algorithm for determining irreducibility of finite dimensional modules over a group algebra have wide ranging applications in computational group theory \cite{Parker84,HR94,Holt98} and are available through symbolic algebra software such as Magma and GAP \cite{MAGMA,GAP}.
+
+The purpose of this paper is to describe and prove correctness of a practical algorithm for computing indecomposable summands of finitely generated modules over a Noetherian commutative ring.
+In particular, our algorithm works over multigraded rings, which enables applications in toric geometry.
+We present multiple examples in the end.
+
+An accompanying implementation in Macaulay2 is available online via the GitHub repository \\
+\centerline{
+  \href{https://github.com/mahrud/DirectSummands}
+  {\texttt{https://github.com/mahrud/DirectSummands}}.}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{The main algorithm}
+
+Throughout, $R$ will be an $\Z^r$-graded ring with $R_0 = k$ a field and homogeneous maximal ideal $m$.
 Note that if $M$ is a finitely generated $R$-module, and $A\in \End_R(M)$, then $A$ acts on the $k$-vector space $M/mM$.
 
+We begin with the observation that if $\phi\in\End_R(M)$\mahrud{why is it $\phi$ here but $A$ everywhere else?
+I feel somewhat inclined to use $A=\End_R(M)$ and $\alpha\in A$ as an endomorphism.} is an idempotent, then $M$ decomposes as $\im\phi \oplus \ker\phi$; if $\phi$ is nontrivial (i.e., neither an isomorphism nor the zero morphism), then $M$ is decomposable.
+The following lemma allows us to check this condition modulo the maximal ideal:
+
 \begin{lem}
-  Let $M$ be an $R$-module, and let $A\in\End_R(M)$.
-If the induced action of $A$ on $M/mM$ is idempotent\mahrud{Maybe you're proving something stronger, but it's possible that action of $A$ on $M$ is either idempotent of nonzero degree or not idempotent, but the induced action on $M/mM$ is}, then $M$ admits a direct sum decomposition $N\oplus M/N$, where $N\cong \im A$.
+  Let $M$ be an $R$-module, and let $A\in\End_R(M)$. If the induced action of $A$ on $M/mM$ is idempotent, then $M$ admits a direct sum decomposition $N\oplus M/N$, where $N\cong \im A$.
 \end{lem}
-
 \begin{proof}
@@
 If $\l$ is an eigenvalue of $A$, then some power of $A-\l I$ acts idempotently.
-Furthermore, if $\l$ is not the only eigenvalue of $A$, then a nonzero power of $A-\l I$ acts idempotently.
+Furthermore, if $\l$ is not the only eigenvalue of $A$, then a power of $A-\l I$ that is nonzero acts idempotently.
+\end{lem}
+\begin{proof}
+  Since all eigenvalues of $A$ are contained in $k$, we can without loss of generality put $A$ in Jordan canonical form, with each Jordan block having the form
+  \[ r_i\Biggl\{
+  \underbrace{\begin{pmatrix}
+      \l_i & 1 & 0 & \dots & 0 \\
+      0 & \l_i & 1  & \dots & 0 \\
+      \vdots & & & & \vdots \\
+      0 & 0 & 0 & \dots & \l_i
+  \end{pmatrix}}_{r_i} \]
+  with each $\l_i$ an eigenvalue of $A$.
+  In this basis, $A-\l I$ will be block-diagonal with blocks of form
+  \[ \begin{pmatrix}
+    \l_i-\l & 1 & 0 & \dots & 0 \\
+    0 & \l_i-\l & 1 & \dots & 0 \\
+    \vdots & & & & \vdots \\
+    0 & 0 & 0 & \dots & \l_i-\l
+  \end{pmatrix} \]
+  Set $\mu_i=\l_i-\l$. Then $(A-\l I)^l$ is block-diagonal with blocks of the form
+  \[ \begin{pmatrix}
+    \mu_i^l & \binom{l}1\mu_i^{l-1} & \binom{l}2\mu_i^{l-2} & \dots & \binom{l}{r_i}\mu_i^{l-r_i} \\
+    0 & \mu_i^l & \binom{l}1\mu_i^{l-1} & \dots & \binom{l}{r_i-1}\mu_i^{l-r_i+1} \\
+    \vdots & & & & \vdots \\
+    0 & 0 & 0 & \dots & \mu_i^l
+  \end{pmatrix} \]
+  If $l$ is $> r_i$ and divisible by $p$, then all nondiagonal terms will vanish, so all blocks will have the form
+  \[ \begin{pmatrix}
+    \mu_i^l & 0 & 0 & \dots & 0 \\
+    0 & \mu_i^l & 0 & \dots & 0 \\
+    \vdots & & & & \vdots \\
+    0 & 0 & 0 & \dots & \mu_i^l
+  \end{pmatrix}. \]
+  Finally, choosing $l$ to be divisible also by $p^e-1$ for some $e>0$, we have that $$\mu_i^{l-1} = (\mu_i^{p^{e}-1})^{l/(p^{e}-1)} =1,$$ since $\mu_i$ is contained in a finite field of characteristic $p$.
+  Thus, $(A-\l I)^l$ is simply a diagonal matrix with 1 or 0 on the diagonal, hence idempotent. Note moreover that if some $\l_i\neq \l$, then $(A-\l I)^l$ is not the zero matrix.
+\end{proof}
+
+This leads to a probabilistic algorithm to find the indecomposable summands of a module $M$, as follows:
+\begin{enumerate}
+\item Take a general element $A_0$ of $[\End M]_0$, the degree-0 part of $\End M$, and consider the resulting endomorphism $A$ of the $k$-vector space $M/mM$.
+\item Find the eigenvalues of $A$.
+\item If $A$ has at least two eigenvalues, choose one eigenvalue $\l$, and compute a sufficiently high power of $A$ (with the power explicitly as in the proof of \ref{}).
+This power will be a nonzero idempotent, and thus produce a splitting of $M$ as $\im A_0 \oplus \coker A_0$ by \ref{}.
+\item Repeat steps (1)--(3) for both $\im A_0$ and $\coker A_0$.
+\end{enumerate}
+
+The following observation implies that if $M$ is indecomposable, then the above algorithm should find the decomposition of $M$:
+
+\begin{lem}
+  If $M$ is not indecomposable, then a general degree-0 endomorphism of $M$ reduces to an endomorphism of $M/mM$ with at least two distinct eigenvalues.
 \end{lem}
 
+\begin{rem}
+  By ``general'' we mean that a general linear combination of a basis for $[\End(M)]_0$ over the algebraic closure of $k$, or equivalently over a sufficiently large algebraic extension of $k$.
+\end{rem}
+
 \begin{proof}
-Since all eigenvalues of $A$ are contained in $k$, we can without loss of generality put $A$ in Jordan canonical form, with each Jordan block having the form
-$$
-r_i\Biggl\{
-\underbrace{\begin{pmatrix}
-\l_i & 1 & 0 &\dots & 0
-\\
-0& \l_i & 1  &\dots & 0
-\\
-\vdots & &&&\vdots
-\\
-0& 0 & 0  &\dots & \l_i
-\end{pmatrix}}_{r_i}
-$$
-with each $\l_i$ an eigenvalue of $A$.
-In this basis, $A-\l I$ will be block-diagonal with blocks of form
-$$
-\begin{pmatrix}
-\l_i-\l & 1 & 0 &\dots & 0
-\\
-0& \l_i-\l & 1  &\dots & 0
-\\
-\vdots & &&&\vdots
-\\
-0& 0 & 0  &\dots & \l_i-\l
-\end{pmatrix}
-$$
-Set $\mu_i=\l_i-\l$. Then $(A-\l I)^l$ is block-diagonal with blocks of the form
-$$
-\begin{pmatrix}
-\mu_i^l & \binom{l}1\mu_i^{l-1} & \binom{l}2\mu_i^{l-2} &\dots & \binom{l}{r_i}\mu_i^{l-r_i}
-\\
-0& \mu_i^l &
-\binom{l}1\mu_i^{l-1}  &\dots & \binom{l}{r_i-1}\mu_i^{l-r_i+1}
-\\
-\vdots & &&&\vdots
-\\
-0& 0 & 0  &\dots & \mu_i^l
-\end{pmatrix}
-$$
-If $l$ is $> r_i$ and divisible by $p$, then all nondiagonal terms will vanish, so all blocks will have the form
-$$
-\begin{pmatrix}
-\mu_i^l & 0 & 0 &\dots & 0
-\\
-0& \mu_i^l & 0  &\dots & 0
-\\
-\vdots & &&&\vdots
-\\
-0& 0 & 0  &\dots & \mu_i^l 
-\end{pmatrix}.
-$$
-Finally, choosing $l$ to be divisible also by $p^e-1$ for some $e>0$, we have that $$\mu_i^{l-1} = (\mu_i^{p^{e}-1})^{l/(p^{e}-1)} =1,$$ since $\mu_i$ is contained in a finite field of characteristic $p$.
-Thus, $(A-\l I)^l$ is simply a diagonal matrix with 1 or 0 on the diagonal, hence idempotent. Note moreover that if some $\l_i\neq \l$, then $(A-\l I)^l$ is not the zero matrix.
+  We may assume that the base field $k$ is algebraically closed.
+  Let $\Phi_1,\dots,\Phi_r$ be a basis for $[\End(M)]_0$, and $\phi_1,\dots,\phi_r$ their images modulo $m$, which we view as matrices with entries in $k$.
+  Let $U\subset \A^r$ be the subset of $r$-tuples $(\l_1,\dots,\l_r)$ such that $\l_1\phi_1+\dots+\l_r\phi_r$ has at least two distinct eigenvalues, i.e., such that $\l_1\Phi_1+\dots+\l_r\Phi_r$ reduces to an endomorphism of $M/mM$ with at least two distinct eigenvalues.
 
+  It suffices to show that $U$ is a nonempty open subset of $\A^r$. First, we show $U$ is nonempty:
+  Say $M=M_1\oplus M_2$, with $M_1$ a nontrivial indecomposable summand.
+We may choose $\Phi_1$ to be the projection to $M_1$, and $\Phi_2$ the projection to $M_2$.
+Then for any $\l_1,\l_2\in k$, $\l_1\phi_1 + \l_2\phi_2$ has eigenvalues $\l_1,\l_2$; thus in particular there is \emph{an} element of $[\End(M)]_0$ reducing to an endomorphism of $M/mM$ with distinct eigenvalues, so $U$ is nonempty.
+
+  Now, we show that $U$ is open.
+This is a purely linear algebraic statement: we claim that given a matrix $\phi$ with at least two distinct eigenvalues, and any $r$ matrices $\phi_1,\dots,\phi_r$, that
+  $$ A_{\l_1,\dots,\l_r}:=\phi+\l_1\phi_1+\dots+\l_r\phi_r $$
+  has at least two distinct eigenvalues for $\l_1,\dots,\l_r$ outside a Zariski-closed subset of $\A^r$.
+  The eigenvalues of $A_{\l_1,\dots,\l_r}$ are the roots of $\det(A_{\l_1,\dots,\l_r}-t I)$, which is of course a polynomial in $t$ with coefficients in $\l_1,\dots,\l_r$.
+$A_{\l_1,\dots,\l_r}$ fails to have at least two distinct eigenvalues exactly when this polynomial factors as a power of a linear term.
+This is a polynomial condition in the coefficients of powers of $t$ in $\det(A_{\l_1,\dots,\l_r}-t I)$ and thus in the $\l_i$; to see this, note that if
+  $ f:=t^n b_n +\dots +t b_1 +b_0 $
+  has an $n$-fold root exactly when
+  $ f,\d f/\d t,\dots, \d^n f/\d t^n $
+  vanish simultaneously; the resultant of these $n$ polynomials in $n$ equations gives polynomial conditions in the $b_i$ for this to occur; in our setting, the $b_i$ are themselves polynomials in the $\l_i$, and thus we have obtained polynomial equations defining the locus where $A_{\l_1,\dots,\l_r}$ fails to have distinct roots.
 \end{proof}
+
+\begin{rem}
+  Note that the above algorithm is quite sensitive to the ground field $k$, because it needs an eigenvalue of the endomorphism $A$ of $M/mM$ to be contained in $k$.
+While theoretically the issue can be avoided by working over an algebraically closed ground field $\bar k$, for practical use on a computer algebra system it is better to extend $k$ to some larger finite field.
+However, the general linear combinations we take in Step \ref{} should be taken with respect to the prime subfield (otherwise, as we increase the size of the finite field $k$, the eigenvalues of a general linear combination will live in higher and higher field extensions).
+\end{rem}
+
+If the above algorithm fails to produce a nontrivial idempotent, it does not certify that $M$ is indecomposable.
+However, there are a few sufficient conditions to be indecomposable, which in practice often (but not always) produce such a certification.
+The following sufficiency condition is immediate, but can be quite useful in practice for verifying indecomposability:
+
+\begin{lem}
+  Let $M$ be a finitely generated $R$-module and let $[\End M]_0$ be the $k$-vector space of degree-0 endomorphisms. Suppose that either:
+  \begin{enumerate}
+  \item $[\End M]_0$ is 1-dimensional and thus spanned by the identity endomorphism, or
+  \item every non-identity element of $[\End M]_0$, viewed as a matrix, has all entries contained in $m$;
+  \end{enumerate}
+  then $M$ is indecomposable.
+\end{lem}
+\begin{proof}
+  If $M$ decomposes nontrivially as $M_1\oplus M_2$, then the projections onto each factor are nontrivial degree-0 endomorphisms not equal to the identity, and which does not have entries contained in $m$.
+\end{proof}
+
+This is the local analog of the following fact about indecomposability of coherent sheaves:
+
+\begin{cor}
+  Let $X$ be a projective variety over a field $k$, and $\F$ a coherent sheaf on $X$.
+  If $ H^0(\End \F) = k, $ then $\F$ is indecomposable.
+\end{cor}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Examples}
+
+While the preceding section was written in the language of modules, by the standard translation to global (multi)projective varieties, it works equally well to find indecomposable decompositions of coherent sheaves.
+In this section, we give examples of the kind of observations the algorithm from the previous section allows us to make.
+
+\renewcommand{\char}{\operatorname{char}}
+
+\begin{exa}(Frobenius pushforward on the projective space $\P^n$).
+  Let $S = k[x_0,\dots,x_n]$ be a polynomial ring with $\char k = p$ and $\deg x_i = 1$ and consider the Frobenius endomorphism
+  \[ F\colon S\to S \quad \text{given by} \quad f \to f^p. \]
+  Hartshorne proved \cite{Hartshorne70} that for any line bundle $L\in\Pic\P^n$, the Frobenius pushforward $F_*L$ splits as a sum of line bundles.
+  \begin{align*}
+    \text{When p=3, n=2:} \\
+    F_*\O_{\P^2} = \O \oplus &\O(-1)^7 \oplus \O(-2). \\
+    \text{When p=2, n=5:} \\
+    F_*\O_{\P^5} = \O \oplus &\O(-1)^{15} \oplus \O(-2)^{15} \oplus \O(-3). \\
+    F_*^2\O_{\P^5} = \O \oplus &\O(-1)^{120} \oplus \O(-2)^{546} \oplus \O(-3)^{336} \oplus \O(-4)^{21}
+  \end{align*}
+\end{exa}
+
+\begin{exa}(Frobenius pushforward on toric varieties).
+  Let $X$ be a smooth toric variety and consider its Cox ring
+  \[ S = \bigoplus\!{}_{[D]\in\Pic{X}} \; \Gamma(X, \O(D)). \]
+  Similar to the case of the projective space, Bogvad and Thomsen showed that $F_*L$ totally splits as a direct sum of line bundles for any line bundle $L\in\Pic X$ \cite{Bogvad98,Thomsen00}.
+
+  As an example, consider the Hirzebruch surface $X$ of type 3 over a field of characteristic 2.
+  \[ F_*\O_X = \O_X \oplus \O_X(-1,0)^2 \oplus \O_X(0,-1)^2 \oplus \O_X(1,-1)^3 \oplus \O_X(2,-1). \]
+
+  In fact, Achinger \cite{Achinger13} showed that the total splitting of $F_*L$ characterizes smooth projective toric varieties.
+
+  % The toric Frobenius is simply multiplication by $p$ on the torus.
+  % The $p$ need not be prime, nor even match the characteristic!
+\end{exa}
+
+\begin{exa}(Frobenius pushforward on Grassmannians).
+  Consider the Grassmannian $X = \mathrm{Gr(2,4)}$. We may work over the Cox ring $S$,
+  %\[ S = \bigoplus\!{}_{[D]\in\Pic{X}} \; \Gamma(X, \O(D)). \]
+  which in this case coincides with the coordinate ring
+  \[ S = \frac{k[p_{0,1},\dots p_{0,2},p_{1,2},p_{0,3},p_{1,3},p_{2,3}]}{p_{1,2}p_{0,3}-p_{0,2}p_{1,3}+p_{0,1}p_{2,3}}. \]
+  Then in characteristic $p=3$ we have:
+  \[ F_*\O_X = \O \oplus \O(-1)^{44} \oplus \O(-2)^{20} \oplus A^4 \oplus B^4, \]
+  where $A$ and $B$ are rank 2 indecomposable bundles (c.f.~\mahrud{cite[Raedschelders--Špenko--Van den Bergh, '17]}).
+\end{exa}
+
+\begin{exa}(Frobenius pushforward on Mori Dream Spaces).
+  Continuing with the theme of computations over the Cox ring, the natural geometric setting is to consider the class of projective varieties known as Mori dream spaces \cite{HuKeel00}.
+
+  For instance, consider $X = \mathrm{Bl}_4\P^2$, the blowup of $\P^2$ at 4 general points. We will working over the $\Z^5$-graded Cox ring
+  \[ S = k[x_1,\dots,x_{10}]/\text{(five quadric Pl\"ucker relations)} \]
+  With degrees
+    \[
+    \left(\!\begin{array}{cccccccccc}
+      0&0&0&0&1&1&1&1&1&1 \\
+      1&0&0&0&-1&-1&-1&0&0&0 \\
+      0&1&0&0&-1&0&0&-1&-1&0 \\
+      0&0&1&0&0&-1&0&-1&0&-1 \\
+      0&0&0&1&0&0&-1&0&-1&-1
+    \end{array}\!\right)
+    \]
+    Then in characteristic 2 we have:
+    \begin{align*}
+      F_*^2\O_X = {\O_{X}^{1}}
+      &\oplus {\O_{X}^{2}\ \left(-2,\,1,\,1,\,1,\,1\right)} \oplus {\O_{X}^{2}\ \left(-1,\,0,\,0,\,0,\,1\right)} \\
+      &\oplus {\O_{X}^{2}\ \left(-1,\,0,\,0,\,1,\,0\right)} \oplus {\O_{X}^{2}\ \left(-1,\,0,\,1,\,0,\,0\right)} \\
+      &\oplus {\O_{X}^{2}\ \left(-1,\,1,\,0,\,0,\,0\right)} \oplus B \oplus G,
+    \end{align*}
+    where $B, G$ are rank 3 and rank 2 indecomposable modules \cite{Hara15}.
+\end{exa}
+
+\begin{exa}
+Cubic surfaces
+\end{exa}
+
+\begin{exa}
+Summands of syzyzgies
+\end{exa}
+
+\end{document}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
   Consider the short exact sequence
-$$
-0\to \ker A \to M \to \im A\to 0.
-$$
+  $$ 0\to \ker A \to M \to \im A\to 0. $$
   The map $M\to M$ given by $-A+\id_M$ sends $M\to \ker A$, since $-A(A-\id_M)(m) = -(A^2-A)(m)=0$.
@@
   Consider the short exact sequence
-$$
-0\to \ker A \to M \to \im A\to 0.
-$$
+  $$ 0\to \ker A \to M \to \im A\to 0. $$
   The map $M\to M$ given by $-A^{d-1}+\id_M$ sends $M\to \ker A$, since $-A(A^{d-1}-\id_M)(m) = -(A^d-A)(m)=0$.
 In fact, this map $M\to \ker A$ is a splitting of $\ker A\to M$, since if $m\in \ker A$ then also $m\in \ker A^{d-1}$ and thus we have $-(A^{d-1}-\id_M)(m) = -A^d(m)+\id_M(m)=m$.
-
 \end{proof}
 
-\end{document}
+% Local Variables:
+% tab-width: 2
+% eval: (visual-line-mode)
+% End:

### 2024-02-16 paper/main.tex
added final MEGA 2024 submission

@@
 
-\title{Computing direct sum decompositions}
+\title{Computing Direct Sum Decompositions}
 \author{Devlin Mallory and Mahrud Sayrafi}
@@
 % for these, authors can choose to submit either an extended abstract of at least six pages (not counting bibliography) that should contain full results, an outline of the proofs and all of the main ideas; or else a full paper with complete proofs.
+
+% Abstract
+% The problems of finding isomorphism classes of indecomposable modules with certain properties, or determining the indecomposable summands of a module, are ubiquitous in commutative algebra, group theory, representation theory, and other fields.
+The purpose of this paper is to describe and prove correctness of a practical algorithm for computing indecomposable summands of finitely generated modules over a finitely generated $k$-algebra, for $k$ a field of positive characteristic.
+Our algorithm works over multigraded rings, which enables the computation of indecomposable summands of coherent sheaves on subvarieties of toric varieties (in particular, for varieties embedded in projective space).
+We also present multiple examples, including some which present previously unknown phenomena regarding the behavior of summands of Frobenius pushforwards and syzygies over Artinian rings.
 
@@
 
-The problem of finding isomorphism classes of indecomposable modules with certain properties is ubiquitous in commutative algebra, group theory, representation theory, and other fields.
-Within commutative algebra, for instance, the classification of Cohen--Macaulay local rings $R$ for which there are only finitely many indecomposable maximal Cohen--Macaulay $R$-modules (the \emph{finite CM-type} property), or determining whether iterated Frobenius pushforwards of a Noetherian ring in positive characteristic have finitely many isomorphism classes of indecomposable summands (the \emph{finite F-representation type} property) are two well established research problems where making and testing conjectures depends on the ability to find all summands of modules or verify their indecomposability.
+The problems of finding isomorphism classes of indecomposable modules with certain properties, or determining the indecomposable summands of a module, are ubiquitous in commutative algebra, group theory, representation theory, and other fields.
+Within commutative algebra, for instance, the classification of Cohen--Macaulay local rings $R$ for which there are only finitely many indecomposable maximal Cohen--Macaulay $R$-modules (the \emph{finite CM-type} property), or determining whether iterated Frobenius pushforwards of a Noetherian ring in positive characteristic have finitely many isomorphism classes of indecomposable summands (the \emph{finite F-representation type} property) are two well-established research problems.
+For both these problems, and many others, making and testing conjectures depends on finding summands of modules and verifying their indecomposability.
 
-Currently there are no widely used algorithms for checking indecomposability or finding summands of modules over commutative rings.
+Currently there are no efficient algorithms available for checking indecomposability or finding summands of modules over commutative rings.
 In contrast, variants of the ``Meat-Axe'' algorithm for determining irreducibility of finite dimensional modules over a group algebra have wide ranging applications in computational group theory \cite{Parker84,HR94,Holt98} and are available through symbolic algebra software such as Magma and GAP \cite{MAGMA,GAP}.
 
-The purpose of this paper is to describe and prove correctness of a practical algorithm for computing indecomposable summands of finitely generated modules over a Noetherian commutative ring.
-In particular, our algorithm works over multigraded rings, which enables applications in toric geometry.
-We present multiple examples in the end.
+The purpose of this paper is to describe and prove correctness of a practical algorithm for computing indecomposable summands of finitely generated modules over a finitely generated $k$-algebra, for $k$ a field of positive characteristic.
+In particular, our algorithm works over multigraded rings, which enables the computation of indecomposable summands of coherent sheaves on subvarieties of toric varieties (in particular, for varieties embedded in projective space).
+After describing the algorithm and proving its correctness, we present multiple examples in the end, including some which present previously unknown phenomena regarding the behavior of summands of Frobenius pushforwards and syzygies over Artinian rings.
 
-An accompanying implementation in Macaulay2 is available online via the GitHub repository \\
+An accompanying implementation in Macaulay2 is available online via the GitHub repository 
 \centerline{
@@
 
+\begin{rem}
+  Although the algorithm described below is only proved to result (probabilistically) in a decomposition into indecomposable summands in positive characteristic, in practice it often does produce nontrivial indecomposable decompositions even in characteristic 0.
+Moreover, if a module over a ring of characteristic 0 is decomposable, its reductions modulo $p$ will be as well; thus, our algorithm provides a heuristic for verifying decomposability in characteristic 0.
+
+  We also point out that while the discussion below, and our implementation in Macaulay2, concerns the case where $R$ is a \emph{commutative} $k$-algebra, many of the techniques extend beyond this case; we plan to extend the results and algorithms to noncommutative rings such as Weyl algebras in future work.
+\end{rem}
+
 \section{The main algorithm}
 
-Throughout, $R$ will be an $\Z^r$-graded ring with $R_0 = k$ a field and homogeneous maximal ideal $m$.
+Throughout, $R$ will be an $\Z^r$-graded ring with $R_0 = k$ a field of positive characteristic and homogeneous maximal ideal $m$.
 Note that if $M$ is a finitely generated $R$-module, and $A\in \End_R(M)$, then $A$ acts on the $k$-vector space $M/mM$.
 
-We begin with the observation that if $\phi\in\End_R(M)$\mahrud{why is it $\phi$ here but $A$ everywhere else?
-I feel somewhat inclined to use $A=\End_R(M)$ and $\alpha\in A$ as an endomorphism.} is an idempotent, then $M$ decomposes as $\im\phi \oplus \ker\phi$; if $\phi$ is nontrivial (i.e., neither an isomorphism nor the zero morphism), then $M$ is decomposable.
-The following lemma allows us to check this condition modulo the maximal ideal:
+We begin with the observation that if $A\in\End_R(M)$ is an idempotent, then $M$ decomposes as $\im A \oplus \ker A$; if $A$ is nontrivial (i.e., neither an isomorphism nor the zero morphism), then $M$ is decomposable.
+The following lemma allows us to check only for idempotents modulo the maximal ideal:
 
-\begin{lem}
+\begin{lem}\label{idem}
   Let $M$ be an $R$-module, and let $A\in\End_R(M)$. If the induced action of $A$ on $M/mM$ is idempotent, then $M$ admits a direct sum decomposition $N\oplus M/N$, where $N\cong \im A$.
@@
 We claim that this composition is surjective.
-We may complete at the maximal ideal to check this, and thus assume $R$ and $M$ are complete\mahrud{out of curiosity, what can go wrong?}.
+We may complete at the maximal ideal to check this, and thus assume $R$ and $M$ are complete.
 Let $n_0\in N$.
@@
   \[ A(n_0) = A^2(m_1) = A(m_1) + n_1 = n_0 + n_1, \]
-  or equivalently \mahrud{should this say $n_1 = A(n_0) - n_0$ instead?}
+  or equivalently
   \[ n_0 = A(n_0) - n_1 \]
@@
   Thus, we can write $n_1 = A(m_2)$ for $m_2\in M$.
-  Now, apply $A$ to both sides: by the assumption that $A$ is idempotent modulo $m$\mahrud{how is this used here?}, we have
+  Now, apply $A$ to both sides: by the assumption that $A$ is idempotent modulo $m$, we have
   \[ A(n_1)=A^2(m_2) = A(m_2) + n_2 = n_1+n_2, \]
@@
 \begin{lem}
+\label{JordanLemma}
   Let $k$ be a finite field of characteristic $p$, and let $A$ be an endomorphism of a $k$-vector space such that all eigenvalues of $A$ are contained in $k$.
-If $\l$ is an eigenvalue of $A$, then some power of $A-\l I$ acts idempotently.
-Furthermore, if $\l$ is not the only eigenvalue of $A$, then a power of $A-\l I$ that is nonzero acts idempotently.
+If $\l$ is an eigenvalue of $A$, then some power of $A-\l I$ is an idempotent.
+Furthermore, if $\l$ is not the only eigenvalue of $A$, then a power of $A-\l I$ that is nonzero is an idempotent.
 \end{lem}
@@
   \end{pmatrix} \]
-  If $l$ is $> r_i$ and divisible by $p$, then all nondiagonal terms will vanish, so all blocks will have the form
+  If $l$ is $> r_i$ and divisible by $p$, then all non-diagonal terms will vanish, so all blocks will have the form
   \[ \begin{pmatrix}
@@
 \begin{enumerate}
-\item Take a general element $A_0$ of $[\End M]_0$, the degree-0 part of $\End M$, and consider the resulting endomorphism $A$ of the $k$-vector space $M/mM$.
+\item \label{1} Take a general element $A_0$ of $[\End M]_0$, the degree-0 part of $\End M$, and consider the resulting endomorphism $A$ of the $k$-vector space $M/mM$.
 \item Find the eigenvalues of $A$.
-\item If $A$ has at least two eigenvalues, choose one eigenvalue $\l$, and compute a sufficiently high power of $A$ (with the power explicitly as in the proof of \ref{}).
-This power will be a nonzero idempotent, and thus produce a splitting of $M$ as $\im A_0 \oplus \coker A_0$ by \ref{}.
+\item If $A$ has at least two eigenvalues, choose one eigenvalue $\l$, and compute a sufficiently high power of $A$ (with the power explicitly as in the proof of Lemma~\ref{JordanLemma}).
+This power will be a nonzero idempotent, and thus produce a splitting of $M$ as $\im A_0 \oplus \coker A_0$ by Lemma~\ref{idem}.
 \item Repeat steps (1)--(3) for both $\im A_0$ and $\coker A_0$.
@@
 While theoretically the issue can be avoided by working over an algebraically closed ground field $\bar k$, for practical use on a computer algebra system it is better to extend $k$ to some larger finite field.
-However, the general linear combinations we take in Step \ref{} should be taken with respect to the prime subfield (otherwise, as we increase the size of the finite field $k$, the eigenvalues of a general linear combination will live in higher and higher field extensions).
+However, the general linear combinations we take in Step \ref{1} should be taken with respect to the prime subfield (otherwise, as we increase the size of the finite field $k$, the eigenvalues of a general linear combination will live in higher and higher field extensions).
+See Example~\ref{algclos} for a demonstration of the necessity of extending the base field.
 \end{rem}
@@
 \begin{proof}
-  If $M$ decomposes nontrivially as $M_1\oplus M_2$, then the projections onto each factor are nontrivial degree-0 endomorphisms not equal to the identity, and which does not have entries contained in $m$.
+  If $M$ decomposes non-trivially as $M_1\oplus M_2$, then the projections onto each factor are nontrivial degree-0 endomorphisms not equal to the identity, and which does not have entries contained in~$m$.
 \end{proof}
@@
 While the preceding section was written in the language of modules, by the standard translation to global (multi)projective varieties, it works equally well to find indecomposable decompositions of coherent sheaves.
-In this section, we give examples of the kind of observations the algorithm from the previous section allows us to make.
+In this section, we give examples of the kind of calculations and observations the algorithm from the previous section allows us to make.
 
@@
 
-\begin{exa}(Frobenius pushforward on the projective space $\P^n$).
+\begin{exa}[Frobenius pushforward on the projective space $\P^n$]
   Let $S = k[x_0,\dots,x_n]$ be a polynomial ring with $\char k = p$ and $\deg x_i = 1$ and consider the Frobenius endomorphism
   \[ F\colon S\to S \quad \text{given by} \quad f \to f^p. \]
-  Hartshorne proved \cite{Hartshorne70} that for any line bundle $L\in\Pic\P^n$, the Frobenius pushforward $F_*L$ splits as a sum of line bundles.
+  Hartshorne \cite{Hartshorne70} proved that for any line bundle $L\in\Pic\P^n$, the Frobenius pushforward $F_*L$ splits as a sum of line bundles.
+While the following calculations are straightforward to do by hand, they are immediately calculated via our algorithm:
   \begin{align*}
-    \text{When p=3, n=2:} \\
-    F_*\O_{\P^2} = \O \oplus &\O(-1)^7 \oplus \O(-2). \\
-    \text{When p=2, n=5:} \\
-    F_*\O_{\P^5} = \O \oplus &\O(-1)^{15} \oplus \O(-2)^{15} \oplus \O(-3). \\
-    F_*^2\O_{\P^5} = \O \oplus &\O(-1)^{120} \oplus \O(-2)^{546} \oplus \O(-3)^{336} \oplus \O(-4)^{21}
+    \text{When }p=3, n=2: \\
+    F_*\O_{\P^2} &= \O \oplus \O(-1)^7 \oplus \O(-2). \\
+    \text{When } p=2, n=5: \\
+    F_*\O_{\P^5} &= \O \oplus \O(-1)^{15} \oplus \O(-2)^{15} \oplus \O(-3). \\
+    F_*^2\O_{\P^5} &= \O \oplus \O(-1)^{120} \oplus \O(-2)^{546} \oplus \O(-3)^{336} \oplus \O(-4)^{21}
   \end{align*}
@@
 
-\begin{exa}(Frobenius pushforward on toric varieties).
+\begin{exa}[Frobenius pushforward on toric varieties]
   Let $X$ be a smooth toric variety and consider its Cox ring
-  \[ S = \bigoplus\!{}_{[D]\in\Pic{X}} \; \Gamma(X, \O(D)). \]
-  Similar to the case of the projective space, Bogvad and Thomsen showed that $F_*L$ totally splits as a direct sum of line bundles for any line bundle $L\in\Pic X$ \cite{Bogvad98,Thomsen00}.
+  \[ S = \bigoplus_{[D]\in\Pic{X}} \; \Gamma(X, \O(D)). \]
+  Similar to the case of the projective space, B{\o}gvad and Thomsen
+\cite{Bogvad98,Thomsen00}
+ showed that $F_*L$ totally splits as a direct sum of line bundles for any line bundle $L\in\Pic X$.
 
-  As an example, consider the Hirzebruch surface $X$ of type 3 over a field of characteristic 2.
-  \[ F_*\O_X = \O_X \oplus \O_X(-1,0)^2 \oplus \O_X(0,-1)^2 \oplus \O_X(1,-1)^3 \oplus \O_X(2,-1). \]
-
-  In fact, Achinger \cite{Achinger13} showed that the total splitting of $F_*L$ characterizes smooth projective toric varieties.
+  As an example, consider the third Hirzebruch surface $X=\P(\O_{\P^1}\oplus \O_{\P^1}(3))$ over a field of characteristic 3. We have, for example, that
+\begin{align*}
+F_*\O_X &= \O_X \oplus \O_X(-1,0)^2 \oplus \O_X(0,-1)^2 \oplus \O_X(1,-1)^3 \oplus \O_X(2,-1)\\
+F_*\O_X(1,1) &= \O_X^3 \oplus \O_X(-1,0) \oplus \O_X(1,-1) \oplus \O_X(1,0)^2 \oplus \O_X(2,-1)^2
+\end{align*}
+  In fact, Achinger \cite{Achinger13} showed that the total splitting of $F_*L$ for every line bundle $L$ characterizes smooth projective toric varieties.
 
@@
 
-\begin{exa}(Frobenius pushforward on Grassmannians).
+\begin{exa}[Frobenius pushforward on elliptic curves]\label{algclos}
+  Consider the elliptic curve $X=\Proj (\mathbb F_7[x,y,z]/(x^3+y^3+z^3))$.
+  %%Weierstrauss form is y^2=x^3-432
+  This is an ordinary elliptic curve, hence $F$-split; thus $\O_X$ is a summand of $F_* \O_X$.
+Over the algebraic closure of $\mathbb F_7$, $F_*\O_X$ will decompose as $\bigoplus_{p=1}^7 \O_X(p_i)$, where $p_1,\dots,p_7$ are the 7-torsion points of $X$.
+
+  However, over $\mathbb F_7$, our algorithm calculates that $F_*\O_X$ decomposes only as 
+  $$ F_* \O_X =\O_X \oplus M_1\oplus M_2\oplus M_3, $$
+  with $M_i$ indecomposable (over $\mathbb F_7$) of rank 2.
+
+  If one extends the ground field to $\mathbb F_{49}$, however, our algorithm calculates the full decomposition 
+  $$ F_* \O_X=\bigoplus_{p=1}^7 \O_X(p_i). $$
+  This reflects the fact that the 7-torsion points $p_i$ of $X$, and thus the sheaves $\O_X(p_i)$, are not defined over $\mathbb F_7$, but they are defined over $\mathbb F_{49}$.
+\end{exa}
+
+\begin{exa}[Frobenius pushforward on Grassmannians]
   Consider the Grassmannian $X = \mathrm{Gr(2,4)}$. We may work over the Cox ring $S$,
@@
   which in this case coincides with the coordinate ring
-  \[ S = \frac{k[p_{0,1},\dots p_{0,2},p_{1,2},p_{0,3},p_{1,3},p_{2,3}]}{p_{1,2}p_{0,3}-p_{0,2}p_{1,3}+p_{0,1}p_{2,3}}. \]
+  \[ S = \frac{k[p_{0,1},p_{0,2},p_{0,3},p_{1,2},p_{1,3},p_{2,3}]}{p_{1,2}p_{0,3}-p_{0,2}p_{1,3}+p_{0,1}p_{2,3}}. \]
   Then in characteristic $p=3$ we have:
   \[ F_*\O_X = \O \oplus \O(-1)^{44} \oplus \O(-2)^{20} \oplus A^4 \oplus B^4, \]
-  where $A$ and $B$ are rank 2 indecomposable bundles (c.f.~\mahrud{cite[Raedschelders--Špenko--Van den Bergh, '17]}).
+  where $A$ and $B$ are rank-2 indecomposable bundles (c.f.~\cite{RSB22}).
 \end{exa}
 
-\begin{exa}(Frobenius pushforward on Mori Dream Spaces).
+\begin{exa}[Frobenius pushforward on Mori Dream Spaces]
   Continuing with the theme of computations over the Cox ring, the natural geometric setting is to consider the class of projective varieties known as Mori dream spaces \cite{HuKeel00}.
@@
   \[ S = k[x_1,\dots,x_{10}]/\text{(five quadric Pl\"ucker relations)} \]
-  With degrees
+  with degrees
     \[
-    \left(\!\begin{array}{cccccccccc}
+    \left(\!\begin{array}{rrrrrrrrrr}
       0&0&0&0&1&1&1&1&1&1 \\
@@
       0&0&0&1&0&0&-1&0&-1&-1
-    \end{array}\!\right)
+    \end{array}\!\right).
     \]
@@
     \end{align*}
-    where $B, G$ are rank 3 and rank 2 indecomposable modules \cite{Hara15}.
+    where $B, G$ are rank-3 and rank-2 indecomposable modules, as calculated in \cite{Hara15}.
 \end{exa}
 
-\begin{exa}
-Cubic surfaces
+\begin{exa}[Frobenius pushforward on cubic surfaces]
+  Let $X$ be a smooth cubic surface. Aside from a single exception in characteristic 0, $X$ will be globally $F$-split, so that any $F^e_*\O_X $ admits $\O_X$ as a direct summand.
+  The other summands of Frobenius pushforwards of $\O_X$ have yet to be studied, and in particular it is not known whether such rings should have the finite $F$-representation type property.
+
+  The use of our algorithm to compute examples in small $p$ and $e$ suggest the following behavior: 
+  $$ F_* \O_X = \O_X\oplus M, $$
+  with $M$ indecomposable, and furthermore $F^*_e M$ remains indecomposable for all $e\geq 0$. In other words, the indecomposable decomposition of $F^e_* \O_X$ is 
+  $$ F_*^e \O_X \cong \O_X\oplus M\oplus F_* M\oplus\dots\oplus F_*^{e-1}M. $$
+  In particular, $\O_X$ will fail to have the finite $F$-representation type property.
+  In fact, we believe a similar description holds true for quartic del Pezzos, which arise as an intersection of quadrics in $P^4$.
 \end{exa}
 
-\begin{exa}
-Summands of syzyzgies
+\begin{exa}[Syzygies over Artinian rings]
+  Let $R = k[x,y]/(x^3,x^2y^3,y^5)$ and consider the (infinite) minimal free resolution of the residue field, which has rank $2^n$ in homological index $n$.
+In forthcoming work based in part on examples calculated using our algorithm, Dao, Eisenbud, and Polini study the indecomposable summands of syzygy modules in examples such as this one, showing unexpected periodicity behavior, and in particular proving that the syzygy modules are direct sums of only three indecomposable modules: the residue field $k$, the maximal ideal $m$, and an additional module $M$.
+
+  For example, the fourth syzygy module decomposes (ignoring the grading) as the direct sum
+  $$ k^3 \oplus m^2 \oplus M^3. $$
+  and the fifth syzygy module as 
+  $$ k^8\oplus m^9 \oplus M^2. $$
+  The use of our algorithm was essential to the observation that only the one additional module $M$ beyond the ``guaranteed'' summands of $k$ and $m$ appealrs.
+  %% nebedsPackage "DirectSummands"
+  %% R = ZZ/101[x,y]/(ideal"x3,x2y3,y5")
+  %% ideal(x^2*y^2,x*y^3)
+  %% F = res (cokernel vars R, LengthLimit => 6)
+  %% netList apply(length F, i -> summands coker F.dd_i)
 \end{exa}
 
+\bibliographystyle{alpha}
+\bibliography{references}
 \end{document}

### 2024-05-11 paper/main.tex
paper changes

@@
-\documentclass{article}
+s\documentclass{article}
 
@@
 
-\begin{lem}\label{idem}
+\begin{lem}\label{lem:idemp}
   Let $M$ be an $R$-module, and let $A\in\End_R(M)$. If the induced action of $A$ on $M/mM$ is idempotent, then $M$ admits a direct sum decomposition $N\oplus M/N$, where $N\cong \im A$.
@@
 
-  By the Artin--Rees lemma, for there's some positive integer $k$ such that for $n\gg0$ we can write
+  By the Artin--Rees lemma, there's some positive integer $k$ such that for $n\gg0$ we can write
   \[ m^n M\cap N = m^{n-k} ( m^kM\cap N)\subset m^{n-k} N. \]
@@
 
-\begin{lem}
-\label{JordanLemma}
+\begin{lem}\label{lem:jordan}
   Let $k$ be a finite field of characteristic $p$, and let $A$ be an endomorphism of a $k$-vector space such that all eigenvalues of $A$ are contained in $k$.
@@
   \end{pmatrix} \]
-  If $l$ is $> r_i$ and divisible by $p$, then all non-diagonal terms will vanish, so all blocks will have the form
+  If $l > r_i$ and $l$ divisible by $p$, then all non-diagonal terms will vanish, so all blocks will have the form
   \[ \begin{pmatrix}
@@
 \item Find the eigenvalues of $A$.
-\item If $A$ has at least two eigenvalues, choose one eigenvalue $\l$, and compute a sufficiently high power of $A$ (with the power explicitly as in the proof of Lemma~\ref{JordanLemma}).
-This power will be a nonzero idempotent, and thus produce a splitting of $M$ as $\im A_0 \oplus \coker A_0$ by Lemma~\ref{idem}.
+\item If $A$ has at least two eigenvalues, choose one eigenvalue $\l$, and compute a sufficiently high power of $A$ (with the power explicitly as in the proof of Lemma~\ref{lem:jordan}).
+This power will be a nonzero idempotent, and thus produce a splitting of $M$ as $\im A_0 \oplus \coker A_0$ by Lemma~\ref{lem:idemp}.
 \item Repeat steps (1)--(3) for both $\im A_0$ and $\coker A_0$.
@@
   \[ F\colon S\to S \quad \text{given by} \quad f \to f^p. \]
-  Hartshorne \cite{Hartshorne70} proved that for any line bundle $L\in\Pic\P^n$, the Frobenius pushforward $F_*L$ splits as a sum of line bundles.
+  Hartshorne \cite{Hartshorne1970} proved that for any line bundle $L\in\Pic\P^n$, the Frobenius pushforward $F_*L$ splits as a sum of line bundles.
 While the following calculations are straightforward to do by hand, they are immediately calculated via our algorithm:
@@
     \text{When } p=2, n=5: \\
-    F_*\O_{\P^5} &= \O \oplus \O(-1)^{15} \oplus \O(-2)^{15} \oplus \O(-3). \\
-    F_*^2\O_{\P^5} &= \O \oplus \O(-1)^{120} \oplus \O(-2)^{546} \oplus \O(-3)^{336} \oplus \O(-4)^{21}
+    F_*\O_{\P^5} &= \O \oplus \O(-1)^{15} \oplus \O(-2)^{15} \oplus \O(-3), \\
+    F_*^2\O_{\P^5} &= \O \oplus \O(-1)^{120} \oplus \O(-2)^{546} \oplus \O(-3)^{336} \oplus \O(-4)^{21}.
   \end{align*}
@@
   \[ S = \bigoplus_{[D]\in\Pic{X}} \; \Gamma(X, \O(D)). \]
-  Similar to the case of the projective space, B{\o}gvad and Thomsen
-\cite{Bogvad98,Thomsen00}
- showed that $F_*L$ totally splits as a direct sum of line bundles for any line bundle $L\in\Pic X$.
+  Similar to the case of the projective space, B{\o}gvad and Thomsen \cite{Bogvad98,Thomsen00} showed that $F_*L$ totally splits as a direct sum of line bundles for any line bundle $L\in\Pic X$.
 
@@
   \begin{align*}
-F_*\O_X &= \O_X \oplus \O_X(-1,0)^2 \oplus \O_X(0,-1)^2 \oplus \O_X(1,-1)^3 \oplus \O_X(2,-1)\\
-F_*\O_X(1,1) &= \O_X^3 \oplus \O_X(-1,0) \oplus \O_X(1,-1) \oplus \O_X(1,0)^2 \oplus \O_X(2,-1)^2
+    F_*\O_X      &= \O_X   \oplus \O_X(-1,0)^2 \oplus \O_X(0,-1)^2 \oplus \O_X(1,-1)^3 \oplus \O_X(2,-1), \\
+    F_*\O_X(1,1) &= \O_X^3 \oplus \O_X(-1,0)   \oplus \O_X(1,-1)   \oplus \O_X(1, 0)^2 \oplus \O_X(2,-1)^2.
   \end{align*}
-  In fact, Achinger \cite{Achinger13} showed that the total splitting of $F_*L$ for every line bundle $L$ characterizes smooth projective toric varieties.
-
+  In fact, Achinger \cite{Achinger15} showed that the total splitting of $F_*L$ for every line bundle $L$ characterizes smooth projective toric varieties.
   % The toric Frobenius is simply multiplication by $p$ on the torus.
@@
 \begin{exa}[Frobenius pushforward on elliptic curves]\label{algclos}
-  Consider the elliptic curve $X=\Proj (\mathbb F_7[x,y,z]/(x^3+y^3+z^3))$.
+  Consider the elliptic curve
+  \[ X = \Proj \mathbb F_7[x,y,z]/(x^3+y^3+z^3). \]
   %%Weierstrauss form is y^2=x^3-432
@@
   \[ F_*\O_X = \O \oplus \O(-1)^{44} \oplus \O(-2)^{20} \oplus A^4 \oplus B^4, \]
-  where $A$ and $B$ are rank-2 indecomposable bundles (c.f.~\cite{RSB22}).
+  where $A$ and $B$ are rank-2 indecomposable bundles (c.f.~\cite{RSVdB22}).
 \end{exa}
@@
 \begin{exa}[Frobenius pushforward on Mori Dream Spaces]
-  Continuing with the theme of computations over the Cox ring, the natural geometric setting is to consider the class of projective varieties known as Mori dream spaces \cite{HuKeel00}.
+  Continuing with the theme of computations over the Cox ring, the natural geometric setting is to consider the class of projective varieties known as Mori dream spaces \cite{HK00}.
 

### 2024-07-28 beamer/JMM/beamer.tex
moved JMM beamer files

--- 2024-01-29 beamer/beamer.tex
@@
   %\pause
-  - Hartshorne: any $L\in\Pic\PPn$, $F_*L$ splits as a sum of line bundles.
+  - Hartshorne: any $L\in\Pic\PPn$, $F_*L$ splits as a sum of line bundles. \\
+  %\pause
+  - Raczka '24: any v.b.\ on $\PPn$, $F_*^e\cE$ splits as a sum of $\OO$'s and $\Omega$'s.
   %\pause
@@
   %\pause
-  \begin{block}{Example: $X = $ Hirzebruch surface of type 3, $F_*\o_X, p = 2$}
+  \begin{block}{Example: $X = $ Hirzebruch surface of type 3, $F_*\o_X, p = 3$}
     \vspace*{-0.2in}
@@
     \vspace*{-0.15in}
-    h\begin{align*}
+    \begin{align*}
       F_*^2\o_X = {\o_{X}^{1}}
@@
     \begin{enumerate}
-    \item compute $\Hom(M, M)$; %\pause
-    \item find element $h$ corresponding to an idempotent; %\pause
-    \item find the image and cokernel of $h$; %\pause
-    \item repeat.
+    \item Compute $\Hom(M, M)$; \\ %\pause
+      Note: the degree zero part only, so \textbf{multigrading helps!} %\pause
+    \item Find element $h$ corresponding to an idempotent; \\ %\pause
+      \begin{itemize}
+      \item Hope that generators of $\Hom$ are idempotents; %\pause
+      \item In finite characteristic, use a general endomorphism. %\pause
+      \end{itemize}
+    \item Find the image and cokernel of $h$;
+    \item Repeat.
     \end{enumerate}
   \end{alertblock}
+
+  \vfill
+  %\pause
+  \begin{lemma}
+    Let $M$ be an $R$-module, and let $A\in\End_R(M)$. If the induced action of $A$ on $M/mM$ is idempotent, then $M$ admits a direct sum decomposition $N\oplus M/N$, where $N\cong \im A$.
+    % Let $A$ be an endomorphism of a $k$-vector space such that all eigenvalues of $A$ are contained in $k$. If $\lambda$ is an eigenvalue of $A$, then some power of $A-\lambda I$ is an idempotent.
+  \end{lemma}
+
 \end{frame}

### 2024-07-30 beamer/MEGA/beamer.tex
added MEGA24 slides

\documentclass[beamer]{paper}

\usepackage{quiver}
\usepackage[italic,symbolgreek]{mathastext}

% for highlighting
\usepackage{xcolor, soul}
%\makeatletter
%\let\HL\hl
%\renewcommand\hl{%
%  \let\set@color\beamerorig@set@color
%  \let\reset@color\beamerorig@reset@color
%  \HL}
%\makeatother

%\usetheme{Warsaw}
\beamertemplatenavigationsymbolsempty
%\setbeamertemplate{footline}[frame number]
%\setbeamercovered{transparent} % gray out hidden stuff
\AtBeginSection{\frame{\sectionpage}}

% Inverted colors
\definecolor{rossred}  {rgb}{1.0, 0.25,0.66}
\definecolor{rossgreen}{rgb}{0.25,0.66,0.25}
\definecolor{rossblue} {rgb}{0.25,0.66,1.0}
\renewcommand{\red}[1]{\color{rossred}  #1\color{white}}
\newcommand{\green}[1]{\color{rossgreen}#1\color{white}}
\newcommand{\blue} [1]{\color{rossblue} #1\color{white}}

\setbeamercolor{title}      {fg=white,     bg=black}
\setbeamercolor{frametitle} {fg=rossgreen, bg=black}
\setbeamercolor{normal text}{fg=white,     bg=black}
\setbeamercolor{structure}  {fg=cyan!90!black}
\setbeamercolor{block title}{fg=black,     bg=rossblue}

%%%%%%%%%%%

\newcommand{\cE}{\mathcal{E}} % bundle
\newcommand{\cH}{\mathcal{H}} % Hirzebruch surface
\newcommand{\cL}{\mathcal{L}} % line bundle

\newcommand{\Om}{\Omega}
\newcommand{\om}{\omega}

%%%%%%%%%%%

\begin{document}

% Use Computer Modern with Serif Roman Font Family
\rmfamily
\setbeamerfont{frametitle}{family=\rmfamily\selectfont}

\title[Computing Direct Sum Decompositions]
      {\textsc{Computing Direct Sum Decompositions}} % \\ over Multigraded Cox Rings

% I acknowledge that the University of Minnesota stands on Miní Sóta Makhóčhe, the homelands of the Dakhóta Oyáte.

% Abstract: It is known that the Frobenius pushforward of the structure sheaf on toric varieties splits as a sum of line bundles, and high enough pushforwards of it generate the derived category. Working over the multigraded Cox ring of a Mori Dream Space, we discuss algorithms for determining the summands of the Frobenius pushforward of the structure sheaf.

\newcommand{\lc}{\MakeLowercase}
\author[Mahrud Sayrafi]{M\lc{ahrud} S\lc{ayrafi} \\
  {\footnotesize MPI MiS}}% \\[1em]
%  {\footnotesize Joint work with Devlin Mallory \\
%    Basque Center for Applied Mathematics}}

%\institute[UMN]{U\lc{niversity} of M\lc{innesota}, T\lc{win} C\lc{ities}}

% https://www.jointmathematicsmeetings.org/meetings/national/jmm2024/2300_progfull.html
\date{{\footnotesize Effective Methods in Algebraic Geometry \\ July 30, 2024}}

\frame[plain]{\titlepage}

\begin{frame}[c,plain]
  Let's start with linear algebra. \\[1em]

  \pause
  \scalebox{0.53}{
    $
    \left(\!\begin{array}{ccccc}
         {.37}&{.28}&{.23}&{.26}&{.36}\\
         {.87}&{.76}&{.11}&{.37}&{.86}\\
         {.52}&{.82}&{.76}&{.26}&{.12}\\
         {.91}&{.91}&{.79}&{.82}&{.48}\\
         {.44}&{.13}&{.28}&{.37}&{.84}
    \end{array}\!\right) % \pause
    =
    %\underbrace{
    \left(\!\begin{array}{ccccc}
    -{.25}&{.03}+{.05}\,\mathbf{i}&{.03}-{.05}\,\mathbf{i}&{.76}&-{.04}\\
    -{.46}&{.30}-{.48}\,\mathbf{i}&{.30}+{.48}\,\mathbf{i}&-{.48}&-{.03}\\
    -{.41}&-{.57}&-{.57}&{.15}&{.44}\\
    -{.67}&-{.35}+{.3}\,\mathbf{i}&-{.35}-{.3}\,\mathbf{i}&-{.33}&-{.83}\\
    -{.32}&{.33}+{.19}\,\mathbf{i}&{.33}-{.19}\,\mathbf{i}&-{.23}&{.35}
    \end{array}\!\right)
    %}_{P}
    \cdot
    \left(\!\begin{array}{ccccc}
         {2.5}&{0}&{0}&{0}&{0}\\
         {0}&{.39}+{.47}\,\mathbf{i}&{0}&{0}&{0}\\
         {0}&{0}&{.39}-{.47}\,\mathbf{i}&{0}&{0}\\
         {0}&{0}&{0}&{.02}&{0}\\
         {0}&{0}&{0}&{0}&{.27}
    \end{array}\!\right)
    \cdot
    P^{-1}
    $
  }

  \vfill
  \pause
  \begin{enumerate}
  \item We know conditions for diagonalizability, and
  \item We have algorithms for diagonalization.
  \end{enumerate}
\end{frame}

\begin{frame}[c,plain]
  \textbf{Nonlinear Algebra:} what if we worked over a polynomial ring? \pause \vfill

  \scalebox{0.36}{
    \hspace*{-0.7in}
    $A = \left(\!\begin{array}{ccccccccccccccc}
    0 & 0 & 0 & 0 & 0 & x_1y_0y_1 & 0 & 0 & x_1y_0y_2 & 0 & 0 & 0 & x_0y_1y_2 & 0 & 0\\
    -x_0y_1+x_1y_1-x_1y_2 & -x_0y_2 & 0 & 0 & 0 & x_1y_2 & x_1y_2 & x_1y_2 & -x_1y_2 & -x_0y_1+x_1y_1-x_1y_2 & -x_1y_2 & 0 & x_1y_2 & x_1y_2 & x_1y_2\\
    0 & 0 & 0 & x_1y_1 & 0 & 0 & 0 & 0 & -x_1y_2 & 0 & -x_1y_2 & 0 & -x_0y_1-x_1y_2 & -x_0y_1 & -x_1y_2\\
    0 & 0 & -x_1y_2 & 0 & x_0y_0-x_1y_0-x_0y_2 & x_1y_2 & x_1y_2 & -x_1y_2 & -x_1y_2 & 0 & x_0y_2-x_1y_2 & x_0y_2 & -x_0y_2+x_1y_2 & x_0y_2+x_1y_2 & 0\\
    0 & 0 & 0 & 0 & 0 & -x_1y_2 & -x_1y_2 & x_1y_2 & x_1y_2 & x_0y_1-x_1y_1+x_1y_2 & x_1y_2 & 0 & -x_1y_2 & -x_1y_2 & -x_1y_2\\
    0 & 0 & x_1y_1 & 0 & x_1y_0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & x_0y_1 & x_1y_2\\
    0 & 0 & 0 & 0 & 0 & -x_1y_0 & 0 & -x_1y_0 & -x_1y_0 & 0 & x_0y_2 & x_0y_0 & 0 & 0 & 0\\
    x_0y_1 & x_0y_0 & 0 & 0 & 0 & 0 & 0 & x_1y_0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
    0 & 0 & 0 & -x_1 & 0 & -x_1 & -x_1 & -x_1 & -x_1 & 0 & x_0-x_1 & x_0 & x_0 & -x_0-x_1 & 0\\
    y_1y_2 & y_0y_2 & 0 & 0 & 0 & -y_0y_1+y_1y_2 & -y_0y_1+y_1y_2 & 0 & 0 & -y_1y_2 & 0 & 0 & -y_1y_2 & y_1y_2 & 0\\
    0 & 0 & y_1y_2 & 0 & y_0y_2 & 0 & y_0y_1 & 0 & y_0y_2 & -y_1y_2 & y_0y_2 & -y_0y_2 & -y_0y_2 & -y_0y_2 & -y_0y_2\\
    -y_1 & -y_0 & 0 & 0 & 0 & 0 & 0 & 0 & y_0-y_2 & y_1 & y_0-y_2 & 0 & -y_0-y_2 & -y_0 & -y_0+y_2\\
    0 & 0 & 0 & -y_1 & 0 & -y_1 & y_1 & 0 & y_2 & 0 & 0 & 0 & y_1+y_2 & y_1 & -y_2\\
    0 & 0 & -y_1 & y_1 & -y_0 & y_1 & y_1 & 0 & y_2 & 0 & y_2 & 0 & -y_1+y_2 & y_1 & -y_2\\
    0 & 0 & 0 & 0 & 0 & -y_0+y_2 & -y_0+y_2 & y_0-y_2 & y_0-y_2 & -y_2 & -y_2 & -y_0 & y_2 & y_2 & 0\\
    0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & -y_1 & -y_0 & -y_0 & y_0 & y_0 & y_0-y_2
    \end{array}\!\right)$
  }
  \pause
  \vfill

  \scalebox{0.406}{
    \hspace*{-0.7in}
    $B = \left(\!\begin{array}{ccccccccccccccc}
    0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
    x_1y_2 & x_0y_1-x_1y_1 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
    -y_0+y_2 & y_0-y_1 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
    0 & 0 & x_0y_1 & x_1y_1-x_1y_2 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
    0 & 0 & y_0 & y_0-y_2 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
    0 & 0 & 0 & 0 & -x_0+x_1 & x_1 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
    0 & 0 & 0 & 0 & y_0y_2-y_1y_2 & y_0y_1-y_1y_2 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
    0 & 0 & 0 & 0 & 0 & 0 & x_1y_0 & x_0y_0-x_0y_2 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
    0 & 0 & 0 & 0 & 0 & 0 & -y_1 & -y_1+y_2 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
    0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & x_0y_1-x_1y_1+x_1y_2 & x_0y_2 & 0 & 0 & 0 & 0 & 0\\
    0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & -y_1 & -y_0 & 0 & 0 & 0 & 0 & 0\\
    0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & x_1y_2 & 0 & x_0y_2 & 0 & x_0y_0-x_1y_0-x_0y_2\\
    0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & x_1y_1 & 0 & x_0y_1 & 0 & -x_1y_0\\
    0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & x_1y_0 & -x_0y_2 & x_0y_0-x_0y_2 & 0\\
    0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & y_1y_2 & y_0y_1 & 0 & y_0y_1-y_0y_2 & -y_0y_2\\
    0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & y_1 & y_1 & y_1-y_2 & y_1-y_2 & -y_0
    \end{array}\!\right)$
  }
  \pause
  \vfill

  This means $\coker A \cong \coker B$ splits as a direct sum of 7 modules.
  \pause
  \vfill

  \begin{center}
    $A$ is block \textbf{\color{rossred}diagonalizable} $\iff$ $\coker A$ \textbf{\color{rossred}splits} as a direct sum.
  \end{center}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame} \centering \LARGE Conditions for Diagonalizability \end{frame}

{
\usebackgroundtemplate{\includegraphics[height=\paperheight]{moses-mohammed.png}}
\setbeamercolor{background canvas}{bg=black}
\begin{frame}[b]
  \scriptsize{\hspace{0.35in} Falnameh, 16th century \hfill Chludov Psalter, 9th century \hspace*{0.12in}}
\end{frame}
}

\begin{frame}[t]{Warm-up: Frobenius pushforward on $\PPn$}
  % Let $\kk$ be an algebraically closed field of char $p>0$.

  Let $S = \kk[x_0,\dots,x_n]$ be a polynomial ring with $\deg x_i = 1$.

  \vfill
  \pause
  Consider the Frobenius endomorphism
  \[ F\colon S\to S \quad \text{given by} \quad x_i \to x_i^p \]

  \pause
  - Hartshorne: any $L\in\Pic\PPn$, $F_*L$ splits as a sum of line bundles. \\
  \pause
  - Rączka '24: any v.b.\ on $\PPn$, $F_*^e\cE$ splits as a sum of $\OO$'s and $\Omega$'s.
  \pause
  \begin{block}{Example: $F_*\o_{\PPn}$}
    \vspace*{-0.2in}
    \begin{align*}
      \text{When $p=3$, $n=2$:} \\
      F_*\o_{\PP2}   = \o \;\oplus\; &\o(-1)^7  \;\;\oplus\; \o(-2). \\
      \text{When $p=2$, $n=5$:} \\
      F_*\o_{\PP5}   = \o \;\oplus\; &\o(-1)^{15} \;\;\oplus\; \o(-2)^{15} \,\;\;\oplus\; \o(-3). \\
      F_*^2\o_{\PP5} = \o \;\oplus\; &\o(-1)^{120}    \;\oplus\; \o(-2)^{546}  \;\oplus\; \o(-3)^{336} \;\oplus\; \o(-4)^{21}
    \end{align*}
  \end{block}
\end{frame}

\begin{frame}[t]{Warm-up: Frobenius pushforward on toric varieties}
  Let $X$ be a smooth toric variety and consider the Cox ring
  \[ S = \bigoplus\!{}_{[D]\in\Pic{X}} \; \Gamma(X, \OO(D)) \]
  This is a \textbf{\color{rossred}polynomial ring} with a {\color{rossred}$\mathbf{\Pic{X}-}$\textbf{grading}}.

  \pause
  \begin{theorem}[Thm.~3.2, \cite{[Cox95]}]
    Every coherent sheaf on a simplicial toric variety $X$ corresponds \\
    to a finitely generated {\color{rossred}$\Pic X$-graded} $S$-module.
  \end{theorem}

  \vfill
  \pause
  - Bogvad '98, Thomsen '00: $F_*L$ totally splits for any $L\in\Pic X$. \\
  \pause
  - Achinger '13: \\ \hfill splitting of $F_*L$ characterizes smooth projective toric varieties.

  \vfill
  \pause
  \begin{block}{Example: $X = $ Hirzebruch surface of type 3, $\;F_*\o_X, \;p = 3$}
    \vspace*{-0.2in}
    \[ F_*\o_X = \o_X \;\oplus\; \o_X(-1,0)^2 \;\oplus\; \o_X(0,-1)^2 \;\oplus\; \o_X(1,-1)^3 \;\oplus\; \o_X(2,-1). \]
  \end{block}

  %% \vfill
  %% \pause
  %% \begin{block}{Remark}
  %%   The toric Frobenius is simply multiplication by $p$ on the torus.
  %% \end{block}
  %% \pause
  %% \begin{alertblock}{Funky Remark}
  %%   The $p$ need not be prime, nor even match the characteristic!
  %% \end{alertblock}
\end{frame}

%% \begin{frame}[t]{Warm-up: splitting of vector bundles on $\PPn$}
%%   Let $S = \kk[x_0,\dots,x_n]$ be a polynomial ring with $\deg x_i = 1$.

%%   \pause
%%   Let $E$ be a vector bundle of rank $r$ on $\PPn = \Proj S$.

%%   \vfill
%%   \pause
%%   \textbf{Fact:} vector bundles on $\PPn$ $\iff$ projective modules over $S$.

%%   \vfill
%%   \pause
%%   \begin{block}{Classical question}
%%     \centering
%%     When does $E$ \emph{split} as a direct sum of line bundles?
%%   \end{block}

%%   \vfill
%%   \pause
%%   - Grothendieck: any vector bundle on $\PP1$ splits as $E = \bigoplus\o_{\PP1}(d_i)$. \\
%%   \pause
%%   - There are indecomposable bundles of rank $n-1$ on $\PPn$, $n\!\geq\!3$. \\
%%   \pause
%%   - Horrocks--Mumford: an indecomposable rank $2$ bundle on $\PP4$ from
%%   \[ \o_{\PP4}^5 \gets \Om_{\PP4}^2(2) \oplus \Om_{\PP4}^2(2) \gets \o_{\PP4}(-1)^5. \]
%%   \pause
%%   - Hartshorne's conjecture: there are no rank 2 bundles on $\PPn, n\geq7$.

%%   %\vfill
%%   %% \pause
%%   %Notation: \\
%%   %- $\o_X$ is the structure sheaf on $X$, $\o_X(2) = \o_X(1) \otimes \o_X(1)$, etc. \\
%%   %- $\Om_X$ is the cotangent sheaf on $X$, $\Om_X^2(2) = \Om_X(1) \wedge \Om_X(1)$, etc.
%%   % TODO: add example of two resolutions on \PP1
%% \end{frame}

\begin{frame}{Warm-up: a splitting criterion for vector bundles on $\PPn$}
  \begin{theorem}[Horrocks '64]
    If for all twists $d\in\ZZ$ and $i\geq 0$, $\HH^i(\PPn, E\otimes\o_{\PPn}(d))$ is
    equal to the cohomology of a direct sum of line bundles, then $E$ splits.
  \end{theorem}

  \vfill
  \pause
  \begin{center}
    \color{rossred}``If it walks like a duck and quacks like a duck, then it's a duck.''
  \end{center}

  %\pause
  %- Computable using \emph{BGG} and \emph{Tate resolutions} over exterior algebras. \\
  %- This is a finite condition: construct the \emph{Beilinson monad} for $E$:
  %\[ B_i = \bigoplus_{j\in\ZZ} \HH^{j-i}\left(\PPn, E\otimes\o_{\PPn}(-j)\right) \otimes \Om_{\PPn}^j(j). \]

  \vfill
  \pause
  - Dolgachev '82: Weighted projective spaces \\
  - Ottaviani '89: $\mathrm{Gr}(k,n)$ and quadric hypersurfaces in $\PPn, n\geq4$ \\
  - Fulger--Marchitan '11: rank 2 bundles on Hirzebruch surfaces \\
  %- Products of projective spaces: Eisenbud--Erman--Schreyer ('15)

  \vfill
  \pause
  \begin{block}{First Goal}
    \centering
    Prove a similar splitting criterion over normal toric varieties, \pause \\
    i.e., a condition for diagonalizability of graded matrices.
  \end{block}
\end{frame}

%% \begin{frame}[t]{Warm-up: Frobenius pushforward on toric varieties}
%%   Let $X$ be a smooth toric variety and consider the Cox ring
%%   \[ S = \bigoplus\!{}_{[D]\in\Pic{X}} \; \Gamma(X, \OO(D)) \]

%%   \pause
%%   \begin{theorem}[Thm.~3.2, \cite{[Cox95]}]
%%     Every coherent sheaf on a simplicial toric variety $X$ corresponds \\
%%     to a finitely generated $\Pic X$-graded $S$-module.
%%   \end{theorem}

%%   \vfill
%%   \pause
%%   Fact: for any toric variety, $S$ is a \textbf{multigraded} polynomial ring.
%%   \pause
%%   Upshot: commutative algebra over a toric variety is easy! \\
%%   \pause
%%   \hfill i.e. we have monomial orderings, Gr\"obner bases, syzygies, etc.
%% \end{frame}

\begin{frame}[t]{What makes a splitting criterion strong?}
  When $X = \PP{n_1}\times\PP{n_2}\times\cdots\times\PP{n_t}$:
  \begin{theorem}[Eisenbud--Erman--Schreyer '15]
    Let $E$ be a vector bundle on $X$ whose sheaf cohomology matches that of
    \( \sum_{i=1}^r \OO(\underline{d}_i)^{e_i}, \) for $\underline{d}_i\in\ZZ^r$.
    {\color{rossred}If $\underline{d}_r \le \cdots \le \underline{d}_1$}, then $E$ splits.
  \end{theorem}
  \pause
  - Holds in arbitrary dimension and Picard rank. \\
  - But {\color{rossred}only when $\Nef X = \Eff X = $ the positive orthant}.

  \vfill
  \pause
  When $X = \PP{}(\o\oplus\o(a_1)\oplus\cdots\oplus\o(a_s)) \to \PP{t}$:
  \pause
  \begin{theorem}[Brown--S. '22]
    Let $E$ be a vector bundle on $X$ whose sheaf cohomology matches that of
    \( \sum_{i=1}^r \OO(b_i, c_i)^{e_i}. \)
    {\color{rossred}If $(b_r, c_r) \le \cdots \le(b_1, c_1)$}, then $E$ splits.
  \end{theorem}
  \pause
  - Holds in arbitrary dimension, but {\color{rossred}only Picard rank = 2}. \\
  - Now $\Nef X = $ the positive quadrant, but {\color{rossgreen}$\Eff X$ can be larger}.
\end{frame}

\begin{frame}[t]{Ingredient list for other toric varieties}
  When $X$ is an arbitrary smooth projective toric variety:
  \begin{theorem}[S. '24]
    Let $E$ be a vector bundle on $X$ whose sheaf cohomology matches that of
    \( \sum_{1}^r \OO(D_i)^{e_i}\!, \) $D_i\in\Pic X$\!.
    {\color{rossred}If $D_{i+1}\!-\!D_i$ is ample}, then $E$ splits.
  \end{theorem}
  \pause
  - Holds in {\color{rossgreen}arbitrary} dimension and Picard rank. \\
  - Holds for {\color{rossgreen}arbitrary} $\Nef X$ and $\Eff X$. \\ \pause
  - But: the additional hypothesis on $E$ is stronger \pause ... probably!

  \vfill
  \pause
  \textbf{\color{rossred} Upshot:} if $\coker A$ satisfies the conditions above, $A$ diagonalizes.

  \vfill
  \pause
  \begin{exampleblock}{}
    \textbf{Key ingredient:} \\
    \centering
    a resolution of the diagonal for $X$, with special properties %such that we can prove \\
    % \hl{for all $D$ ample, if $E = \o(-D)$ the ``page'' is ``upper triangular''.}
  \end{exampleblock}

  \pause
  \textbf{Candidate:} Hanlon--Hicks--Lazarev's resolution of the diagonal.%\\ \pause
  % \textbf{New tool:} vanishing theorems on line bundles and $\QQ$-divisors.
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
  \centering
  \LARGE Algorithm for Diagonalizability \\[1em]
  \footnotesize Joint work with Devlin Mallory \\
  Basque Center for Applied Mathematics
\end{frame}

{
\usebackgroundtemplate{\includegraphics[height=\paperheight]{odysseymonkey}}
\setbeamercolor{background canvas}{bg=black}
\begin{frame}
\end{frame}
}

\begin{frame}{Frobenius pushforward on Grassmannians}
  We can still consider the Cox ring
  \[ S = \bigoplus\!{}_{[D]\in\Pic{X}} \; \Gamma(X, \OO(D)) \]
  For Grassmannians this coincides with the coordinate ring.

  \pause
  \begin{block}{Example: $X = \mathrm{Gr(1,3)}, \;F_*\o_X, \;\char \kk = p = 3$}
    Working over the ring:
    \[ S = \frac{\kk[P_{01},P_{02},P_{12},P_{03},P_{13},P_{23}]}{P_{12}P_{03}-P_{02}P_{13}+P_{01}P_{23}} \]
    When $p = 3$:
    \[ F_*\o_X = \o \oplus \o(-1)^{44} \oplus \o(-2)^{20} \oplus A^4 \oplus B^4, \]
    where $A$ and $B$ are rank 2 indecomposable bundles. \\
    \hfill (c.f. [Raedschelders--Špenko--Van den Bergh, '17])
  \end{block}
\end{frame}

\begin{frame}{Frobenius pushforward on Mori Dream Spaces}
  This is the natural setting to work over the Cox ring
  \[ S = \bigoplus\!{}_{[D]\in\Pic{X}} \; \Gamma(X, \OO(D)). \]
  Hu--Keel: $X$ is an $MDS$ $\iff$ the Cox ring is finitely generated.

  \vfill
  \pause
  \begin{exampleblock}{Examples}
  \; - Projective toric varieties \\
  \; - Grassmannians and flag varieties \\
  \; - Determinantal varieties \\
  \; - Many complete intersection rings \\
  \; - Smooth Fano varieties over $\CC$ \\
  \; - $\overline{M}_{0,n}$ for $n\leq 6$ (7,8,9 are unknown)
  \end{exampleblock}
\end{frame}

\begin{frame}{Frobenius pushforward on Mori Dream Spaces}
  \begin{block}{Example: $X = \mathrm{Bl}_4\PP2, \; F_*^2\o_X, \; \char \kk = p = 2$}
    Working over the $\ZZ^5$-graded Cox ring\vspace*{-0.15in}
    \[ S = \kk[x_1,\dots,x_{10}]/\text{(five quadric Pl\"ucker relations)} \]
    \pause
    With degrees \vspace*{-0.15in}
    \[
    \left(\!\begin{array}{cccccccccc}
      0&0&0&0&1&1&1&1&1&1 \\
      1&0&0&0&-1&-1&-1&0&0&0 \\
      0&1&0&0&-1&0&0&-1&-1&0 \\
      0&0&1&0&0&-1&0&-1&0&-1 \\
      0&0&0&1&0&0&-1&0&-1&-1
    \end{array}\!\right)
    \]
    In the basis:
    \[ \Pic X \cong \langle H \rangle \oplus \langle E_1 \rangle \oplus \langle E_2 \rangle \oplus \langle E_3 \rangle \oplus \langle E_4 \rangle. \]
  \end{block}
\end{frame}

\begin{frame}{Frobenius pushforward on Mori Dream Spaces}
  \begin{block}{Example: $X = \mathrm{Bl}_4\PP2, \; F_*^2\o_X, \; \char \kk = p = 2$}
    \vspace*{-0.15in}
    \begin{align*}
      F_*^2\o_X = {\o_{X}^{1}}
      &\oplus {\o_{X}^{2}\ \left(-1,\,1,\,0,\,0,\,0\right)} \oplus {\o_{X}^{2}\ \left(-1,\,0,\,1,\,0,\,0\right)} \\ 
      &\oplus {\o_{X}^{2}\ \left(-1,\,0,\,0,\,1,\,0\right)} \oplus {\o_{X}^{2}\ \left(-1,\,0,\,0,\,0,\,1\right)} \\
      &\oplus {\o_{X}^{2}\ \left(-2,\,1,\,1,\,1,\,1\right)} \oplus B \oplus G,
    \end{align*}
    where $B, G$ are rank 3 and rank 2 indecomposables. \\
    \hfill (c.f. [Hara, '15])
  \end{block}
\end{frame}

{
\usebackgroundtemplate{\includegraphics[width=\paperwidth]{higgs}}
\setbeamercolor{background canvas}{bg=black}
\begin{frame}
\end{frame}
}

\begin{frame}{The splitting algorithm}
  \begin{alertblock}{The ``Meataxe'' algorithm for graded $S$-modules:}
    Input: finitely generated graded module $M$ over an algebra.
    \pause
    \begin{enumerate}
    \item Compute $\Hom(M, M)_0$ \\ \pause
      Note: the degree zero part only, so \textbf{\color{rossgreen}multigrading helps!} \pause
    \item Find element $h$ corresponding to an idempotent \\ \pause
      \begin{itemize}
      \item Often a generator of $\Hom$ is already an idempotents! \pause
      \item In characteristic $p > 0$, we find a general endomorphism $A$
        and prove that $(A-\lambda I)^\ell$ is idempotent for $\ell = kp(p^e-1)$. \pause
      \end{itemize}
    \item Find the image and cokernel of $h$;
    \item Repeat.
    \end{enumerate}
  \end{alertblock}

  %% \vfill
  %% \pause
  %% \begin{lemma}
  %%   Let $M$ be an $R$-module, and let $A\in\End_R(M)$. If the induced action of $A$ on $M/mM$ is idempotent, then $M$ admits a direct sum decomposition $N\oplus M/N$, where $N\cong \im A$.
  %%   % Let $A$ be an endomorphism of a $k$-vector space such that all eigenvalues of $A$ are contained in $k$. If $\lambda$ is an eigenvalue of $A$, then some power of $A-\lambda I$ is an idempotent.
  %% \end{lemma}
\end{frame}

\begin{frame}{Other Applications}
  Even in absence of a multigrading, explicit computations of direct
  sum decompositions is useful for mathematical experiments: \\[1em]
  \; - Finite Cohen--Macaulay-type property of singularities; \\
  \; - $F$-representation type property of cubic surfaces; \\
  \; - Structure of syzygies over Artinian rings. \\

  \vfill
  \pause
  \begin{block}{Question for the road:}
    \centering
    What are applications of diagonalization of polynomial \\
    matrices in algebraic geometry and nonlinear algebra?
  \end{block}
\end{frame}

%% \begin{frame}[t]{Frobenius pushforward on determinantal varieties}
%%   \textbf{Fact:} a determinantal variety $X$ is a Mori dream space. \\
%%   \hfill (i.e. the Cox ring is a finitely generated algebra.)

%%   \vfill
%%   Singh: does $F^e_*\OO_X$ have finitely many classes of summands? \\
%%   This is known as having \emph{globally finite $F$-representation type}.

%%   \begin{block}{Problem}
%%     Classify MDS with globally finite $F$-representation type.
%%   \end{block}

%%   \begin{alertblock}{Upshot}
%%     Determining line bundle summands is easy over the Cox ring.
%%   \end{alertblock}

%%   \vfill
%% \end{frame}

\begin{frame}[t]{References}
  \scriptsize{
    - Hartshorne '70, \hfill\textit{Ample subvarieties of algebraic varieties} \\[1.5em]
    - Bogvad '98, \hfill\textit{Splitting of the direct image of sheaves under the Frobenius} \\[1.5em]
    - Thomsen '00, \hfill\textit{Frobenius direct images of line bundles on toric varieties} \\[1.5em]
    - Achinger '13, \hfill\textit{A characterization of toric varieties in characteristic $p$} \\[1.5em]
    - Hara '15, \hfill\textit{Looking out for Frobenius summands on a blown-up surface of $\PP2$} \\[1.5em]
    - Raedschelders, Špenko, Van den Bergh '17, \\ \hfill\textit{The Frobenius morphism in invariant theory} \\[0.5em]
    - Rączka '24, \hfill\textit{Frobenius pushforwards of vector bundles on projective spaces}
  }

  \renewcommand{\section}{\subsection}
  \printbibliography
\end{frame}

\end{document}

% Local Variables:
% tab-width: 2
% eval: (visual-line-mode)
% End:

### 2024-11-15 paper/main.tex
version of paper submitted to JSC

@@
-s\documentclass{article}
+\documentclass[12pt]{article}
 
@@
 \renewtheorem{thm}{Theorem}
+\renewtheorem{alg}[thm]{Algorithm}
 \renewtheorem{lem}[thm]{Lemma}
@@
 
-% Regular papers, aimed to contributed talks:
-% for these, authors can choose to submit either an extended abstract of at least six pages (not counting bibliography) that should contain full results, an outline of the proofs and all of the main ideas; or else a full paper with complete proofs.
-
-% Abstract
-% The problems of finding isomorphism classes of indecomposable modules with certain properties, or determining the indecomposable summands of a module, are ubiquitous in commutative algebra, group theory, representation theory, and other fields.
-The purpose of this paper is to describe and prove correctness of a practical algorithm for computing indecomposable summands of finitely generated modules over a finitely generated $k$-algebra, for $k$ a field of positive characteristic.
+\begin{abstract}
+  The problems of finding isomorphism classes of indecomposable modules with certain properties, or determining the indecomposable summands of a module, are ubiquitous in commutative algebra, group theory, representation theory, and other fields.
+The purpose of this work is to describe and prove correctness of a practical algorithm for computing indecomposable summands of finitely generated modules over a finitely generated $k$-algebra, for $k$ a field of positive characteristic.
 Our algorithm works over multigraded rings, which enables the computation of indecomposable summands of coherent sheaves on subvarieties of toric varieties (in particular, for varieties embedded in projective space).
 We also present multiple examples, including some which present previously unknown phenomena regarding the behavior of summands of Frobenius pushforwards and syzygies over Artinian rings.
+\end{abstract}
 
@@
 
-%% The introduction of the paper must explicitly address the following questions in succinct and informal manner:
-%% – What is the problem?
-%% – Why is the problem important?
-%% – What has been done so far on the problem?
-%% – What is the main contribution of the paper on the problem?
-%% – Why is the contribution original? (see below for clarification)
-%% – Why is the contribution non-trivial?
-%% – How does the journal paper differ from the conference paper (if relevant)?
-
 The problems of finding isomorphism classes of indecomposable modules with certain properties, or determining the indecomposable summands of a module, are ubiquitous in commutative algebra, group theory, representation theory, and other fields.
-Within commutative algebra, for instance, the classification of Cohen--Macaulay local rings $R$ for which there are only finitely many indecomposable maximal Cohen--Macaulay $R$-modules (the \emph{finite CM-type} property), or determining whether iterated Frobenius pushforwards of a Noetherian ring in positive characteristic have finitely many isomorphism classes of indecomposable summands (the \emph{finite F-representation type} property) are two well-established research problems.
-For both these problems, and many others, making and testing conjectures depends on finding summands of modules and verifying their indecomposability.
+Within commutative algebra, for instance, the classification of rings $R$ for which there are only finitely many isomorphism classes of indecomposable maximal Cohen--Macaulay $R$-modules (the \emph{finite CM-type} property), or determining whether iterated Frobenius pushforwards of a Noetherian ring in positive characteristic have finitely many isomorphism classes of indecomposable summands (the \emph{finite F-representation type} property) are two well-established research problems.
+For both these problems, and many others, making and testing conjectures depends on computing summands of modules and verifying their indecomposability.
 
 Currently there are no efficient algorithms available for checking indecomposability or finding summands of modules over commutative rings.
-In contrast, variants of the ``Meat-Axe'' algorithm for determining irreducibility of finite dimensional modules over a group algebra have wide ranging applications in computational group theory \cite{Parker84,HR94,Holt98} and are available through symbolic algebra software such as Magma and GAP \cite{MAGMA,GAP}.
+In contrast, variants of the ``Meat-Axe'' algorithm for determining irreducibility of finite-dimensional modules over a group algebra have wide ranging applications in computational group theory \cite{Parker84,HR94,Holt98} and are available through symbolic algebra software such as Magma and GAP \cite{MAGMA,GAP}.
 
-The purpose of this paper is to describe and prove correctness of a practical algorithm for computing indecomposable summands of finitely generated modules over a finitely generated $k$-algebra, for $k$ a field of positive characteristic.
-In particular, our algorithm works over multigraded rings, which enables the computation of indecomposable summands of coherent sheaves on subvarieties of toric varieties (in particular, for varieties embedded in projective space).
-After describing the algorithm and proving its correctness, we present multiple examples in the end, including some which present previously unknown phenomena regarding the behavior of summands of Frobenius pushforwards and syzygies over Artinian rings.
+The purpose of this paper is to describe and prove correctness of a practical algorithm for computing indecomposable summands of finitely generated modules over a finitely generated local or graded $k$-algebra, for $k$ a field of positive characteristic.
+We also describe applications of our algorithm for computations in characteristic 0.
+Further, our algorithm works over multigraded rings, which enables the computation of indecomposable summands of coherent sheaves on subvarieties of toric varieties (in particular, for varieties embedded in projective space).
 
-An accompanying implementation in Macaulay2 is available online via the GitHub repository 
+After describing the algorithm and proving its correctness, we present multiple examples, including some which present previously unknown phenomena regarding the behavior of summands of Frobenius pushforwards and syzygies over Artinian rings.
+In particular, we highlight the results of \cite{CDE+24}, which shows a recurrence formula for indecomposable summands of high syzygies of the residue field of Golod rings, made possible through experiments and observations using our algorithm.
+
+An accompanying implementation in Macaulay2 is available online via the GitHub repository \\
 \centerline{
@@
 
-  We also point out that while the discussion below, and our implementation in Macaulay2, concerns the case where $R$ is a \emph{commutative} $k$-algebra, many of the techniques extend beyond this case; we plan to extend the results and algorithms to noncommutative rings such as Weyl algebras in future work.
+  We note that while the discussion below, and our implementation in Macaulay2, concerns the case where $R$ is a \emph{commutative} ring, the basic techniques extend beyond this case.
+We plan to extend the results and algorithms to non-commutative rings, such as Weyl algebras, in future work.
 \end{rem}
 
-\section{The main algorithm}
+\subsection*{Acknowledgements}
+The authors would like to thank David Eisenbud for several very useful conversations and ideas.
+We also thank the American Institute of Mathematics for hosting the ``Macaulay2: expanded functionality and improved efficiency'' workshop, during which some of this work took place.
+The work of the first author was supported in by the National Science Foundation under Grant No.~1840190.
+The second author was supported in part by the Doctoral Dissertation Fellowship at the University of Minnesota and the National Science Foundation under Grant No.~2001101.
 
-Throughout, $R$ will be an $\Z^r$-graded ring with $R_0 = k$ a field of positive characteristic and homogeneous maximal ideal $m$.
-Note that if $M$ is a finitely generated $R$-module, and $A\in \End_R(M)$, then $A$ acts on the $k$-vector space $M/mM$.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-We begin with the observation that if $A\in\End_R(M)$ is an idempotent, then $M$ decomposes as $\im A \oplus \ker A$; if $A$ is nontrivial (i.e., neither an isomorphism nor the zero morphism), then $M$ is decomposable.
-The following lemma allows us to check only for idempotents modulo the maximal ideal:
+\section{The main algorithm}\label{sec:main-alg}
+
+Throughout this section, $(R,m)$ will be either local with maximal ideal $m$, or a (multi)graded ring with $R_0 = k \subset \overline{\mathbb F_p}$ a field of positive characteristic and homogeneous maximal ideal $m$.
+
+We begin with the observation that if $M$ is a finitely generated $R$-module and $A\in\End_R(M)$ is an idempotent, then $M$ decomposes as $\im A \oplus \coker A$.
+If $A$ is neither an isomorphism nor the zero morphism, then both factors are nonzero and $M$ is decomposable.
+Note that $A$ also acts on the $k$-vector space $M/mM$.
+
+The following lemma allows us to check only for idempotents modulo the maximal ideal.
 
 \begin{lem}\label{lem:idemp}
-  Let $M$ be an $R$-module, and let $A\in\End_R(M)$. If the induced action of $A$ on $M/mM$ is idempotent, then $M$ admits a direct sum decomposition $N\oplus M/N$, where $N\cong \im A$.
+  Let $M$ be a finitely generated $R$-module, and let $A\in\End_R(M)$. If the induced action of $A$ on $M/mM$ is idempotent, then $M$ admits a direct sum decomposition $\im A \oplus \coker A$.
 \end{lem}
@@
 
-  Let $N = \im A$.
-Consider the composition $N\inc M \xra{A} \im(A) = N$.
+  Let $N = \im A$. We want to show that $0 \to N\inc M$ splits. Consider the composition
+  \[ N\inc M \xra{A} \im A = N.\]
   We claim that this composition is surjective.
-We may complete at the maximal ideal to check this, and thus assume $R$ and $M$ are complete.
+Since a surjective endomorphism of finitely generated modules is invertible, we would then conclude that this composition is an isomorphism on $N$; say $\a$.
+  Therefore the inclusion \( 0 \to N \inc M \) is split by the map of $R$-modules $M\xra{A} N\xra{\a\inv} N$, and thus $M$ decomposes as claimed.
+
+  To check the surjectivity of $N\inc M \xra{A} N$,  we may complete at the maximal ideal and thus assume $R$, $M$, and $N$ are complete.
 Let $n_0\in N$.
@@
   \[ m^n M\cap N = m^{n-k} ( m^kM\cap N)\subset m^{n-k} N. \]
-  That is, the terms of $n_0-n_1+n_2-\dots$ are going to 0 in the $m$-adic topology on $N$. Thus, we can write
-  \[ n_0=A(n_0-n_1+n_2-\dots), \]
-  with $n_0-n_1+n_2-\dots\in N$. Thus, we have that $A$ is surjective as a map $N\to N$.
+  That is, the terms of $n_0-n_1+n_2-\cdots$ are going to 0 in the $m$-adic topology on $N$. Hence we can write
+  \[ n_0=A(n_0-n_1+n_2-\cdots), \]
+  with $n_0-n_1+n_2-\cdots\in N$. We conclude that $A$ is surjective as a map $N\to N$.
+\end{proof}
 
-  Since a surjective endomorphism of finitely generated modules is invertible, we have that this composition is an isomorphism on $N$; say $\a$.
-  Finally, we then have that
-  \[ 0 \inc N \inc M \]
-  is split by the map of $R$-modules $M\xra A N\xra{\a\inv} N$.
-\end{proof}
+Thus, if we produce an element $A\in \End_R(M)$ that is an idempotent modulo $m$, we obtain a splitting of $M$.
+The following lemma allows us to produce idempotents modulo $m$.
 
@@
   Let $k$ be a finite field of characteristic $p$, and let $A$ be an endomorphism of a $k$-vector space such that all eigenvalues of $A$ are contained in $k$.
-If $\l$ is an eigenvalue of $A$, then some power of $A-\l I$ is an idempotent.
-Furthermore, if $\l$ is not the only eigenvalue of $A$, then a power of $A-\l I$ that is nonzero is an idempotent.
+If $\l$ is an eigenvalue of $A$, then some power of $A-\l I$ is idempotent.
+Furthermore, if $\l$ is not the only eigenvalue of $A$, then a  nonzero power of $A-\l I$ is idempotent.
 \end{lem}
@@
   \end{pmatrix} \]
-  Set $\mu_i=\l_i-\l$. Then $(A-\l I)^l$ is block-diagonal with blocks of the form
+  Set $\mu_i=\l_i-\l$. Then for any $l\geq 1$, the $l$-th power $(A-\l I)^l$ is block-diagonal with blocks of the form
   \[ \begin{pmatrix}
@@
   \end{pmatrix} \]
-  If $l > r_i$ and $l$ divisible by $p$, then all non-diagonal terms will vanish, so all blocks will have the form
+  If we choose $l > r_i$ and $l$ divisible by $p$, then all non-diagonal terms will vanish, so all blocks will have the form
   \[ \begin{pmatrix}
@@
   \end{pmatrix}. \]
-  Finally, choosing $l$ to be divisible also by $p^e-1$ for some $e>0$, we have that $$\mu_i^{l-1} = (\mu_i^{p^{e}-1})^{l/(p^{e}-1)} =1,$$ since $\mu_i$ is contained in a finite field of characteristic $p$.
-  Thus, $(A-\l I)^l$ is simply a diagonal matrix with 1 or 0 on the diagonal, hence idempotent. Note moreover that if some $\l_i\neq \l$, then $(A-\l I)^l$ is not the zero matrix.
+  Finally, say that $k=\mathbb F_p$. Choosing $l$ to be divisible also by $p^e-1$, we have that $$\mu_i^{l-1} = (\mu_i^{p^{e}-1})^{l/(p^{e}-1)} =1.$$
+  Thus, $(A-\l I)^l$ is a diagonal matrix with diagonal entries 1 or 0, hence idempotent. Note moreover that if some $\l_i\neq \l$, then $(A-\l I)^l$ is not the zero matrix.
 \end{proof}
 
-This leads to a probabilistic algorithm to find the indecomposable summands of a module $M$, as follows:
+This leads to a probabilistic algorithm to find the indecomposable summands of a finitely generated $R$-module $M$ in our setting, as follows:
+
+\begin{alg} Given a finitely generated graded $R$-module $M$, the indecomposable summands of $M$ are computed as follows.
   \begin{enumerate}
-\item \label{1} Take a general element $A_0$ of $[\End M]_0$, the degree-0 part of $\End M$, and consider the resulting endomorphism $A$ of the $k$-vector space $M/mM$.
-\item Find the eigenvalues of $A$.
-\item If $A$ has at least two eigenvalues, choose one eigenvalue $\l$, and compute a sufficiently high power of $A$ (with the power explicitly as in the proof of Lemma~\ref{lem:jordan}).
-This power will be a nonzero idempotent, and thus produce a splitting of $M$ as $\im A_0 \oplus \coker A_0$ by Lemma~\ref{lem:idemp}.
-\item Repeat steps (1)--(3) for both $\im A_0$ and $\coker A_0$.
+  \item \label{1} In the graded case, take a general element $A_0$ of $[\End_R(M)]_0$, the degree-0 part of $\End_R(M)$; in the local case, take a general element $A_0$ of $\End_R(M) \setminus m \End_R(M)$.
+  \item Consider the resulting endomorphism $A$ of the $k$-vector space $M/mM$ and find the eigenvalues of $A$.
+  \item If $A$ has at least two eigenvalues, choose one eigenvalue $\l$, and compute a sufficiently high power $B=A^{p^e(p^{e}-1)}$ of $A$ (with the power explicitly as in the proof of Lemma~\ref{lem:jordan}).
+This power will be a nonzero idempotent, and thus produce a splitting of $M$ as $\im B \oplus \coker B$ by Lemma~\ref{lem:idemp}.
+  \item Repeat steps (1)--(3) for both $\im B$ and $\coker B$.
   \end{enumerate}
+\end{alg}
 
-The following observation implies that if $M$ is indecomposable, then the above algorithm should find the decomposition of $M$:
+As a consequence of the algorithm above, we have:
 
 \begin{lem}
+If $M$ has an endomorphism with at least two distinct eigenvalues modulo $m$, then $M$ is not indecomposable.
+\end{lem}
+
+The converse observation implies that if $M$ is not indecomposable, then the above algorithm will find the direct sum decomposition of $M$:
+
+\begin{lem}
+Retain the notation above.
   If $M$ is not indecomposable, then a general degree-0 endomorphism of $M$ reduces to an endomorphism of $M/mM$ with at least two distinct eigenvalues.
@@
 \begin{rem}
-  By ``general'' we mean that a general linear combination of a basis for $[\End(M)]_0$ over the algebraic closure of $k$, or equivalently over a sufficiently large algebraic extension of $k$.
+  By ``general'' we mean that a general linear combination of a basis for $\End_R(M)$ over the algebraic closure of $k$, or equivalently over a sufficiently large algebraic extension of $k$.
 \end{rem}
@@
   We may assume that the base field $k$ is algebraically closed.
-  Let $\Phi_1,\dots,\Phi_r$ be a basis for $[\End(M)]_0$, and $\phi_1,\dots,\phi_r$ their images modulo $m$, which we view as matrices with entries in $k$.
+  Let $\Phi_1,\dots,\Phi_r$ be a basis for $\End_R(M)$, and $\phi_1,\dots,\phi_r$ their images modulo $m$, which we view as matrices with entries in $k$.
   Let $U\subset \A^r$ be the subset of $r$-tuples $(\l_1,\dots,\l_r)$ such that $\l_1\phi_1+\dots+\l_r\phi_r$ has at least two distinct eigenvalues, i.e., such that $\l_1\Phi_1+\dots+\l_r\Phi_r$ reduces to an endomorphism of $M/mM$ with at least two distinct eigenvalues.
@@
   It suffices to show that $U$ is a nonempty open subset of $\A^r$. First, we show $U$ is nonempty:
-  Say $M=M_1\oplus M_2$, with $M_1$ a nontrivial indecomposable summand.
-We may choose $\Phi_1$ to be the projection to $M_1$, and $\Phi_2$ the projection to $M_2$.
-Then for any $\l_1,\l_2\in k$, $\l_1\phi_1 + \l_2\phi_2$ has eigenvalues $\l_1,\l_2$; thus in particular there is \emph{an} element of $[\End(M)]_0$ reducing to an endomorphism of $M/mM$ with distinct eigenvalues, so $U$ is nonempty.
+  Say $M=M_1\oplus M_2$ with $M_1,M_2$  nonzero  summands.
+Choose $\Phi_1$ to be the projection to $M_1$, and $\Phi_2$ the projection to $M_2$.
+Then for any $\l_1,\l_2\in k$, $\l_1\phi_1 + \l_2\phi_2$ has eigenvalues $\l_1,\l_2$; thus in particular there is \emph{an} element of $[\End_R(M)]_0$ reducing to an endomorphism of $M/mM$ with distinct eigenvalues, so $U$ is nonempty.
 
@@
   has at least two distinct eigenvalues for $\l_1,\dots,\l_r$ outside a Zariski-closed subset of $\A^r$.
-  The eigenvalues of $A_{\l_1,\dots,\l_r}$ are the roots of $\det(A_{\l_1,\dots,\l_r}-t I)$, which is of course a polynomial in $t$ with coefficients in $\l_1,\dots,\l_r$.
+  The eigenvalues of $A_{\l_1,\dots,\l_r}$ are the roots of $\det(A_{\l_1,\dots,\l_r}-t I)$, which is a polynomial in $t$ with coefficients in $\l_1,\dots,\l_r$.
 $A_{\l_1,\dots,\l_r}$ fails to have at least two distinct eigenvalues exactly when this polynomial factors as a power of a linear term.
-This is a polynomial condition in the coefficients of powers of $t$ in $\det(A_{\l_1,\dots,\l_r}-t I)$ and thus in the $\l_i$; to see this, note that if
-  $ f:=t^n b_n +\dots +t b_1 +b_0 $
+
+This condition is polynomial in the coefficients of powers of $t$ in $\det(A_{\l_1,\dots,\l_r}-t I)$ and thus in the $\l_i$; to see this, note that
+if we write
+  $$ f:=\det(A_{\l_1,\dots,\l_r}-t I)=t^n  +t^{n-1}b_{n-1}+\dots +t b_1 +b_0, $$
+then $f$
   has an $n$-fold root exactly when
-  $ f,\d f/\d t,\dots, \d^n f/\d t^n $
-  vanish simultaneously; the resultant of these $n$ polynomials in $n$ equations gives polynomial conditions in the $b_i$ for this to occur; in our setting, the $b_i$ are themselves polynomials in the $\l_i$, and thus we have obtained polynomial equations defining the locus where $A_{\l_1,\dots,\l_r}$ fails to have distinct roots.
+  $ f,\d f/\d t,\dots, \d^{n-1} f/\d t^{n-1} $
+  vanish simultaneously; the resultant of these $n$ polynomials in the $n$ variables $b_i$ gives polynomial conditions in the $b_i$ for this to occur.
+%\devlin{TODO: this is $n+1$ equations, so make sure it's okay}
+ In our setting, the $b_i$ are themselves polynomials in the $\l_i$, and thus we have obtained polynomial equations defining the locus where $A_{\l_1,\dots,\l_r}$ fails to have distinct roots, and thus the complement $U$ is open.
 \end{proof}
@@
 \begin{rem}
-  Note that the above algorithm is quite sensitive to the ground field $k$, because it needs an eigenvalue of the endomorphism $A$ of $M/mM$ to be contained in $k$.
+  Note that the above algorithm is quite sensitive to the ground field $k$, because it needs all the eigenvalues of the endomorphism $A$ of $M/mM$ to be contained in $k$.
 While theoretically the issue can be avoided by working over an algebraically closed ground field $\bar k$, for practical use on a computer algebra system it is better to extend $k$ to some larger finite field.
@@
 \begin{lem}
-  Let $M$ be a finitely generated $R$-module and let $[\End M]_0$ be the $k$-vector space of degree-0 endomorphisms. Suppose that either:
+  Suppose $R$ is graded, and let $M$ be a finitely generated $R$-module and let $[\End_R(M)]_0$ be the $k$-vector space of degree-0 endomorphisms. Suppose that either:
   \begin{enumerate}
-  \item $[\End M]_0$ is 1-dimensional and thus spanned by the identity endomorphism, or
-  \item every non-identity element of $[\End M]_0$, viewed as a matrix, has all entries contained in $m$;
+  \item $[\End_R(M)]_0$ is 1-dimensional and thus spanned by the identity endomorphism, or
+  \item every non-identity element of $[\End_R(M)]_0$, viewed as a matrix, has all entries contained in $m$;
   \end{enumerate}
@@
 \end{lem}
+
 \begin{proof}
-  If $M$ decomposes non-trivially as $M_1\oplus M_2$, then the projections onto each factor are nontrivial degree-0 endomorphisms not equal to the identity, and which does not have entries contained in~$m$.
+  If $M$ decomposes non-trivially as $M_1\oplus M_2$, then the projections onto each factor are nontrivial degree-0 endomorphisms not equal to the identity, and which do not have entries contained in~$m$.
 \end{proof}
 
-This is the local analog of the following fact about indecomposability of coherent sheaves:
+This is the algebraic analog of the following fact about indecomposability of coherent sheaves:
 
@@
 
+\begin{rem}[Characteristic zero]
+If $R$ is a local or graded ring over a field $k\subset \C$ of characteristic 0, and $M$ a finitely generated $R$-module, it would be useful to have an algorithm to find a direct sum decomposition of $M$.
+At the moment, we do not have such an algorithm.
+However, we note that the algorithm above can be used to test indecomposability of $M$ via reduction modulo $p$, as follows:
+One can choose a finitely generated $\Z$-algebra $A$ and an $A$-algebra $R_A$ such that $R_A\otimes _A k=R$, and likewise an $R_A$-module $M_A$ such that $M_A\otimes _A k = R$ and such that $M_A$ is flat over $A$.
+If $n$ is a maximal ideal of $A$, one can check that $A/n \cong \mathbb F_{p^e}$ for some prime $p$ and $e$.
+The key point is that if $M$ is decomposable, then $M_A$ also can be taken to be decomposable. Thus, the various reductions  $M\otimes A/n$ will also be decomposable for all $n$.
+If our algorithm does not detect an indecomposable summand of $M\otimes A/n$ for an appropriate choice of $n$, then, the original module $M$ must have been indecomposable.
+However, if $M$ is \emph{decomposable}, we do not have a way of patching the decompositions of various $M_A\otimes_A A/n$ into a decomposition of $M$.
+
+We also point out that it is possible to ``guess'' an idempotent for $M$, even when there is no algorithm to produce one.
+In practice, if $M$ does have nontrivial idempotents, Macaulay2 often chooses them as some of the generators of $[\End_R(M)]_0$.
+By checking if the generators of $[\End_R(M)]_0$ are idempotent, and trying to construct idempotents by subtracting eigenvalues off of general endomorphisms and checking idempotency, we can often produce nontrivial direct sum decompositions in characteristic 0.
+See Examples~\ref{example:QQ} for such a case.
+\end{rem}
+
+%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Decomposing coherent sheaves}
+
+While the preceding section was written in the language of modules, by the standard translation to global (multi)projective varieties, the algorithm works equally well to find indecomposable decompositions of coherent sheaves on (multi)projective varieties.
+In this section, we make a few notes regarding the relation between the eigenvalues discussed in the preceding section (in this section, called \emph{irrelevant eigenvalues}) with the notion of eigenvalue of an endomorphism of a vector bundle.
+
+Let $X\subset \P^n$ be a projective variety, with ample line bundle $\O_X(1)$.
+Throughout, let $E$ be a vector bundle (i.e., a locally free coherent sheaf) on $X$.
+
+The following is well-known:
+
+\begin{lem}
+A morphism $f:E\to E$ is injective if and only if it is an isomorphism.
+\end{lem}
+
+%(It is also true that $f$ is surjective if and only if it's an isomorphism, but this is more well-known.)
+
+For completeness, we give the standard proof:
+
+\begin{proof}
+This requires only that $E$ is a coherent sheaf:
+% since $\End_{\O_X}(E)$ is finite-dimensional, $f$ satisfies some minimal monic polynomial $\phi(x)$; since $f$ is injective, the constant term of $\phi(x)$ is nonzero.
+We claim that $f\otimes k(x)$ is injective for any $k(x)$. This then implies that $f\otimes k(x)$ is surjective for every $x\in X$ and thus $f$ is surjective.
+To see this, note that $f$ satisfies some minimal-degree monic polynomial, since $\End_{\O_X}(E)$ is finite-dimensional over $k$, and that this monic polynomial has nonzero constant term, since $f$ is injective.
+Now, $f\otimes k(x)$ satisfies this same polynomial, so must be injective.
+\end{proof}
+
+\begin{dfn}
+For $f\in \End_{\O_X}(E)$ there an associated map
+$$ \bigwedge^{\rank E} f: \det E\to \det E; $$
+since $\End_{\O_X}(\det E) = H^0(\O_X)=k$, $\bigwedge^{\rank E} f$ is multiplication by $\l\in k$; we write $\det f$ for this scalar $\l$.
+\end{dfn}
+
+\begin{lem}
+Let $x \in X$ be any point (not necessarily closed). Then $\det(f\otimes k(x)) =\det f$.
+%In other words, if we restrict $f$ to the fiber $E_x$, the determinant is the same as $\det f$.
+\end{lem}
+
+\begin{proof}
+Taking the determinant commutes with restriction to a fiber, so $\det(f\otimes k(x)) =\det(f)\otimes k(x)$, but $\det(f)\in k$ already so is unaffected by going modulo $m_x$.
+\end{proof}
+
+\begin{lem}
+$\det f \neq 0 $ if and only if $f$ is injective (if and only $f$ is an isomorphism).
+\end{lem}
+
+\begin{proof}
+  Let $\ker f \neq 0$. %Then we have $0\to K\to E \to \im f\to0$;
+ Then $\det(f\otimes k(X))= 0$ (since $\ker f$ is torsion-free), and in particular  $\det f = 0$ as well.
+If $\ker f = 0$, then
+$f$ is an isomorphism, hence an isomorphism on fibers, and hence
+$\det(f\otimes k(x)) \neq 0$ for any $x$. Thus $\det f\neq 0$.
+\end{proof}
+
+\begin{dfn}
+$\l \in k$ is an eigenvalue of $f \in \End_{\O_X}(E)$ if $\det(f-\l \id_E)= 0$.
+In other words, the eigenvalues of $f$ are the zeroes of the univariate polynomial $\det(f-t\id _E)$ with coefficients in $k$.
+\end{dfn}
+
+Note that $f$ is an isomorphism if and only if $\l$ is not an eigenvalue of $f$, just as for ordinary linear operators on vector spaces.
+
+\begin{dfn}
+Let $M$ be a homogeneous module over a graded ring $R$ with $R_0=k$.
+Let $g \in \End_R(M)$. The \emph{irrelevant eigenvalues} of $g$ are defined to be the eigenvalues of the map of vector spaces $g\otimes R/m:M/mM \to M/mM$.
+\end{dfn}
+
+In other words, the irrelevant eigenvalues are the eigenvalues modulo the maximal ideal discussed in the previous section.
+
+Now, let $R$ be the homogeneous coordinate ring of $X\subset \P^n$, and let $M$ be a graded $R$-module such that $\wtilde M=E$.
+
+\begin{lem}
+Let $f:E\to E$ arise from a map $g:M\to M$ (i.e., $\wtilde M = E$ and $\wtilde g = f$).
+The eigenvalues of $f$ are a subset of the irrelevant eigenvalues of $g$.
+\end{lem}
+
+Note that any endomorphism of $E$ corresponds to an endomorphism of $M=\Gamma_*(E)$; we may want the freedom to work with other module representatives of $E$ though.
+
+\begin{proof}
+  All that needs to be shown is that if $\det(f-\l \id_E) =0$, then $\det((g-\l \id_M)\otimes R/m)=0$.
+Replacing $f$ by $f-\l \id_E$ and likewise $g$ by $g-\l\id_M$, we just need to show that $\det(f)=0$ implies $\det(g\otimes R/m)=0$.
+
+Say $\det(f)=0$.
+If $\det(g\otimes R/m)\neq 0$, then $g$ induces a surjection $M/mM\to M/mM$, and thus $g$ induces a surjection $M\to M$ by Nakayama's lemma.
+This then forces $f$ to be a surjection, hence an isomorphism, and thus $\det(f)\neq 0$.
+\end{proof}
+
+\begin{exa}
+By giving $M$ irrelevant summands, one can always add ``extraneous'' irrelevant eigenvalues, so the containment of eigenvalues may be proper.
+%However, if $M=\Gamma_*(E)$, set of values will be the same: in this case, if $g\otimes R/m$ is not an isomorphism, then $g$ cannot be surjective.
+Note that in general the multiplicities will never be equal, even if one takes $M=\Gamma_*(E)$: if $f$ is multiplication by $\l$, then $f$ will have $\rank E$ many eigenvalues, while $g$ will have $\mu(M)$ irrelevant eigenvalues.
+\end{exa}
+
+By combining the discussion above with the results of the preceding section, we thus have:
+
+\begin{prop}
+If $X$ is a variety over a field $k\subset \overline{\mathbb F_{p}}$,
+$E$ a vector bundle on $X$, and $f:E\to E$ has two distinct eigenvalues $\l_1,\l_2$, then $E$ is a nontrivial direct sum, with one summand given by $(f-\l_1)^{p^e(p^e-1)}$ for some $e\geq 1$.
+\end{prop}
+
+\begin{proof}
+Take $M= \Gamma_*(E)$; then since the eigenvalues $\l_1\neq \l_2$ are also irrelevant eigenvalues of $M$, our algorithm produces a nontrivial direct sum decomposition $M_1\oplus M_2$, and since $M$ is torsion-free we know that $\widetilde M_i\neq 0$, and thus $E=\widetilde M_1\oplus \widetilde M_2$ is a nontrivial direct summand of sheaves.
+\end{proof}
+
+\begin{rem}
+We note that it is known already by \cite[Proposition~15]{Atiyah} that an endomorphism of an indecomposable vector $E$ cannot have two distinct eigenvalues.
+The utility of the preceding lemma is in producing an explicit direct sum decomposition of $E$.
+\end{rem}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-While the preceding section was written in the language of modules, by the standard translation to global (multi)projective varieties, it works equally well to find indecomposable decompositions of coherent sheaves.
 In this section, we give examples of the kind of calculations and observations the algorithm from the previous section allows us to make.
@@
   $$ F_* \O_X=\bigoplus_{p=1}^7 \O_X(p_i). $$
-  This reflects the fact that the 7-torsion points $p_i$ of $X$, and thus the sheaves $\O_X(p_i)$, are not defined over $\mathbb F_7$, but they are defined over $\mathbb F_{49}$.
+  This reflects the fact that the 7-torsion points $p_i$ of $X$, and thus the sheaves $\O_X(p_i)$, are not defined over $\mathbb F_7$, but rather are defined over $\mathbb F_{49}$.
 \end{exa}
@@
   $$ F_*^e \O_X \cong \O_X\oplus M\oplus F_* M\oplus\dots\oplus F_*^{e-1}M. $$
-  In particular, $\O_X$ will fail to have the finite $F$-representation type property.
-  In fact, we believe a similar description holds true for quartic del Pezzos, which arise as an intersection of quadrics in $P^4$.
+  In particular, this suggests $\O_X$ will fail to have the finite $F$-representation type property.
+  In fact, we believe a similar description holds true for quartic del Pezzos.
 \end{exa}
@@
 \begin{exa}[Syzygies over Artinian rings]
+  In recent work suggested by examples calculated using our algorithm, \cite{CDE+24} studies the indecomposable summands of syzygy modules over a Golod ring $(R,m,k)$.
+In particular, they find previously unexpected recurring behavior, specifically that the syzygy modules of the residue field are direct sums of only three indecomposable modules: the residue field $k$, the maximal ideal $m$, and an additional module $N = \Hom_R(m, R)$.
+
+  Here, we give a concrete example of one such ring:
   Let $R = k[x,y]/(x^3,x^2y^3,y^5)$ and consider the (infinite) minimal free resolution of the residue field, which has rank $2^n$ in homological index $n$.
-In forthcoming work based in part on examples calculated using our algorithm, Dao, Eisenbud, and Polini study the indecomposable summands of syzygy modules in examples such as this one, showing unexpected periodicity behavior, and in particular proving that the syzygy modules are direct sums of only three indecomposable modules: the residue field $k$, the maximal ideal $m$, and an additional module $M$.
 
-  For example, the fourth syzygy module decomposes (ignoring the grading) as the direct sum
-  $$ k^3 \oplus m^2 \oplus M^3. $$
+  The fourth syzygy module of the residue field decomposes (ignoring the grading) as the direct sum
+  $$ k^3 \oplus m^2 \oplus N^3. $$
   and the fifth syzygy module as
-  $$ k^8\oplus m^9 \oplus M^2. $$
-  The use of our algorithm was essential to the observation that only the one additional module $M$ beyond the ``guaranteed'' summands of $k$ and $m$ appealrs.
-  %% nebedsPackage "DirectSummands"
+  $$ k^8\oplus m^9 \oplus N^2. $$
+  The use of our algorithm was essential to the observation that beyond the ``guaranteed'' summands of $k$ and $m$ (which were known to appear by work of \cite{DE23}) only the one additional indecomposable module $N$ appears in the summands of syzygies of $k$.
+  %% needsPackage "DirectSummands"
   %% R = ZZ/101[x,y]/(ideal"x3,x2y3,y5")
@@
 \end{exa}
+
+\begin{exa}[Symbolic diagonalization]\label{example:QQ}
+  An interesting application of our algorithm, suggested by Bernd Sturmfels,  is automated diagonalization of symbolically parameterized matrices.
+As a simple demonstration, let $R = K[a,b,c,d]$ and consider the following matrix
+  \[ A = \begin{pmatrix}
+    a&b&c&d\\
+    d&a&b&c\\
+    c&d&a&b\\
+    b&c&d&a
+  \end{pmatrix} \]
+  Then the splittings of $\coker A$ over $K = \Q$ and $K = \Q(i)$ have the following presentations, respectively:
+  \[\medmuskip1.5mu \begin{pmatrix}
+    a+b+c+d&0&0&0\\
+    0&a-b+c-d&0&0\\
+    0&0&a-c&b-d\\
+    0&0&b-d&c-a
+  \end{pmatrix},
+\quad
+  \begin{pmatrix}
+    a+b+c+d&0&0&0\\
+    0&a-b+c-d&0&0\\
+    0&0&a+i\,b-c-i\,d&0\\
+    0&0&0&a-i\,b-c+i\,d
+  \end{pmatrix}.
+  \]
+
+\end{exa}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \begin{lem}
-  Let $A\in \End_R M$ be idempotent. Then $M=\im  A\oplus \ker A$.
+  Let $A\in \End_R M$ be idempotent. Then $M = \im A\oplus \ker A$.
 \end{lem}

### 2024-11-15 paper/titlepage.tex
version of paper submitted to JSC

\documentclass[12pt]{article}
\usepackage{notes,hyperref}
\renewcommand{\thefootnote}{\alph{footnote}}

\begin{document}
\subsubsection*{Title}
Computing Direct Sum Decompositions
\subsubsection*{Authors} Devlin Mallory\footnote{Basque Center for Applied Mathematics, Plaza Euskadi 5, 48009 Bilbao, Basque Country,  Spain}~and Mahrud Sayrafi\footnote{MPI für Mathematik in den Naturwissenschaften, Inselstraße 22, 04103 Leipzig, Germany}
\subsubsection*{Corresponding author} Mahrud Sayrafi

\end{document}

### 2024-12-27 paper/main.tex
version submitted to arXiv

@@
 \title{Computing Direct Sum Decompositions}
+
 \author{Devlin Mallory and Mahrud Sayrafi}
+
+%% \author{Devlin Mallory}
+%% \affiliation{Basque Center for Applied Mathematics, Bilbao, Basque Country, Spain}
+%% \email{\href{mailto:dmallory@bcamath.org}{dmallory@bcamath.org}}
+
+%% \author{Mahrud Sayrafi}
+%% \affiliation{Max Planck Institute for Mathematics in the Sciences, Leipzig, Germany}
+%% \email{\href{mailto:mahrud@mis.mpg.de}{mahrud@mis.mpg.de}}
+
+%% \subjclass[2020]{16D70; 14F06, 13A35}
+
 \begin{document}
@@
 
-An accompanying implementation in Macaulay2 is available online via the GitHub repository \\
+An accompanying implementation in Macaulay2 \cite{M2} is available via the GitHub repository \\
 \centerline{
@@
   vanish simultaneously; the resultant of these $n$ polynomials in the $n$ variables $b_i$ gives polynomial conditions in the $b_i$ for this to occur.
-%\devlin{TODO: this is $n+1$ equations, so make sure it's okay}
  In our setting, the $b_i$ are themselves polynomials in the $\l_i$, and thus we have obtained polynomial equations defining the locus where $A_{\l_1,\dots,\l_r}$ fails to have distinct roots, and thus the complement $U$ is open.
@@
 \begin{rem}
-We note that it is known already by \cite[Proposition~15]{Atiyah} that an endomorphism of an indecomposable vector $E$ cannot have two distinct eigenvalues.
+We note that it is known already by \cite[Proposition~15]{Atiyah57} that an endomorphism of an indecomposable vector $E$ cannot have two distinct eigenvalues.
 The utility of the preceding lemma is in producing an explicit direct sum decomposition of $E$.
@@
 \begin{exa}[Frobenius pushforward on Grassmannians]
+  % https://arxiv.org/abs/1901.10956
   Consider the Grassmannian $X = \mathrm{Gr(2,4)}$. We may work over the Cox ring $S$,
@@
 
+%\begin{exa}[Frobenius pushforward on invariant rings of finite groups]
+  % https://arxiv.org/abs/2312.11786
+%  \mahrud{TODO: add an invariant ring example from \cite{HS24}}
+%\end{exa}
+
 \begin{exa}[Syzygies over Artinian rings]
+  % https://arxiv.org/abs/2208.05427
   In recent work suggested by examples calculated using our algorithm, \cite{CDE+24} studies the indecomposable summands of syzygy modules over a Golod ring $(R,m,k)$.

### 2024-12-31 paper/main.tex
uncommitted changes

@@
 %% \author{Devlin Mallory}
-%% \affiliation{Basque Center for Applied Mathematics, Bilbao, Basque Country, Spain}
+%% \address{Basque Center for Applied Mathematics, Bilbao, Basque Country, Spain}
 %% \email{\href{mailto:dmallory@bcamath.org}{dmallory@bcamath.org}}
@@
 %% \author{Mahrud Sayrafi}
-%% \affiliation{Max Planck Institute for Mathematics in the Sciences, Leipzig, Germany}
+%% \address{Max Planck Institute for Mathematics in the Sciences, Leipzig, Germany}
 %% \email{\href{mailto:mahrud@mis.mpg.de}{mahrud@mis.mpg.de}}
@@
 After describing the algorithm and proving its correctness, we present multiple examples, including some which present previously unknown phenomena regarding the behavior of summands of Frobenius pushforwards and syzygies over Artinian rings.
-In particular, we highlight the results of \cite{CDE+24}, which shows a recurrence formula for indecomposable summands of high syzygies of the residue field of Golod rings, made possible through experiments and observations using our algorithm.
+In particular, we highlight the results of \cite{CDE24}, which shows a recurrence formula for indecomposable summands of high syzygies of the residue field of Golod rings, made possible through experiments and observations using our algorithm.
 
@@
   % https://arxiv.org/abs/2208.05427
-  In recent work suggested by examples calculated using our algorithm, \cite{CDE+24} studies the indecomposable summands of syzygy modules over a Golod ring $(R,m,k)$.
+  In recent work suggested by examples calculated using our algorithm, \cite{CDE24} studies the indecomposable summands of syzygy modules over a Golod ring $(R,m,k)$.
 In particular, they find previously unexpected recurring behavior, specifically that the syzygy modules of the residue field are direct sums of only three indecomposable modules: the residue field $k$, the maximal ideal $m$, and an additional module $N = \Hom_R(m, R)$.

### 2025-01-07 paper/main.tex
added remark about complexity and references to TDA

@@
 
-Currently there are no efficient algorithms available for checking indecomposability or finding summands of modules over commutative rings.
+Currently there are no efficient algorithms available for checking indecomposability or finding summands of arbitrary modules over commutative rings.
 In contrast, variants of the ``Meat-Axe'' algorithm for determining irreducibility of finite-dimensional modules over a group algebra have wide ranging applications in computational group theory \cite{Parker84,HR94,Holt98} and are available through symbolic algebra software such as Magma and GAP \cite{MAGMA,GAP}.
+
+Meanwhile, effective computation of indecomposable components of multiparameter persistence modules is essential in topological data analysis, which has been applied to computational chemistry, materials science, neuroscience, and many other areas \cite{BL23}.
 
@@
 We also describe applications of our algorithm for computations in characteristic 0.
-Further, our algorithm works over multigraded rings, which enables the computation of indecomposable summands of coherent sheaves on subvarieties of toric varieties (in particular, for varieties embedded in projective space).
+Further, our algorithm takes advantage of multigradings, which enables the computation of indecomposable summands of coherent sheaves on subvarieties of toric varieties (in particular, for varieties embedded in $\P^n$).
 
@@
 \subsection*{Acknowledgements}
-The authors would like to thank David Eisenbud for several very useful conversations and ideas.
+The authors would like to thank David Eisenbud for several very useful conversations and ideas, and Ezra Miller for pointing out the relevance of this algorithm in topological data analysis.
 We also thank the American Institute of Mathematics for hosting the ``Macaulay2: expanded functionality and improved efficiency'' workshop, during which some of this work took place.
@@
   \begin{enumerate}
-  \item \label{1} In the graded case, take a general element $A_0$ of $[\End_R(M)]_0$, the degree-0 part of $\End_R(M)$; in the local case, take a general element $A_0$ of $\End_R(M) \setminus m \End_R(M)$.
+  \item\label{item:End0} In the graded case, take a general element $A_0$ of $[\End_R(M)]_0$, the degree-0 part of $\End_R(M)$; in the local case, take a general element $A_0$ of $\End_R(M) \setminus m \End_R(M)$.
   \item Consider the resulting endomorphism $A$ of the $k$-vector space $M/mM$ and find the eigenvalues of $A$.
@@
 \begin{rem}
-  By ``general'' we mean that a general linear combination of a basis for $\End_R(M)$ over the algebraic closure of $k$, or equivalently over a sufficiently large algebraic extension of $k$.
+  By ``general'' we mean that a general linear combination of a basis for $[\End_R(M)]_0$ over the algebraic closure of $k$, or equivalently over a sufficiently large algebraic extension of $k$.
 \end{rem}
@@
 While theoretically the issue can be avoided by working over an algebraically closed ground field $\bar k$, for practical use on a computer algebra system it is better to extend $k$ to some larger finite field.
-However, the general linear combinations we take in Step \ref{1} should be taken with respect to the prime subfield (otherwise, as we increase the size of the finite field $k$, the eigenvalues of a general linear combination will live in higher and higher field extensions).
+However, the general linear combinations we take in Step \ref{item:End0} should be taken with respect to the prime subfield (otherwise, as we increase the size of the finite field $k$, the eigenvalues of a general linear combination will live in higher and higher field extensions).
 See Example~\ref{algclos} for a demonstration of the necessity of extending the base field.
@@
   \item $[\End_R(M)]_0$ is 1-dimensional and thus spanned by the identity endomorphism, or
-  \item every non-identity element of $[\End_R(M)]_0$, viewed as a matrix, has all entries contained in $m$;
+  \item every non-identity element of $[\End_R(M)]_0$, viewed as a matrix, has entries in $m$;
   \end{enumerate}
@@
 
-This is the algebraic analog of the following fact about indecomposability of coherent sheaves:
+This is the algebraic analog of the following fact about coherent sheaves:
 
@@
 See Examples~\ref{example:QQ} for such a case.
+\end{rem}
+
+\begin{rem}
+  We omit a computational complexity analysis here because without any assumptions about the module there are too many parameters that in any given application may become prominent.
+We note, however, that $[\End_R(M)]_0$ can be computed using degree limited syzygy algorithms without computing all of $\End_R(M)$.
+While in many applications this significantly reduces the computational complexity, in general the complexity of the algorithm presented here is at least bounded by the complexity of computing syzygies.
+Compare with \cite{DX22}, which provides a complexity analysis for a decomposition algorithm with the additional assumption that the generators of $R$ as well as generators and relations of $M$ have distinct degrees.
+This assumption is exceedingly rare in commutative algebra and algebraic geometry applications, but it is relevant, though not guaranteed, in the context of multiparameter persistence in topological data analysis.
 \end{rem}
@@
   \]
+\end{exa}
 
-\end{exa}
+%\begin{exa}[Multiparameter persistence modules]
+%  \mahrud{TODO: add an invariant ring example from \cite{BL23} or \cite{DX22}}
+%\end{exa}
 

### 2025-04-16 paper/main.tex
mostly Devlin's changes, a little of mine

@@
 \usepackage{notes,hyperref}
+\usepackage{cleveref}
+\let\b\beta
 \let\l\lambda
@@
 \let\bar\overline
-\def\O{\mathcal O}
 \def\sing{{\mathrm{sing}}}
-\let\b\beta
-\def\F{\mathcal F}
+
+\def\cF{\mathcal F}
+\def\OO{\mathcal O}
+\def\FF{\mathbb F}
+\def\PP{\mathbb P}
+\def\ZZ{\mathbb Z}
+\def\Gr{\mathrm Gr}
+\def\Bl{\mathrm Bl}
+\makeop{Eff}
 
@@
 
-\numberwithin{equation}{section}
 \let\d\partial
@@
 \theoremstyle{theorem}
-\renewtheorem{thm}{Theorem}
-\renewtheorem{alg}[thm]{Algorithm}
-\renewtheorem{lem}[thm]{Lemma}
-\renewtheorem{cor}[thm]{Corollary}
-\renewtheorem{prop}[thm]{Proposition}
-\renewtheorem{conj}[thm]{Conjecture}
+\newtheorem{thm}{Theorem}
 \numberwithin{thm}{section}
+\newtheorem{algm}[thm]{Algorithm}
+\newtheorem{lem}[thm]{Lemma}
+\newtheorem{cor}[thm]{Corollary}
+\newtheorem{prop}[thm]{Proposition}
+\newtheorem{conj}[thm]{Conjecture}
 \theoremstyle{definition}
-\renewtheorem{dfn}[thm]{Definition}
-\renewtheorem{exa}[thm]{Example}
-\renewtheorem{note}[thm]{Note}
-\renewtheorem{nota}[thm]{Notation}
-\renewtheorem{quest}[thm]{Question}
-\renewtheorem{rem}[thm]{Remark}
-
-\def\L{\mathcal L}
-\def\M{\mathcal M}
+\newtheorem{dfn}[thm]{Definition}
+\newtheorem{exa}[thm]{Example}
+\newtheorem{note}[thm]{Note}
+\newtheorem{nota}[thm]{Notation}
+\newtheorem{quest}[thm]{Question}
+\newtheorem{rem}[thm]{Remark}
+\def\defn#1{{\it #1}}
 
@@
 \begin{abstract}
-  The problems of finding isomorphism classes of indecomposable modules with certain properties, or determining the indecomposable summands of a module, are ubiquitous in commutative algebra, group theory, representation theory, and other fields.
-The purpose of this work is to describe and prove correctness of a practical algorithm for computing indecomposable summands of finitely generated modules over a finitely generated $k$-algebra, for $k$ a field of positive characteristic.
-Our algorithm works over multigraded rings, which enables the computation of indecomposable summands of coherent sheaves on subvarieties of toric varieties (in particular, for varieties embedded in projective space).
+  The problems of finding isomorphism classes of indecomposable modules with certain properties, or determining the indecomposable summands of a module, are ubiquitous in commutative algebra, representation theory, topological data analysis, and other fields.\linebreak The purpose of this work is to describe and prove correctness of two practical algorithms for computing indecomposable summands of finitely generated modules over a finitely generated $k$-algebra $R$: the first algorithm applies in the (multi)graded case, which enables the computation of indecomposable summands of coherent sheaves on subvarieties of toric varieties (in particular, for varieties embedded in projective space); the second algorithm applies when $R$ is local and $k$ is a finite field, which opens the door to computing decompositions in singularity theory.
 We also present multiple examples, including some which present previously unknown phenomena regarding the behavior of summands of Frobenius pushforwards and syzygies over Artinian rings.
@@
 For both these problems, and many others, making and testing conjectures depends on computing summands of modules and verifying their indecomposability.
+\devlin{Reference 15.10.9 of \cite{Eisenbud95}}
 
@@
 
+We also thank the referees for their careful reading and suggestions, which significantly improved the exposition of the paper.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\section{The main algorithm}\label{sec:main-alg}
+\section{Notations}
+
+\begin{itemize}
+\item For a field $k$, we will write $\overline k$ for a choice of algebraic closure of $k$. In particular, $\overline{\mathbb F_p}$ is the algebraic closure of the finite field $\mathbb F_p$.
+\item For an algebraic variety $X$ and a point $x\in X$, we write $k(x)=\O_{X,x}/m_x$ for the residue field of $X$ at $x$.
+\item We write $k(X)$ for the function field of $X$.
+\item If $X$ is a projective variety, and $\O_X(1)$ a fixed very ample line bundle on $X$, then for any coherent $\O_X$-module $\cF$ we write 
+$\Gamma_* \cF = \bigoplus_{n\geq 0} H(X,\cF(n))$. This is a module over the graded ring $S=\bigoplus_{n\geq 0} H^0(X,\O_X(n))$.
+\item Conversely, if $S$ is a graded ring defining a projective variety $X$, and $M$ is a finitely generated $S$-module, we write $\wtilde M$ for the coherent sheaf on $X$ associated to $M$.
+\item For a graded or local ring $(R,m)$ and an $R$-module $M$, we write $\mu(M):=\dim_{R/m} M/mM$ for the minimal number of generators of $M$ as an $R$-module.
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{The main algorithms} 
+Here, we describe the main algorithms for finding the indecomposable summands of a finitely generated $R$-module $M$, when $R$ is either a graded ring or a local ring with residue field contained in $\overline{\mathbb F_p}$.
+
+\subsection{The graded algorithm}\label{sec:main-alg-new}
+Let $R$ be a $\N$-graded ring, with $R_0=k$ an algebraically closed field.
+%Assume furthermore that the degree support of $R$ is strongly convex, i.e., that 
+%%the effective cone
+%$\Eff R = \{ \deg r : r\in R\}$ is a strongly convex cone in $\Z^n$.
+Let $M$ be a finitely generated graded $R$-module, and let $\phi \in \End M$ be a degree-0 endomorphism.
+
+%Let $P\subset \Z^n$ be the degrees of a set of minimal generators of $M$; 
+%this is a poset with respect to the partial order $\a\leq \b$ if and only if $\b-\a\in \Eff R$.
+%Because $P$ is finite, we can order the elements of $P$ such that 
+%$\deg m_1\not\leq \deg m_2\not \leq \dots\not\leq \deg m_r$.
+
+\begin{lem}
+\label{graded_algorithm}
+Let $n=\mu(M)$ and set $\psi=\phi^N$. Then $M=\im \psi \oplus \coker \psi$.
+\end{lem}
+
+\begin{proof}
+Fix homogeneous generators $m_1,\dots,m_n$ of $M$, with $\deg m_i$ nonincreasing.
+The matrix for $\phi$ is then ``block upper triangular'' with diagonal blocks with entries in $k$, i.e., of the form
+$$
+\phi=
+\begin{pmatrix}
+  B_1 & * & \dots& * \\
+  0 & B_2 & \dots &* \\
+\vdots &\vdots & \ddots &\vdots \\
+0 & 0 & \dots &B_l
+\end{pmatrix},
+$$
+where all blocks labeled $B_i$ are arbitrary matrices \emph{with entries in $k$} and the $*$ are arbitrary.
+To see this, just note that since $\phi$ has degree 0, $\phi(m_i)$ is a $k$-linear combination of the $m_j$ with $\deg m_j=\deg m_i$ plus an $R$-linear combination of the $m_j$ with $\deg m_j<\deg m_i$.
+
+Thus, since each matrix $B_i$ has entries in the algebraically closed field $k$, by performing row and column operations over $k$ (which correspond to replacing each $m_i$ with a $k$-linear combination of $m_j$ with $\deg m_i = \deg m_j$, which is an endomorphism of $M$) we can put each $B_i$ into Jordan canonical form $J_i$, with each Jordan block having entries in $k$.
+By doing so starting with $B_1$ and proceeding to $B_l$, the resulting matrix has the form
+\begin{equation}
+\label{eq:Jordan}
+\phi
+=
+\begin{pmatrix}
+  J_1 & ** & \dots& ** \\
+  0 & J_2 & \dots &** \\
+\vdots &\vdots & \ddots &\vdots \\
+0 & 0 & \dots & J_l
+\end{pmatrix},
+\end{equation}
+with each $J_i$ a Jordan block with entries in $k$ and eigenvalue $\l_i$, and $**$ arbitrary.
+Relabeling, we may assume that $\l_i\neq 0$ for $i=1,\dots,l_0$ and $\l_i=0$ for $i=l_0+1,\dots,l$.
+
+If we take $N_0  $ to be at least the sum of the sizes of the nilpotent Jordan blocks $J_{l_0+1},\dots,J_l$, then for $N\geq N_0$ we have
+$$
+\phi^N = 
+\begin{pmatrix}
+  J_1^N & *** &\dots & \dots& \dots  &\dots & *** \\
+  0 & J_2^N & \dots &\dots&\dots & \dots& *** \\
+\vdots &\vdots & \ddots &\vdots \\
+0 & 0 & \dots & J_{l_0}^N & *** & \dots & ***\\
+0 & 0 & \dots & 0 & 0 & \dots & 0\\
+\vdots &\vdots & \ddots &\vdots \\
+0 & 0 & \dots & 0 & 0 & \dots & 0
+\end{pmatrix},
+$$
+where each nonzero $J_i^N$ has the units $\l_i^N$ on the diagonal. 
+($\mu(M)$ is an upper bound on the sum of the sizes of the nilpotent Jordan blocks, so the above holds for $N\geq \mu(M)$.)
+
+Fix $N$ as above.
+Now, consider the map $ \phi^N: \im(\phi^N) \to \im(\phi^N)$.
+If this map is surjective, then $\im(\phi^N)$ is a direct summand of $M$, with complementary summand $\coker \phi^N$: to see this, observe that the injection $\im(\phi^N) \inc M$ would then  compose with $\phi^N$ to be a surjection, and since $\im(\phi^N)$ is a finitely generated module, a surjection must be an isomorphism $\a:\im \phi^N \to \im \phi^N$.
+Thus, we obtain a splitting of the inclusion $\im \phi^N \inc M$ via $\a\inv\circ \phi^N$; since $\a\inv$ is an isomorphism, we have that $\im \phi^N$ is (isomorphic to) a direct summand of $M$, with complementary summand $\coker \phi^N$.
+
+So, all that needs to be seen is that $\psi:=\psi^N$ is surjective.
+To see this, note that after replacing the original generators for $M$ by the generators that put $\psi$ into the form above,  a generating set for $\im(\phi^N)$ is given by $f_i = \psi(m_i) = u_i\cdot e_i + \sum_{j<i } a_{ij} e_j$ for $i = 1,\dots,n-N_0$, where $u_i$ is a unit (explicitly, $u_i$ is a power of some nonzero $\l_{i'}\in k$).
+That is, a generating set is given by 
+$$u_1 e_1,\quad  u_2 e_2+a_{21}e_1,\quad \dots ,\quad u_{n-n_0}e_{n-N_0}+a_{n-N_0,n-1}e_{n-N_0-1}+\dots + a_{N-n_0,1}e_1.$$ 
+Since the $u_i$ are units, the $R$-span of these generators is clearly the same as the $R$-span of $e_1,\dots,e_{n-N_0}$, and thus after applying $\psi$ we get back the $R$-span of $f_1,\dots,f_{n-N_0} $, which is $\im \psi^N$.
+Thus, $\phi: \im \psi^N \to \im \psi^N$ is surjective, and the result follows.
+\end{proof}
+
+\begin{rem}[how to do it in multigraded case]
+Let $R$ be an $\Z^n$-graded ring, with $R_0=k$ an algebraically closed field.
+Assume furthermore that the degree support of $R$ is strongly convex, i.e., that 
+the effective cone
+$\Eff R = \{ \deg r : r\in R\}$ is a strongly convex cone in $\Z^n$.
+Let $M$ be a finitely generated graded $R$-module, and let $\phi \in \End M$ be a degree-0 endomorphism.
+
+Let $P\subset \Z^n$ be the degrees of a set of minimal generators of $M$; 
+this is a poset with respect to the partial order $\a\leq \b$ if and only if $\b-\a\in \Eff R$.
+Because $P$ is finite, we can order the elements of $P$ such that 
+$\deg m_1\not\leq \deg m_2\not \leq \dots\not\leq \deg m_r$.
+
+\end{rem}
+
+The above can of course produce the trivial direct sum decomposition: if the eigenvalues of $\phi$ are all nonzero, the proof yields that $\im \psi = M$ and $\coker \psi = 0$.
+The following corollary of the proof of \cref{graded_algorithm} ensures that we can always assume $\phi$ has zero as an eigenvalue:
+
+\begin{lem}
+The eigenvalues of $\phi$ lie in $k$.
+In particular,
+for any  $\phi \in \End M$, there is $\l \in k$ such that $\phi - \l \id_M \in \End M$ has zero as an eigenvalue.
+\end{lem}
+
+Note that by ``eigenvalues of $\phi$'' we mean the roots of the characteristic polynomial $\det(\phi - t \id_M)$, which is a univariate polynomial with coefficients in $R$ and thus a priori has solutions in $t$ only in the algebraic closure of the total ring of fractions of $R$.
+
+\begin{proof}
+Note that the expression of $\psi$ obtained in \cref{eq:Jordan} shows that the determinant of $\phi - t \id_M$ is just the product of $\mu_i -t$, and thus the eigenvalues are just the $\mu_i$, which lie in~$k$.
+\end{proof}
+
+\begin{algm} Given a finitely generated graded $R$-module $M$, the indecomposable summands of $M$ are computed as follows.
+  \begin{enumerate}
+  \item\label{item:End0} Take a random element $\phi_0$ of $[\End_R(M)]_0$, the degree-0 part of $\End_R(M)$.
+  \item For an eigenvalue $\mu $ of $\phi_0$, let $\phi=\phi_0-\mu \id_M$.
+  \item If $\phi$ has at least two eigenvalues, let $\psi=\phi^{\mu(M)}$.
+By \Cref{graded_algorithm}, $M\cong \im \psi \oplus \coker \psi$.
+\item Repeat steps (1)--(3) for both $\im \psi$ and $\coker \psi$.
+  \end{enumerate}
+\end{algm}
+
+We will see via \Cref{lemma_term} that this will result in an indecomposable decomposition of $M$.
+
+\subsection{The local algorithm}\label{sec:main-alg}
 
@@
     0 & 0 & 0 & \dots & \l_i-\l
-  \end{pmatrix} \]
+  \end{pmatrix}. \]
   Set $\mu_i=\l_i-\l$. Then for any $l\geq 1$, the $l$-th power $(A-\l I)^l$ is block-diagonal with blocks of the form
@@
     0 & 0 & 0 & \dots & \mu_i^l
-  \end{pmatrix} \]
-  If we choose $l > r_i$ and $l$ divisible by $p$, then all non-diagonal terms will vanish, so all blocks will have the form
+  \end{pmatrix}. \]
+  If we choose $l > r_i$ and $l$ a power of $p$, then all non-diagonal terms will vanish, so all blocks will have the form
   \[ \begin{pmatrix}
@@
   \end{pmatrix}. \]
-  Finally, say that $k=\mathbb F_p$. Choosing $l$ to be divisible also by $p^e-1$, we have that $$\mu_i^{l-1} = (\mu_i^{p^{e}-1})^{l/(p^{e}-1)} =1.$$
+  Finally, since $k=\mathbb F_{p^e}$ for some $e$, if we choose $l$ to be divisible also by $p^e-1$, we have that $$\mu_i^{l-1} = (\mu_i^{p^{e}-1})^{l/(p^{e}-1)} =1.$$
   Thus, $(A-\l I)^l$ is a diagonal matrix with diagonal entries 1 or 0, hence idempotent. Note moreover that if some $\l_i\neq \l$, then $(A-\l I)^l$ is not the zero matrix.
@@
 
-\begin{alg} Given a finitely generated graded $R$-module $M$, the indecomposable summands of $M$ are computed as follows.
+\begin{algm}\label{local_algorithm} 
+Given a finitely generated graded $R$-module $M$, the indecomposable summands of $M$ are computed as follows.
   \begin{enumerate}
-  \item\label{item:End0} In the graded case, take a general element $A_0$ of $[\End_R(M)]_0$, the degree-0 part of $\End_R(M)$; in the local case, take a general element $A_0$ of $\End_R(M) \setminus m \End_R(M)$.
+%In the graded case, take a general element $A_0$ of $[\End_R(M)]_0$, the degree-0 part of $\End_R(M)$; in the local case, 
+\item Take a general element $A_0$ of $\End_R(M) \setminus m \End_R(M)$.
   \item Consider the resulting endomorphism $A$ of the $k$-vector space $M/mM$ and find the eigenvalues of $A$.
-  \item If $A$ has at least two eigenvalues, choose one eigenvalue $\l$, and compute a sufficiently high power $B=A^{p^e(p^{e}-1)}$ of $A$ (with the power explicitly as in the proof of Lemma~\ref{lem:jordan}).
-This power will be a nonzero idempotent, and thus produce a splitting of $M$ as $\im B \oplus \coker B$ by Lemma~\ref{lem:idemp}.
+  \item If $A$ has at least two eigenvalues, choose one eigenvalue $\l$, and compute a sufficiently high power $B=A^{p^e(p^{e}-1)}$ of $A$ (with the power explicitly as in the proof of \Cref{lem:jordan}).
+This power will be a nonzero idempotent, and thus produce a splitting of $M$ as $\im B \oplus \coker B$ by \Cref{lem:idemp}.
   \item Repeat steps (1)--(3) for both $\im B$ and $\coker B$.
   \end{enumerate}
-\end{alg}
+\end{algm}
 
-As a consequence of the algorithm above, we have:
+\subsection{Termination of the algorithms}
+
+As a consequence of \Cref{graded_algorithm} and \Cref{local_algorithm} above, we have:
 
 \begin{lem}
+Assume that either $(R,\m)$ is a graded ring with $R_0=k$, or $(R,m)$ is a local ring with $k=R/m$ a field of positive characteristic, and that $M$ is a finitely generated $R$-module (homogeneous in the graded case).
 If $M$ has an endomorphism with at least two distinct eigenvalues modulo $m$, then $M$ is not indecomposable.
@@
 
-The converse observation implies that if $M$ is not indecomposable, then the above algorithm will find the direct sum decomposition of $M$:
+The converse observation implies that if $M$ is not indecomposable, then these algorithms will find the direct sum decomposition of $M$:
 
 \begin{lem}
+\label{lemma_term}
 Retain the notation above.
-  If $M$ is not indecomposable, then a general degree-0 endomorphism of $M$ reduces to an endomorphism of $M/mM$ with at least two distinct eigenvalues.
+  If $M$ is not indecomposable, then a general endomorphism of $M$ (of degree 0 in the graded case) reduces to an endomorphism of $M/mM$ with at least two distinct eigenvalues.
+\devlin{need to separate into graded and local cases maybe}
 \end{lem}
+
+In particular, if $M$ is not indecomposable, then \Cref{graded_algorithm} and \Cref{local_algorithm} produce a nontrivial direct sum decomposition of $M$, because a general element of $\End M$ will have distinct eigenvalues.
 
@@
 While theoretically the issue can be avoided by working over an algebraically closed ground field $\bar k$, for practical use on a computer algebra system it is better to extend $k$ to some larger finite field.
-However, the general linear combinations we take in Step \ref{item:End0} should be taken with respect to the prime subfield (otherwise, as we increase the size of the finite field $k$, the eigenvalues of a general linear combination will live in higher and higher field extensions).
-See Example~\ref{algclos} for a demonstration of the necessity of extending the base field.
+However, the general linear combinations we take in \Cref{item:End0} should be taken with respect to the prime subfield (otherwise, as we increase the size of the finite field $k$, the eigenvalues of a general linear combination will live in higher and higher field extensions).
+See \Cref{algclos} for a demonstration of the necessity of extending the base field.
 \end{rem}
@@
 
-This is the algebraic analog of the following fact about coherent sheaves:
+The following lemma is the algebraic analogue of the following fact about coherent sheaves:
 
 \begin{cor}
-  Let $X$ be a projective variety over a field $k$, and $\F$ a coherent sheaf on $X$.
-  If $ H^0(\End \F) = k, $ then $\F$ is indecomposable.
+  Let $X$ be a projective variety over a field $k$, and $\cF$ a coherent sheaf on $X$.
+  If $ H^0(\End \cF) = k, $ then $\cF$ is indecomposable.
 \end{cor}
 
+\begin{proof}
+If there is a nontrivial direct sum decomposition $\cF = \cF_1\oplus \cF_2$, then the composing the projection and inclusions we obtain $\cF\to \cF_1\inc \cF$ yields a nonzero endomorphism of $\cF$ that is a scalar multiple of $\id_\cF$, and thus $H^0(\End \cF) \neq k$.
+\end{proof}
+
 \begin{rem}[Characteristic zero]
-If $R$ is a local or graded ring over a field $k\subset \C$ of characteristic 0, and $M$ a finitely generated $R$-module, it would be useful to have an algorithm to find a direct sum decomposition of $M$.
-At the moment, we do not have such an algorithm.
+If $R$ is a local ring over a field $k\subset \C$ of characteristic 0, and $M$ a finitely generated $R$-module, it would be useful to have an algorithm to find a direct sum decomposition of $M$.
+At the moment, we do not have such an algorithm outside the graded case.
 However, we note that the algorithm above can be used to test indecomposability of $M$ via reduction modulo $p$, as follows:
@@
 We also point out that it is possible to ``guess'' an idempotent for $M$, even when there is no algorithm to produce one.
-In practice, if $M$ does have nontrivial idempotents, Macaulay2 often chooses them as some of the generators of $[\End_R(M)]_0$.
-By checking if the generators of $[\End_R(M)]_0$ are idempotent, and trying to construct idempotents by subtracting eigenvalues off of general endomorphisms and checking idempotency, we can often produce nontrivial direct sum decompositions in characteristic 0.
-See Examples~\ref{example:QQ} for such a case.
+In practice, if $M$ does have nontrivial idempotents, Macaulay2 often chooses them as some of the generators of $\End_R(M)$.
+By checking if the generators of $\End_R(M)$ are idempotent, and trying to construct idempotents by subtracting eigenvalues off of general endomorphisms and checking idempotency, we can often produce nontrivial direct sum decompositions in characteristic 0.
+%See \Cref{example:QQ} for such a case.
+
+\devlin{example of this?}
 \end{rem}
@@
 
-\begin{lem}
+\begin{lem}{{\cite[Exercise~4.1]{Friedman}}}
 A morphism $f:E\to E$ is injective if and only if it is an isomorphism.
@@
 
-For completeness, we give the standard proof:
+%For completeness, we give the standard proof (see, e.g., \cite[Exercise~4.1]{Friedman}).
 
-\begin{proof}
-This requires only that $E$ is a coherent sheaf:
-% since $\End_{\O_X}(E)$ is finite-dimensional, $f$ satisfies some minimal monic polynomial $\phi(x)$; since $f$ is injective, the constant term of $\phi(x)$ is nonzero.
-We claim that $f\otimes k(x)$ is injective for any $k(x)$. This then implies that $f\otimes k(x)$ is surjective for every $x\in X$ and thus $f$ is surjective.
-To see this, note that $f$ satisfies some minimal-degree monic polynomial, since $\End_{\O_X}(E)$ is finite-dimensional over $k$, and that this monic polynomial has nonzero constant term, since $f$ is injective.
+%\begin{proof}
+%This requires only that $E$ is a coherent sheaf:
+%% since $\End_{\O_X}(E)$ is finite-dimensional, $f$ satisfies some minimal monic polynomial $\phi(x)$; since $f$ is injective, the constant term of $\phi(x)$ is nonzero.
+%We claim that $f\otimes k(x)$ is injective for any $k(x)$. This then implies that $f\otimes k(x)$ is surjective for every $x\in X$ and thus $f$ is surjective.
+%To see this, note that $f$ satisfies some minimal-degree monic polynomial, since $\End_{\O_X}(E)$ is finite-dimensional over $k$, and that this monic polynomial has nonzero constant term, since $f$ is injective.
 Now, $f\otimes k(x)$ satisfies this same polynomial, so must be injective.
-\end{proof}
+%\end{proof}
 
 \begin{dfn}
-For $f\in \End_{\O_X}(E)$ there an associated map
-$$ \bigwedge^{\rank E} f: \det E\to \det E; $$
-since $\End_{\O_X}(\det E) = H^0(\O_X)=k$, $\bigwedge^{\rank E} f$ is multiplication by $\l\in k$; we write $\det f$ for this scalar $\l$.
+For $f\in \End_{\O_X}(E)$, taking top exterior powers yields the map
+$$ \bigwedge^{\rank E} f: \det E  \to \det E, $$
+where $\det E := \bigwedge^{\rank E} E$.
+Since $\End_{\O_X}(\det E) = H^0(\O_X)=k$, $\bigwedge^{\rank E} f$ is multiplication by $\l\in k$; we write $\det f$ for this scalar $\l$.
 \end{dfn}
@@
 \begin{lem}
-Let $x \in X$ be any point (not necessarily closed). Then $\det(f\otimes k(x)) =\det f$.
+Let $x \in X$ be any point (not necessarily closed), with residue field $k(x) = \O_{X,x}/m_x$. Then $\det(f\otimes k(x)) =\det f$.
 %In other words, if we restrict $f$ to the fiber $E_x$, the determinant is the same as $\det f$.
@@
 \begin{proof}
-Taking the determinant commutes with restriction to a fiber, so $\det(f\otimes k(x)) =\det(f)\otimes k(x)$, but $\det(f)\in k$ already so is unaffected by going modulo $m_x$.
+Since $(\bigwedge^{\rank E} f)\otimes k(x) = \bigwedge^{\rank( E\otimes k(x))}( f\otimes k(x))$ as $k(x)$-linear maps on $E\otimes k(x)$,
+ we have $\det(f\otimes k(x)) =\det(f)\otimes k(x)$, but $\det(f)\in k$ and so is unaffected by going modulo $m_x$.
 \end{proof}
@@
 \begin{dfn}
-$\l \in k$ is an eigenvalue of $f \in \End_{\O_X}(E)$ if $\det(f-\l \id_E)= 0$.
+$\l \in k$ is an \defn{eigenvalue} of $f \in \End_{\O_X}(E)$ if $\det(f-\l \id_E)= 0$.
 In other words, the eigenvalues of $f$ are the zeroes of the univariate polynomial $\det(f-t\id _E)$ with coefficients in $k$.
@@
 
-Note that $f$ is an isomorphism if and only if $\l$ is not an eigenvalue of $f$, just as for ordinary linear operators on vector spaces.
+Note that $f$ is an isomorphism if and only if $\l=0$ is not an eigenvalue of $f$, just as for ordinary linear operators on vector spaces.
 
@@
 Let $M$ be a homogeneous module over a graded ring $R$ with $R_0=k$.
-Let $g \in \End_R(M)$. The \emph{irrelevant eigenvalues} of $g$ are defined to be the eigenvalues of the map of vector spaces $g\otimes R/m:M/mM \to M/mM$.
+Let $g \in \End_R(M)$. The \defn{irrelevant eigenvalues} of $g$ are defined to be the eigenvalues of the map of vector spaces $g\otimes R/m:M/mM \to M/mM$.
 \end{dfn}
@@
 \begin{rem}
-We note that it is known already by \cite[Proposition~15]{Atiyah57} that an endomorphism of an indecomposable vector $E$ cannot have two distinct eigenvalues.
+We note that it is known already by \cite[Proposition~15]{Atiyah57} that an endomorphism of an indecomposable vector bundle $E$ cannot have two distinct eigenvalues.
 The utility of the preceding lemma is in producing an explicit direct sum decomposition of $E$.
@@
   % https://arxiv.org/abs/1901.10956
-  Consider the Grassmannian $X = \mathrm{Gr(2,4)}$. We may work over the Cox ring $S$,
+  Consider the Grassmannian $X = \Gr(2,4)$. We may work over the Cox ring $S$,
   %\[ S = \bigoplus\!{}_{[D]\in\Pic{X}} \; \Gamma(X, \O(D)). \]
@@
 
-  For instance, consider $X = \mathrm{Bl}_4\P^2$, the blowup of $\P^2$ at 4 general points. We will working over the $\Z^5$-graded Cox ring
+  For instance, consider $X = \Bl_4\P^2$, the blowup of $\P^2$ at 4 general points. We will be working over the $\Z^5$-graded Cox ring
   \[ S = k[x_1,\dots,x_{10}]/\text{(five quadric Pl\"ucker relations)} \]
@@
   $$ F_* \O_X = \O_X\oplus M, $$
-  with $M$ indecomposable, and furthermore $F^*_e M$ remains indecomposable for all $e\geq 0$. In other words, the indecomposable decomposition of $F^e_* \O_X$ is
+  with $M$ indecomposable, and furthermore $F_*^e M$ remains indecomposable for all $e\geq 0$. In other words, the indecomposable decomposition of $F^e_* \O_X$ is
   $$ F_*^e \O_X \cong \O_X\oplus M\oplus F_* M\oplus\dots\oplus F_*^{e-1}M. $$
@@
 \end{exa}
+
+%\begin{exa}[]
+%  \mahrud{TODO: example about Auslander-Ritten quivers}
+%\end{exa}
 

### 2025-04-16 paper/main.tex
macro changes

@@
 We also describe applications of our algorithm for computations in characteristic 0.
-Further, our algorithm takes advantage of multigradings, which enables the computation of indecomposable summands of coherent sheaves on subvarieties of toric varieties (in particular, for varieties embedded in $\P^n$).
+Further, our algorithm takes advantage of multigradings, which enables the computation of indecomposable summands of coherent sheaves on subvarieties of toric varieties (in particular, for varieties embedded in $\PP^n$).
 
@@
 \begin{itemize}
-\item For a field $k$, we will write $\overline k$ for a choice of algebraic closure of $k$. In particular, $\overline{\mathbb F_p}$ is the algebraic closure of the finite field $\mathbb F_p$.
-\item For an algebraic variety $X$ and a point $x\in X$, we write $k(x)=\O_{X,x}/m_x$ for the residue field of $X$ at $x$.
+\item For a field $k$, we will write $\overline k$ for a choice of algebraic closure of $k$. In particular, $\overline{\FF_p}$ is the algebraic closure of the finite field $\FF_p$.
+\item For an algebraic variety $X$ and a point $x\in X$, we write $k(x)=\OO_{X,x}/m_x$ for the residue field of $X$ at $x$.
 \item We write $k(X)$ for the function field of $X$.
-\item If $X$ is a projective variety, and $\O_X(1)$ a fixed very ample line bundle on $X$, then for any coherent $\O_X$-module $\cF$ we write 
-$\Gamma_* \cF = \bigoplus_{n\geq 0} H(X,\cF(n))$. This is a module over the graded ring $S=\bigoplus_{n\geq 0} H^0(X,\O_X(n))$.
+\item If $X$ is a projective variety, and $\OO_X(1)$ a fixed very ample line bundle on $X$, then for any coherent $\OO_X$-module $\cF$ we write
+$\Gamma_* \cF = \bigoplus_{n\geq 0} H(X,\cF(n))$. This is a module over the graded ring $S=\bigoplus_{n\geq 0} H^0(X,\OO_X(n))$.
 \item Conversely, if $S$ is a graded ring defining a projective variety $X$, and $M$ is a finitely generated $S$-module, we write $\wtilde M$ for the coherent sheaf on $X$ associated to $M$.
@@
 \section{The main algorithms}
-Here, we describe the main algorithms for finding the indecomposable summands of a finitely generated $R$-module $M$, when $R$ is either a graded ring or a local ring with residue field contained in $\overline{\mathbb F_p}$.
+Here, we describe the main algorithms for finding the indecomposable summands of a finitely generated $R$-module $M$, when $R$ is either a graded ring or a local ring with residue field contained in $\overline{\FF_p}$.
 
@@
 %%the effective cone
-%$\Eff R = \{ \deg r : r\in R\}$ is a strongly convex cone in $\Z^n$.
+%$\Eff R = \{ \deg r : r\in R\}$ is a strongly convex cone in $\ZZ^n$.
 Let $M$ be a finitely generated graded $R$-module, and let $\phi \in \End M$ be a degree-0 endomorphism.
 
-%Let $P\subset \Z^n$ be the degrees of a set of minimal generators of $M$; 
+%Let $P\subset \ZZ^n$ be the degrees of a set of minimal generators of $M$;
 %this is a poset with respect to the partial order $\a\leq \b$ if and only if $\b-\a\in \Eff R$.
@@
 \begin{rem}[how to do it in multigraded case]
-Let $R$ be an $\Z^n$-graded ring, with $R_0=k$ an algebraically closed field.
+Let $R$ be an $\ZZ^n$-graded ring, with $R_0=k$ an algebraically closed field.
 Assume furthermore that the degree support of $R$ is strongly convex, i.e., that
 the effective cone
-$\Eff R = \{ \deg r : r\in R\}$ is a strongly convex cone in $\Z^n$.
+$\Eff R = \{ \deg r : r\in R\}$ is a strongly convex cone in $\ZZ^n$.
 Let $M$ be a finitely generated graded $R$-module, and let $\phi \in \End M$ be a degree-0 endomorphism.
 
-Let $P\subset \Z^n$ be the degrees of a set of minimal generators of $M$; 
+Let $P\subset \ZZ^n$ be the degrees of a set of minimal generators of $M$;
 this is a poset with respect to the partial order $\a\leq \b$ if and only if $\b-\a\in \Eff R$.
@@
 
-Throughout this section, $(R,m)$ will be either local with maximal ideal $m$, or a (multi)graded ring with $R_0 = k \subset \overline{\mathbb F_p}$ a field of positive characteristic and homogeneous maximal ideal $m$.
+Throughout this section, $(R,m)$ will be either local with maximal ideal $m$, or a (multi)graded ring with $R_0 = k \subset \overline{\FF_p}$ a field of positive characteristic and homogeneous maximal ideal $m$.
 
@@
   \end{pmatrix}. \]
-  Finally, since $k=\mathbb F_{p^e}$ for some $e$, if we choose $l$ to be divisible also by $p^e-1$, we have that $$\mu_i^{l-1} = (\mu_i^{p^{e}-1})^{l/(p^{e}-1)} =1.$$
+  Finally, since $k=\FF_{p^e}$ for some $e$, if we choose $l$ to be divisible also by $p^e-1$, we have that $$\mu_i^{l-1} = (\mu_i^{p^{e}-1})^{l/(p^{e}-1)} =1.$$
   Thus, $(A-\l I)^l$ is a diagonal matrix with diagonal entries 1 or 0, hence idempotent. Note moreover that if some $\l_i\neq \l$, then $(A-\l I)^l$ is not the zero matrix.
@@
 One can choose a finitely generated $\Z$-algebra $A$ and an $A$-algebra $R_A$ such that $R_A\otimes _A k=R$, and likewise an $R_A$-module $M_A$ such that $M_A\otimes _A k = R$ and such that $M_A$ is flat over $A$.
-If $n$ is a maximal ideal of $A$, one can check that $A/n \cong \mathbb F_{p^e}$ for some prime $p$ and $e$.
+If $n$ is a maximal ideal of $A$, one can check that $A/n \cong \FF_{p^e}$ for some prime $p$ and $e$.
 The key point is that if $M$ is decomposable, then $M_A$ also can be taken to be decomposable. Thus, the various reductions  $M\otimes A/n$ will also be decomposable for all $n$.
@@
 
-Let $X\subset \P^n$ be a projective variety, with ample line bundle $\O_X(1)$.
+Let $X\subset \P^n$ be a projective variety, with ample line bundle $\OO_X(1)$.
 Throughout, let $E$ be a vector bundle (i.e., a locally free coherent sheaf) on $X$.
@@
 %This requires only that $E$ is a coherent sheaf:
-%% since $\End_{\O_X}(E)$ is finite-dimensional, $f$ satisfies some minimal monic polynomial $\phi(x)$; since $f$ is injective, the constant term of $\phi(x)$ is nonzero.
+%% since $\End_{\OO_X}(E)$ is finite-dimensional, $f$ satisfies some minimal monic polynomial $\phi(x)$; since $f$ is injective, the constant term of $\phi(x)$ is nonzero.
 %We claim that $f\otimes k(x)$ is injective for any $k(x)$. This then implies that $f\otimes k(x)$ is surjective for every $x\in X$ and thus $f$ is surjective.
-%To see this, note that $f$ satisfies some minimal-degree monic polynomial, since $\End_{\O_X}(E)$ is finite-dimensional over $k$, and that this monic polynomial has nonzero constant term, since $f$ is injective.
+%To see this, note that $f$ satisfies some minimal-degree monic polynomial, since $\End_{\OO_X}(E)$ is finite-dimensional over $k$, and that this monic polynomial has nonzero constant term, since $f$ is injective.
 Now, $f\otimes k(x)$ satisfies this same polynomial, so must be injective.
@@
 \begin{dfn}
-For $f\in \End_{\O_X}(E)$, taking top exterior powers yields the map
+For $f\in \End_{\OO_X}(E)$, taking top exterior powers yields the map
 $$ \bigwedge^{\rank E} f: \det E  \to \det E, $$
 where $\det E := \bigwedge^{\rank E} E$.
-Since $\End_{\O_X}(\det E) = H^0(\O_X)=k$, $\bigwedge^{\rank E} f$ is multiplication by $\l\in k$; we write $\det f$ for this scalar $\l$.
+Since $\End_{\OO_X}(\det E) = H^0(\OO_X)=k$, $\bigwedge^{\rank E} f$ is multiplication by $\l\in k$; we write $\det f$ for this scalar $\l$.
 \end{dfn}
@@
 \begin{lem}
-Let $x \in X$ be any point (not necessarily closed), with residue field $k(x) = \O_{X,x}/m_x$. Then $\det(f\otimes k(x)) =\det f$.
+Let $x \in X$ be any point (not necessarily closed), with residue field $k(x) = \OO_{X,x}/m_x$. Then $\det(f\otimes k(x)) =\det f$.
 %In other words, if we restrict $f$ to the fiber $E_x$, the determinant is the same as $\det f$.
@@
 \begin{dfn}
-$\l \in k$ is an \defn{eigenvalue} of $f \in \End_{\O_X}(E)$ if $\det(f-\l \id_E)= 0$.
+$\l \in k$ is an \defn{eigenvalue} of $f \in \End_{\OO_X}(E)$ if $\det(f-\l \id_E)= 0$.
 In other words, the eigenvalues of $f$ are the zeroes of the univariate polynomial $\det(f-t\id _E)$ with coefficients in $k$.
@@
 \begin{prop}
-If $X$ is a variety over a field $k\subset \overline{\mathbb F_{p}}$,
+If $X$ is a variety over a field $k\subset \overline{\FF_{p}}$,
 $E$ a vector bundle on $X$, and $f:E\to E$ has two distinct eigenvalues $\l_1,\l_2$, then $E$ is a nontrivial direct sum, with one summand given by $(f-\l_1)^{p^e(p^e-1)}$ for some $e\geq 1$.
@@
     \text{When }p=3, n=2: \\
-    F_*\O_{\P^2} &= \O \oplus \O(-1)^7 \oplus \O(-2). \\
+    F_*\OO_{\P^2} &= \OO \oplus \OO(-1)^7 \oplus \OO(-2). \\
     \text{When } p=2, n=5: \\
-    F_*\O_{\P^5} &= \O \oplus \O(-1)^{15} \oplus \O(-2)^{15} \oplus \O(-3), \\
-    F_*^2\O_{\P^5} &= \O \oplus \O(-1)^{120} \oplus \O(-2)^{546} \oplus \O(-3)^{336} \oplus \O(-4)^{21}.
+    F_*\OO_{\P^5} &= \OO \oplus \OO(-1)^{15} \oplus \OO(-2)^{15} \oplus \OO(-3), \\
+    F_*^2\OO_{\P^5} &= \OO \oplus \OO(-1)^{120} \oplus \OO(-2)^{546} \oplus \OO(-3)^{336} \oplus \OO(-4)^{21}.
   \end{align*}
@@
   Let $X$ be a smooth toric variety and consider its Cox ring
-  \[ S = \bigoplus_{[D]\in\Pic{X}} \; \Gamma(X, \O(D)). \]
+  \[ S = \bigoplus_{[D]\in\Pic{X}} \; \Gamma(X, \OO(D)). \]
   Similar to the case of the projective space, B{\o}gvad and Thomsen \cite{Bogvad98,Thomsen00} showed that $F_*L$ totally splits as a direct sum of line bundles for any line bundle $L\in\Pic X$.
 
-  As an example, consider the third Hirzebruch surface $X=\P(\O_{\P^1}\oplus \O_{\P^1}(3))$ over a field of characteristic 3. We have, for example, that
+  As an example, consider the third Hirzebruch surface $X=\P(\OO_{\P^1}\oplus \OO_{\P^1}(3))$ over a field of characteristic 3. We have, for example, that
   \begin{align*}
-    F_*\O_X      &= \O_X   \oplus \O_X(-1,0)^2 \oplus \O_X(0,-1)^2 \oplus \O_X(1,-1)^3 \oplus \O_X(2,-1), \\
-    F_*\O_X(1,1) &= \O_X^3 \oplus \O_X(-1,0)   \oplus \O_X(1,-1)   \oplus \O_X(1, 0)^2 \oplus \O_X(2,-1)^2.
+    F_*\OO_X      &= \OO_X   \oplus \OO_X(-1,0)^2 \oplus \OO_X(0,-1)^2 \oplus \OO_X(1,-1)^3 \oplus \OO_X(2,-1), \\
+    F_*\OO_X(1,1) &= \OO_X^3 \oplus \OO_X(-1,0)   \oplus \OO_X(1,-1)   \oplus \OO_X(1, 0)^2 \oplus \OO_X(2,-1)^2.
   \end{align*}
@@
   Consider the elliptic curve
-  \[ X = \Proj \mathbb F_7[x,y,z]/(x^3+y^3+z^3). \]
+  \[ X = \Proj \FF_7[x,y,z]/(x^3+y^3+z^3). \]
   %%Weierstrauss form is y^2=x^3-432
-  This is an ordinary elliptic curve, hence $F$-split; thus $\O_X$ is a summand of $F_* \O_X$.
-Over the algebraic closure of $\mathbb F_7$, $F_*\O_X$ will decompose as $\bigoplus_{p=1}^7 \O_X(p_i)$, where $p_1,\dots,p_7$ are the 7-torsion points of $X$.
+  This is an ordinary elliptic curve, hence $F$-split; thus $\OO_X$ is a summand of $F_* \OO_X$.
+Over the algebraic closure of $\FF_7$, $F_*\OO_X$ will decompose as $\bigoplus_{p=1}^7 \OO_X(p_i)$, where $p_1,\dots,p_7$ are the 7-torsion points of $X$.
 
-  However, over $\mathbb F_7$, our algorithm calculates that $F_*\O_X$ decomposes only as
-  $$ F_* \O_X =\O_X \oplus M_1\oplus M_2\oplus M_3, $$
-  with $M_i$ indecomposable (over $\mathbb F_7$) of rank 2.
+  However, over $\FF_7$, our algorithm calculates that $F_*\OO_X$ decomposes only as
+  $$ F_* \OO_X =\OO_X \oplus M_1\oplus M_2\oplus M_3, $$
+  with $M_i$ indecomposable (over $\FF_7$) of rank 2.
 
-  If one extends the ground field to $\mathbb F_{49}$, however, our algorithm calculates the full decomposition
-  $$ F_* \O_X=\bigoplus_{p=1}^7 \O_X(p_i). $$
-  This reflects the fact that the 7-torsion points $p_i$ of $X$, and thus the sheaves $\O_X(p_i)$, are not defined over $\mathbb F_7$, but rather are defined over $\mathbb F_{49}$.
+  If one extends the ground field to $\FF_{49}$, however, our algorithm calculates the full decomposition
+  $$ F_* \OO_X=\bigoplus_{p=1}^7 \OO_X(p_i). $$
+  This reflects the fact that the 7-torsion points $p_i$ of $X$, and thus the sheaves $\OO_X(p_i)$, are not defined over $\FF_7$, but rather are defined over $\FF_{49}$.
 \end{exa}
@@
   Consider the Grassmannian $X = \Gr(2,4)$. We may work over the Cox ring $S$,
-  %\[ S = \bigoplus\!{}_{[D]\in\Pic{X}} \; \Gamma(X, \O(D)). \]
+  %\[ S = \bigoplus\!{}_{[D]\in\Pic{X}} \; \Gamma(X, \OO(D)). \]
   which in this case coincides with the coordinate ring
@@
   Then in characteristic $p=3$ we have:
-  \[ F_*\O_X = \O \oplus \O(-1)^{44} \oplus \O(-2)^{20} \oplus A^4 \oplus B^4, \]
+  \[ F_*\OO_X = \OO \oplus \OO(-1)^{44} \oplus \OO(-2)^{20} \oplus A^4 \oplus B^4, \]
   where $A$ and $B$ are rank-2 indecomposable bundles (c.f.~\cite{RSVdB22}).
@@
 
-  For instance, consider $X = \Bl_4\P^2$, the blowup of $\P^2$ at 4 general points. We will be working over the $\Z^5$-graded Cox ring
+  For instance, consider $X = \Bl_4\P^2$, the blowup of $\P^2$ at 4 general points. We will be working over the $\ZZ^5$-graded Cox ring
   \[ S = k[x_1,\dots,x_{10}]/\text{(five quadric Pl\"ucker relations)} \]
@@
     \begin{align*}
-      F_*^2\O_X = {\O_{X}^{1}}
-      &\oplus {\O_{X}^{2}\ \left(-2,\,1,\,1,\,1,\,1\right)} \oplus {\O_{X}^{2}\ \left(-1,\,0,\,0,\,0,\,1\right)} \\
-      &\oplus {\O_{X}^{2}\ \left(-1,\,0,\,0,\,1,\,0\right)} \oplus {\O_{X}^{2}\ \left(-1,\,0,\,1,\,0,\,0\right)} \\
-      &\oplus {\O_{X}^{2}\ \left(-1,\,1,\,0,\,0,\,0\right)} \oplus B \oplus G,
+      F_*^2\OO_X = {\OO_{X}^{1}}
+      &\oplus {\OO_{X}^{2}\ \left(-2,\,1,\,1,\,1,\,1\right)} \oplus {\OO_{X}^{2}\ \left(-1,\,0,\,0,\,0,\,1\right)} \\
+      &\oplus {\OO_{X}^{2}\ \left(-1,\,0,\,0,\,1,\,0\right)} \oplus {\OO_{X}^{2}\ \left(-1,\,0,\,1,\,0,\,0\right)} \\
+      &\oplus {\OO_{X}^{2}\ \left(-1,\,1,\,0,\,0,\,0\right)} \oplus B \oplus G,
     \end{align*}
@@
 \begin{exa}[Frobenius pushforward on cubic surfaces]
-  Let $X$ be a smooth cubic surface. Aside from a single exception in characteristic 0, $X$ will be globally $F$-split, so that any $F^e_*\O_X $ admits $\O_X$ as a direct summand.
-  The other summands of Frobenius pushforwards of $\O_X$ have yet to be studied, and in particular it is not known whether such rings should have the finite $F$-representation type property.
+  Let $X$ be a smooth cubic surface. Aside from a single exception in characteristic 0, $X$ will be globally $F$-split, so that any $F^e_*\OO_X $ admits $\OO_X$ as a direct summand.
+  The other summands of Frobenius pushforwards of $\OO_X$ have yet to be studied, and in particular it is not known whether such rings should have the finite $F$-representation type property.
 
   The use of our algorithm to compute examples in small $p$ and $e$ suggest the following behavior:
-  $$ F_* \O_X = \O_X\oplus M, $$
-  with $M$ indecomposable, and furthermore $F_*^e M$ remains indecomposable for all $e\geq 0$. In other words, the indecomposable decomposition of $F^e_* \O_X$ is
-  $$ F_*^e \O_X \cong \O_X\oplus M\oplus F_* M\oplus\dots\oplus F_*^{e-1}M. $$
-  In particular, this suggests $\O_X$ will fail to have the finite $F$-representation type property.
+  $$ F_* \OO_X = \OO_X\oplus M, $$
+  with $M$ indecomposable, and furthermore $F_*^e M$ remains indecomposable for all $e\geq 0$. In other words, the indecomposable decomposition of $F^e_* \OO_X$ is
+  $$ F_*^e \OO_X \cong \OO_X\oplus M\oplus F_* M\oplus\dots\oplus F_*^{e-1}M. $$
+  In particular, this suggests $\OO_X$ will fail to have the finite $F$-representation type property.
   In fact, we believe a similar description holds true for quartic del Pezzos.

### 2025-04-21 paper/main.tex
paper updates

@@
 \usepackage{notes,hyperref}
+\usepackage{mathtools}
 \usepackage{cleveref}
@@
 \def\ZZ{\mathbb Z}
-\def\Gr{\mathrm Gr}
-\def\Bl{\mathrm Bl}
+\makeop{Gr}
+\makeop{Bl}
 \makeop{Eff}
@@
 \newcommand{\devlin}[1]{{\color{red} \sf $\clubsuit$ Devlin: [#1]}}
+\def\n{\mathfrak n}
 
@@
 \begin{abstract}
-  The problems of finding isomorphism classes of indecomposable modules with certain properties, or determining the indecomposable summands of a module, are ubiquitous in commutative algebra, representation theory, topological data analysis, and other fields.\linebreak The purpose of this work is to describe and prove correctness of two practical algorithms for computing indecomposable summands of finitely generated modules over a finitely generated $k$-algebra $R$: the first algorithm applies in the (multi)graded case, which enables the computation of indecomposable summands of coherent sheaves on subvarieties of toric varieties (in particular, for varieties embedded in projective space); the second algorithm applies when $R$ is local and $k$ is a finite field, which opens the door to computing decompositions in singularity theory.
-We also present multiple examples, including some which present previously unknown phenomena regarding the behavior of summands of Frobenius pushforwards and syzygies over Artinian rings.
+  %The problems of finding isomorphism classes of indecomposable modules with certain properties, or determining the indecomposable summands of a module, are ubiquitous in commutative algebra, representation theory, topological data analysis, and other fields.
+  \devlin{We had the commented-out sentence in both the abstract and intro; I think we can just remove it from the abstract? If not, at least rephrase}
+  The purpose of this work is to describe and prove correctness of two practical algorithms for computing indecomposable summands of finitely generated modules over a finitely generated $k$-algebra $R$: the first algorithm applies in the (multi)graded case, which enables the computation of indecomposable summands of coherent sheaves on subvarieties of toric varieties (in particular, for varieties embedded in projective space); the second algorithm applies when $R$ is local and $k$ is a finite field, which opens the door to computing decompositions in singularity theory.
+We also present multiple examples, including some which present previously unknown phenomena regarding the behavior of summands of Frobenius pushforwards (including in the non-graded case) and syzygies over Artinian rings.
 \end{abstract}
@@
 \section{Introduction}
+\label{sec:intro}
 
 The problems of finding isomorphism classes of indecomposable modules with certain properties, or determining the indecomposable summands of a module, are ubiquitous in commutative algebra, group theory, representation theory, and other fields.
-Within commutative algebra, for instance, the classification of rings $R$ for which there are only finitely many isomorphism classes of indecomposable maximal Cohen--Macaulay $R$-modules (the \emph{finite CM-type} property), or determining whether iterated Frobenius pushforwards of a Noetherian ring in positive characteristic have finitely many isomorphism classes of indecomposable summands (the \emph{finite F-representation type} property) are two well-established research problems.
+Within commutative algebra, for instance, the classification of rings $R$ for which there are only finitely many isomorphism classes of indecomposable maximal Cohen--Macaulay $R$-modules (the \emph{finite CM-type} property), or determining whether iterated Frobenius pushforwards of a Noetherian ring in positive characteristic have finitely many isomorphism classes of indecomposable summands (the \emph{finite F-representation type} property) are two well-established research problems (see, for example, \cite{Yoshino90,LW12} for the former and \cite{SVdB97,Hara15,TT08}, among many others, for the latter).
 For both these problems, and many others, making and testing conjectures depends on computing summands of modules and verifying their indecomposability.
-\devlin{Reference 15.10.9 of \cite{Eisenbud95}}
+The need for, and absence of, an algorithm to do so is noted, for example, in
+\cite[15.10.9]{Eisenbud95}.
 
-Currently there are no efficient algorithms available for checking indecomposability or finding summands of arbitrary modules over commutative rings.
+Currently there are no efficient \devlin{would it be too bold to remove ``efficient''?} algorithms available for checking indecomposability or finding summands of arbitrary modules over commutative rings.
 In contrast, variants of the ``Meat-Axe'' algorithm for determining irreducibility of finite-dimensional modules over a group algebra have wide ranging applications in computational group theory \cite{Parker84,HR94,Holt98} and are available through symbolic algebra software such as Magma and GAP \cite{MAGMA,GAP}.
@@
 
-The purpose of this paper is to describe and prove correctness of a practical algorithm for computing indecomposable summands of finitely generated modules over a finitely generated local or graded $k$-algebra, for $k$ a field of positive characteristic.
-We also describe applications of our algorithm for computations in characteristic 0.
-Further, our algorithm takes advantage of multigradings, which enables the computation of indecomposable summands of coherent sheaves on subvarieties of toric varieties (in particular, for varieties embedded in $\PP^n$).
+The purpose of this paper is to describe and prove correctness of two practical algorithms for computing indecomposable summands of finitely generated modules over a finitely generated $k$-algebra $R$, when either $R$ is (multi)graded and $k$ an arbitrary field, or $R$ local and $k$ a field of positive characteristic.
+In the graded case, our algorithm is often capable of completely decomposing a module in a single pass, while in the local case it proceeds by recursively finding summands and then reapplying the algorithm to each summand.
+%We also describe applications of our algorithm for computations in characteristic 0.
+The multigraded case of our algorithm enables the computation of indecomposable summands of coherent \emph{sheaves} on subvarieties of toric varieties (in particular, for varieties embedded in $\PP^n$).
 
-After describing the algorithm and proving its correctness, we present multiple examples, including some which present previously unknown phenomena regarding the behavior of summands of Frobenius pushforwards and syzygies over Artinian rings.
+After describing the algorithms and proving their correctness, we present multiple examples, including some which present previously unknown phenomena regarding the behavior of summands of Frobenius pushforwards and syzygies over Artinian rings.
 In particular, we highlight the results of \cite{CDE24}, which shows a recurrence formula for indecomposable summands of high syzygies of the residue field of Golod rings, made possible through experiments and observations using our algorithm.
 
-An accompanying implementation in Macaulay2 \cite{M2} is available via the GitHub repository \\
+An implementation in Macaulay2 \cite{M2} is available via the GitHub repository \\
 \centerline{
@@
 \begin{rem}
-Although the algorithm described below is only proved to result (probabilistically) in a decomposition into indecomposable summands in positive characteristic, in practice it often does produce nontrivial indecomposable decompositions even in characteristic 0.
+  There is a ``missing'' case of our algorithm: when $R$ is only local, not graded, and $k$ has characteristic 0.
+  If a module over a ring of characteristic 0 is decomposable, its reductions modulo $p$ will be as well; thus, our algorithm at least provides a heuristic for verifying decomposability in characteristic 0.
+See \Cref{rem:char}.
+  %Although the local algorithm described below is only proved to result (probabilistically) in a decomposition into indecomposable summands in positive characteristic, in practice it often does produce nontrivial indecomposable decompositions even in characteristic 0.
+  %Although the algorithm described below is only proved to result (probabilistically) in a decomposition into indecomposable summands in positive characteristic, in practice it often does produce nontrivial indecomposable decompositions even in characteristic 0.
 Moreover, if a module over a ring of characteristic 0 is decomposable, its reductions modulo $p$ will be as well; thus, our algorithm provides a heuristic for verifying decomposability in characteristic 0.
 
-  We note that while the discussion below, and our implementation in Macaulay2, concerns the case where $R$ is a \emph{commutative} ring, the basic techniques extend beyond this case.
+  We note also that while the discussion below, and our implementation in Macaulay2, concerns the case where $R$ is a \emph{commutative} ring, the basic techniques extend beyond this case.
 We plan to extend the results and algorithms to non-commutative rings, such as Weyl algebras, in future work.
@@
 We also thank the American Institute of Mathematics for hosting the ``Macaulay2: expanded functionality and improved efficiency'' workshop, during which some of this work took place.
-The work of the first author was supported in by the National Science Foundation under Grant No.~1840190.
+The work of the first author was supported in by the National Science Foundation RTG Grant No.~1840190.
 The second author was supported in part by the Doctoral Dissertation Fellowship at the University of Minnesota and the National Science Foundation under Grant No.~2001101.
-
 We also thank the referees for their careful reading and suggestions, which significantly improved the exposition of the paper.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Notation}
+\label{sec:notation}
 
-\section{Notations}
+Throughout we will work over a field $k$ and write $\overline k$ for a choice of algebraic closure of $k$.
+\linebreak In particular, $\overline{\FF_p}$ is the algebraic closure of the finite field $\ZZ/p\ZZ$.
 
-\begin{itemize}
-\item For a field $k$, we will write $\overline k$ for a choice of algebraic closure of $k$. In particular, $\overline{\FF_p}$ is the algebraic closure of the finite field $\FF_p$.
-\item For an algebraic variety $X$ and a point $x\in X$, we write $k(x)=\OO_{X,x}/m_x$ for the residue field of $X$ at $x$.
-\item We write $k(X)$ for the function field of $X$.
-\item If $X$ is a projective variety, and $\OO_X(1)$ a fixed very ample line bundle on $X$, then for any coherent $\OO_X$-module $\cF$ we write
-$\Gamma_* \cF = \bigoplus_{n\geq 0} H(X,\cF(n))$. This is a module over the graded ring $S=\bigoplus_{n\geq 0} H^0(X,\OO_X(n))$.
-\item Conversely, if $S$ is a graded ring defining a projective variety $X$, and $M$ is a finitely generated $S$-module, we write $\wtilde M$ for the coherent sheaf on $X$ associated to $M$.
-\item For a graded or local ring $(R,m)$ and an $R$-module $M$, we write $\mu(M):=\dim_{R/m} M/mM$ for the minimal number of generators of $M$ as an $R$-module.
-\end{itemize}
+For a graded or local ring $R$ with maximal ideal $\m$ and an $R$-module $M$, we write $\mu(M):=\dim_{R/\m}(M/\m M)$ for the minimal number of generators of $M$ as an $R$-module.
+
+% TODO: \mahrud{k(x) and k(X) are used only in one place, move there?}
+For an algebraic variety $X$ and a point $x\in X$, we write $k(x)=\OO_{X,x}/\m_x$ for the residue field of $X$ at $x$ and $k(X)$ for the function field of $X$.
+%
+When $X$ is a projective variety and $\OO_X(1)$ is a fixed very ample line bundle on $X$, the twisted global sections functor $\Gamma_*$ is the functor given by $\cF \mapsto \bigoplus_{n\in \Z} H^0(X, \cF(n))$, sending coherent sheaves to modules over the graded ring $S = \Gamma_*(\OO_X)$.
+Conversely, if $M$ is a finitely generated, graded $S$-module, we write $\wtilde M$ for the coherent sheaf associated to $M$.
 
@@
 \section{The main algorithms}
-Here, we describe the main algorithms for finding the indecomposable summands of a finitely generated $R$-module $M$, when $R$ is either a graded ring or a local ring with residue field contained in $\overline{\FF_p}$.
+\label{sec:main-algs}
+
+We begin by describing the main algorithms for finding the indecomposable summands of a finitely generated module over a graded ring or a local ring with residue field contained in $\overline{\FF_p}$.
 
 \subsection{The graded algorithm}\label{sec:main-alg-new}
-Let $R$ be a $\N$-graded ring, with $R_0=k$ an algebraically closed field.
+
+Let $R$ be a $\N$-graded ring, with $R_0 = k$, an algebraically closed field.
 %Assume furthermore that the degree support of $R$ is strongly convex, i.e., that
@@
 %$\Eff R = \{ \deg r : r\in R\}$ is a strongly convex cone in $\ZZ^n$.
-Let $M$ be a finitely generated graded $R$-module, and let $\phi \in \End M$ be a degree-0 endomorphism.
+Let $M$ be a finitely generated, graded $R$-module and consider a degree zero endomorphism $\phi \in [\End M]_0$.
 
@@
 
-\begin{lem}
-\label{graded_algorithm}
-Let $n=\mu(M)$ and set $\psi=\phi^N$. Then $M=\im \psi \oplus \coker \psi$.
+\begin{lem}\label{graded_algorithm}
+  Let $N = \mu(M)$ and set $\psi = \phi^N$. Then $M = \im\psi \oplus \coker\psi$.
 \end{lem}
+\begin{proof}
+  Fix homogeneous generators $m_1,\dots, m_n$ of $M$, with $\deg m_i$ nonincreasing.
+  The matrix for $\phi$ is then ``block upper triangular'' with diagonal blocks with entries in $k$, i.e., of the form
+  \[ \phi = \begin{pmatrix}
+    B_1    & *      & \dots  & *      \\
+    0      & B_2    & \dots  & *      \\
+    \vdots & \vdots & \ddots & \vdots \\
+    0      & 0      & \dots  & B_l
+  \end{pmatrix}, \]
+  where the blocks $B_i$ are square matrices \emph{with entries in $k$}, and the remaining entries are in $R$.
+To see this, note that since $\phi$ has degree zero, $\phi(m_i)$ is a $k$-linear combination of the $m_j$ with $\deg m_j = \deg m_i$ plus an $R$-linear combination of the $m_j$ with $\deg m_j < \deg m_i$.
 
-\begin{proof}
-Fix homogeneous generators $m_1,\dots,m_n$ of $M$, with $\deg m_i$ nonincreasing.
-The matrix for $\phi$ is then ``block upper triangular'' with diagonal blocks with entries in $k$, i.e., of the form
-$$
-\phi=
-\begin{pmatrix}
-  B_1 & * & \dots& * \\
-  0 & B_2 & \dots &* \\
-\vdots &\vdots & \ddots &\vdots \\
-0 & 0 & \dots &B_l
-\end{pmatrix},
-$$
-where all blocks labeled $B_i$ are arbitrary matrices \emph{with entries in $k$} and the $*$ are arbitrary.
-To see this, just note that since $\phi$ has degree 0, $\phi(m_i)$ is a $k$-linear combination of the $m_j$ with $\deg m_j=\deg m_i$ plus an $R$-linear combination of the $m_j$ with $\deg m_j<\deg m_i$.
-
-Thus, since each matrix $B_i$ has entries in the algebraically closed field $k$, by performing row and column operations over $k$ (which correspond to replacing each $m_i$ with a $k$-linear combination of $m_j$ with $\deg m_i = \deg m_j$, which is an endomorphism of $M$) we can put each $B_i$ into Jordan canonical form $J_i$, with each Jordan block having entries in $k$.
+  Thus, since each matrix $B_i$ has entries in the algebraically closed field $k$, by performing row and column operations over $k$ we can put each $B_i$ into Jordan canonical form $J_i$, with each Jordan block having entries in $k$.
+This corresponds to replacing each generator $m_i$ with a $k$-linear combination of generators $m_j$ with $\deg m_i = \deg m_j$, which is an endomorphism of $M$.
 By doing so starting with $B_1$ and proceeding to $B_l$, the resulting matrix has the form
@@
     \begin{pmatrix}
-  J_1 & ** & \dots& ** \\
-  0 & J_2 & \dots &** \\
-\vdots &\vdots & \ddots &\vdots \\
+      J_1    & *      & \dots  & *      \\
+      0      & J_2    & \dots  & *      \\
+      \vdots & \vdots & \ddots & \vdots \\
       0      & 0      & \dots  & J_l
@@
   \end{equation}
-with each $J_i$ a Jordan block with entries in $k$ and eigenvalue $\l_i$, and $**$ arbitrary.
+  where each $J_i$ is a Jordan block over $k$ with eigenvalue $\l_i$, and the remaining entries are arbitrary.
   Relabeling, we may assume that $\l_i\neq 0$ for $i=1,\dots,l_0$ and $\l_i=0$ for $i=l_0+1,\dots,l$.
@@
   Now, consider the map $ \phi^N: \im(\phi^N) \to \im(\phi^N)$.
-If this map is surjective, then $\im(\phi^N)$ is a direct summand of $M$, with complementary summand $\coker \phi^N$: to see this, observe that the injection $\im(\phi^N) \inc M$ would then  compose with $\phi^N$ to be a surjection, and since $\im(\phi^N)$ is a finitely generated module, a surjection must be an isomorphism $\a:\im \phi^N \to \im \phi^N$.
+If this map is surjective, then $\im(\phi^N)$ is a direct summand of $M$, with complementary summand $\coker \phi^N$: to see this, observe that the injection $\im(\phi^N) \inc M$ would then  compose with $\phi^N$ to be a surjection, and since $\im(\phi^N)$ is a finitely generated module, a surjection must be an isomorphism $\a:\im \phi^N \to \im \phi^N$ by \cite[Corollary~4.4]{Eisenbud95}.
 Thus, we obtain a splitting of the inclusion $\im \phi^N \inc M$ via $\a\inv\circ \phi^N$; since $\a\inv$ is an isomorphism, we have that $\im \phi^N$ is (isomorphic to) a direct summand of $M$, with complementary summand $\coker \phi^N$.
@@
   $\Eff R = \{ \deg r : r\in R\}$ is a strongly convex cone in $\ZZ^n$.
-Let $M$ be a finitely generated graded $R$-module, and let $\phi \in \End M$ be a degree-0 endomorphism.
+  Let $M$ be a finitely generated graded $R$-module, and let $\phi \in \End M$ be a degree zero endomorphism.
 
@@
 
-\begin{algm} Given a finitely generated graded $R$-module $M$, the indecomposable summands of $M$ are computed as follows.
+\begin{algm}
+  \label{alg:graded}
+  Given a finitely generated graded $R$-module $M$, the indecomposable summands of $M$ are computed as follows.
   \begin{enumerate}
-  \item\label{item:End0} Take a random element $\phi_0$ of $[\End_R(M)]_0$, the degree-0 part of $\End_R(M)$.
+  \item\label{item:End0} Take a random element $\phi_0$ of $[\End_R(M)]_0$, the degree zero part of $\End_R(M)$.
   \item For an eigenvalue $\mu $ of $\phi_0$, let $\phi=\phi_0-\mu \id_M$.
@@
 
-We will see via \Cref{lemma_term} that this will result in an indecomposable decomposition of $M$.
+We will see via \Cref{lem:term} that this will result in an indecomposable decomposition of $M$.
 
@@
 
-Throughout this section, $(R,m)$ will be either local with maximal ideal $m$, or a (multi)graded ring with $R_0 = k \subset \overline{\FF_p}$ a field of positive characteristic and homogeneous maximal ideal $m$.
+Throughout this section, $(R,\m)$ will be either local with maximal ideal $\m$, or a (multi)graded ring with $R_0 = k \subset \overline{\FF_p}$ a field of positive characteristic and homogeneous maximal ideal $\m$.
 
@@
 If $A$ is neither an isomorphism nor the zero morphism, then both factors are nonzero and $M$ is decomposable.
-Note that $A$ also acts on the $k$-vector space $M/mM$.
+Note that $A$ also acts on the $k$-vector space $M/\m M$.
 
@@
 \begin{lem}\label{lem:idemp}
-  Let $M$ be a finitely generated $R$-module, and let $A\in\End_R(M)$. If the induced action of $A$ on $M/mM$ is idempotent, then $M$ admits a direct sum decomposition $\im A \oplus \coker A$.
+  Let $M$ be a finitely generated $R$-module, and let $A\in\End_R(M)$. If the induced action of $A$ on $M/\m M$ is idempotent, then $M$ admits a direct sum decomposition $\im A \oplus \coker A$.
 \end{lem}
 \begin{proof}
-  By assumption, we can write $A^2 = A + B$, where $B\in\End_R(M)$ with $B(M)\subset mM$.
-  Note that if $n \in m^kM$, then $A^2(n) - A(n) = B(n)$ lies in $m^{k+1}M$.
+  By assumption, we can write $A^2 = A + B$, where $B\in\End_R(M)$ with $B(M)\subset \m M$.
+  Note that if $x \in \m^kM$, then $A^2(x) - A(x) = B(x)$ lies in $\m^{k+1}M$.
 
@@
   \[ n_0 = A(n_0) - n_1 \]
-  for some $n_1\in mM$. In fact, since $n_0$ and $A(n_0)$ are both in $N$, we have $n_1\in N$ also, so $n_1\in mM\cap N$.
+  for some $n_1\in \m M$. In fact, since $n_0$ and $A(n_0)$ are both in $N$, we have $n_1\in N$ also, so $n_1\in \m M\cap N$.
 
   Thus, we can write $n_1 = A(m_2)$ for $m_2\in M$.
-  Now, apply $A$ to both sides: by the assumption that $A$ is idempotent modulo $m$, we have
+  Now, apply $A$ to both sides: by the assumption that $A$ is idempotent modulo $\m$, we have
   \[ A(n_1)=A^2(m_2) = A(m_2) + n_2 = n_1+n_2, \]
-  Thus, $n_2=A(n_1)-n_1$, so $n_2\in m^2M$; clearly also $n_2\in N$ as well.
+  Thus, $n_2=A(n_1)-n_1$, so $n_2\in \m^2M$; clearly also $n_2\in N$ as well.
   Combining the previous equations, we can write
   \[ n_0 = A(n_0) - n_1 = A(n_0) - A(n_1) + n_2 = A(n_0 - n_1) + n_2, \]
-  with $n_1\in mM\cap N$ and $n_2\in m^2M\cap N$.
+  with $n_1\in \m M\cap N$ and $n_2\in \m^2M\cap N$.
 
@@
   \[ n_0=A(n_0-n_1+\dots \pm n_k) \mp n_{k+1}, \]
-  with $n_i \in m^i M\cap N$.
+  with $n_i \in \m^i M\cap N$.
 
-  By the Artin--Rees lemma, there's some positive integer $k$ such that for $n\gg0$ we can write
-  \[ m^n M\cap N = m^{n-k} ( m^kM\cap N)\subset m^{n-k} N. \]
-  That is, the terms of $n_0-n_1+n_2-\cdots$ are going to 0 in the $m$-adic topology on $N$. Hence we can write
+  By the Artin--Rees lemma \cite[Lemma~5.1]{Eisenbud95}, there's some positive integer $k$ such that for $n\gg0$ we can write
+  \[ \m^n M\cap N = \m^{n-k} ( \m^kM\cap N)\subset \m^{n-k} N. \]
+  That is, the terms of $n_0-n_1+n_2-\cdots$ are going to 0 in the $\m$-adic topology on $N$. Hence we can write
   \[ n_0=A(n_0-n_1+n_2-\cdots), \]
@@
 
-Thus, if we produce an element $A\in \End_R(M)$ that is an idempotent modulo $m$, we obtain a splitting of $M$.
-The following lemma allows us to produce idempotents modulo $m$.
+Thus, if we produce an element $A\in \End_R(M)$ that is an idempotent modulo $\m$, we obtain a splitting of $M$.
+The following lemma allows us to produce idempotents modulo $\m$.
 
@@
   \end{pmatrix}. \]
-  If we choose $l > r_i$ and $l$ a power of $p$, then all non-diagonal terms will vanish, so all blocks will have the form
+  If we choose $l > r_i$ and $l$ a power of $p$, say $p^{e_0}$, then all non-diagonal terms will vanish, so all blocks will have the form
   \[ \begin{pmatrix}
@@
   \end{pmatrix}. \]
-  Finally, since $k=\FF_{p^e}$ for some $e$, if we choose $l$ to be divisible also by $p^e-1$, we have that $$\mu_i^{l-1} = (\mu_i^{p^{e}-1})^{l/(p^{e}-1)} =1.$$
-  Thus, $(A-\l I)^l$ is a diagonal matrix with diagonal entries 1 or 0, hence idempotent. Note moreover that if some $\l_i\neq \l$, then $(A-\l I)^l$ is not the zero matrix.
+  Finally, since $k\subset \FF_{p^e}$ for some $e$, if we choose $l$ to be divisible also by $p^e-1$, we have that $$\mu_i^{l-1} = (\mu_i^{p^{e}-1})^{l/(p^{e}-1)} =1.$$
+  In particular, taking $l=p^{e_0}(p^e-1)$,
+  $(A-\l I)^{p^{e_0}(p^e-1)}$ is a diagonal matrix with diagonal entries 1 or 0, hence idempotent. Note moreover that if some $\l_i\neq \l$, then $(A-\l I)^l$ is not the zero matrix.
 \end{proof}
@@
 
-\begin{algm}\label{local_algorithm}
+\begin{algm}\label{alg:local}
   Given a finitely generated graded $R$-module $M$, the indecomposable summands of $M$ are computed as follows.
   \begin{enumerate}
-%In the graded case, take a general element $A_0$ of $[\End_R(M)]_0$, the degree-0 part of $\End_R(M)$; in the local case,
-\item Take a general element $A_0$ of $\End_R(M) \setminus m \End_R(M)$.
-  \item Consider the resulting endomorphism $A$ of the $k$-vector space $M/mM$ and find the eigenvalues of $A$.
-  \item If $A$ has at least two eigenvalues, choose one eigenvalue $\l$, and compute a sufficiently high power $B=A^{p^e(p^{e}-1)}$ of $A$ (with the power explicitly as in the proof of \Cref{lem:jordan}).
+    %In the graded case, take a general element $A_0$ of $[\End_R(M)]_0$, the degree zero part of $\End_R(M)$; in the local case,
+  \item Take a general element $A_0$ of $\End_R(M) \setminus \m \End_R(M)$.
+  \item Consider the resulting endomorphism $A$ of the $k$-vector space $M/\m M$ and find the eigenvalues of $A$.
+  \item If $A$ has at least two eigenvalues, choose one eigenvalue $\l$, and compute a sufficiently high power $B=A^{p^{e_0}(p^{e}-1)}$ of $A$ (with the power explicitly as in the proof of \Cref{lem:jordan}).
 This power will be a nonzero idempotent, and thus produce a splitting of $M$ as $\im B \oplus \coker B$ by \Cref{lem:idemp}.
@@
 
-As a consequence of \Cref{graded_algorithm} and \Cref{local_algorithm} above, we have:
+As a consequence of \Cref{alg:graded} and \Cref{alg:local} above, we have:
 
 \begin{lem}
-Assume that either $(R,\m)$ is a graded ring with $R_0=k$, or $(R,m)$ is a local ring with $k=R/m$ a field of positive characteristic, and that $M$ is a finitely generated $R$-module (homogeneous in the graded case).
-If $M$ has an endomorphism with at least two distinct eigenvalues modulo $m$, then $M$ is not indecomposable.
+  \label{lem:term}
+  Assume that either $(R,\m)$ is a graded ring with $R_0=k$, or $(R,\m)$ is a local ring with $k=R/\m$ a field of positive characteristic, and that $M$ is a finitely generated $R$-module (homogeneous in the graded case).
+  If $M$ has an endomorphism with at least two distinct eigenvalues modulo $\m$, then $M$ is not indecomposable.
 \end{lem}
@@
 \begin{lem}
-\label{lemma_term}
-Retain the notation above.
-  If $M$ is not indecomposable, then a general endomorphism of $M$ (of degree 0 in the graded case) reduces to an endomorphism of $M/mM$ with at least two distinct eigenvalues.
-\devlin{need to separate into graded and local cases maybe}
+  \label{lem:distinct}
+  Retain the notation of \Cref{lem:term}.
+  \begin{itemize}
+  \item If $R$ is graded and $M$ is not indecomposable, then a general endomorphism of $M$ of degree zero has at least two distinct eigenvalues (necessarily in $k$)
+  \item If $R$ is local and $M$ is not indecomposable, then a general endomorphism of $M$ of degree zero reduces to an endomorphism of $M/\m M$ with at least two distinct eigenvalues.
+  \end{itemize}
 \end{lem}
 
-In particular, if $M$ is not indecomposable, then \Cref{graded_algorithm} and \Cref{local_algorithm} produce a nontrivial direct sum decomposition of $M$, because a general element of $\End M$ will have distinct eigenvalues.
+\begin{rem}
+  By ``general'' we mean that a general linear combination of a basis for $[\End_R(M)]_0$ (in the graded case) or of minimal generators of $\End_R(M)$ (in the local case) over the algebraic closure of $k$, or equivalently over a sufficiently large algebraic extension of~$k$.
+\end{rem}
 
-\begin{rem}
-  By ``general'' we mean that a general linear combination of a basis for $[\End_R(M)]_0$ over the algebraic closure of $k$, or equivalently over a sufficiently large algebraic extension of $k$.
-\end{rem}
+In particular, if $M$ is not indecomposable, then \Cref{alg:graded} and \Cref{alg:local} produce a nontrivial direct sum decomposition of $M$, because a general element of $\End M$ will have distinct eigenvalues and thus the algorithms will produce a nontrivial decomposition.
 
@@
   We may assume that the base field $k$ is algebraically closed.
-  Let $\Phi_1,\dots,\Phi_r$ be a basis for $\End_R(M)$, and $\phi_1,\dots,\phi_r$ their images modulo $m$, which we view as matrices with entries in $k$.
-  Let $U\subset \A^r$ be the subset of $r$-tuples $(\l_1,\dots,\l_r)$ such that $\l_1\phi_1+\dots+\l_r\phi_r$ has at least two distinct eigenvalues, i.e., such that $\l_1\Phi_1+\dots+\l_r\Phi_r$ reduces to an endomorphism of $M/mM$ with at least two distinct eigenvalues.
+  Let $\Phi_1,\dots,\Phi_r$ be a basis for $[\End_R(M)]_0$ in the graded case, or minimal generators of $\End_R(M)$ in the local case, and $\phi_1,\dots,\phi_r$ their images modulo $\m$, which we view as matrices with entries in $k$.
+  Note that if $M$ is decomposable, say $M=M_1\oplus M_2$, then we may always take $\Phi_1,\Phi_2$ to be projectors onto each summand.
+  Let $U\subset \A^r$ be the subset of $r$-tuples $(\l_1,\dots,\l_r)$ such that $\l_1\phi_1+\dots+\l_r\phi_r$ has at least two distinct eigenvalues, i.e., such that $\l_1\Phi_1+\dots+\l_r\Phi_r$ reduces to an endomorphism of $M/\m M$ with at least two distinct eigenvalues.
+(Note that in the graded case the eigenvalues of the reduction of an endomorphism modulo $\m$ are the same as the eigenvalues of endomorphism, so there is no harm in looking only at $M/\m M$.)
 
@@
 Choose $\Phi_1$ to be the projection to $M_1$, and $\Phi_2$ the projection to $M_2$.
-Then for any $\l_1,\l_2\in k$, $\l_1\phi_1 + \l_2\phi_2$ has eigenvalues $\l_1,\l_2$; thus in particular there is \emph{an} element of $[\End_R(M)]_0$ reducing to an endomorphism of $M/mM$ with distinct eigenvalues, so $U$ is nonempty.
+Then for any $\l_1,\l_2\in k$, $\l_1\phi_1 + \l_2\phi_2$ has eigenvalues $\l_1,\l_2$; thus in particular there is \emph{an} element of $\End_R(M)$ reducing to an endomorphism of $M/\m M$ with distinct eigenvalues, so $U$ is nonempty.
 
@@
 \begin{rem}
-  Note that the above algorithm is quite sensitive to the ground field $k$, because it needs all the eigenvalues of the endomorphism $A$ of $M/mM$ to be contained in $k$.
+  Note that the above algorithm is quite sensitive to the ground field $k$, because it needs all the eigenvalues of the endomorphism $A$ of $M/\m M$ to be contained in $k$.
 While theoretically the issue can be avoided by working over an algebraically closed ground field $\bar k$, for practical use on a computer algebra system it is better to extend $k$ to some larger finite field.
@@
 \begin{lem}
-  Suppose $R$ is graded, and let $M$ be a finitely generated $R$-module and let $[\End_R(M)]_0$ be the $k$-vector space of degree-0 endomorphisms. Suppose that either:
+  Suppose $R$ is graded, and let $M$ be a finitely generated $R$-module and let $[\End_R(M)]_0$ be the $k$-vector space of degree zero endomorphisms. Suppose that either:
   \begin{enumerate}
   \item $[\End_R(M)]_0$ is 1-dimensional and thus spanned by the identity endomorphism, or
-  \item every non-identity element of $[\End_R(M)]_0$, viewed as a matrix, has entries in $m$;
+  \item every non-identity element of $[\End_R(M)]_0$, viewed as a matrix, has entries in $\m$;
   \end{enumerate}
@@
 \begin{proof}
-  If $M$ decomposes non-trivially as $M_1\oplus M_2$, then the projections onto each factor are nontrivial degree-0 endomorphisms not equal to the identity, and which do not have entries contained in~$m$.
+  If $M$ decomposes non-trivially as $M_1\oplus M_2$, then the projections onto each factor are nontrivial degree zero endomorphisms not equal to the identity, and which do not have entries contained in~$\m$.
 \end{proof}
@@
 
-\begin{rem}[Characteristic zero]
+\begin{rem}
+  \label{rem:char}
   If $R$ is a local ring over a field $k\subset \C$ of characteristic 0, and $M$ a finitely generated $R$-module, it would be useful to have an algorithm to find a direct sum decomposition of $M$.
@@
   One can choose a finitely generated $\Z$-algebra $A$ and an $A$-algebra $R_A$ such that $R_A\otimes _A k=R$, and likewise an $R_A$-module $M_A$ such that $M_A\otimes _A k = R$ and such that $M_A$ is flat over $A$.
-If $n$ is a maximal ideal of $A$, one can check that $A/n \cong \FF_{p^e}$ for some prime $p$ and $e$.
-The key point is that if $M$ is decomposable, then $M_A$ also can be taken to be decomposable. Thus, the various reductions  $M\otimes A/n$ will also be decomposable for all $n$.
-If our algorithm does not detect an indecomposable summand of $M\otimes A/n$ for an appropriate choice of $n$, then, the original module $M$ must have been indecomposable.
-However, if $M$ is \emph{decomposable}, we do not have a way of patching the decompositions of various $M_A\otimes_A A/n$ into a decomposition of $M$.
+If $\n$ is a maximal ideal of $A$, one can check that $A/\n \cong \FF_{p^e}$ for some prime $p$ and $e$.
+  The key point is that if $M$ is decomposable, then $M_A$ also can be taken to be decomposable. Thus, the various reductions  $M\otimes A/\n$ will also be decomposable for all $\n$.
+  If our algorithm does not detect an indecomposable summand of $M\otimes A/\n$ for an appropriate choice of $\n$, then, the original module $M$ must have been indecomposable.
+  However, if $M$ is \emph{decomposable}, we do not have a way of patching the decompositions of various $M_A\otimes_A A/\n$ into a decomposition of $M$.
 
@@
 By checking if the generators of $\End_R(M)$ are idempotent, and trying to construct idempotents by subtracting eigenvalues off of general endomorphisms and checking idempotency, we can often produce nontrivial direct sum decompositions in characteristic 0.
-%See \Cref{example:QQ} for such a case.
-
-\devlin{example of this?}
 \end{rem}
@@
   We omit a computational complexity analysis here because without any assumptions about the module there are too many parameters that in any given application may become prominent.
-We note, however, that $[\End_R(M)]_0$ can be computed using degree limited syzygy algorithms without computing all of $\End_R(M)$.
+We note, however, that $[\End_R(M)]_0$ can be computed using degree-limited syzygy algorithms without computing all of $\End_R(M)$.
 While in many applications this significantly reduces the computational complexity, in general the complexity of the algorithm presented here is at least bounded by the complexity of computing syzygies.
@@
 \section{Decomposing coherent sheaves}
+\label{sec:coherent}
 
 While the preceding section was written in the language of modules, by the standard translation to global (multi)projective varieties, the algorithm works equally well to find indecomposable decompositions of coherent sheaves on (multi)projective varieties.
-In this section, we make a few notes regarding the relation between the eigenvalues discussed in the preceding section (in this section, called \emph{irrelevant eigenvalues}) with the notion of eigenvalue of an endomorphism of a vector bundle.
+In this section, we make a few notes in the single-graded case regarding the relation between the eigenvalues discussed in the preceding section (in this section, called \emph{module-theoretic eigenvalues}) with the notion of eigenvalue of an endomorphism of a vector bundle.
 
@@
 
-\begin{lem}{{\cite[Exercise~4.1]{Friedman}}}
+\begin{lem}{{\cite[Exercise~4.1]{Friedman98}}}
   A morphism $f:E\to E$ is injective if and only if it is an isomorphism.
@@
 \begin{lem}
-Let $x \in X$ be any point (not necessarily closed), with residue field $k(x) = \OO_{X,x}/m_x$. Then $\det(f\otimes k(x)) =\det f$.
+  Let $x \in X$ be any point (not necessarily closed), with residue field $k(x) = \OO_{X,x}/\m_x$. Then $\det(f\otimes k(x)) =\det f$.
   %In other words, if we restrict $f$ to the fiber $E_x$, the determinant is the same as $\det f$.
@@
   Since $(\bigwedge^{\rank E} f)\otimes k(x) = \bigwedge^{\rank( E\otimes k(x))}( f\otimes k(x))$ as $k(x)$-linear maps on $E\otimes k(x)$,
- we have $\det(f\otimes k(x)) =\det(f)\otimes k(x)$, but $\det(f)\in k$ and so is unaffected by going modulo $m_x$.
+  we have $\det(f\otimes k(x)) =\det(f)\otimes k(x)$, but $\det(f)\in k$ and so is unaffected by going modulo $\m_x$.
 \end{proof}
@@
 \begin{proof}
-  Let $\ker f \neq 0$. %Then we have $0\to K\to E \to \im f\to0$;
- Then $\det(f\otimes k(X))= 0$ (since $\ker f$ is torsion-free), and in particular  $\det f = 0$ as well.
-If $\ker f = 0$, then
+  Let $\ker f \neq 0$. Since localization is exact
+  we have
+  $\ker(f\otimes k(X)) = (\ker f)\otimes k(X)$.
+  Since $\ker f \subset E$ is torsionfree, the localization map $\ker f \to (\ker f)\otimes k(X)$ is injective and thus $(\ker f) \otimes k(X)\neq 0$.
+  Since the $k(X)$-vector space map $f\otimes k(X)$ is not injective, $\det(f\otimes k(X))= 0$.
+  In particular  $\det f = 0$ as well.
+  Conversely,
+  if $\ker f = 0$, then
   $f$ is an isomorphism, hence an isomorphism on fibers, and hence
@@
   Let $M$ be a homogeneous module over a graded ring $R$ with $R_0=k$.
-Let $g \in \End_R(M)$. The \defn{irrelevant eigenvalues} of $g$ are defined to be the eigenvalues of the map of vector spaces $g\otimes R/m:M/mM \to M/mM$.
+  Let $g \in \End_R(M)$. The \defn{module-theoretic eigenvalues} of $g$ are defined to be the eigenvalues of the map of vector spaces $g\otimes R/\m:M/\m M \to M/\m M$.
 \end{dfn}
 
-In other words, the irrelevant eigenvalues are the eigenvalues modulo the maximal ideal discussed in the previous section.
+In other words, the module-theoretic eigenvalues are the eigenvalues modulo the maximal ideal discussed in the previous section.
 
@@
   Let $f:E\to E$ arise from a map $g:M\to M$ (i.e., $\wtilde M = E$ and $\wtilde g = f$).
-The eigenvalues of $f$ are a subset of the irrelevant eigenvalues of $g$.
+  The eigenvalues of $f$ are a subset of the module-theoretic eigenvalues of $g$.
 \end{lem}
 
-Note that any endomorphism of $E$ corresponds to an endomorphism of $M=\Gamma_*(E)$; we may want the freedom to work with other module representatives of $E$ though.
+Note that any endomorphism of $E$ corresponds to an endomorphism of $\Gamma_*(E)$; we may want the freedom to work with other module representatives of $E$ though.
 
 \begin{proof}
-  All that needs to be shown is that if $\det(f-\l \id_E) =0$, then $\det((g-\l \id_M)\otimes R/m)=0$.
-Replacing $f$ by $f-\l \id_E$ and likewise $g$ by $g-\l\id_M$, we just need to show that $\det(f)=0$ implies $\det(g\otimes R/m)=0$.
+  All that needs to be shown is that if $\det(f-\l \id_E) =0$, then $\det((g-\l \id_M)\otimes R/\m)=0$.
+  Replacing
+  $f-\l \id_E$
+  by
+  $f$
+  and likewise
+  $g-\l\id_M$
+  by
+  $g$,
+  we just need to show that $\det(f)=0$ implies $\det(g\otimes R/\m)=0$.
 
   Say $\det(f)=0$.
-If $\det(g\otimes R/m)\neq 0$, then $g$ induces a surjection $M/mM\to M/mM$, and thus $g$ induces a surjection $M\to M$ by Nakayama's lemma.
-This then forces $f$ to be a surjection, hence an isomorphism, and thus $\det(f)\neq 0$.
+  If $\det(g\otimes R/\m)\neq 0$, then $g$ induces a surjection $M/\m M\to M/\m M$, and thus $g$ induces a surjection $M\to M$ by Nakayama's lemma.
+The endomorphism $f$ is thus an isomorphism \cite[Corollary~4.4]{Eisenbud95}, and so $\det(f)\neq 0$.
 \end{proof}
@@
 \begin{exa}
-By giving $M$ irrelevant summands, one can always add ``extraneous'' irrelevant eigenvalues, so the containment of eigenvalues may be proper.
+  By adding  irrelevant summands to $M$, one can always add extraneous module-theoretic eigenvalues, so the containment of eigenvalues may be proper.
   %However, if $M=\Gamma_*(E)$, set of values will be the same: in this case, if $g\otimes R/m$ is not an isomorphism, then $g$ cannot be surjective.
-Note that in general the multiplicities will never be equal, even if one takes $M=\Gamma_*(E)$: if $f$ is multiplication by $\l$, then $f$ will have $\rank E$ many eigenvalues, while $g$ will have $\mu(M)$ irrelevant eigenvalues.
+  Note that in general the multiplicities will never be equal, even if one takes $M=\Gamma_*(E)$: if $f$ is multiplication by $\l$, then $f$ will have $\rank E$ many eigenvalues, while $g$ will have $\mu(M)$ module-theoretic eigenvalues.
 \end{exa}
 
-By combining the discussion above with the results of the preceding section, we thus have:
+By combining the discussion above with the results of \Cref{sec:main-alg-new}, we thus have:
 
 \begin{prop}
-If $X$ is a variety over a field $k\subset \overline{\FF_{p}}$,
-$E$ a vector bundle on $X$, and $f:E\to E$ has two distinct eigenvalues $\l_1,\l_2$, then $E$ is a nontrivial direct sum, with one summand given by $(f-\l_1)^{p^e(p^e-1)}$ for some $e\geq 1$.
+  If $X$ is a variety over a field $k$,
+  $E$ a vector bundle on $X$, and $f:E\to E$ has two distinct eigenvalues $\l_1,\l_2$, then $E$ is a nontrivial direct sum, with one summand given by $(f-\l_1)^{N}$ for some $N$.
 \end{prop}
@@
 \begin{proof}
-Take $M= \Gamma_*(E)$; then since the eigenvalues $\l_1\neq \l_2$ are also irrelevant eigenvalues of $M$, our algorithm produces a nontrivial direct sum decomposition $M_1\oplus M_2$, and since $M$ is torsion-free we know that $\widetilde M_i\neq 0$, and thus $E=\widetilde M_1\oplus \widetilde M_2$ is a nontrivial direct summand of sheaves.
+  Take $M= \Gamma_*(E)$; then since the eigenvalues $\l_1\neq \l_2$ are also module-theoretic eigenvalues of $M$, \Cref{alg:graded} produces a nontrivial direct sum decomposition $M_1\oplus M_2$, and since $M$ is torsion-free we know that $\widetilde M_i\neq 0$, and thus $E=\widetilde M_1\oplus \widetilde M_2$ is a nontrivial direct summand of sheaves.
 \end{proof}
@@
 \section{Examples}
+\label{sec:examples}
 
-In this section, we give examples of the kind of calculations and observations the algorithm from the previous section allows us to make.
+In this section, we give examples of the kind of calculations and observations \Cref{alg:graded} and \Cref{alg:local} allow us to make.
 
@@
 
-  For instance, consider $X = \Bl_4\P^2$, the blowup of $\P^2$ at 4 general points. We will be working over the $\ZZ^5$-graded Cox ring
+  For instance, consider $X = \Bl_4\P^2$, the blowup of $\P^2$ at 4 general points. We work over the $\ZZ^5$-graded Cox ring
   \[ S = k[x_1,\dots,x_{10}]/\text{(five quadric Pl\"ucker relations)} \]
@@
 
+\begin{exa}[The local algorithm]
+  Let $R=\F_2[x,y,z]_{(x,y,z)}/(x^2y+xy^2+xyz+z^2)$.
+In the notation of \cite{Artin77}, this is the $D_4^1$ singularity, which is an exceptional version of the usual rational double point/Du Val singularities appearing in characteristic 2.
+In particular, note that $R$ is not homogeneous.
+If $F_* R$ denotes the Frobenius pushforward of $R$, then the algorithm of \Cref{sec:main-alg} computes the following indecomposable summands:
+  $$
+  F_*R = R\oplus
+  \coker
+  \begin{pmatrix}
+    x+y+z&z\\
+    z&xy
+  \end{pmatrix}\oplus\coker \begin{pmatrix}
+    y&z\\
+    z&x^{2}+xy+xz
+  \end{pmatrix}\oplus \coker\begin{pmatrix}
+  x&z\\
+  z&xy+y^{2}+yz
+  \end{pmatrix};
+  $$
+  there is one free summand and three reflexive modules of rank 1.
+\end{exa}
+
 \begin{exa}[Syzygies over Artinian rings]
   % https://arxiv.org/abs/2208.05427
-  In recent work suggested by examples calculated using our algorithm, \cite{CDE24} studies the indecomposable summands of syzygy modules over a Golod ring $(R,m,k)$.
-In particular, they find previously unexpected recurring behavior, specifically that the syzygy modules of the residue field are direct sums of only three indecomposable modules: the residue field $k$, the maximal ideal $m$, and an additional module $N = \Hom_R(m, R)$.
+  In recent work suggested by examples calculated using our algorithm, \cite{CDE24} studies the indecomposable summands of syzygy modules over a Golod ring $(R,\m,k)$.
+In particular, they find previously unexpected recurring behavior, specifically that the syzygy modules of the residue field are direct sums of only three indecomposable modules: the residue field $k$, the maximal ideal $\m$, and an additional module $N = \Hom_R(\m, R)$.
 
   Here, we give a concrete example of one such ring:
-  Let $R = k[x,y]/(x^3,x^2y^3,y^5)$ and consider the (infinite) minimal free resolution of the residue field, which has rank $2^n$ in homological index $n$.
+  Let $k$ be any field and let
+  $R = k[x,y]/(x^3,x^2y^3,y^5)$ and consider the (infinite) minimal free resolution of the residue field, which has rank $2^n$ in homological index $n$.
 
   The fourth syzygy module of the residue field decomposes (ignoring the grading) as the direct sum
-  $$ k^3 \oplus m^2 \oplus N^3. $$
+  $$ k^3 \oplus \m^2 \oplus N^3. $$
   and the fifth syzygy module as
-  $$ k^8\oplus m^9 \oplus N^2. $$
-  The use of our algorithm was essential to the observation that beyond the ``guaranteed'' summands of $k$ and $m$ (which were known to appear by work of \cite{DE23}) only the one additional indecomposable module $N$ appears in the summands of syzygies of $k$.
+  $$ k^8\oplus \m^9 \oplus N^2. $$
+  The use of our algorithm was essential to the observation that beyond the ``guaranteed'' summands of $k$ and $\m$ (which were known to appear by work of \cite{DE23}) only the one additional indecomposable module $N$ appears in the summands of syzygies of $k$.
+
+  In this example, the module $N$ can be explicitly presented as
+  $$N=\coker
+  \begin{pmatrix}
+    x^{2}&0&0&0&y^{4}&xy^{3}&0&0\\
+    -y&x&y^{3}&0&0&0&0&0\\
+    0&0&0&y&-x&0&0&0\\
+    0&0&0&0&0&-y&x&y^{2}
+  \end{pmatrix}
+  $$
+
   %% needsPackage "DirectSummands"
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{lem}
-  Let $A\in \End_R M$ be idempotent. Then $M = \im A\oplus \ker A$.
-\end{lem}
-
-\begin{proof}
-  Consider the short exact sequence
-  $$ 0\to \ker A \to M \to \im A\to 0. $$
-  The map $M\to M$ given by $-A+\id_M$ sends $M\to \ker A$, since $-A(A-\id_M)(m) = -(A^2-A)(m)=0$.
-In fact, this map $M\to \ker A$ is a splitting of $\ker A\to M$, since if $m\in \ker A$ we have $-(A-\id_M)(m) = -A(m)+\id_M(m)=m$.
-\end{proof}
-
-\begin{lem}
-  Let $A\in \End_R M$ satisfy $A^d=A$. Then $M=\im  A\oplus \ker A$.
-\end{lem}
-
-\begin{proof}
-  Consider the short exact sequence
-  $$ 0\to \ker A \to M \to \im A\to 0. $$
-  The map $M\to M$ given by $-A^{d-1}+\id_M$ sends $M\to \ker A$, since $-A(A^{d-1}-\id_M)(m) = -(A^d-A)(m)=0$.
-In fact, this map $M\to \ker A$ is a splitting of $\ker A\to M$, since if $m\in \ker A$ then also $m\in \ker A^{d-1}$ and thus we have $-(A^{d-1}-\id_M)(m) = -A^d(m)+\id_M(m)=m$.
-\end{proof}
-
 % Local Variables:

### 2025-04-22 paper/main.tex
changes before fixing a mistake

@@
 \let\oldb\b
-\let\oldl\l
 \usepackage[dvipsnames]{xcolor}
@@
 \let\b\beta
-\let\l\lambda
+
 \let\wtilde\widetilde
@@
 \begin{abstract}
-  %The problems of finding isomorphism classes of indecomposable modules with certain properties, or determining the indecomposable summands of a module, are ubiquitous in commutative algebra, representation theory, topological data analysis, and other fields.
-  \devlin{We had the commented-out sentence in both the abstract and intro; I think we can just remove it from the abstract? If not, at least rephrase}
-  The purpose of this work is to describe and prove correctness of two practical algorithms for computing indecomposable summands of finitely generated modules over a finitely generated $k$-algebra $R$: the first algorithm applies in the (multi)graded case, which enables the computation of indecomposable summands of coherent sheaves on subvarieties of toric varieties (in particular, for varieties embedded in projective space); the second algorithm applies when $R$ is local and $k$ is a finite field, which opens the door to computing decompositions in singularity theory.
+  We describe and prove correctness of two practical algorithms for computing indecomposable summands of finitely generated modules over a finitely generated $k$-algebra $R$: the first algorithm applies in the (multi)graded case, which enables the computation of indecomposable summands of coherent sheaves on subvarieties of toric varieties (in particular, for varieties embedded in projective space); the second algorithm applies when $R$ is local and $k$ is a finite field, which opens the door to computing decompositions in singularity theory.
 We also present multiple examples, including some which present previously unknown phenomena regarding the behavior of summands of Frobenius pushforwards (including in the non-graded case) and syzygies over Artinian rings.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Introduction}
 
-\section{Introduction}
-\label{sec:intro}
-
-The problems of finding isomorphism classes of indecomposable modules with certain properties, or determining the indecomposable summands of a module, are ubiquitous in commutative algebra, group theory, representation theory, and other fields.
+The problems of finding isomorphism classes of indecomposable modules with certain properties, or determining the indecomposable summands of a module, are ubiquitous in commutative algebra, representation theory, topological data analysis, and other fields.
 Within commutative algebra, for instance, the classification of rings $R$ for which there are only finitely many isomorphism classes of indecomposable maximal Cohen--Macaulay $R$-modules (the \emph{finite CM-type} property), or determining whether iterated Frobenius pushforwards of a Noetherian ring in positive characteristic have finitely many isomorphism classes of indecomposable summands (the \emph{finite F-representation type} property) are two well-established research problems (see, for example, \cite{Yoshino90,LW12} for the former and \cite{SVdB97,Hara15,TT08}, among many others, for the latter).
@@
 
-Currently there are no efficient \devlin{would it be too bold to remove ``efficient''?} algorithms available for checking indecomposability or finding summands of arbitrary modules over commutative rings.
+Currently there are no algorithms available for checking indecomposability or finding summands of arbitrary modules over commutative rings.
 In contrast, variants of the ``Meat-Axe'' algorithm for determining irreducibility of finite-dimensional modules over a group algebra have wide ranging applications in computational group theory \cite{Parker84,HR94,Holt98} and are available through symbolic algebra software such as Magma and GAP \cite{MAGMA,GAP}.
@@
 The purpose of this paper is to describe and prove correctness of two practical algorithms for computing indecomposable summands of finitely generated modules over a finitely generated $k$-algebra $R$, when either $R$ is (multi)graded and $k$ an arbitrary field, or $R$ local and $k$ a field of positive characteristic.
-In the graded case, our algorithm is often capable of completely decomposing a module in a single pass, while in the local case it proceeds by recursively finding summands and then reapplying the algorithm to each summand.
+In the graded case, our algorithm is often capable of completely decomposing a module in a single pass, while in the local case it proceeds by splitting the module into two summands and continuing recursively.
 %We also describe applications of our algorithm for computations in characteristic 0.
@@
 
-After describing the algorithms and proving their correctness, we present multiple examples, including some which present previously unknown phenomena regarding the behavior of summands of Frobenius pushforwards and syzygies over Artinian rings.
+Finally, we present multiple examples, including some which present previously unknown phenomena regarding the behavior of summands of Frobenius pushforwards and syzygies over Artinian rings.
 In particular, we highlight the results of \cite{CDE24}, which shows a recurrence formula for indecomposable summands of high syzygies of the residue field of Golod rings, made possible through experiments and observations using our algorithm.
@@
   There is a ``missing'' case of our algorithm: when $R$ is only local, not graded, and $k$ has characteristic 0.
-  If a module over a ring of characteristic 0 is decomposable, its reductions modulo $p$ will be as well; thus, our algorithm at least provides a heuristic for verifying decomposability in characteristic 0.
+If a module over such a ring is decomposable, its reductions modulo $p$ will be as well; thus, our algorithm at least provides a heuristic for verifying decomposability in characteristic 0.
 See \Cref{rem:char}.
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection*{Acknowledgements}
+
 The authors would like to thank David Eisenbud for several very useful conversations and ideas, and Ezra Miller for pointing out the relevance of this algorithm in topological data analysis.
@@
 \section{Notation}
-\label{sec:notation}
 
@@
 \section{The main algorithms}
-\label{sec:main-algs}
 
@@
 
-\subsection{The graded algorithm}\label{sec:main-alg-new}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{The graded case}\label{sec:graded-alg}
 
@@
 %$\Eff R = \{ \deg r : r\in R\}$ is a strongly convex cone in $\ZZ^n$.
-Let $M$ be a finitely generated, graded $R$-module and consider a degree zero endomorphism $\phi \in [\End M]_0$.
+Suppose $M = \bigoplus M_d$ is a finitely generated, graded $R$-module and consider a degree zero endomorphism $\phi \in [\End M]_0$.
 
@@
 
-\begin{lem}\label{graded_algorithm}
+\begin{lem}\label{lem:graded_algorithm}
   Let $N = \mu(M)$ and set $\psi = \phi^N$. Then $M = \im\psi \oplus \coker\psi$.
@@
 \begin{proof}
-  Fix homogeneous generators $m_1,\dots, m_n$ of $M$, with $\deg m_i$ nonincreasing.
-  The matrix for $\phi$ is then ``block upper triangular'' with diagonal blocks with entries in $k$, i.e., of the form
+  Fix homogeneous generators $m_1,\dots,m_N$ of $M$, with $\deg m_i$ nonincreasing, and let $d_1,\dots,d_n$ be the decreasing list of unique degrees of generators of $M$.
+The matrix for $\phi$ is then ``block upper triangular'' with $n$ diagonal blocks with entries in $k$, i.e., of the form
   \[ \phi = \begin{pmatrix}
@@
     \vdots & \vdots & \ddots & \vdots \\
-    0      & 0      & \dots  & B_l
+    0      & 0      & \dots  & B_n
   \end{pmatrix}, \]
@@
 To see this, note that since $\phi$ has degree zero, $\phi(m_i)$ is a $k$-linear combination of the $m_j$ with $\deg m_j = \deg m_i$ plus an $R$-linear combination of the $m_j$ with $\deg m_j < \deg m_i$.
+In particular, $B_i$ is the induced endomorphism on the submodule generated by all generators of degree $d_i$.
 
-  Thus, since each matrix $B_i$ has entries in the algebraically closed field $k$, by performing row and column operations over $k$ we can put each $B_i$ into Jordan canonical form $J_i$, with each Jordan block having entries in $k$.
-This corresponds to replacing each generator $m_i$ with a $k$-linear combination of generators $m_j$ with $\deg m_i = \deg m_j$, which is an endomorphism of $M$.
-By doing so starting with $B_1$ and proceeding to $B_l$, the resulting matrix has the form
+  Since the matrices $B_i$ have entries in the algebraically closed field $k$, by performing row and column operations over $k$ we can put each $B_i$ into Jordan canonical form $J_i$ with entries in $k$.
+In particular, since the generators of degree $d_i$ are linearly independent, a $k$-linear change of basis on this set induces an automorphism $\alpha_i$ of $M$.
+
+  Let $\alpha$ be the composition of these automorphisms for each block $B_i$, then $\phi' = \alpha^{-1}\circ\phi\circ\alpha$ corresponds to performing row and column operations on $\phi$ over $k$ and has the form
   \begin{equation}
     \label{eq:Jordan}
-    \phi
-    =
-    \begin{pmatrix}
+    \phi' = \begin{pmatrix}
       J_1    & *      & \dots  & *      \\
@@
       \vdots & \vdots & \ddots & \vdots \\
-      0      & 0      & \dots  & J_l
+      0      & 0      & \dots  & J_n
     \end{pmatrix},
   \end{equation}
-  where each $J_i$ is a Jordan block over $k$ with eigenvalue $\l_i$, and the remaining entries are arbitrary.
-  Relabeling, we may assume that $\l_i\neq 0$ for $i=1,\dots,l_0$ and $\l_i=0$ for $i=l_0+1,\dots,l$.
+  where each $J_i$ is a Jordan block over $k$ with eigenvalue $\lambda_i$, and the remaining entries are in $R$.
+  Relabeling, we may assume that $\lambda_i\neq 0$ for $i=1,\dots,n_0$ and $\lambda_i=0$ for $i=n_0+1,\dots,n$.
 
-  If we take $N_0  $ to be at least the sum of the sizes of the nilpotent Jordan blocks $J_{l_0+1},\dots,J_l$, then for $N\geq N_0$ we have
+  If we take $N_0$ to be at least the sum of the sizes of the nilpotent Jordan blocks $J_{n_0+1},\dots,J_n$, then for $N\geq N_0$ we have
   $$
@@
     \vdots &\vdots & \ddots &\vdots \\
-    0 & 0 & \dots & J_{l_0}^N & *** & \dots & ***\\
+    0 & 0 & \dots & J_{n_0}^N & *** & \dots & ***\\
     0 & 0 & \dots & 0 & 0 & \dots & 0\\
@@
   $$
-  where each nonzero $J_i^N$ has the units $\l_i^N$ on the diagonal.
+  where each nonzero $J_i^N$ has the units $\lambda_i^N$ on the diagonal.
   ($\mu(M)$ is an upper bound on the sum of the sizes of the nilpotent Jordan blocks, so the above holds for $N\geq \mu(M)$.)
@@
   So, all that needs to be seen is that $\psi:=\psi^N$ is surjective.
-  To see this, note that after replacing the original generators for $M$ by the generators that put $\psi$ into the form above,  a generating set for $\im(\phi^N)$ is given by $f_i = \psi(m_i) = u_i\cdot e_i + \sum_{j<i } a_{ij} e_j$ for $i = 1,\dots,n-N_0$, where $u_i$ is a unit (explicitly, $u_i$ is a power of some nonzero $\l_{i'}\in k$).
+  To see this, note that after replacing the original generators for $M$ by the generators that put $\psi$ into the form above,  a generating set for $\im(\phi^N)$ is given by $f_i = \psi(m_i) = u_i\cdot e_i + \sum_{j<i } a_{ij} e_j$ for $i = 1,\dots,n-N_0$, where $u_i$ is a unit (explicitly, $u_i$ is a power of some nonzero $\lambda_{i'}\in k$).
   That is, a generating set is given by
@@
 The above can of course produce the trivial direct sum decomposition: if the eigenvalues of $\phi$ are all nonzero, the proof yields that $\im \psi = M$ and $\coker \psi = 0$.
-The following corollary of the proof of \cref{graded_algorithm} ensures that we can always assume $\phi$ has zero as an eigenvalue:
+The following corollary of the proof of \cref{lem:graded_algorithm} ensures that we can always assume $\phi$ has zero as an eigenvalue:
 
@@
   In particular,
-  for any  $\phi \in \End M$, there is $\l \in k$ such that $\phi - \l \id_M \in \End M$ has zero as an eigenvalue.
+  for any  $\phi \in \End M$, there is $\lambda \in k$ such that $\phi - \lambda \id_M \in \End M$ has zero as an eigenvalue.
 \end{lem}
@@
 
-\begin{algm}
-  \label{alg:graded}
+\begin{algm}\label{alg:graded}
   Given a finitely generated graded $R$-module $M$, the indecomposable summands of $M$ are computed as follows.
@@
   \item If $\phi$ has at least two eigenvalues, let $\psi=\phi^{\mu(M)}$.
-By \Cref{graded_algorithm}, $M\cong \im \psi \oplus \coker \psi$.
+By \Cref{lem:graded_algorithm}, $M\cong \im \psi \oplus \coker \psi$.
 \item Repeat steps (1)--(3) for both $\im \psi$ and $\coker \psi$.
@@
 
-\subsection{The local algorithm}\label{sec:main-alg}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{The local case}\label{sec:local-alg}
 
@@
   Let $k$ be a finite field of characteristic $p$, and let $A$ be an endomorphism of a $k$-vector space such that all eigenvalues of $A$ are contained in $k$.
-If $\l$ is an eigenvalue of $A$, then some power of $A-\l I$ is idempotent.
-  Furthermore, if $\l$ is not the only eigenvalue of $A$, then a  nonzero power of $A-\l I$ is idempotent.
+If $\lambda$ is an eigenvalue of $A$, then some power of $A-\lambda I$ is idempotent.
+  Furthermore, if $\lambda$ is not the only eigenvalue of $A$, then a  nonzero power of $A-\lambda I$ is idempotent.
 \end{lem}
@@
   \underbrace{\begin{pmatrix}
-      \l_i & 1 & 0 & \dots & 0 \\
-      0 & \l_i & 1  & \dots & 0 \\
+      \lambda_i & 1 & 0 & \dots & 0 \\
+      0 & \lambda_i & 1  & \dots & 0 \\
       \vdots & & & & \vdots \\
-      0 & 0 & 0 & \dots & \l_i
+      0 & 0 & 0 & \dots & \lambda_i
   \end{pmatrix}}_{r_i} \]
-  with each $\l_i$ an eigenvalue of $A$.
-  In this basis, $A-\l I$ will be block-diagonal with blocks of form
+  with each $\lambda_i$ an eigenvalue of $A$.
+  In this basis, $A-\lambda I$ will be block-diagonal with blocks of form
   \[ \begin{pmatrix}
-    \l_i-\l & 1 & 0 & \dots & 0 \\
-    0 & \l_i-\l & 1 & \dots & 0 \\
+    \lambda_i-\lambda & 1 & 0 & \dots & 0 \\
+    0 & \lambda_i-\lambda & 1 & \dots & 0 \\
     \vdots & & & & \vdots \\
-    0 & 0 & 0 & \dots & \l_i-\l
+    0 & 0 & 0 & \dots & \lambda_i-\lambda
   \end{pmatrix}. \]
-  Set $\mu_i=\l_i-\l$. Then for any $l\geq 1$, the $l$-th power $(A-\l I)^l$ is block-diagonal with blocks of the form
+  Set $\mu_i=\lambda_i-\lambda$. Then for any $l\geq 1$, the $l$-th power $(A-\lambda I)^l$ is block-diagonal with blocks of the form
   \[ \begin{pmatrix}
@@
   In particular, taking $l=p^{e_0}(p^e-1)$,
-  $(A-\l I)^{p^{e_0}(p^e-1)}$ is a diagonal matrix with diagonal entries 1 or 0, hence idempotent. Note moreover that if some $\l_i\neq \l$, then $(A-\l I)^l$ is not the zero matrix.
+  $(A-\lambda I)^{p^{e_0}(p^e-1)}$ is a diagonal matrix with diagonal entries 1 or 0, hence idempotent.
+Note moreover that if some $\lambda_i\neq \lambda$, then $(A-\lambda I)^l$ is not the zero matrix.
 \end{proof}
@@
   \item Consider the resulting endomorphism $A$ of the $k$-vector space $M/\m M$ and find the eigenvalues of $A$.
-  \item If $A$ has at least two eigenvalues, choose one eigenvalue $\l$, and compute a sufficiently high power $B=A^{p^{e_0}(p^{e}-1)}$ of $A$ (with the power explicitly as in the proof of \Cref{lem:jordan}).
+  \item If $A$ has at least two eigenvalues, choose one eigenvalue $\lambda$, and compute a sufficiently high power $B=A^{p^{e_0}(p^{e}-1)}$ of $A$ (with the power explicitly as in the proof of \Cref{lem:jordan}).
 This power will be a nonzero idempotent, and thus produce a splitting of $M$ as $\im B \oplus \coker B$ by \Cref{lem:idemp}.
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Termination of the algorithms}
@@
 
-\begin{lem}
-  \label{lem:term}
+\begin{lem}\label{lem:term}
   Assume that either $(R,\m)$ is a graded ring with $R_0=k$, or $(R,\m)$ is a local ring with $k=R/\m$ a field of positive characteristic, and that $M$ is a finitely generated $R$-module (homogeneous in the graded case).
@@
 
-\begin{lem}
-  \label{lem:distinct}
+\begin{lem}\label{lem:distinct}
   Retain the notation of \Cref{lem:term}.
@@
   Note that if $M$ is decomposable, say $M=M_1\oplus M_2$, then we may always take $\Phi_1,\Phi_2$ to be projectors onto each summand.
-  Let $U\subset \A^r$ be the subset of $r$-tuples $(\l_1,\dots,\l_r)$ such that $\l_1\phi_1+\dots+\l_r\phi_r$ has at least two distinct eigenvalues, i.e., such that $\l_1\Phi_1+\dots+\l_r\Phi_r$ reduces to an endomorphism of $M/\m M$ with at least two distinct eigenvalues.
+  Let $U\subset \A^r$ be the subset of $r$-tuples $(\lambda_1,\dots,\lambda_r)$ such that $\lambda_1\phi_1+\dots+\lambda_r\phi_r$ has at least two distinct eigenvalues, i.e., such that $\lambda_1\Phi_1+\dots+\lambda_r\Phi_r$ reduces to an endomorphism of $M/\m M$ with at least two distinct eigenvalues.
 (Note that in the graded case the eigenvalues of the reduction of an endomorphism modulo $\m$ are the same as the eigenvalues of endomorphism, so there is no harm in looking only at $M/\m M$.)
@@
 Choose $\Phi_1$ to be the projection to $M_1$, and $\Phi_2$ the projection to $M_2$.
-Then for any $\l_1,\l_2\in k$, $\l_1\phi_1 + \l_2\phi_2$ has eigenvalues $\l_1,\l_2$; thus in particular there is \emph{an} element of $\End_R(M)$ reducing to an endomorphism of $M/\m M$ with distinct eigenvalues, so $U$ is nonempty.
+Then for any $\lambda_1,\lambda_2\in k$, $\lambda_1\phi_1 + \lambda_2\phi_2$ has eigenvalues $\lambda_1,\lambda_2$; thus in particular there is \emph{an} element of $\End_R(M)$ reducing to an endomorphism of $M/\m M$ with distinct eigenvalues, so $U$ is nonempty.
 
@@
 This is a purely linear algebraic statement: we claim that given a matrix $\phi$ with at least two distinct eigenvalues, and any $r$ matrices $\phi_1,\dots,\phi_r$, that
-  $$ A_{\l_1,\dots,\l_r}:=\phi+\l_1\phi_1+\dots+\l_r\phi_r $$
-  has at least two distinct eigenvalues for $\l_1,\dots,\l_r$ outside a Zariski-closed subset of $\A^r$.
-  The eigenvalues of $A_{\l_1,\dots,\l_r}$ are the roots of $\det(A_{\l_1,\dots,\l_r}-t I)$, which is a polynomial in $t$ with coefficients in $\l_1,\dots,\l_r$.
-  $A_{\l_1,\dots,\l_r}$ fails to have at least two distinct eigenvalues exactly when this polynomial factors as a power of a linear term.
+  $$ A_{\lambda_1,\dots,\lambda_r}:=\phi+\lambda_1\phi_1+\dots+\lambda_r\phi_r $$
+  has at least two distinct eigenvalues for $\lambda_1,\dots,\lambda_r$ outside a Zariski-closed subset of $\A^r$.
+  The eigenvalues of $A_{\lambda_1,\dots,\lambda_r}$ are the roots of $\det(A_{\lambda_1,\dots,\lambda_r}-t I)$, which is a polynomial in $t$ with coefficients in $\lambda_1,\dots,\lambda_r$.
+  $A_{\lambda_1,\dots,\lambda_r}$ fails to have at least two distinct eigenvalues exactly when this polynomial factors as a power of a linear term.
 
-  This condition is polynomial in the coefficients of powers of $t$ in $\det(A_{\l_1,\dots,\l_r}-t I)$ and thus in the $\l_i$; to see this, note that
+  This condition is polynomial in the coefficients of powers of $t$ in $\det(A_{\lambda_1,\dots,\lambda_r}-t I)$ and thus in the $\lambda_i$; to see this, note that
   if we write
-  $$ f:=\det(A_{\l_1,\dots,\l_r}-t I)=t^n  +t^{n-1}b_{n-1}+\dots +t b_1 +b_0, $$
+  $$ f:=\det(A_{\lambda_1,\dots,\lambda_r}-t I)=t^n  +t^{n-1}b_{n-1}+\dots +t b_1 +b_0, $$
   then $f$
@@
   vanish simultaneously; the resultant of these $n$ polynomials in the $n$ variables $b_i$ gives polynomial conditions in the $b_i$ for this to occur.
-  In our setting, the $b_i$ are themselves polynomials in the $\l_i$, and thus we have obtained polynomial equations defining the locus where $A_{\l_1,\dots,\l_r}$ fails to have distinct roots, and thus the complement $U$ is open.
+  In our setting, the $b_i$ are themselves polynomials in the $\lambda_i$, and thus we have obtained polynomial equations defining the locus where $A_{\lambda_1,\dots,\lambda_r}$ fails to have distinct roots, and thus the complement $U$ is open.
 \end{proof}
@@
 However, the general linear combinations we take in \Cref{item:End0} should be taken with respect to the prime subfield (otherwise, as we increase the size of the finite field $k$, the eigenvalues of a general linear combination will live in higher and higher field extensions).
-  See \Cref{algclos} for a demonstration of the necessity of extending the base field.
+  See \Cref{ex:elliptic} for a demonstration of the necessity of extending the base field.
 \end{rem}
@@
 
-\begin{rem}
-  \label{rem:char}
+\begin{rem}\label{rem:char}
   If $R$ is a local ring over a field $k\subset \C$ of characteristic 0, and $M$ a finitely generated $R$-module, it would be useful to have an algorithm to find a direct sum decomposition of $M$.
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Decomposing coherent sheaves}
-\label{sec:coherent}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Decomposing coherent sheaves}\label{sec:coherent}
 
@@
   where $\det E := \bigwedge^{\rank E} E$.
-  Since $\End_{\OO_X}(\det E) = H^0(\OO_X)=k$, $\bigwedge^{\rank E} f$ is multiplication by $\l\in k$; we write $\det f$ for this scalar $\l$.
+  Since $\End_{\OO_X}(\det E) = H^0(\OO_X)=k$, $\bigwedge^{\rank E} f$ is multiplication by $\lambda\in k$; we write $\det f$ for this scalar $\lambda$.
 \end{dfn}
@@
 \begin{dfn}
-  $\l \in k$ is an \defn{eigenvalue} of $f \in \End_{\OO_X}(E)$ if $\det(f-\l \id_E)= 0$.
+  $\lambda \in k$ is an \defn{eigenvalue} of $f \in \End_{\OO_X}(E)$ if $\det(f-\lambda \id_E)= 0$.
 In other words, the eigenvalues of $f$ are the zeroes of the univariate polynomial $\det(f-t\id _E)$ with coefficients in $k$.
@@
 
-Note that $f$ is an isomorphism if and only if $\l=0$ is not an eigenvalue of $f$, just as for ordinary linear operators on vector spaces.
+Note that $f$ is an isomorphism if and only if $\lambda=0$ is not an eigenvalue of $f$, just as for ordinary linear operators on vector spaces.
 
@@
 \begin{proof}
-  All that needs to be shown is that if $\det(f-\l \id_E) =0$, then $\det((g-\l \id_M)\otimes R/\m)=0$.
+  All that needs to be shown is that if $\det(f-\lambda \id_E) =0$, then $\det((g-\lambda \id_M)\otimes R/\m)=0$.
   Replacing
-  $f-\l \id_E$
+  $f-\lambda \id_E$
   by
@@
   and likewise
-  $g-\l\id_M$
+  $g-\lambda\id_M$
   by
@@
   %However, if $M=\Gamma_*(E)$, set of values will be the same: in this case, if $g\otimes R/m$ is not an isomorphism, then $g$ cannot be surjective.
-  Note that in general the multiplicities will never be equal, even if one takes $M=\Gamma_*(E)$: if $f$ is multiplication by $\l$, then $f$ will have $\rank E$ many eigenvalues, while $g$ will have $\mu(M)$ module-theoretic eigenvalues.
+  Note that in general the multiplicities will never be equal, even if one takes $M=\Gamma_*(E)$: if $f$ is multiplication by $\lambda$, then $f$ will have $\rank E$ many eigenvalues, while $g$ will have $\mu(M)$ module-theoretic eigenvalues.
 \end{exa}
 
-By combining the discussion above with the results of \Cref{sec:main-alg-new}, we thus have:
+By combining the discussion above with the results of \Cref{sec:graded-alg}, we thus have:
 
@@
   If $X$ is a variety over a field $k$,
-  $E$ a vector bundle on $X$, and $f:E\to E$ has two distinct eigenvalues $\l_1,\l_2$, then $E$ is a nontrivial direct sum, with one summand given by $(f-\l_1)^{N}$ for some $N$.
+  $E$ a vector bundle on $X$, and $f:E\to E$ has two distinct eigenvalues $\lambda_1,\lambda_2$, then $E$ is a nontrivial direct sum, with one summand given by $(f-\lambda_1)^{N}$ for some $N$.
 \end{prop}
@@
 \begin{proof}
-  Take $M= \Gamma_*(E)$; then since the eigenvalues $\l_1\neq \l_2$ are also module-theoretic eigenvalues of $M$, \Cref{alg:graded} produces a nontrivial direct sum decomposition $M_1\oplus M_2$, and since $M$ is torsion-free we know that $\widetilde M_i\neq 0$, and thus $E=\widetilde M_1\oplus \widetilde M_2$ is a nontrivial direct summand of sheaves.
+  Take $M= \Gamma_*(E)$; then since the eigenvalues $\lambda_1\neq \lambda_2$ are also module-theoretic eigenvalues of $M$, \Cref{alg:graded} produces a nontrivial direct sum decomposition $M_1\oplus M_2$, and since $M$ is torsion-free we know that $\widetilde M_i\neq 0$, and thus $E=\widetilde M_1\oplus \widetilde M_2$ is a nontrivial direct summand of sheaves.
 \end{proof}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Examples}
-\label{sec:examples}
+\section{Examples}\label{sec:examples}
 
@@
 
-\begin{exa}[Frobenius pushforward on elliptic curves]\label{algclos}
+\begin{exa}[Frobenius pushforward on elliptic curves]\label{ex:elliptic}
   Consider the elliptic curve
@@
 In particular, note that $R$ is not homogeneous.
-If $F_* R$ denotes the Frobenius pushforward of $R$, then the algorithm of \Cref{sec:main-alg} computes the following indecomposable summands:
+If $F_* R$ denotes the Frobenius pushforward of $R$, then the algorithm of \Cref{sec:local-alg} computes the following indecomposable summands:
   $$
@@
 
-\begin{exa}[Symbolic diagonalization]\label{example:QQ}
+\begin{exa}[Symbolic diagonalization]\label{ex:diagonalization}
   An interesting application of our algorithm, suggested by Bernd Sturmfels,  is automated diagonalization of symbolically parameterized matrices.
@@
   Then the splittings of $\coker A$ over $K = \Q$ and $K = \Q(i)$ have the following presentations, respectively:
-  \[\medmuskip1.5mu \begin{pmatrix}
+  \[\medmuskip0mu \begin{pmatrix}
     a+b+c+d&0&0&0\\
@@
   \end{pmatrix},
-  \quad
+  \,
   \begin{pmatrix}

### 2025-04-22 paper/main.tex
converted to algorithm environments

@@
 \usepackage{cleveref}
+\crefformat{equation}{(#2#1#3)} % omit eq.
 \let\b\beta
+
+%algorithms
+\usepackage{algorithm}
+\usepackage{algpseudocode}
+\algdef{SE}[SUBALG]{Indent}{EndIndent}{}{\algorithmicend\ }%
+\algtext*{Indent}
+\algtext*{EndIndent}
+\renewcommand{\algorithmicrequire}{\textbf{Input:}} % Changing "Require" to "Input"
+\renewcommand{\algorithmicensure}{\textbf{Output:}} % Changing "Ensure" to "Output"
+
+%\crefname{algocf}{alg.}{algs.}
+%\Crefname{algocf}{Algorithm}{Algorithms}
 
@@
   $\deg m_1\not\leq \deg m_2\not \leq \dots\not\leq \deg m_r$.
-
+  \mahrud{finish}
 \end{rem}
@@
 \begin{lem}
-  The eigenvalues of $\phi$ lie in $k$.
-  In particular,
-  for any  $\phi \in \End M$, there is $\lambda \in k$ such that $\phi - \lambda \id_M \in \End M$ has zero as an eigenvalue.
+  The eigenvalues of $\phi$ lie in $k$. In particular, for any $\phi \in [\End M]_0$, there is $\lambda \in k$ such that $\phi - \lambda \id_M \in [\End M]_0$ has zero as an eigenvalue.
 \end{lem}
 
-Note that by ``eigenvalues of $\phi$'' we mean the roots of the characteristic polynomial $\det(\phi - t \id_M)$, which is a univariate polynomial with coefficients in $R$ and thus a priori has solutions in $t$ only in the algebraic closure of the total ring of fractions of $R$.
+\mahrud{move this above?}
+By ``eigenvalues of $\phi$'' we mean the roots of the characteristic polynomial $\det(\phi - \lambda \id_M)$\mahrud{define determinant over a module, or say determinant of its cover?}, which is a univariate polynomial in $\lambda$ with coefficients in $R$ and thus a priori has solutions only in the algebraic closure of the total ring of fractions of $R$.
 
 \begin{proof}
-  Note that the expression of $\psi$ obtained in \cref{eq:Jordan} shows that the determinant of $\phi - t \id_M$ is just the product of $\mu_i -t$, and thus the eigenvalues are just the $\mu_i$, which lie in~$k$.
+  The expression of $\phi$ obtained in \cref{eq:Jordan} shows that the determinant of $\phi - \lambda \id_M$ is the product of $\lambda_i - \lambda$, thus the eigenvalues are the $\lambda_i$, which all lie in $k$.
 \end{proof}
 
-\begin{algm}\label{alg:graded}
-  Given a finitely generated graded $R$-module $M$, the indecomposable summands of $M$ are computed as follows.
-  \begin{enumerate}
-  \item\label{item:End0} Take a random element $\phi_0$ of $[\End_R(M)]_0$, the degree zero part of $\End_R(M)$.
-  \item For an eigenvalue $\mu $ of $\phi_0$, let $\phi=\phi_0-\mu \id_M$.
-  \item If $\phi$ has at least two eigenvalues, let $\psi=\phi^{\mu(M)}$.
-By \Cref{lem:graded_algorithm}, $M\cong \im \psi \oplus \coker \psi$.
-\item Repeat steps (1)--(3) for both $\im \psi$ and $\coker \psi$.
-  \end{enumerate}
-\end{algm}
+\begin{algorithm}[H]
+  \caption{(Indecomposable summands of a graded module over a commutative ring)}\label{alg:graded}
+  \begin{algorithmic}[1]
+    \smallskip
+    \Require graded module $M$ over a positively graded commutative ring $R$.
+    \Ensure  list $\textproc{Summands}(M)$ of indecomposable summands of $M$.
+    \State Initialize a list $L$.
+    \State Take a general element $\phi$ in $[\End_R(M)]_0$,
+           the degree zero part of $\End_R(M)$. \label{item:End0}
+    \ForAll { eigenvalues $\lambda_i$ of $\phi$ }
+      \State Let $\mu_i$ be the multiplicity of $\lambda_i$.
+      \State Let $\psi_i = (\phi - \lambda_i \id_M)^{\mu_i}$.
+      \State Append \Call{Summands}{$\coker\psi_i$} to $L$.
+    \EndFor
+    \State Let $\psi$ be the composition of all $\psi_i$ above.
+    \If {$\psi$ is nonzero}
+      \State Append \Call{Summands}{$\im\psi$} to $L$.
+    \EndIf
+    \State \Return $L$.
+  \end{algorithmic}
+\end{algorithm}
 
@@
 
-\begin{algm}\label{alg:local}
-  Given a finitely generated graded $R$-module $M$, the indecomposable summands of $M$ are computed as follows.
-  \begin{enumerate}
-    %In the graded case, take a general element $A_0$ of $[\End_R(M)]_0$, the degree zero part of $\End_R(M)$; in the local case,
-  \item Take a general element $A_0$ of $\End_R(M) \setminus \m \End_R(M)$.
-  \item Consider the resulting endomorphism $A$ of the $k$-vector space $M/\m M$ and find the eigenvalues of $A$.
-  \item If $A$ has at least two eigenvalues, choose one eigenvalue $\lambda$, and compute a sufficiently high power $B=A^{p^{e_0}(p^{e}-1)}$ of $A$ (with the power explicitly as in the proof of \Cref{lem:jordan}).
-This power will be a nonzero idempotent, and thus produce a splitting of $M$ as $\im B \oplus \coker B$ by \Cref{lem:idemp}.
-  \item Repeat steps (1)--(3) for both $\im B$ and $\coker B$.
-  \end{enumerate}
-\end{algm}
+\begin{algorithm}[H]
+  \caption{(Indecomposable summands of a module over a commutative local ring)}\label{alg:local}
+  \begin{algorithmic}[1]
+    \smallskip
+    % TODO: finitely generated M? noetherian R?
+    \Require module $M$ over a commutative local ring $R$.
+    \Ensure  list $\textproc{Summands}(M)$ of indecomposable summands of $M$.
+    \State Initialize a list $L$.
+    \State Take a general element $\phi$ in $\End_R(M) \setminus \m \End_R(M)$.
+    \State Let $\phi_0$ be the induced endomorphism of the $k$-vector space $M/\m M$.
+    % TODO: Why did we need to assume $A$ has at least two eigenvalues?
+    \State Let $\lambda$ be any eigenvalue of $\phi_0$.
+    \State Let $\psi = (\phi - \lambda \id_M)^{p^{e_0} (p^{e} - 1)}$
+    (with the power explicitly as in the proof of \Cref{lem:jordan}).
+    \State By \Cref{lem:idemp}, $\psi$ is an idempotent, hence producing a splitting of $M$.
+    \If { both $\coker\psi$ and $\im\psi$ are nonzero }
+      \State Append \Call{Summands}{$\coker\psi$} to $L$.
+      \State Append \Call{Summands}{$\im\psi$} to $L$.
+    \Else
+      \State Append $M$ to $L$.
+    \EndIf
+    \State \Return $L$.
+  \end{algorithmic}
+\end{algorithm}
 
@@
 While theoretically the issue can be avoided by working over an algebraically closed ground field $\bar k$, for practical use on a computer algebra system it is better to extend $k$ to some larger finite field.
-However, the general linear combinations we take in \Cref{item:End0} should be taken with respect to the prime subfield (otherwise, as we increase the size of the finite field $k$, the eigenvalues of a general linear combination will live in higher and higher field extensions).
+However, the general linear combinations we take in Step~\ref{item:End0} of \Cref{alg:graded} should be taken with respect to the prime subfield (otherwise, as we increase the size of the finite field $k$, the eigenvalues of a general linear combination will live in higher and higher field extensions).
   See \Cref{ex:elliptic} for a demonstration of the necessity of extending the base field.

### 2025-04-23 paper/main.tex
finished the remark

@@
 
-\begin{lem}\label{lem:graded_algorithm}
+\begin{lem}\label{lem:graded}
   Let $N = \mu(M)$ and set $\psi = \phi^N$. Then $M = \im\psi \oplus \coker\psi$.
@@
   where each $J_i$ is a Jordan block over $k$ with eigenvalue $\lambda_i$, and the remaining entries are in $R$.
-  Relabeling, we may assume that $\lambda_i\neq 0$ for $i=1,\dots,n_0$ and $\lambda_i=0$ for $i=n_0+1,\dots,n$.
+  Let $\mu_i$ be the multiplicity of $\lambda_i$ and let $\psi_i = (\phi' - \lambda_i \id_M)^{\mu_i}$.
+  \mahrud{TODO}
 
-  If we take $N_0$ to be at least the sum of the sizes of the nilpotent Jordan blocks $J_{n_0+1},\dots,J_n$, then for $N\geq N_0$ we have
-  $$
-  \phi^N =
-  \begin{pmatrix}
-    J_1^N & *** &\dots & \dots& \dots  &\dots & *** \\
-    0 & J_2^N & \dots &\dots&\dots & \dots& *** \\
-    \vdots &\vdots & \ddots &\vdots \\
-    0 & 0 & \dots & J_{n_0}^N & *** & \dots & ***\\
-    0 & 0 & \dots & 0 & 0 & \dots & 0\\
-    \vdots &\vdots & \ddots &\vdots \\
-    0 & 0 & \dots & 0 & 0 & \dots & 0
-  \end{pmatrix},
-  $$
-  where each nonzero $J_i^N$ has the units $\lambda_i^N$ on the diagonal.
-  ($\mu(M)$ is an upper bound on the sum of the sizes of the nilpotent Jordan blocks, so the above holds for $N\geq \mu(M)$.)
+  %% Relabeling, we may assume that $\lambda_i\neq 0$ for $i=1,\dots,n_0$ and $\lambda_i=0$ for $i=n_0+1,\dots,n$.
 
-  Fix $N$ as above.
-  Now, consider the map $ \phi^N: \im(\phi^N) \to \im(\phi^N)$.
+  %% If we take $N_0$ to be at least the sum of the sizes of the nilpotent Jordan blocks $J_{n_0+1},\dots,J_n$, then for $N\geq N_0$ we have
+  %% $$
+  %% \phi^N =
+  %% \begin{pmatrix}
+  %%   J_1^N & *** &\dots & \dots& \dots  &\dots & *** \\
+  %%   0 & J_2^N & \dots &\dots&\dots & \dots& *** \\
+  %%   \vdots &\vdots & \ddots &\vdots \\
+  %%   0 & 0 & \dots & J_{n_0}^N & *** & \dots & ***\\
+  %%   0 & 0 & \dots & 0 & 0 & \dots & 0\\
+  %%   \vdots &\vdots & \ddots &\vdots \\
+  %%   0 & 0 & \dots & 0 & 0 & \dots & 0
+  %% \end{pmatrix},
+  %% $$
+  %% where each nonzero $J_i^N$ has the units $\lambda_i^N$ on the diagonal.
+  %% ($\mu(M)$ is an upper bound on the sum of the sizes of the nilpotent Jordan blocks, so the above holds for $N\geq \mu(M)$.)
+
+  %% Fix $N$ as above.
+  %% Now, consider the map $ \phi^N: \im(\phi^N) \to \im(\phi^N)$.
 If this map is surjective, then $\im(\phi^N)$ is a direct summand of $M$, with complementary summand $\coker \phi^N$: to see this, observe that the injection $\im(\phi^N) \inc M$ would then  compose with $\phi^N$ to be a surjection, and since $\im(\phi^N)$ is a finitely generated module, a surjection must be an isomorphism $\a:\im \phi^N \to \im \phi^N$ by \cite[Corollary~4.4]{Eisenbud95}.
@@
 
-  So, all that needs to be seen is that $\psi:=\psi^N$ is surjective.
-  To see this, note that after replacing the original generators for $M$ by the generators that put $\psi$ into the form above,  a generating set for $\im(\phi^N)$ is given by $f_i = \psi(m_i) = u_i\cdot e_i + \sum_{j<i } a_{ij} e_j$ for $i = 1,\dots,n-N_0$, where $u_i$ is a unit (explicitly, $u_i$ is a power of some nonzero $\lambda_{i'}\in k$).
-  That is, a generating set is given by
-  $$u_1 e_1,\quad  u_2 e_2+a_{21}e_1,\quad \dots ,\quad u_{n-n_0}e_{n-N_0}+a_{n-N_0,n-1}e_{n-N_0-1}+\dots + a_{N-n_0,1}e_1.$$
-  Since the $u_i$ are units, the $R$-span of these generators is clearly the same as the $R$-span of $e_1,\dots,e_{n-N_0}$, and thus after applying $\psi$ we get back the $R$-span of $f_1,\dots,f_{n-N_0} $, which is $\im \psi^N$.
+  %% So, all that needs to be seen is that $\psi:=\psi^N$ is surjective.
+  %% To see this, note that after replacing the original generators for $M$ by the generators that put $\psi$ into the form above,  a generating set for $\im(\phi^N)$ is given by $f_i = \psi(m_i) = u_i\cdot e_i + \sum_{j<i } a_{ij} e_j$ for $i = 1,\dots,n-N_0$, where $u_i$ is a unit (explicitly, $u_i$ is a power of some nonzero $\lambda_{i'}\in k$).
+  %% That is, a generating set is given by
+  %% $$u_1 e_1,\quad  u_2 e_2+a_{21}e_1,\quad \dots ,\quad u_{n-n_0}e_{n-N_0}+a_{n-N_0,n-1}e_{n-N_0-1}+\dots + a_{N-n_0,1}e_1.$$
+  %% Since the $u_i$ are units, the $R$-span of these generators is clearly the same as the $R$-span of $e_1,\dots,e_{n-N_0}$, and thus after applying $\psi$ we get back the $R$-span of $f_1,\dots,f_{n-N_0} $, which is $\im \psi^N$.
 Thus, $\phi: \im \psi^N \to \im \psi^N$ is surjective, and the result follows.
@@
 
-\begin{rem}[how to do it in multigraded case]
-  Let $R$ be an $\ZZ^n$-graded ring, with $R_0=k$ an algebraically closed field.
-  Assume furthermore that the degree support of $R$ is strongly convex, i.e., that
-  the effective cone
-  $\Eff R = \{ \deg r : r\in R\}$ is a strongly convex cone in $\ZZ^n$.
-  Let $M$ be a finitely generated graded $R$-module, and let $\phi \in \End M$ be a degree zero endomorphism.
+\begin{rem}
+  The hypotheses that $R$ is $\N$-graded and $k$ is an algebraically closed field may both be weakened.
+  In the proof of \Cref{lem:graded}, we only need a partial order on the degrees of the generators of $M$ in order to present $\phi$ in a block upper triangular form.
+Suppose $R$ is $\ZZ^r$-graded, then the effective cone $\Eff(R)\subset\Q^n$ generated by the elements $\deg m$ for each monomial $m \in R$ induces a partial order on the degrees of generators of $M$ if it is strongly convex (contains no positive dimensional subspace of $\Q^n$) and $R_0 = k$.
+% TODO: can R_0 be finite dimensional?
 
-  Let $P\subset \ZZ^n$ be the degrees of a set of minimal generators of $M$;
-  this is a poset with respect to the partial order $\a\leq \b$ if and only if $\b-\a\in \Eff R$.
-  Because $P$ is finite, we can order the elements of $P$ such that
-  $\deg m_1\not\leq \deg m_2\not \leq \dots\not\leq \deg m_r$.
-  \mahrud{finish}
+  Similarly, suppose $k$ is not algebraically closed, then there may be blocks $B_i$ which cannot be written in Jordan canonical form.
+Nevertheless, as long as $\phi$ has at least one eigenvalue $\lambda$ in $k$, the corresponding block may be written in Jordan canonical form, and thus the image and cokernel of $(\phi - \lambda_i\id_M)^{\mu_i}$ form summands of $M$.
 \end{rem}
 
-The above can of course produce the trivial direct sum decomposition: if the eigenvalues of $\phi$ are all nonzero, the proof yields that $\im \psi = M$ and $\coker \psi = 0$.
-The following corollary of the proof of \cref{lem:graded_algorithm} ensures that we can always assume $\phi$ has zero as an eigenvalue:
+% The above can of course produce the trivial direct sum decomposition: if the eigenvalues of $\phi$ are all nonzero, the proof yields that $\im \psi = M$ and $\coker \psi = 0$.
+The following corollary of the proof of \Cref{lem:graded} ensures that we can always assume $\phi$ has zero as an eigenvalue:
 
-\begin{lem}
-  The eigenvalues of $\phi$ lie in $k$. In particular, for any $\phi \in [\End M]_0$, there is $\lambda \in k$ such that $\phi - \lambda \id_M \in [\End M]_0$ has zero as an eigenvalue.
-\end{lem}
+%% \begin{lem}
+%%   The eigenvalues of $\phi$ lie in $k$. In particular, for any $\phi \in [\End M]_0$, there is $\lambda \in k$ such that $\phi - \lambda \id_M \in [\End M]_0$ has zero as an eigenvalue.
+%% \end{lem}
 
-\mahrud{move this above?}
 By ``eigenvalues of $\phi$'' we mean the roots of the characteristic polynomial $\det(\phi - \lambda \id_M)$\mahrud{define determinant over a module, or say determinant of its cover?}, which is a univariate polynomial in $\lambda$ with coefficients in $R$ and thus a priori has solutions only in the algebraic closure of the total ring of fractions of $R$.
 
-\begin{proof}
-  The expression of $\phi$ obtained in \cref{eq:Jordan} shows that the determinant of $\phi - \lambda \id_M$ is the product of $\lambda_i - \lambda$, thus the eigenvalues are the $\lambda_i$, which all lie in $k$.
-\end{proof}
+%% \begin{proof}
+%%   The expression of $\phi$ obtained in \cref{eq:Jordan} shows that the determinant of $\phi - \lambda \id_M$ is the product of $\lambda_i - \lambda$, thus the eigenvalues are the $\lambda_i$, which all lie in $k$.
+%% \end{proof}
 
@@
 
-\begin{exa}[The local algorithm]
+\begin{exa}[Local singularities]
   Let $R=\F_2[x,y,z]_{(x,y,z)}/(x^2y+xy^2+xyz+z^2)$.

### 2025-04-23 paper/main.tex
changed proof of graded case

@@
 
-Let $R$ be a $\N$-graded ring, with $R_0 = k$, an algebraically closed field.
+Let $R$ be a $\N$-graded ring, with $R_0 = k$ a field.
 %Assume furthermore that the degree support of $R$ is strongly convex, i.e., that
@@
 %$\Eff R = \{ \deg r : r\in R\}$ is a strongly convex cone in $\ZZ^n$.
-Suppose $M = \bigoplus M_d$ is a finitely generated, graded $R$-module and consider a degree zero endomorphism $\phi \in [\End M]_0$.
+Suppose $M = \bigoplus M_d$ is a finitely generated, graded $R$-module and consider a general degree zero endomorphism $\phi \in [\End M]_0$.
 
@@
 \begin{lem}\label{lem:graded}
-  Let $N = \mu(M)$ and set $\psi = \phi^N$. Then $M = \im\psi \oplus \coker\psi$.
+  \mahrud{rewrite statement} Let $N = \mu(M)$ and set $\psi = \phi^N$. Then $M = \im\psi \oplus \coker\psi$.
 \end{lem}
@@
 The matrix for $\phi$ is then ``block upper triangular'' with $n$ diagonal blocks with entries in $k$, i.e., of the form
-  \[ \phi = \begin{pmatrix}
+  \begin{equation}\label{eq:blocks}
+    \phi = \begin{pmatrix}
     B_1    & *      & \dots  & *      \\
@@
     0      & 0      & \dots  & B_n
-  \end{pmatrix}, \]
+    \end{pmatrix},
+  \end{equation}
   where the blocks $B_i$ are square matrices \emph{with entries in $k$}, and the remaining entries are in $R$.
@@
 
-  Since the matrices $B_i$ have entries in the algebraically closed field $k$, by performing row and column operations over $k$ we can put each $B_i$ into Jordan canonical form $J_i$ with entries in $k$.
-In particular, since the generators of degree $d_i$ are linearly independent, a $k$-linear change of basis on this set induces an automorphism $\alpha_i$ of $M$.
+  Note that the characteristic polynomial $\det(\phi - \lambda \id_M) = \prod \det(B_i - \lambda \id)$ is a univariate polynomial in $\lambda$ with coefficients in $k$ which may have solutions in the algebraic closure of $k$.
+We will refer to the collection of $k$-eigenvalues of $B_i$ as the ``eigenvalues of $\phi$''.
 
-  Let $\alpha$ be the composition of these automorphisms for each block $B_i$, then $\phi' = \alpha^{-1}\circ\phi\circ\alpha$ corresponds to performing row and column operations on $\phi$ over $k$ and has the form
-  \begin{equation}
-    \label{eq:Jordan}
-    \phi' = \begin{pmatrix}
-      J_1    & *      & \dots  & *      \\
-      0      & J_2    & \dots  & *      \\
-      \vdots & \vdots & \ddots & \vdots \\
-      0      & 0      & \dots  & J_n
-    \end{pmatrix},
+  By generality of $\phi$ suppose it has eigenvalues $\lambda_1,\dots,\lambda_r$ in $k$.
+Let $\mu(\lambda_j)$ be the maximum multiplicity of $\lambda$ in each diagonal block and set $\psi_j = (\phi - \lambda_j \id_M)^{\mu(\lambda_j)}$.
+We will show that $\psi_j$ is a split surjection, hence $M = \im\psi_j \oplus \coker\psi_j$.
+Furthermore, note that if $\phi_j$ and $\phi_k$ are different then $M = \coker\psi_j \oplus \coker\psi_k \oplus \im(\psi_j\circ\psi_k)$, thus we have a decomposition
+  \begin{equation}\label{eq:decomposition}
+    M = \coker\psi_1 \oplus \cdots \oplus \coker\psi_r \oplus \im\psi.
   \end{equation}
-  where each $J_i$ is a Jordan block over $k$ with eigenvalue $\lambda_i$, and the remaining entries are in $R$.
-  Let $\mu_i$ be the multiplicity of $\lambda_i$ and let $\psi_i = (\phi' - \lambda_i \id_M)^{\mu_i}$.
-  \mahrud{TODO}
 
-  %% Relabeling, we may assume that $\lambda_i\neq 0$ for $i=1,\dots,n_0$ and $\lambda_i=0$ for $i=n_0+1,\dots,n$.
+  To see that each $\psi_j$ is a split surjection, note that the set of generators of $M$ with degree $d_i$ are linearly independent and we can view each $B_i$ as a $k$-linear transformation of the $k$-vector space spanned by these generators.
+Thus, performing $k$-linear row and column operations on $(B_i - \lambda_j \id)^{\mu(\lambda_j)}$, we may write it as a diagonal matrix with ones and zeros on the diagonal.
+In particular, this is a $k$-linear change of basis on the degree $d_i$ generators of $M$ and induces an automorphism $\alpha_{i,j}$ of $M$.
 
-  %% If we take $N_0$ to be at least the sum of the sizes of the nilpotent Jordan blocks $J_{n_0+1},\dots,J_n$, then for $N\geq N_0$ we have
-  %% $$
-  %% \phi^N =
-  %% \begin{pmatrix}
-  %%   J_1^N & *** &\dots & \dots& \dots  &\dots & *** \\
-  %%   0 & J_2^N & \dots &\dots&\dots & \dots& *** \\
-  %%   \vdots &\vdots & \ddots &\vdots \\
-  %%   0 & 0 & \dots & J_{n_0}^N & *** & \dots & ***\\
-  %%   0 & 0 & \dots & 0 & 0 & \dots & 0\\
-  %%   \vdots &\vdots & \ddots &\vdots \\
-  %%   0 & 0 & \dots & 0 & 0 & \dots & 0
-  %% \end{pmatrix},
-  %% $$
-  %% where each nonzero $J_i^N$ has the units $\lambda_i^N$ on the diagonal.
-  %% ($\mu(M)$ is an upper bound on the sum of the sizes of the nilpotent Jordan blocks, so the above holds for $N\geq \mu(M)$.)
-
-  %% Fix $N$ as above.
-  %% Now, consider the map $ \phi^N: \im(\phi^N) \to \im(\phi^N)$.
-If this map is surjective, then $\im(\phi^N)$ is a direct summand of $M$, with complementary summand $\coker \phi^N$: to see this, observe that the injection $\im(\phi^N) \inc M$ would then  compose with $\phi^N$ to be a surjection, and since $\im(\phi^N)$ is a finitely generated module, a surjection must be an isomorphism $\a:\im \phi^N \to \im \phi^N$ by \cite[Corollary~4.4]{Eisenbud95}.
-Thus, we obtain a splitting of the inclusion $\im \phi^N \inc M$ via $\a\inv\circ \phi^N$; since $\a\inv$ is an isomorphism, we have that $\im \phi^N$ is (isomorphic to) a direct summand of $M$, with complementary summand $\coker \phi^N$.
-
-  %% So, all that needs to be seen is that $\psi:=\psi^N$ is surjective.
-  %% To see this, note that after replacing the original generators for $M$ by the generators that put $\psi$ into the form above,  a generating set for $\im(\phi^N)$ is given by $f_i = \psi(m_i) = u_i\cdot e_i + \sum_{j<i } a_{ij} e_j$ for $i = 1,\dots,n-N_0$, where $u_i$ is a unit (explicitly, $u_i$ is a power of some nonzero $\lambda_{i'}\in k$).
-  %% That is, a generating set is given by
-  %% $$u_1 e_1,\quad  u_2 e_2+a_{21}e_1,\quad \dots ,\quad u_{n-n_0}e_{n-N_0}+a_{n-N_0,n-1}e_{n-N_0-1}+\dots + a_{N-n_0,1}e_1.$$
-  %% Since the $u_i$ are units, the $R$-span of these generators is clearly the same as the $R$-span of $e_1,\dots,e_{n-N_0}$, and thus after applying $\psi$ we get back the $R$-span of $f_1,\dots,f_{n-N_0} $, which is $\im \psi^N$.
-Thus, $\phi: \im \psi^N \to \im \psi^N$ is surjective, and the result follows.
+  Let $\alpha$ be the composition of these automorphisms for each $B_i$ and $\lambda_j$, then $\alpha^{-1}\circ\psi_j\circ\alpha$ corresponds to performing $k$-linear row and column operations and has ones and zeros on the diagonal.
+Finally, performing $R$-linear row and column operations, we may clear the rows and columns with a one on the diagonal to get $\psi'_j$, whose image and cokernel are summands and are isomorphic to the image and cokernel of $\psi_j$.
 \end{proof}
@@
 \end{rem}
-
-% The above can of course produce the trivial direct sum decomposition: if the eigenvalues of $\phi$ are all nonzero, the proof yields that $\im \psi = M$ and $\coker \psi = 0$.
-The following corollary of the proof of \Cref{lem:graded} ensures that we can always assume $\phi$ has zero as an eigenvalue:
-
-%% \begin{lem}
-%%   The eigenvalues of $\phi$ lie in $k$. In particular, for any $\phi \in [\End M]_0$, there is $\lambda \in k$ such that $\phi - \lambda \id_M \in [\End M]_0$ has zero as an eigenvalue.
-%% \end{lem}
-
-By ``eigenvalues of $\phi$'' we mean the roots of the characteristic polynomial $\det(\phi - \lambda \id_M)$\mahrud{define determinant over a module, or say determinant of its cover?}, which is a univariate polynomial in $\lambda$ with coefficients in $R$ and thus a priori has solutions only in the algebraic closure of the total ring of fractions of $R$.
-
-%% \begin{proof}
-%%   The expression of $\phi$ obtained in \cref{eq:Jordan} shows that the determinant of $\phi - \lambda \id_M$ is the product of $\lambda_i - \lambda$, thus the eigenvalues are the $\lambda_i$, which all lie in $k$.
-%% \end{proof}
 

### 2025-04-25 paper/main.tex
updated proof, nearly done

@@
 For both these problems, and many others, making and testing conjectures depends on computing summands of modules and verifying their indecomposability.
-The need for, and absence of, an algorithm to do so is noted, for example, in
+The need for and absence of an algorithm to do so is noted, for example, in
 \cite[15.10.9]{Eisenbud95}.
@@
 
-Finally, we present multiple examples, including some which present previously unknown phenomena regarding the behavior of summands of Frobenius pushforwards and syzygies over Artinian rings.
+Finally, we present multiple examples, including  previously unknown phenomena regarding the behavior of summands of Frobenius pushforwards and syzygies over Artinian rings.
 In particular, we highlight the results of \cite{CDE24}, which shows a recurrence formula for indecomposable summands of high syzygies of the residue field of Golod rings, made possible through experiments and observations using our algorithm.
@@
 For a graded or local ring $R$ with maximal ideal $\m$ and an $R$-module $M$, we write $\mu(M):=\dim_{R/\m}(M/\m M)$ for the minimal number of generators of $M$ as an $R$-module.
+If $R$ is graded and $M$ is homogeneous, we write $[M]_d$ for the $k$-vector space of degree-$d$ elements of $M$.
 
@@
 
-Let $R$ be a $\N$-graded ring, with $R_0 = k$ a field.
-%Assume furthermore that the degree support of $R$ is strongly convex, i.e., that
-%%the effective cone
-%$\Eff R = \{ \deg r : r\in R\}$ is a strongly convex cone in $\ZZ^n$.
-Suppose $M = \bigoplus M_d$ is a finitely generated, graded $R$-module and consider a general degree zero endomorphism $\phi \in [\End M]_0$.
+Let $R$ be a $\N$-graded ring with $R_0 = k$ a field.
+Suppose $M = \bigoplus M_d$ is a finitely generated, graded $R$-module and consider a general degree-zero endomorphism $\phi \in [\End M]_0$.
+Fix a set of minimal homogeneous generators $m_1,\dots,m_{\mu(M)}$ of $M$ so that $\phi$ may be presented as an $\mu(M)\times\mu(M)$ matrix with entries in $R$.
+We will refer to $\chi(\phi) = \det(\phi-\lambda\id_M)$ as the ``characteristic polynomial of $\phi$'', which is a univariate polynomial with coefficients in $R$ and thus a priori has solutions in $\lambda$ only in the algebraic closure of the total ring of fractions of $R$.
 
-%Let $P\subset \ZZ^n$ be the degrees of a set of minimal generators of $M$;
-%this is a poset with respect to the partial order $\a\leq \b$ if and only if $\b-\a\in \Eff R$.
-%Because $P$ is finite, we can order the elements of $P$ such that
-%$\deg m_1\not\leq \deg m_2\not \leq \dots\not\leq \deg m_r$.
-
-\begin{lem}\label{lem:graded}
-  \mahrud{rewrite statement} Let $N = \mu(M)$ and set $\psi = \phi^N$. Then $M = \im\psi \oplus \coker\psi$.
-\end{lem}
+\begin{prop}\label{prop:eigenvalues}
+  The solutions of the characteristic polynomial $\chi(\phi)$ are in $\bar k$.
+\end{prop}
 \begin{proof}
-  Fix homogeneous generators $m_1,\dots,m_N$ of $M$, with $\deg m_i$ nonincreasing, and let $d_1,\dots,d_n$ be the decreasing list of unique degrees of generators of $M$.
+  Since $M$ is $\N$-graded, assume its generators are ordered with $\deg m_i$ nonincreasing, and let $d_1,\dots,d_n$ be the decreasing list of unique degrees of generators of $M$.
 The matrix for $\phi$ is then ``block upper triangular'' with $n$ diagonal blocks with entries in $k$, i.e., of the form
@@
 To see this, note that since $\phi$ has degree zero, $\phi(m_i)$ is a $k$-linear combination of the $m_j$ with $\deg m_j = \deg m_i$ plus an $R$-linear combination of the $m_j$ with $\deg m_j < \deg m_i$.
-In particular, $B_i$ is the induced endomorphism on the submodule generated by all generators of degree $d_i$.
+%In particular, $B_i$ is an $k$-endomorphism on the $k$-vector space spanned by all generators of degree $d_i$.
 
-  Note that the characteristic polynomial $\det(\phi - \lambda \id_M) = \prod \det(B_i - \lambda \id)$ is a univariate polynomial in $\lambda$ with coefficients in $k$ which may have solutions in the algebraic closure of $k$.
-We will refer to the collection of $k$-eigenvalues of $B_i$ as the ``eigenvalues of $\phi$''.
-
-  By generality of $\phi$ suppose it has eigenvalues $\lambda_1,\dots,\lambda_r$ in $k$.
-Let $\mu(\lambda_j)$ be the maximum multiplicity of $\lambda$ in each diagonal block and set $\psi_j = (\phi - \lambda_j \id_M)^{\mu(\lambda_j)}$.
-We will show that $\psi_j$ is a split surjection, hence $M = \im\psi_j \oplus \coker\psi_j$.
-Furthermore, note that if $\phi_j$ and $\phi_k$ are different then $M = \coker\psi_j \oplus \coker\psi_k \oplus \im(\psi_j\circ\psi_k)$, thus we have a decomposition
-  \begin{equation}\label{eq:decomposition}
-    M = \coker\psi_1 \oplus \cdots \oplus \coker\psi_r \oplus \im\psi.
-  \end{equation}
-
-  To see that each $\psi_j$ is a split surjection, note that the set of generators of $M$ with degree $d_i$ are linearly independent and we can view each $B_i$ as a $k$-linear transformation of the $k$-vector space spanned by these generators.
-Thus, performing $k$-linear row and column operations on $(B_i - \lambda_j \id)^{\mu(\lambda_j)}$, we may write it as a diagonal matrix with ones and zeros on the diagonal.
-In particular, this is a $k$-linear change of basis on the degree $d_i$ generators of $M$ and induces an automorphism $\alpha_{i,j}$ of $M$.
-
-  Let $\alpha$ be the composition of these automorphisms for each $B_i$ and $\lambda_j$, then $\alpha^{-1}\circ\psi_j\circ\alpha$ corresponds to performing $k$-linear row and column operations and has ones and zeros on the diagonal.
-Finally, performing $R$-linear row and column operations, we may clear the rows and columns with a one on the diagonal to get $\psi'_j$, whose image and cokernel are summands and are isomorphic to the image and cokernel of $\psi_j$.
+  Then the characteristic polynomial $\det(\phi - \lambda \id_M) = \prod \det(B_i - \lambda \id)$ is a univariate polynomial in $\lambda$ with coefficients in $k$, thus has solutions in the algebraic closure of $k$.
 \end{proof}
 
-\begin{rem}
-  The hypotheses that $R$ is $\N$-graded and $k$ is an algebraically closed field may both be weakened.
-  In the proof of \Cref{lem:graded}, we only need a partial order on the degrees of the generators of $M$ in order to present $\phi$ in a block upper triangular form.
+We will refer to the collection of $k$-eigenvalues of $B_i$ as the ``eigenvalues of $\phi$''. Further, let $\mu(\lambda_j)$ be the sum of the multiplicities of $\lambda_j$ in each $B_i$.
+
+\begin{rem}\label{rem:grading}
+  The hypotheses that $R$ is $\N$-graded may be weakened.
+In the proof of \Cref{prop:eigenvalues}, we only need a partial order on the degrees of the generators of $M$ in order to present $\phi$ in a block upper triangular form \eqref{eq:blocks}.
 Suppose $R$ is $\ZZ^r$-graded, then the effective cone $\Eff(R)\subset\Q^n$ generated by the elements $\deg m$ for each monomial $m \in R$ induces a partial order on the degrees of generators of $M$ if it is strongly convex (contains no positive dimensional subspace of $\Q^n$) and $R_0 = k$.
+We will call such rings \emph{positively graded}.
 % TODO: can R_0 be finite dimensional?
+  This is satisfied, for instance, in the setting of multiparameter persistence modules or when $R$ is the Cox ring of a toric variety or Mori dream space graded by the class group of the variety.
+\end{rem}
 
-  Similarly, suppose $k$ is not algebraically closed, then there may be blocks $B_i$ which cannot be written in Jordan canonical form.
-Nevertheless, as long as $\phi$ has at least one eigenvalue $\lambda$ in $k$, the corresponding block may be written in Jordan canonical form, and thus the image and cokernel of $(\phi - \lambda_i\id_M)^{\mu_i}$ form summands of $M$.
-\end{rem}
+We will use nonzero eigenvalues of $\phi$ to get nontrivial splittings of $M$. As the following lemma shows, this approach allows for decomposing multiple summands at once.
+
+\begin{lem}\label{lem:decomposition}
+  Suppose $\psi_1,\dots,\psi_r$ are endomorphisms of $M$ such that $M\to\im\psi_i$ is a split surjection for all $i$ and set $\psi = \psi_r\circ\dots\circ\psi_1$.
+  %If $\ker\psi_i\cap\ker(\psi_{i-1}\circ\dots\circ\psi_1) = 0$ for all $i$ then $M$ has a direct sum decomposition \( M = \ker\psi_r \oplus \cdots \oplus \ker\psi_1 \oplus \im\psi. \)
+  If $\ker\psi_i\cap\ker \psi_j = 0$ for all $i\neq j$ then $M$ has a direct sum decomposition \( M = \ker\psi_r \oplus \cdots \oplus \ker\psi_1 \oplus \im\psi. \)
+\end{lem}
+\begin{proof}
+  Since each $\psi_i$ is a split surjection, there are direct sum decompositions $M=\im \psi_i \oplus \ker \psi_i$, and thus idempotents $e_i$ with $\im e_i = \im \psi_i$ and $\ker e_i = \ker \psi_i$.
+
+  The $e_i$ must commute: let $f_i = 1- e_i$ be the complementary idempotent (projection onto $\ker \psi_i$). Since $\ker \psi_i \cap \ker \psi_j=0$, $f_if_j=0$. In particular,
+  \[ e_ie_j =(1-f_i)(1-f_j) = 1-f_i-f_j= 1-f_j-f_i = e_j e_i. \]
+
+  Since the $e_i$ are commuting idempotents, $e_r\cdots e_1$ is also an idempotent.
+It's clear that $\im(e_r\cdots e_1) = \im (\psi_r\circ \dots\circ \psi_1) = \im \psi$.
+Moreover, we claim that $\ker(e_r\cdots e_1) = \ker \psi_r \oplus \cdots \oplus \ker \psi_1$.
+To see this, note that since $f_if_j=0$ we get
+  \[ 1-e_r\cdots e_1 = 1-(1-f_r)\cdots (1-f_1) = \sum f_i. \]
+  In other words, the kernel of $e_r\cdots e_1$ is exactly $\bigoplus \ker e_i =\bigoplus \ker \psi_i$.
+
+  Thus, the idempotent $e_r\cdots e_1$ corresponds to a direct sum decomposition
+  \[ M = \ker(e_r\cdots e_1) \oplus \im(e_r\cdots e_1) = \ker\psi_r \oplus \cdots \oplus \ker\psi_1 \oplus \im\psi, \]
+  as desired.
+\end{proof}
+
+Note that as long as each $\psi_i$ in \Cref{lem:decomposition} is nontrivial (not zero or identity endomorphisms), the kernel summands will be nontrivial, however, they may not necessarily be indecomposable.
+Also note that if the $\psi_i$ are complementary surjections, $\im\psi$ will be zero.
+
+\begin{prop}\label{prop:split-surj}
+  Let $\lambda_1,\dots,\lambda_r\in k$ be the eigenvalues of $\phi$ and set $\psi_j = (\phi - \lambda_j\id_M)^{\mu(\lambda_j)}$ and $\psi = \psi_1\circ\cdots\circ\psi_r$.
+Then $M$ has a direct sum decomposition
+  \( M = \ker\psi_1 \oplus \cdots \oplus \ker\psi_r \oplus \im\psi. \)
+\end{prop}
+\begin{proof}
+  First we show that each $\psi_j$ is a split surjection $M\to\im\psi_j$, hence $M = \ker\psi_j \oplus \im\psi_j$.
+To see this, note that $R$-linear row and column operations on endomorphisms of $M$ correspond to pre- and post-composition by $R$-linear maps on $M$.
+In particular, performing row and column operations only involving elements of $R_0 = k$ on $\phi$ we may transform each $B_i$ into Jordan canonical form
+  \begin{equation}
+    J_i = \begin{pmatrix}
+      J_{i,1} & 0       & \dots  & 0      \\
+      0       & J_{i,2} & \dots  & 0      \\
+      \vdots  & \vdots  & \ddots & \vdots \\
+      0       & 0       & \dots  & J_{i,t},
+    \end{pmatrix}
+  \end{equation}
+  where each Jordan block has entries over $k$ and a fixed eigenvalue.
+Suppose $J_{i,j}$ is the block with eigenvalue $\lambda_j$, then $N_i = J_{i,j} - \lambda_j$ is a nilpotent matrix whose size is the multiplicity of $\lambda-\lambda_j$ in $\chi(B_i)$.
+Since $\mu(\lambda_j)$ is the sum of the multiplicities of $\lambda_j$ in each $B_i$, the matrix $N_i^{\mu(\lambda_j)}$ vanishes, hence $(J_i-\lambda_j)^{\mu(\lambda_j)}$ is a matrix with some 1s along the diagonal and 0s elsewhere.
+This implies that there are auto morphisms $\alpha$ and $\beta$ of $M$ (corresponding to column operations clearing the entries above and to the right of each nonzero diagonal) such that $\alpha \circ (\phi - \lambda_j\id_M)^{\mu(\lambda_j)} \circ \beta: M\to M$ is represented by a diagonal matrix $\psi'_j$ with 1s and 0s along the diagonal, which is a split surjection.
+
+  Finally, if $\lambda_j$ and $\lambda_k$ are different then the nilpotent blocks in $J_i-\lambda_j$ and $J_i-\lambda_k$ are distinct, therefore $\ker\psi_j \cap \ker\psi_k = 0$, thus by \Cref{lem:decomposition} we have the desired decomposition.
+\end{proof}
+
+We are ready to state the algorithm for decomposition of a graded module. Recall from \Cref{rem:grading} that we only need the ring $R$ to be positively graded over an arbitrary field.
 
@@
     \smallskip
-    \Require graded module $M$ over a positively graded commutative ring $R$.
+    \Require graded module $M$ over a positively graded commutative $k$-algebra $R$.
     \Ensure  list $\textproc{Summands}(M)$ of indecomposable summands of $M$.
@@
       \State Let $\psi_i = (\phi - \lambda_i \id_M)^{\mu_i}$.
-      \State Append \Call{Summands}{$\coker\psi_i$} to $L$.
+      \State Append \Call{Summands}{$\ker\psi_i$} to $L$.
     \EndFor
@@
 
-We will see via \Cref{lem:term} that this will result in an indecomposable decomposition of $M$.
+Note that \Cref{alg:graded} is characteristic independent.
+In particular, since over a large enough field the eigenvalues of a general endomorphism $\phi$ are distinct, the summands $\ker\psi_i$ and $\im\psi$ will be indecomposable, thus \Cref{alg:graded} will produce indecomposable summands in one non-recursive iteration.
+We will see via \Cref{lem:term}\mahrud{or lem:distinct?} that this will result in an indecomposable decomposition of $M$ over any field.
 
@@
   % https://arxiv.org/abs/2208.05427
-  In recent work suggested by examples calculated using our algorithm, \cite{CDE24} studies the indecomposable summands of syzygy modules over a Golod ring $(R,\m,k)$.
-In particular, they find previously unexpected recurring behavior, specifically that the syzygy modules of the residue field are direct sums of only three indecomposable modules: the residue field $k$, the maximal ideal $\m$, and an additional module $N = \Hom_R(\m, R)$.
+  In recent work suggested by examples calculated using our algorithm, \cite{CDE24} studied the indecomposable summands of syzygy modules over a Golod ring $(R,\m,k)$ and found previously unexpected recurring behavior.
+Specifically, the syzygy modules of the residue field are direct sums of only three indecomposable modules: the residue field $k$, the maximal ideal $\m$, and an additional module $N = \Hom_R(\m, R)$.
 
-  Here, we give a concrete example of one such ring:
-  Let $k$ be any field and let
-  $R = k[x,y]/(x^3,x^2y^3,y^5)$ and consider the (infinite) minimal free resolution of the residue field, which has rank $2^n$ in homological index $n$.
-
+  Here, we give a concrete example of one such ring.
+Let $k$ be any field and let $R = k[x,y]/(x^3,x^2y^3,y^5)$ and consider the (infinite) minimal free resolution of the residue field, which has rank $2^n$ in homological index $n$.
 The fourth syzygy module of the residue field decomposes (ignoring the grading) as the direct sum
-  $$ k^3 \oplus \m^2 \oplus N^3. $$
+  $$ k^3 \oplus \m^2 \oplus N^3, $$
   and the fifth syzygy module as
-  $$ k^8\oplus \m^9 \oplus N^2. $$
-  The use of our algorithm was essential to the observation that beyond the ``guaranteed'' summands of $k$ and $\m$ (which were known to appear by work of \cite{DE23}) only the one additional indecomposable module $N$ appears in the summands of syzygies of $k$.
-
-  In this example, the module $N$ can be explicitly presented as
-  $$N=\coker
+  $$ k^8\oplus \m^9 \oplus N^2, $$
+  where the module $N$ can be explicitly presented as
+  \[ N = \coker
   \begin{pmatrix}
@@
     0&0&0&0&0&-y&x&y^{2}
-  \end{pmatrix}
-  $$
+  \end{pmatrix} \]
 
+  The use of our algorithm was essential to the observation that beyond the ``guaranteed'' summands of $k$ and $\m$ (which were known to appear by work of \cite{DE23}) only the one additional indecomposable module $N$ appears in the summands of syzygies of $k$.
   %% needsPackage "DirectSummands"

### 2025-04-27 paper/main.tex
paper edits before final push

@@
 \makeop{Eff}
+\def\wpsi{\widehat\psi}
 
@@
 \begin{abstract}
-  We describe and prove correctness of two practical algorithms for computing indecomposable summands of finitely generated modules over a finitely generated $k$-algebra $R$: the first algorithm applies in the (multi)graded case, which enables the computation of indecomposable summands of coherent sheaves on subvarieties of toric varieties (in particular, for varieties embedded in projective space); the second algorithm applies when $R$ is local and $k$ is a finite field, which opens the door to computing decompositions in singularity theory.
+  We describe and prove correctness of two practical algorithms for finding indecomposable summands of finitely generated modules over a finitely generated $k$-algebra $R$.
+The first algorithm applies in the (multi)graded case, which enables the computation of indecomposable summands of coherent sheaves on subvarieties of toric varieties (in particular, for varieties embedded in projective space); the second algorithm applies when $R$ is local and $k$ is a finite field, which opens the door to computing decompositions in singularity theory.
 We also present multiple examples, including some which present previously unknown phenomena regarding the behavior of summands of Frobenius pushforwards (including in the non-graded case) and syzygies over Artinian rings.
@@
 
-The problems of finding isomorphism classes of indecomposable modules with certain properties, or determining the indecomposable summands of a module, are ubiquitous in commutative algebra, representation theory, topological data analysis, and other fields.
+The problems of finding isomorphism classes of indecomposable modules with a given property, or determining the indecomposable summands of a module, are ubiquitous in commutative algebra and representation theory and even in areas such as topological data analysis.
 Within commutative algebra, for instance, the classification of rings $R$ for which there are only finitely many isomorphism classes of indecomposable maximal Cohen--Macaulay $R$-modules (the \emph{finite CM-type} property), or determining whether iterated Frobenius pushforwards of a Noetherian ring in positive characteristic have finitely many isomorphism classes of indecomposable summands (the \emph{finite F-representation type} property) are two well-established research problems (see, for example, \cite{Yoshino90,LW12} for the former and \cite{SVdB97,Hara15,TT08}, among many others, for the latter).
@@
 Suppose $M = \bigoplus M_d$ is a finitely generated, graded $R$-module and consider a general degree-zero endomorphism $\phi \in [\End M]_0$.
-Fix a set of minimal homogeneous generators $m_1,\dots,m_{\mu(M)}$ of $M$ so that $\phi$ may be presented as an $\mu(M)\times\mu(M)$ matrix with entries in $R$.
-We will refer to $\chi(\phi) = \det(\phi-\lambda\id_M)$ as the ``characteristic polynomial of $\phi$'', which is a univariate polynomial with coefficients in $R$ and thus a priori has solutions in $\lambda$ only in the algebraic closure of the total ring of fractions of $R$.
+Fix a set of minimal homogeneous generators $m_1,\dots,m_{\mu(M)}$ of $M$ so that $\phi$ may be presented as a $\mu(M)\times\mu(M)$ matrix with entries in $R$.
+We will refer to $\chi_\phi(\lambda) = \det(\phi-\lambda\id_M)$ as the ``characteristic polynomial of $\phi$'', which is a univariate polynomial with coefficients in $R$ and thus a priori has solutions only in the algebraic closure of the total ring of fractions of $R$.
 
 \begin{prop}\label{prop:eigenvalues}
-  The solutions of the characteristic polynomial $\chi(\phi)$ are in $\bar k$.
+  Let $A\colon M/\m M\to M/\m M$ be the $k$-linear map induced by $\phi\in[\End M]_0$.
+The solutions of the characteristic polynomial $\chi_\phi(\lambda)$ are the same as the eigenvalues of $A$, thus they lie in the algebraic closure $\bar k$.
 \end{prop}
@@
   \end{equation}
-  where the blocks $B_i$ are square matrices \emph{with entries in $k$}, and the remaining entries are in $R$.
+  where the blocks $B_i$ are square matrices \emph{with entries in $k$}, and the remaining entries are in $\m$.
 To see this, note that since $\phi$ has degree zero, $\phi(m_i)$ is a $k$-linear combination of the $m_j$ with $\deg m_j = \deg m_i$ plus an $R$-linear combination of the $m_j$ with $\deg m_j < \deg m_i$.
@@
 
-  Then the characteristic polynomial $\det(\phi - \lambda \id_M) = \prod \det(B_i - \lambda \id)$ is a univariate polynomial in $\lambda$ with coefficients in $k$, thus has solutions in the algebraic closure of $k$.
+  Then the characteristic polynomial $\det(\phi - \lambda \id_M) = \prod \det(B_i - \lambda \id) = \det(\phi - \lambda \id_{M/\m M})$ is a univariate polynomial in $\lambda$ with coefficients in $k$, thus has solutions in $\bar k$.
 \end{proof}
 
-We will refer to the collection of $k$-eigenvalues of $B_i$ as the ``eigenvalues of $\phi$''. Further, let $\mu(\lambda_j)$ be the sum of the multiplicities of $\lambda_j$ in each $B_i$.
+We will refer to the collection of $k$-eigenvalues of $B_i$ as the ``eigenvalues of $\phi$''.
+Further, let $\mu_j \coloneqq \mu(\lambda_j)$ be the sum of the multiplicities of $\lambda_j$ in each $B_i$.
+Note that the multiplicity of $\lambda_j$ in $\chi_\phi(\lambda)$ is the sum of the multiplicities of $\lambda_j$ in each $\chi_{B_i}(\lambda)$.
 
@@
 Note that as long as each $\psi_i$ in \Cref{lem:decomposition} is nontrivial (not zero or identity endomorphisms), the kernel summands will be nontrivial, however, they may not necessarily be indecomposable.
-Also note that if the $\psi_i$ are complementary surjections, $\im\psi$ will be zero.
+Also note that $\im\psi$ may be zero.
 
+\pagebreak
 \begin{prop}\label{prop:split-surj}
-  Let $\lambda_1,\dots,\lambda_r\in k$ be the eigenvalues of $\phi$ and set $\psi_j = (\phi - \lambda_j\id_M)^{\mu(\lambda_j)}$ and $\psi = \psi_1\circ\cdots\circ\psi_r$.
+  Let $\lambda_1,\dots,\lambda_r$ be the eigenvalues of $\phi$ over $k$ and set $\psi_j = (\phi - \lambda_j\id_M)^{\mu_j}$ and $\psi = \psi_1\circ\cdots\circ\psi_r$.
 Then $M$ has a direct sum decomposition
@@
 \begin{proof}
-  First we show that each $\psi_j$ is a split surjection $M\to\im\psi_j$, hence $M = \ker\psi_j \oplus \im\psi_j$.
+  First, suppose $\lambda_{r+1},\dots,\lambda_t$ are the other eigenvalues of $\phi$ over the algebraic closure $\bar k$ and set $\rho = \psi_{r+1}\circ\dots\circ\psi_t$, then by the Cayley--Hamilton theorem $\chi_\phi(\phi) = \psi_1\circ\dots\circ\psi_r\circ\rho = \psi\circ\rho = 0$.
+Since the factors in $\chi_\phi(\phi)$ can be reordered, we can write $0 = \psi\circ\rho = \rho \circ \psi$.
+In particular, suppose the desired direct sum decomposition existed over $\bar k$, then we have \mahrud{WHY??}
+  \begin{align*}
+    M = \ker\psi_1\oplus\dots\oplus\ker\psi_r\oplus(\ker\psi_{r+1}\oplus\dots\oplus\ker\psi_t)
+    &= \ker\psi_1\oplus\dots\oplus\ker\psi_r\oplus\ker\rho \\
+    &= \ker\psi_1\oplus\dots\oplus\ker\psi_r\oplus\im\psi.
+  \end{align*}
+  Thus it suffices to prove the direct sum decomposition only over the algebraic closure.
+
+  If $r = 1$ then $\im\psi_1=0$ is zero, hence we get the trivial decomposition $M = M \oplus 0$.
+
+  %  We will find a right inverse to $\psi_1$ and $\psi_2$ using $\chi_\phi(\phi) = 0$
+  If $r = 2$, let $A\colon M/\m M\to M/\m M$ be the $k$-linear map induced by $\phi$ and $A_i\coloneqq(A - \lambda_i)^{\mu_i}$ the map induced by $\psi_i$ for $i=1,2$. Observe that
+  \begin{align*}
+  \chi_\phi(\phi) = (\phi - \lambda_1\id_M)^{\mu_1} (\phi - \lambda_2\id_M)^{\mu_2}
+  &= \left(\phi^{\mu_1} - \dots + c_1\phi - c_0\id_M\right)\circ\psi_2 \\
+  &= \left(\phi^{\mu_1} - \dots + c_1\phi\right)\circ\psi_2 - c_0\psi_2 = 0,
+  \end{align*}
+  where $c_0 = -(-\lambda_1)^{\mu_1}$ and $c_1 = (-\lambda_1)^{\mu_1-1}$, and therefore we have a left pseudoinverse
+  \[
+  \psi_2
+  = c_0^{-1} \left(\phi^{\mu_2} - \dots + c_1\phi \right) \circ\psi_2
+  % = \phi\circ c_0^{-1} \left(\phi^{\mu_2-1} - \dots + c_1 \right) \circ\psi_2.
+  \]
+  This means we have a short exact sequence $0 \to \im\psi_2 \hookrightarrow M \to \coker\psi_2 \to 0$ which is split by the map $c_0^{-1} \left(\phi^{\mu_2} - \dots + c_1\phi \right)$, % which is to say $M \to \im\psi_2$ is a split surjection.
+
+  \begin{align*}
+  \chi_\phi(\phi) = (\phi - \lambda_2\id_M)^{\mu_2} (\phi - \lambda_1\id_M)^{\mu_1}
+  &= \psi_2\circ\left(\phi^{\mu_1} - \dots + c_1\phi - c_0\id_M\right) \\
+  &= \psi_2 c_0 - \psi_2\circ \left(\phi^{\mu_1} - \dots + c_1\phi\right) = 0,
+  \end{align*}
+
+  % $\lambda_1^{-1}A_1$ is an idempotent and by \Cref{lem:idemp} (and scaling by $\lambda_1$ for simplicity) we get $M = \ker\psi_1 \oplus \im\psi_1$.
+% \devlin{But $\lambda_1\inv A_1$ is not guaranteed to be an idempotent: e.g., if $\lambda_1 = 1$ with multiplicity 2, then $A_1$ has the block $\begin{pmatrix}1 & 2 \\ 0 & 1\end{pmatrix}$ (i.e., a $2\times 2$ Jordan block squared), and it's not an idempotent.}
+
+  Otherwise, observe that $\wpsi_j\coloneqq \psi_1\circ\dots\circ\psi_{j-1}\circ\psi_{j+1}\circ\dots\circ\psi_r $ has only one nonzero eigenvalue over $k$, namely $\widehat\lambda_j \coloneqq \prod_{i\neq j}(\lambda_j-\lambda_i)$: Using \Cref{prop:eigenvalues}, eigenvalues of $\phi$ are the same as the eigenvalues of $A$.
+Say $A v = \lambda_\ell v$ for $v\in M/\m M$ and apply the $k$-linear map induced by $\psi_1\circ\dots\circ\psi_r$ to get
+  \begin{align}\label{eq:eigenvalues-psi}
+    \prod(A-\lambda_i)v
+    &= (A - \lambda_1)\cdots(A - \lambda_{r-1})(\lambda_\ell - \lambda_r)v \nonumber \\
+    &= (A - \lambda_1)\cdots(\lambda_\ell - \lambda_{r-1})(\lambda_\ell - \lambda_r)v
+    = \dots = \prod(\lambda_\ell - \lambda_i)v,
+  \end{align}
+  where the product is over $i=1,\dots,r$.
+In particular, since the $\lambda_i$ are distict the $k$-linear map $\widehat A_j\coloneqq (A - \lambda_1)\dots(A-\lambda_{j-1})(A - \lambda_{j+1})\dots(A - \lambda_r)$ induced by $\wpsi_j$ removes the $(\lambda_\ell-\lambda_j)$ factor from \eqref{eq:eigenvalues-psi}, hence the eigenvalue of $\widehat A_j$ corresponding to $v$ is $\prod_{i\neq j}(\lambda_j-\lambda_i)$ if $\lambda_\ell=\lambda_j$ and 0 otherwise, with multiplicities $\mu_j$ and $\mu(M) - \mu_j$ respectively.
+
+  In particular, $\sfrac{\widehat A_j}{\widehat\lambda_j}$ is an idempotent, so by \Cref{lem:idemp} (and scaling by $\widehat\lambda_j$ for simplicity) we have $M = \ker\wpsi_j \oplus \im\wpsi_j$ and further each $\psi_j$ is a split surjection $M\to\im\psi_j = \ker\wpsi_j$.
+Finally, note that elements of $\ker\psi_i = \im\wpsi_i$ must be generated by columns of $\phi$ corresponding to the eigenvalue $\lambda_i$, hence $\ker\psi_i \cap \ker\psi_j = 0$ for $i\neq j$, thus by \Cref{lem:decomposition} we have the desired decomposition.
+\end{proof}
+
+%% \begin{proof}
+%%   First we show that each $\psi_j$ is a split surjection $M\to\im\psi_j$, hence $M = \ker\psi_j \oplus \im\psi_j$.
 To see this, note that $R$-linear row and column operations on endomorphisms of $M$ correspond to pre- and post-composition by $R$-linear maps on $M$.
 In particular, performing row and column operations only involving elements of $R_0 = k$ on $\phi$ we may transform each $B_i$ into Jordan canonical form
-  \begin{equation}
-    J_i = \begin{pmatrix}
-      J_{i,1} & 0       & \dots  & 0      \\
-      0       & J_{i,2} & \dots  & 0      \\
-      \vdots  & \vdots  & \ddots & \vdots \\
-      0       & 0       & \dots  & J_{i,t},
-    \end{pmatrix}
-  \end{equation}
-  where each Jordan block has entries over $k$ and a fixed eigenvalue.
-Suppose $J_{i,j}$ is the block with eigenvalue $\lambda_j$, then $N_i = J_{i,j} - \lambda_j$ is a nilpotent matrix whose size is the multiplicity of $\lambda-\lambda_j$ in $\chi(B_i)$.
-Since $\mu(\lambda_j)$ is the sum of the multiplicities of $\lambda_j$ in each $B_i$, the matrix $N_i^{\mu(\lambda_j)}$ vanishes, hence $(J_i-\lambda_j)^{\mu(\lambda_j)}$ is a matrix with some 1s along the diagonal and 0s elsewhere.
-This implies that there are auto morphisms $\alpha$ and $\beta$ of $M$ (corresponding to column operations clearing the entries above and to the right of each nonzero diagonal) such that $\alpha \circ (\phi - \lambda_j\id_M)^{\mu(\lambda_j)} \circ \beta: M\to M$ is represented by a diagonal matrix $\psi'_j$ with 1s and 0s along the diagonal, which is a split surjection.
+%%   \begin{equation}
+%%     J_i = \begin{pmatrix}
+%%       J_{i,1} & 0       & \dots  & 0      \\
+%%       0       & J_{i,2} & \dots  & 0      \\
+%%       \vdots  & \vdots  & \ddots & \vdots \\
+%%       0       & 0       & \dots  & J_{i,t},
+%%     \end{pmatrix}
+%%   \end{equation}
+%%   where each Jordan block has entries over $k$ and a fixed eigenvalue.
+Suppose $J_{i,j}$ is the block with eigenvalue $\lambda_j$, then $N_i = J_{i,j} - \lambda_j$ is a nilpotent matrix whose size is the multiplicity of $\lambda-\lambda_j$ in $\chi_{B_i}(\lambda)$.
+Since $\mu_j$ is the sum of the multiplicities of $\lambda_j$ in each $B_i$, the matrix $N_i^{\mu_j}$ vanishes, hence $(J_i-\lambda_j)^{\mu_j}$ is a matrix with some 1s along the diagonal and 0s elsewhere.
+This implies that there are auto morphisms $\alpha$ and $\beta$ of $M$ (corresponding to column operations clearing the entries above and to the right of each nonzero diagonal) such that $\alpha \circ (\phi - \lambda_j\id_M)^{\mu_j} \circ \beta: M\to M$ is represented by a diagonal matrix $\psi'_j$ with 1s and 0s along the diagonal, which is a split surjection.
 
-  Finally, if $\lambda_j$ and $\lambda_k$ are different then the nilpotent blocks in $J_i-\lambda_j$ and $J_i-\lambda_k$ are distinct, therefore $\ker\psi_j \cap \ker\psi_k = 0$, thus by \Cref{lem:decomposition} we have the desired decomposition.
-\end{proof}
+%%   Finally, if $\lambda_j$ and $\lambda_k$ are different then the nilpotent blocks in $J_i-\lambda_j$ and $J_i-\lambda_k$ are distinct, therefore $\ker\psi_j \cap \ker\psi_k = 0$, 
+%% \end{proof}
 
@@
 
-Note that \Cref{alg:graded} is characteristic independent.
-In particular, since over a large enough field the eigenvalues of a general endomorphism $\phi$ are distinct, the summands $\ker\psi_i$ and $\im\psi$ will be indecomposable, thus \Cref{alg:graded} will produce indecomposable summands in one non-recursive iteration.
-We will see via \Cref{lem:term}\mahrud{or lem:distinct?} that this will result in an indecomposable decomposition of $M$ over any field.
+We will see via \Cref{lem:distinct} that this will result in an indecomposable decomposition of $M$ over any field.
+
+\begin{rem}
+Note that \Cref{alg:graded} does not depend on the characteristic of the field.
+Since over a large enough field (i.e., when $|k|$ is larger than the number of summands) the eigenvalues of a general endomorphism $\phi$ are distinct (by a variant of the argument of \Cref{lem:term}),
+the summands $\ker\psi_i$ and $\im\psi$ will be indecomposable; thus \Cref{alg:graded} will produce indecomposable summands in one non-recursive iteration. 
+When $|k|$ is less than or approximately the number of summands, the algorithm will be recursive.
+\end{rem}
 
@@
 If $A$ is neither an isomorphism nor the zero morphism, then both factors are nonzero and $M$ is decomposable.
+
 Note that $A$ also acts on the $k$-vector space $M/\m M$.
@@
   \begin{itemize}
-  \item If $R$ is graded and $M$ is not indecomposable, then a general endomorphism of $M$ of degree zero has at least two distinct eigenvalues (necessarily in $k$)
+  \item If $R$ is graded and $M$ is not indecomposable, then a general endomorphism of $M$ of degree zero has at least two distinct eigenvalues (in $\bar k$)
   \item If $R$ is local and $M$ is not indecomposable, then a general endomorphism of $M$ of degree zero reduces to an endomorphism of $M/\m M$ with at least two distinct eigenvalues.
@@
 \begin{rem}
-  Note that the above algorithm is quite sensitive to the ground field $k$, because it needs all the eigenvalues of the endomorphism $A$ of $M/\m M$ to be contained in $k$.
+  Note that 
+\Cref{alg:graded} and \Cref{alg:local} are 
+ quite sensitive to the ground field $k$, because they needs all the eigenvalues of the endomorphism 
+%$A$ of $M/\m M$ 
+to be contained in $k$.
 While theoretically the issue can be avoided by working over an algebraically closed ground field $\bar k$, for practical use on a computer algebra system it is better to extend $k$ to some larger finite field.
@@
   See \Cref{ex:elliptic} for a demonstration of the necessity of extending the base field.
+
+We also note that the algorithm detects the necessity of a field extension: if the characteristic polynomial of the endomorphism does not factor over $k$, the algorithm will calculate the splitting field of $k$; the user can then opt to extend the ground field to this splitting field.
 \end{rem}

### 2025-04-28 paper/main.tex
with any luck, everything is correct

@@
 
-We will refer to the collection of $k$-eigenvalues of $B_i$ as the ``eigenvalues of $\phi$''.
+We will refer to the collection of eigenvalues of $B_i$, extending the field if necessary, as the ``eigenvalues of $\phi$''.
 Further, let $\mu_j \coloneqq \mu(\lambda_j)$ be the sum of the multiplicities of $\lambda_j$ in each $B_i$.
@@
 
-\pagebreak
 \begin{prop}\label{prop:split-surj}
@@
 \begin{proof}
-  First, suppose $\lambda_{r+1},\dots,\lambda_t$ are the other eigenvalues of $\phi$ over the algebraic closure $\bar k$ and set $\rho = \psi_{r+1}\circ\dots\circ\psi_t$, then by the Cayley--Hamilton theorem $\chi_\phi(\phi) = \psi_1\circ\dots\circ\psi_r\circ\rho = \psi\circ\rho = 0$.
-Since the factors in $\chi_\phi(\phi)$ can be reordered, we can write $0 = \psi\circ\rho = \rho \circ \psi$.
-In particular, suppose the desired direct sum decomposition existed over $\bar k$, then we have \mahrud{WHY??}
+  First note that if $\psi_i$ is injective on $\im\psi_i$, then $\im\psi_i$ is a summand: the composition $\im\psi_i\to M\to \im\psi_i$ is injective, hence it is an isomorphism, which means the inclusion $\im\psi_i\to M$ is split.
+Therefore using the short exact sequence $0\to\ker\psi_i\to M\to\im\psi_i\to 0$ we get the decomposition $M = \ker\psi_i\oplus\im\psi_i$.
+
+  To see that $\psi_i$ is injective on $\im\psi_i$, first for simplicity replace $\phi$ by $\phi-\lambda_i\id_M$ so $\lambda_i = 0$ is an eigenvalue of $\phi$ and $\psi_i = \phi^{\mu_i}$.
+By the Cayley--Hamilton theorem $\psi_i$ divides $\chi_\phi(\phi)$, hence
   \begin{align*}
-    M = \ker\psi_1\oplus\dots\oplus\ker\psi_r\oplus(\ker\psi_{r+1}\oplus\dots\oplus\ker\psi_t)
-    &= \ker\psi_1\oplus\dots\oplus\ker\psi_r\oplus\ker\rho \\
-    &= \ker\psi_1\oplus\dots\oplus\ker\psi_r\oplus\im\psi.
+    \chi_\phi(\phi) = \left(\phi^{\mu(M) - \mu_i} - \dots + c_1\phi - c_0\right) \circ \psi_i
+    % &= \left(\phi^{\mu(M) - \mu_i} - \dots + c_1\phi\right) \circ \psi_i - c_0 \psi_i \\
+    &= \left(\phi^{\mu(M) - \mu_i - 1} - \dots + c_1\right) \circ \phi \circ \psi_i - c_0 \psi_i,
   \end{align*}
-  Thus it suffices to prove the direct sum decomposition only over the algebraic closure.
+  for some $c_0,c_1$ in $k$, and therefore
+  \( \psi_i = c_0^{-1} \left(\phi^{\mu(M) - \mu_i - 1} - \dots + c_1\right) \circ \phi \circ \psi_i. \)
+  This means that $\phi$ is injective on $\im\psi_i$, therefore $\psi_i=\phi^{\mu_i}$ is injective on $\im\psi_i$.
 
-  If $r = 1$ then $\im\psi_1=0$ is zero, hence we get the trivial decomposition $M = M \oplus 0$.
+  Moreover, using the standard proof for vector spaces, we know $\ker\psi_i\cap\ker\psi_j = 0$ if $i\neq j$: suppose $(\phi - \lambda_i)^m(v) = 0$ for some $v\in M$ and $m$ is minimal so that $w = (\phi-\lambda_i)^{m-1}(v) \neq 0$ satisfies $(\phi-\lambda_i)(w) = 0$, but then since $(\phi - \lambda_j)(\phi - \lambda_i) = (\phi - \lambda_i)(\phi - \lambda_j)$ we get
+  \[
+  (\phi - \lambda_j)(w)
+  = (\phi - \lambda_j)(\phi - \lambda_i)^{m-1}(v)
+  =  (\phi - \lambda_i)^{m-1}(\phi - \lambda_j)(v) = 0,
+  \]
+  which implies
+  \(
+  0 = (\phi - \lambda_j)(w) = (\phi - \lambda_i + \lambda_i - \lambda_j)(w) = (\lambda_i - \lambda_j)(w),
+  \)
+  which is a contradiction.
 
-  %  We will find a right inverse to $\psi_1$ and $\psi_2$ using $\chi_\phi(\phi) = 0$
-  If $r = 2$, let $A\colon M/\m M\to M/\m M$ be the $k$-linear map induced by $\phi$ and $A_i\coloneqq(A - \lambda_i)^{\mu_i}$ the map induced by $\psi_i$ for $i=1,2$. Observe that
-  \begin{align*}
-  \chi_\phi(\phi) = (\phi - \lambda_1\id_M)^{\mu_1} (\phi - \lambda_2\id_M)^{\mu_2}
-  &= \left(\phi^{\mu_1} - \dots + c_1\phi - c_0\id_M\right)\circ\psi_2 \\
-  &= \left(\phi^{\mu_1} - \dots + c_1\phi\right)\circ\psi_2 - c_0\psi_2 = 0,
-  \end{align*}
-  where $c_0 = -(-\lambda_1)^{\mu_1}$ and $c_1 = (-\lambda_1)^{\mu_1-1}$, and therefore we have a left pseudoinverse
-  \[
-  \psi_2
-  = c_0^{-1} \left(\phi^{\mu_2} - \dots + c_1\phi \right) \circ\psi_2
-  % = \phi\circ c_0^{-1} \left(\phi^{\mu_2-1} - \dots + c_1 \right) \circ\psi_2.
-  \]
-  This means we have a short exact sequence $0 \to \im\psi_2 \hookrightarrow M \to \coker\psi_2 \to 0$ which is split by the map $c_0^{-1} \left(\phi^{\mu_2} - \dots + c_1\phi \right)$, % which is to say $M \to \im\psi_2$ is a split surjection.
-
-  \begin{align*}
-  \chi_\phi(\phi) = (\phi - \lambda_2\id_M)^{\mu_2} (\phi - \lambda_1\id_M)^{\mu_1}
-  &= \psi_2\circ\left(\phi^{\mu_1} - \dots + c_1\phi - c_0\id_M\right) \\
-  &= \psi_2 c_0 - \psi_2\circ \left(\phi^{\mu_1} - \dots + c_1\phi\right) = 0,
-  \end{align*}
-
-  % $\lambda_1^{-1}A_1$ is an idempotent and by \Cref{lem:idemp} (and scaling by $\lambda_1$ for simplicity) we get $M = \ker\psi_1 \oplus \im\psi_1$.
-% \devlin{But $\lambda_1\inv A_1$ is not guaranteed to be an idempotent: e.g., if $\lambda_1 = 1$ with multiplicity 2, then $A_1$ has the block $\begin{pmatrix}1 & 2 \\ 0 & 1\end{pmatrix}$ (i.e., a $2\times 2$ Jordan block squared), and it's not an idempotent.}
-
-  Otherwise, observe that $\wpsi_j\coloneqq \psi_1\circ\dots\circ\psi_{j-1}\circ\psi_{j+1}\circ\dots\circ\psi_r $ has only one nonzero eigenvalue over $k$, namely $\widehat\lambda_j \coloneqq \prod_{i\neq j}(\lambda_j-\lambda_i)$: Using \Cref{prop:eigenvalues}, eigenvalues of $\phi$ are the same as the eigenvalues of $A$.
-Say $A v = \lambda_\ell v$ for $v\in M/\m M$ and apply the $k$-linear map induced by $\psi_1\circ\dots\circ\psi_r$ to get
-  \begin{align}\label{eq:eigenvalues-psi}
-    \prod(A-\lambda_i)v
-    &= (A - \lambda_1)\cdots(A - \lambda_{r-1})(\lambda_\ell - \lambda_r)v \nonumber \\
-    &= (A - \lambda_1)\cdots(\lambda_\ell - \lambda_{r-1})(\lambda_\ell - \lambda_r)v
-    = \dots = \prod(\lambda_\ell - \lambda_i)v,
-  \end{align}
-  where the product is over $i=1,\dots,r$.
-In particular, since the $\lambda_i$ are distict the $k$-linear map $\widehat A_j\coloneqq (A - \lambda_1)\dots(A-\lambda_{j-1})(A - \lambda_{j+1})\dots(A - \lambda_r)$ induced by $\wpsi_j$ removes the $(\lambda_\ell-\lambda_j)$ factor from \eqref{eq:eigenvalues-psi}, hence the eigenvalue of $\widehat A_j$ corresponding to $v$ is $\prod_{i\neq j}(\lambda_j-\lambda_i)$ if $\lambda_\ell=\lambda_j$ and 0 otherwise, with multiplicities $\mu_j$ and $\mu(M) - \mu_j$ respectively.
-
-  In particular, $\sfrac{\widehat A_j}{\widehat\lambda_j}$ is an idempotent, so by \Cref{lem:idemp} (and scaling by $\widehat\lambda_j$ for simplicity) we have $M = \ker\wpsi_j \oplus \im\wpsi_j$ and further each $\psi_j$ is a split surjection $M\to\im\psi_j = \ker\wpsi_j$.
-Finally, note that elements of $\ker\psi_i = \im\wpsi_i$ must be generated by columns of $\phi$ corresponding to the eigenvalue $\lambda_i$, hence $\ker\psi_i \cap \ker\psi_j = 0$ for $i\neq j$, thus by \Cref{lem:decomposition} we have the desired decomposition.
+  Finally, \Cref{lem:decomposition} implies the desired decomposition.
 \end{proof}
-
-%% \begin{proof}
-%%   First we show that each $\psi_j$ is a split surjection $M\to\im\psi_j$, hence $M = \ker\psi_j \oplus \im\psi_j$.
-To see this, note that $R$-linear row and column operations on endomorphisms of $M$ correspond to pre- and post-composition by $R$-linear maps on $M$.
-In particular, performing row and column operations only involving elements of $R_0 = k$ on $\phi$ we may transform each $B_i$ into Jordan canonical form
-%%   \begin{equation}
-%%     J_i = \begin{pmatrix}
-%%       J_{i,1} & 0       & \dots  & 0      \\
-%%       0       & J_{i,2} & \dots  & 0      \\
-%%       \vdots  & \vdots  & \ddots & \vdots \\
-%%       0       & 0       & \dots  & J_{i,t},
-%%     \end{pmatrix}
-%%   \end{equation}
-%%   where each Jordan block has entries over $k$ and a fixed eigenvalue.
-Suppose $J_{i,j}$ is the block with eigenvalue $\lambda_j$, then $N_i = J_{i,j} - \lambda_j$ is a nilpotent matrix whose size is the multiplicity of $\lambda-\lambda_j$ in $\chi_{B_i}(\lambda)$.
-Since $\mu_j$ is the sum of the multiplicities of $\lambda_j$ in each $B_i$, the matrix $N_i^{\mu_j}$ vanishes, hence $(J_i-\lambda_j)^{\mu_j}$ is a matrix with some 1s along the diagonal and 0s elsewhere.
-This implies that there are auto morphisms $\alpha$ and $\beta$ of $M$ (corresponding to column operations clearing the entries above and to the right of each nonzero diagonal) such that $\alpha \circ (\phi - \lambda_j\id_M)^{\mu_j} \circ \beta: M\to M$ is represented by a diagonal matrix $\psi'_j$ with 1s and 0s along the diagonal, which is a split surjection.
-
-%%   Finally, if $\lambda_j$ and $\lambda_k$ are different then the nilpotent blocks in $J_i-\lambda_j$ and $J_i-\lambda_k$ are distinct, therefore $\ker\psi_j \cap \ker\psi_k = 0$, 
-%% \end{proof}
 
@@
 
-We will see via \Cref{lem:distinct} that this will result in an indecomposable decomposition of $M$ over any field.
+We will see via \Cref{lem:distinct} that this will result in an indecomposable decomposition of $M$.
 
@@
   Note that \Cref{alg:graded} does not depend on the characteristic of the field.
-Since over a large enough field (i.e., when $|k|$ is larger than the number of summands) the eigenvalues of a general endomorphism $\phi$ are distinct (by a variant of the argument of \Cref{lem:term}),
-the summands $\ker\psi_i$ and $\im\psi$ will be indecomposable; thus \Cref{alg:graded} will produce indecomposable summands in one non-recursive iteration. 
+Since over a large enough field (i.e., when $|k|$ is larger than the number of summands) the eigenvalues of a general endomorphism $\phi$ are distinct (by a variant of the argument of \Cref{lem:term}), the summands $\ker\psi_i$ and $\im\psi$ will be indecomposable; thus \Cref{alg:graded} will produce indecomposable summands in one non-recursive iteration.
 When $|k|$ is less than or approximately the number of summands, the algorithm will be recursive.

### 2025-04-28 paper/main.tex
final changes

@@
 \let\b\beta
+
+%\usepackage[notref,notcite]{showkeys}
 
@@
 For both these problems, and many others, making and testing conjectures depends on computing summands of modules and verifying their indecomposability.
-The need for and absence of an algorithm to do so is noted, for example, in
-\cite[15.10.9]{Eisenbud95}.
 
-Currently there are no algorithms available for checking indecomposability or finding summands of arbitrary modules over commutative rings.
-In contrast, variants of the ``Meat-Axe'' algorithm for determining irreducibility of finite-dimensional modules over a group algebra have wide ranging applications in computational group theory \cite{Parker84,HR94,Holt98} and are available through symbolic algebra software such as Magma and GAP \cite{MAGMA,GAP}.
+Currently there are no algorithms available for performing these tasks on an arbitrary module over a commutative ring.
+The need for and absence of such algorithms is noted, for example, in \cite[\S15.10.9]{Eisenbud95} and \cite[\S8.5]{BL23}.
+In contrast, variants of the ``Meat-Axe'' algorithm for determining irreducibility of finite-dimensional modules over a group algebra have wide ranging applications in computational group theory \cite{Parker84,HR94,Holt98,IL00} and are available through symbolic algebra software such as Magma and GAP \cite{MAGMA,GAP}.
 
@@
 
-The purpose of this paper is to describe and prove correctness of two practical algorithms for computing indecomposable summands of finitely generated modules over a finitely generated $k$-algebra $R$, when either $R$ is (multi)graded and $k$ an arbitrary field, or $R$ local and $k$ a field of positive characteristic.
-In the graded case, our algorithm is often capable of completely decomposing a module in a single pass, while in the local case it proceeds by splitting the module into two summands and continuing recursively.
-%We also describe applications of our algorithm for computations in characteristic 0.
-The multigraded case of our algorithm enables the computation of indecomposable summands of coherent \emph{sheaves} on subvarieties of toric varieties (in particular, for varieties embedded in $\PP^n$).
+The purpose of this paper is to describe and prove correctness of two practical algorithms for computing indecomposable summands of finitely generated modules over a $k$-algebra $R$: \Cref{alg:graded} if $M$ is a homogeneous module over a (multi)graded ring $R$ with $k$ a field of arbitrary characteristic, and \Cref{alg:local} if $R$ is local and $k$ a field of positive characteristic.
+Over a large enough field, our algorithms are often capable of completely decomposing a module in a single non-recursive iteration.
+
+In the graded case, \Cref{alg:graded} enables the computation of indecomposable summands of coherent sheaves on subvarieties of the projective space $\PP^n$ as well as other toric varieties, while in the local case \Cref{alg:local} enables the study of germs of inhomogeneous singularities in positive characteristic.
 
@@
 
-\begin{rem}
+\begin{rem}\label{rem:missing}
   There is a ``missing'' case of our algorithm: when $R$ is only local, not graded, and $k$ has characteristic 0.
@@
 See \Cref{rem:char}.
-  %Although the local algorithm described below is only proved to result (probabilistically) in a decomposition into indecomposable summands in positive characteristic, in practice it often does produce nontrivial indecomposable decompositions even in characteristic 0.
-  %Although the algorithm described below is only proved to result (probabilistically) in a decomposition into indecomposable summands in positive characteristic, in practice it often does produce nontrivial indecomposable decompositions even in characteristic 0.
-Moreover, if a module over a ring of characteristic 0 is decomposable, its reductions modulo $p$ will be as well; thus, our algorithm provides a heuristic for verifying decomposability in characteristic 0.
 
-  We note also that while the discussion below, and our implementation in Macaulay2, concerns the case where $R$ is a \emph{commutative} ring, the basic techniques extend beyond this case.
+  While this paper, and our implementation in Macaulay2, concerns the case where $R$ is a \emph{commutative} ring, the basic techniques may extend beyond this case.
 We plan to extend the results and algorithms to non-commutative rings, such as Weyl algebras, in future work.
@@
 
-For a graded or local ring $R$ with maximal ideal $\m$ and an $R$-module $M$, we write $\mu(M):=\dim_{R/\m}(M/\m M)$ for the minimal number of generators of $M$ as an $R$-module.
-If $R$ is graded and $M$ is homogeneous, we write $[M]_d$ for the $k$-vector space of degree-$d$ elements of $M$.
+For a local or graded ring $R$ with maximal ideal $\m$ and an $R$-module $M$, we write $\mu(M)\coloneqq\dim_{R/\m}(M/\m M)$ for the minimal number of generators of $M$ as an $R$-module.
+If $R$ is graded and $M$ is homogeneous, we write $[M]_d$ for the $k$-span of its degree-$d$ elements.
 
-% TODO: \mahrud{k(x) and k(X) are used only in one place, move there?}
 For an algebraic variety $X$ and a point $x\in X$, we write $k(x)=\OO_{X,x}/\m_x$ for the residue field of $X$ at $x$ and $k(X)$ for the function field of $X$.
@@
 Conversely, if $M$ is a finitely generated, graded $S$-module, we write $\wtilde M$ for the coherent sheaf associated to $M$.
+
+We will use Greek letters for endomorphisms of $M$ and capital Latin letters for the induced $k$-linear maps on $M/\m M$.
 
@@
 Let $R$ be a $\N$-graded ring with $R_0 = k$ a field.
-Suppose $M = \bigoplus M_d$ is a finitely generated, graded $R$-module and consider a general degree-zero endomorphism $\phi \in [\End M]_0$.
+Suppose $M = \bigoplus [M]_d$ is a finitely generated, graded $R$-module and consider a general degree-zero endomorphism $\phi \in [\End_R(M)]_0$.
 Fix a set of minimal homogeneous generators $m_1,\dots,m_{\mu(M)}$ of $M$ so that $\phi$ may be presented as a $\mu(M)\times\mu(M)$ matrix with entries in $R$.
@@
 \begin{prop}\label{prop:eigenvalues}
-  Let $A\colon M/\m M\to M/\m M$ be the $k$-linear map induced by $\phi\in[\End M]_0$.
+  Let $A\colon M/\m M\to M/\m M$ be the $k$-linear map induced by $\phi\in[\End_R(M)]_0$.
 The solutions of the characteristic polynomial $\chi_\phi(\lambda)$ are the same as the eigenvalues of $A$, thus they lie in the algebraic closure $\bar k$.
@@
 
-  Then the characteristic polynomial $\det(\phi - \lambda \id_M) = \prod \det(B_i - \lambda \id) = \det(\phi - \lambda \id_{M/\m M})$ is a univariate polynomial in $\lambda$ with coefficients in $k$, thus has solutions in $\bar k$.
+  Then the characteristic polynomial $\det(\phi - \lambda \id_M) = \prod \det(B_i - \lambda) = \det(A - \lambda \id_{M/\m M})$ is a univariate polynomial in $\lambda$ with coefficients in $k$, thus has solutions in $\bar k$.
 \end{proof}
 
-We will refer to the collection of eigenvalues of $B_i$, extending the field if necessary, as the ``eigenvalues of $\phi$''.
-Further, let $\mu_j \coloneqq \mu(\lambda_j)$ be the sum of the multiplicities of $\lambda_j$ in each $B_i$.
-Note that the multiplicity of $\lambda_j$ in $\chi_\phi(\lambda)$ is the sum of the multiplicities of $\lambda_j$ in each $\chi_{B_i}(\lambda)$.
+\begin{dfn}\label{def:eigenvalues}
+  The collection of eigenvalues of $B_i$ in the splitting field of $\chi_\phi(\lambda)$ are the \emph{eigenvalues of $\phi$}.
+We will use $\mu_j \coloneqq \mu(\lambda_j)$ for the sum of the (geometric) multiplicities of $\lambda_j$ in each $B_i$.
+Note that the (algebraic) multiplicity of $\lambda_j$ in $\chi_\phi(\lambda)$ is the sum of the multiplicities of $\lambda_j$ in each $\chi_{B_i}(\lambda)$ and may be larger than $\mu_j$.
+\end{dfn}
 
@@
 \begin{proof}
-  First note that if $\psi_i$ is injective on $\im\psi_i$, then $\im\psi_i$ is a summand: the composition $\im\psi_i\to M\to \im\psi_i$ is injective, hence it is an isomorphism, which means the inclusion $\im\psi_i\to M$ is split.
-Therefore using the short exact sequence $0\to\ker\psi_i\to M\to\im\psi_i\to 0$ we get the decomposition $M = \ker\psi_i\oplus\im\psi_i$.
+  If $\psi_i$ is injective on $\im\psi_i$, then $\im\psi_i$ is a summand: in this case, the composition $\im\psi_i\hookrightarrow M\to \im\psi_i$ is injective, but a degree-zero endomorphism of a finitely generated $R$-module is an isomorphism (it must be injective, hence surjective, on each finite-dimensional $[M]_d$), and thus the inclusion $\im \psi \to M$ is split by the surjection $M\to \im \psi$.
+Therefore the short exact sequence $0\to\ker\psi_i\to M\to\im\psi_i\to 0$ is split and $M = \ker\psi_i\oplus\im\psi_i$.
 
-  To see that $\psi_i$ is injective on $\im\psi_i$, first for simplicity replace $\phi$ by $\phi-\lambda_i\id_M$ so $\lambda_i = 0$ is an eigenvalue of $\phi$ and $\psi_i = \phi^{\mu_i}$.
+  To see that $\psi_i$ is injective on $\im\psi_i$, first for simplicity replace $\phi$ by $\phi-\lambda_i\id_M$ so $\psi_i = \phi^{\mu_i}$.
 By the Cayley--Hamilton theorem $\psi_i$ divides $\chi_\phi(\phi)$, hence
@@
     % &= \left(\phi^{\mu(M) - \mu_i} - \dots + c_1\phi\right) \circ \psi_i - c_0 \psi_i \\
-    &= \left(\phi^{\mu(M) - \mu_i - 1} - \dots + c_1\right) \circ \phi \circ \psi_i - c_0 \psi_i,
+    &= \left(\phi^{\mu(M) - \mu_i - 1} - \dots + c_1\right) \circ \phi \circ \psi_i - c_0 \psi_i = 0,
   \end{align*}
@@
     \ForAll { eigenvalues $\lambda_i$ of $\phi$ }
-      \State Let $\mu_i$ be the multiplicity of $\lambda_i$.
-      \State Let $\psi_i = (\phi - \lambda_i \id_M)^{\mu_i}$.
+      \State Let $\mu_i$ be the (geometric) multiplicity of $\lambda_i$.
+      \State Let $\psi_i = (\phi - \lambda_i \id_M)^{\mu_i}$ (or, for simplicity, raised to $\mu(M)$).
+      \State By \Cref{prop:split-surj}, $\psi_i$ is a split surjection, hence producing a splitting of $M$.
       \State Append \Call{Summands}{$\ker\psi_i$} to $L$.
@@
 
-We will see via \Cref{lem:distinct} that this will result in an indecomposable decomposition of $M$.
+We will see via \Cref{lem:distinct} that this results in an indecomposable decomposition of $M$.
 
-\begin{rem}
+\begin{rem}\label{rem:graded}
   Note that \Cref{alg:graded} does not depend on the characteristic of the field.
-Since over a large enough field (i.e., when $|k|$ is larger than the number of summands) the eigenvalues of a general endomorphism $\phi$ are distinct (by a variant of the argument of \Cref{lem:term}), the summands $\ker\psi_i$ and $\im\psi$ will be indecomposable; thus \Cref{alg:graded} will produce indecomposable summands in one non-recursive iteration.
-When $|k|$ is less than or approximately the number of summands, the algorithm will be recursive.
+Further, since over a large enough field (i.e., when $|k|$ is larger than the number of summands) the eigenvalues of a general endomorphism $\phi$ are distinct (e.g., by a variant of \Cref{lem:distinct}), the summands $\ker\psi_i$ and $\im\psi$ will be indecomposable; thus \Cref{alg:graded} will produce an indecomposable decomposition of $M$ in one non-recursive iteration.
+When $|k|$ is less than or approximately the same as the number of summands, the algorithm will be recursive.
 \end{rem}
@@
 
-Throughout this section, $(R,\m)$ will be either local with maximal ideal $\m$, or a (multi)graded ring with $R_0 = k \subset \overline{\FF_p}$ a field of positive characteristic and homogeneous maximal ideal $\m$.
+Throughout this section, $R$ will be a local ring over a field of positive characteristic $k \subset \overline{\FF_p}$ with maximal ideal $\m$ and $M$ a finitely generated $R$-module.
+Recall that if $\psi\in\End_R(M)$ is an idempotent, then $M$ decomposes as $\im \psi \oplus \coker \psi$.
+If $\psi$ is nontrivial (not zero or an isomorphism), then both factors are nonzero and $M$ is decomposable.
 
-We begin with the observation that if $M$ is a finitely generated $R$-module and $A\in\End_R(M)$ is an idempotent, then $M$ decomposes as $\im A \oplus \coker A$.
-If $A$ is neither an isomorphism nor the zero morphism, then both factors are nonzero and $M$ is decomposable.
-
-Note that $A$ also acts on the $k$-vector space $M/\m M$.
-
-The following lemma allows us to check only for idempotents modulo the maximal ideal.
+We begin with the observation that $\psi$ also acts on the $k$-vector space $M/\m M$. The following lemma allows us to check only for idempotents modulo the maximal ideal.
 
 \begin{lem}\label{lem:idemp}
-  Let $M$ be a finitely generated $R$-module, and let $A\in\End_R(M)$. If the induced action of $A$ on $M/\m M$ is idempotent, then $M$ admits a direct sum decomposition $\im A \oplus \coker A$.
+  Let $A\colon M/\m M\to M/\m M$ be the induced action of $\psi\in\End_R(M)$ on $M/\m M$. If $A$ is an idempotent, then $M$ admits a direct sum decomposition $\ker\psi \oplus \im\psi$.
 \end{lem}
 \begin{proof}
-  By assumption, we can write $A^2 = A + B$, where $B\in\End_R(M)$ with $B(M)\subset \m M$.
-  Note that if $x \in \m^kM$, then $A^2(x) - A(x) = B(x)$ lies in $\m^{k+1}M$.
+  Let $N \coloneqq \im\psi$. We want to show that $0 \to N \inc M$ splits. Consider the composition
+  \begin{equation}\label{eq:N-M-N}
+    N \inc M \xra{\psi} N.
+  \end{equation}
+  We claim that this composition is surjective. Since a surjective endomorphism of finitely generated modules is invertible, we conclude that this composition is an isomorphism $\alpha$ on $N$.
+  Therefore the inclusion $ 0 \to N \inc M $ is split by the composition
+  \[ M \xra{\psi} N \xra{\alpha\inv} N, \]
+  and thus $M$ decomposes as claimed.
 
-  Let $N = \im A$. We want to show that $0 \to N\inc M$ splits. Consider the composition
-  \[ N\inc M \xra{A} \im A = N.\]
-  We claim that this composition is surjective.
-Since a surjective endomorphism of finitely generated modules is invertible, we would then conclude that this composition is an isomorphism on $N$; say $\a$.
-  Therefore the inclusion \( 0 \to N \inc M \) is split by the map of $R$-modules $M\xra{A} N\xra{\a\inv} N$, and thus $M$ decomposes as claimed.
+  To check the surjectivity of \eqref{eq:N-M-N}, we may complete at the maximal ideal and assume $R$, $M$, and $N$ are complete.
+Observe that by assumption we can write $\psi^2 = \psi + \psi'$ with $\psi'(M)\subset \m M$.
+  Note that if $x \in \m^\ell M$, then $\psi'(x) = \psi^2(x) - \psi(x)$ lies in $\m^{\ell+1}M$.
 
-  To check the surjectivity of $N\inc M \xra{A} N$,  we may complete at the maximal ideal and thus assume $R$, $M$, and $N$ are complete.
-Let $n_0\in N$.
-By assumption, $n_0=A(m_1)$ for some $m_1\in M$.
-Applying $A$ again, we get
-  \[ A(n_0) = A^2(m_1) = A(m_1) + n_1 = n_0 + n_1, \]
-  or equivalently
-  \[ n_0 = A(n_0) - n_1 \]
-  for some $n_1\in \m M$. In fact, since $n_0$ and $A(n_0)$ are both in $N$, we have $n_1\in N$ also, so $n_1\in \m M\cap N$.
-
-  Thus, we can write $n_1 = A(m_2)$ for $m_2\in M$.
-  Now, apply $A$ to both sides: by the assumption that $A$ is idempotent modulo $\m$, we have
-  \[ A(n_1)=A^2(m_2) = A(m_2) + n_2 = n_1+n_2, \]
-  Thus, $n_2=A(n_1)-n_1$, so $n_2\in \m^2M$; clearly also $n_2\in N$ as well.
-  Combining the previous equations, we can write
-  \[ n_0 = A(n_0) - n_1 = A(n_0) - A(n_1) + n_2 = A(n_0 - n_1) + n_2, \]
-  with $n_1\in \m M\cap N$ and $n_2\in \m^2M\cap N$.
-
-  Continuing in this fashion, for any $k$ we can write
-  \[ n_0=A(n_0-n_1+\dots \pm n_k) \mp n_{k+1}, \]
+  Let $n_0\in N$. By assumption, $n_0 = \psi(m_1)$ for some $m_1\in M$. Applying $\psi$ again, we get
+  \[ \psi(n_0) = \psi^2(m_1) = \psi(m_1) + \psi'(m_1) = n_0 + n_1, \]
+  where $n_1 = \psi^2(m_1) - \psi(m_1) = \psi(n_0) - n_0 \in \m M$.
+In fact, since $\psi(n_0)$ and $n_0$ are both in $N$, we have $n_1\in N$ as well, so $n_1\in \m M\cap N$.
+Thus, we can write $n_1 = \psi(m_2)$ for $m_2\in M$.
+  Now, apply $\psi$ to both sides: by the assumption that $\psi$ is idempotent modulo $\m$, we have
+  \[ \psi(n_1)=\psi^2(m_2) = \psi(m_2) + \psi'(m_2) = n_1 + n_2, \]
+  where, similarly, $n_2 = \psi(n_1) - n_1$, so $n_2 \in \m^2M\cap N$ as well.
+  Combining, we can write
+  \[ n_0 = \psi(n_0) - n_1 = \psi(n_0) - \psi(n_1) + n_2 = \psi(n_0 - n_1) + n_2, \]
+  with $n_1\in \m M\cap N$ and $n_2\in \m^2M\cap N$. Continuing in this fashion, for any $\ell$ we can write
+  \[ n_0=\psi(n_0-n_1+\dots \pm n_{\ell}) \mp n_{\ell+1}, \]
   with $n_i \in \m^i M\cap N$.
 
-  By the Artin--Rees lemma \cite[Lemma~5.1]{Eisenbud95}, there's some positive integer $k$ such that for $n\gg0$ we can write
-  \[ \m^n M\cap N = \m^{n-k} ( \m^kM\cap N)\subset \m^{n-k} N. \]
+  By the Artin--Rees lemma \cite[Lemma~5.1]{Eisenbud95}, there is a number $\ell>0$ such that for any $i\gg0$ we get
+  \[ \m^i M\cap N = \m^{i-\ell} ( \m^i M\cap N)\subset \m^{i-\ell} N. \]
   That is, the terms of $n_0-n_1+n_2-\cdots$ are going to 0 in the $\m$-adic topology on $N$. Hence we can write
-  \[ n_0=A(n_0-n_1+n_2-\cdots), \]
-  with $n_0-n_1+n_2-\cdots\in N$. We conclude that $A$ is surjective as a map $N\to N$.
+  \[ n_0 = \psi(n_0-n_1+n_2-\cdots), \]
+  with $n_0-n_1+n_2-\cdots\in N$. We conclude that $\psi$ is surjective as a map $N\to N$.
 \end{proof}
 
-Thus, if we produce an element $A\in \End_R(M)$ that is an idempotent modulo $\m$, we obtain a splitting of $M$.
+Thus, if we produce an element $\psi \in \End_R(M)$ that is an idempotent modulo $\m$, we obtain a splitting of $M$.
 The following lemma allows us to produce idempotents modulo $\m$.
@@
   Let $k$ be a finite field of characteristic $p$, and let $A$ be an endomorphism of a $k$-vector space such that all eigenvalues of $A$ are contained in $k$.
-If $\lambda$ is an eigenvalue of $A$, then some power of $A-\lambda I$ is idempotent.
-  Furthermore, if $\lambda$ is not the only eigenvalue of $A$, then a  nonzero power of $A-\lambda I$ is idempotent.
+If $\lambda$ is an eigenvalue of $A$, then some power of $A-\lambda$ is idempotent.
+  Furthermore, if $\lambda$ is not the only eigenvalue of $A$, then a  nonzero power of $A-\lambda$ is idempotent.
 \end{lem}
 \begin{proof}
-  Since all eigenvalues of $A$ are contained in $k$, we can without loss of generality put $A$ in Jordan canonical form, with each Jordan block having the form
-  \[ r_i\Biggl\{
-  \underbrace{\begin{pmatrix}
+  Since all eigenvalues of $A$ are contained in $k$, we can without loss of generality put $A$ in Jordan canonical form, with each Jordan block being an $r_i\times r_i$ matrix of the form
+  \[ \begin{pmatrix}
         \lambda_i & 1 & 0 & \dots & 0 \\
         0 & \lambda_i & 1  & \dots & 0 \\
-      \vdots & & & & \vdots \\
+        %0 & 0 & \lambda_i & \dots & 0 \\
+        \vdots & \vdots & \vdots & \ddots & \vdots \\
         0 & 0 & 0 & \dots & \lambda_i
-  \end{pmatrix}}_{r_i} \]
-  with each $\lambda_i$ an eigenvalue of $A$.
-  In this basis, $A-\lambda I$ will be block-diagonal with blocks of form
+    \end{pmatrix}, \]
+  where each $\lambda_i$ is an eigenvalue of $A$.
+  In this basis, $A-\lambda$ will be block-diagonal with blocks
   \[ \begin{pmatrix}
@@
     0 & \lambda_i-\lambda & 1 & \dots & 0 \\
-    \vdots & & & & \vdots \\
+    \vdots & \vdots & \vdots & \ddots & \vdots \\
     0 & 0 & 0 & \dots & \lambda_i-\lambda
   \end{pmatrix}. \]
-  Set $\mu_i=\lambda_i-\lambda$. Then for any $l\geq 1$, the $l$-th power $(A-\lambda I)^l$ is block-diagonal with blocks of the form
+  Set $\nu_i=\lambda_i-\lambda$. Then for any $n\geq 1$,
+  the $n$-th power $(A-\lambda)^n$ is block-diagonal with blocks
   \[ \begin{pmatrix}
-    \mu_i^l & \binom{l}1\mu_i^{l-1} & \binom{l}2\mu_i^{l-2} & \dots & \binom{l}{r_i}\mu_i^{l-r_i} \\
-    0 & \mu_i^l & \binom{l}1\mu_i^{l-1} & \dots & \binom{l}{r_i-1}\mu_i^{l-r_i+1} \\
-    \vdots & & & & \vdots \\
-    0 & 0 & 0 & \dots & \mu_i^l
+    \nu_i^n & \binom{n}1\nu_i^{n-1} & \binom{n}2\nu_i^{n-2} & \dots & \binom{n}{r_i}\nu_i^{n-r_i} \\
+    0 & \nu_i^n & \binom{n}1\nu_i^{n-1} & \dots & \binom{n}{r_i-1}\nu_i^{n-r_i+1} \\
+    \vdots & \vdots & \vdots & \ddots & \vdots \\
+    0 & 0 & 0 & \dots & \nu_i^n
   \end{pmatrix}. \]
-  If we choose $l > r_i$ and $l$ a power of $p$, say $p^{e_0}$, then all non-diagonal terms will vanish, so all blocks will have the form
+  If we choose $n > r_i$ and $n$ a power of $p$, say $p^{e_0}$, then all non-diagonal terms will vanish, so all blocks will have the form
   \[ \begin{pmatrix}
-    \mu_i^l & 0 & 0 & \dots & 0 \\
-    0 & \mu_i^l & 0 & \dots & 0 \\
-    \vdots & & & & \vdots \\
-    0 & 0 & 0 & \dots & \mu_i^l
+    \nu_i^n & 0 & 0 & \dots & 0 \\
+    0 & \nu_i^n & 0 & \dots & 0 \\
+    \vdots & \vdots & \vdots & \ddots & \vdots \\
+    0 & 0 & 0 & \dots & \nu_i^n
   \end{pmatrix}. \]
-  Finally, since $k\subset \FF_{p^e}$ for some $e$, if we choose $l$ to be divisible also by $p^e-1$, we have that $$\mu_i^{l-1} = (\mu_i^{p^{e}-1})^{l/(p^{e}-1)} =1.$$
-  In particular, taking $l=p^{e_0}(p^e-1)$,
-  $(A-\lambda I)^{p^{e_0}(p^e-1)}$ is a diagonal matrix with diagonal entries 1 or 0, hence idempotent.
-Note moreover that if some $\lambda_i\neq \lambda$, then $(A-\lambda I)^l$ is not the zero matrix.
+  Finally, since $k\subset \FF_{p^e}$ for some $e$, if we choose $n$ to be divisible also by $p^e-1$, we have that $$\nu_i^{n-1} = (\nu_i^{p^{e}-1})^{n/(p^{e}-1)} =1.$$
+  In particular, taking $n=p^{e_0}(p^e-1)$, the matrix $(A-\lambda)^{p^{e_0}(p^e-1)}$ is a diagonal matrix with diagonal entries 1 or 0, hence idempotent.
+Note moreover that if some $\lambda_i\neq \lambda$, then $(A-\lambda)^n$ is not the zero matrix.
+\end{proof}
+
+\begin{prop}\label{prop:split-idems}
+  Let $\lambda_1,\dots,\lambda_r$ be the eigenvalues of $\phi\otimes R/\m$ defined over $\FF_{p^e}$ and set $\psi_j = (\phi - \lambda_j \id_M)^{p^{e_i} (p^{e} - 1)}$ with $e_i = \lceil \log_p(\mu_i) \rceil$ and $\psi = \psi_1\circ\cdots\circ\psi_r$.
+Then $M$ has a direct sum decomposition
+  \( M = \ker\psi_1 \oplus \cdots \oplus \ker\psi_r \oplus \im\psi. \)
+\end{prop}
+\begin{proof}
+  The proof is analogous to \Cref{prop:split-surj}.
+Specifically, using \Cref{lem:jordan} and \Cref{lem:idemp}, each $\psi_i$ is an idempotent, and in particular, a split surjection.
+Further, $\ker\psi_i\cap\ker\psi_j = 0$ by the same argument as in the proof of \Cref{prop:split-surj}.
+Therefore, using \Cref{lem:decomposition}, we have the desired direct sum decomposition.
 \end{proof}
@@
     % TODO: finitely generated M? noetherian R?
-    \Require module $M$ over a commutative local ring $R$.
+    \Require module $M$ over a commutative local ring $R$ such that $R/\m \subset\bar\FF_p$
     \Ensure  list $\textproc{Summands}(M)$ of indecomposable summands of $M$.
@@
     \State Take a general element $\phi$ in $\End_R(M) \setminus \m \End_R(M)$.
-    \State Let $\phi_0$ be the induced endomorphism of the $k$-vector space $M/\m M$.
-    % TODO: Why did we need to assume $A$ has at least two eigenvalues?
-    \State Let $\lambda$ be any eigenvalue of $\phi_0$.
-    \State Let $\psi = (\phi - \lambda \id_M)^{p^{e_0} (p^{e} - 1)}$
-    (with the power explicitly as in the proof of \Cref{lem:jordan}).
-    \State By \Cref{lem:idemp}, $\psi$ is an idempotent, hence producing a splitting of $M$.
-    \If { both $\coker\psi$ and $\im\psi$ are nonzero }
-      \State Append \Call{Summands}{$\coker\psi$} to $L$.
+    \State Let $A$ be the induced endomorphism of the $k$-vector space $M/\m M$.
+    \ForAll { eigenvalues $\lambda_i$ of $A$ }
+      \State Let $\mu_i$ be the (geometric) multiplicity of $\lambda_i$.
+      \State Let $e_i = \lceil \log_p(\mu_i) \rceil$ (or, for simplicity, $\lceil \log_p(\mu(M)) \rceil$).
+      \State Let $e$ be the smallest integer such that $\lambda_i\in \FF_{p^e}$.
+      \State Let $\psi_i = (\phi - \lambda \id_M)^{p^{e_0} (p^{e} - 1)}$.
+      \State By \Cref{prop:split-idems}, $\psi_i$ is an idempotent, hence producing a splitting of $M$.
+      \State Append \Call{Summands}{$\ker\psi_i$} to $L$.
+    \EndFor
+    \State Let $\psi$ be the composition of all $\psi_i$ above.
+    \If {$\psi$ is nonzero}
       \State Append \Call{Summands}{$\im\psi$} to $L$.
-    \Else
-      \State Append $M$ to $L$.
     \EndIf
@@
 
+\begin{rem}\label{rem:local}
+  Similar to \Cref{rem:graded}, over a large enough field \Cref{alg:local} will produce indecomposable summands in one non-recursive iteration, and otherwise it will recursively split the summands until it finds an indecomposable decomposition of $M$.
+\end{rem}
+
+\begin{rem}\label{rem:char}
+  If $R$ is a local ring over a field $k\subset \C$ of characteristic 0, and $M$ a finitely generated $R$-module, it would be useful to have an algorithm to find a direct sum decomposition of $M$.
+  At the moment, we do not have such an algorithm outside the graded case.
+  However, we note that \Cref{alg:local} can be used to test indecomposability of $M$ via reduction modulo $p$, as follows.
+  One can choose a finitely generated $\Z$-algebra $A$ and an $A$-algebra $R_A$ such that $R_A\otimes _A k=R$, and likewise an $R_A$-module $M_A$ such that $M_A\otimes _A k = R$ and such that $M_A$ is flat over $A$.
+If $\n$ is a maximal ideal of $A$, one can check that $A/\n \cong \FF_{p^e}$ for some prime $p$ and $e$.
+  The key point is that if $M$ is decomposable, then $M_A$ also can be taken to be decomposable. Thus, the various reductions  $M\otimes A/\n$ will also be decomposable for all $\n$.
+  If our algorithm does not detect an indecomposable summand of $M\otimes A/\n$ for an appropriate choice of $\n$, then, the original module $M$ must have been indecomposable.
+  However,  we do not have a way of patching the decompositions of various $M_A\otimes_A A/\n$ into a decomposition of $M$.
+\looseness-1
+
+  We also point out that it is possible to ``guess'' an idempotent for $M$, even when there is no algorithm to produce one.
+In practice, if $M$ does have nontrivial idempotents, Macaulay2 often chooses them as some of the generators of $\End_R(M)$.
+By checking if the generators of $\End_R(M)$ are idempotent, and trying to construct idempotents by subtracting eigenvalues off of general endomorphisms and checking idempotency, we can often produce nontrivial direct sum decompositions in the local case, even when the characteristic is 0.
+\end{rem}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-As a consequence of \Cref{alg:graded} and \Cref{alg:local} above, we have:
+As a consequence of \Cref{alg:graded} and \Cref{alg:local}, we have:
 
-\begin{lem}\label{lem:term}
+\begin{lem}\label{lem:terminates}
   Assume that either $(R,\m)$ is a graded ring with $R_0=k$, or $(R,\m)$ is a local ring with $k=R/\m$ a field of positive characteristic, and that $M$ is a finitely generated $R$-module (homogeneous in the graded case).
-  If $M$ has an endomorphism with at least two distinct eigenvalues modulo $\m$, then $M$ is not indecomposable.
+  If $M$ has an endomorphism with at least two distinct eigenvalues modulo $\m$, then $M$ is decomposable.
 \end{lem}
 
-The converse observation implies that if $M$ is not indecomposable, then these algorithms will find the direct sum decomposition of $M$:
+The converse observation implies that if $M$ is decomposable, then these algorithms will find the direct sum decomposition of $M$:
 
 \begin{lem}\label{lem:distinct}
-  Retain the notation of \Cref{lem:term}.
+  Retain the notation of \Cref{lem:terminates}.
   \begin{itemize}
-  \item If $R$ is graded and $M$ is not indecomposable, then a general endomorphism of $M$ of degree zero has at least two distinct eigenvalues (in $\bar k$)
-  \item If $R$ is local and $M$ is not indecomposable, then a general endomorphism of $M$ of degree zero reduces to an endomorphism of $M/\m M$ with at least two distinct eigenvalues.
+  \item If $R$ is graded and $M$ is decomposable, then a general endomorphism of $M$ of degree zero has at least two distinct eigenvalues (in $\bar k$)
+  \item If $R$ is local and $M$ is decomposable, then a general endomorphism of $M$ of degree zero reduces to an endomorphism of $M/\m M$ with at least two distinct eigenvalues.
   \end{itemize}
@@
 
-\begin{rem}
-  By ``general'' we mean that a general linear combination of a basis for $[\End_R(M)]_0$ (in the graded case) or of minimal generators of $\End_R(M)$ (in the local case) over the algebraic closure of $k$, or equivalently over a sufficiently large algebraic extension of~$k$.
+\begin{rem}\label{rem:general}
+  By ``general'' we mean a general $k$-linear combination of a $k$-basis for $[\End_R(M)]_0$ (in the graded case) or of minimal generators of $\End_R(M)$ (in the local case) over the algebraic closure of $k$, or equivalently over a large enough algebraic extension of $k$.
 \end{rem}
 
-In particular, if $M$ is not indecomposable, then \Cref{alg:graded} and \Cref{alg:local} produce a nontrivial direct sum decomposition of $M$, because a general element of $\End M$ will have distinct eigenvalues and thus the algorithms will produce a nontrivial decomposition.
+In particular, if $M$ is decomposable, then \Cref{alg:graded} and \Cref{alg:local} produce a nontrivial direct sum decomposition of $M$, because a general element of $\End_R(M)$ will have distinct eigenvalues and thus the algorithms will produce a nontrivial decomposition.
+We emphasize that even if $M$ splits over $k$, a general element of $\End_R(M)$ may have eigenvalues over an extension of $k$ (see \Cref{rem:extension}).
 
-\begin{proof}
+\begin{proof}[Proof of \Cref{lem:distinct}]
   We may assume that the base field $k$ is algebraically closed.
-  Let $\Phi_1,\dots,\Phi_r$ be a basis for $[\End_R(M)]_0$ in the graded case, or minimal generators of $\End_R(M)$ in the local case, and $\phi_1,\dots,\phi_r$ their images modulo $\m$, which we view as matrices with entries in $k$.
-  Note that if $M$ is decomposable, say $M=M_1\oplus M_2$, then we may always take $\Phi_1,\Phi_2$ to be projectors onto each summand.
-  Let $U\subset \A^r$ be the subset of $r$-tuples $(\lambda_1,\dots,\lambda_r)$ such that $\lambda_1\phi_1+\dots+\lambda_r\phi_r$ has at least two distinct eigenvalues, i.e., such that $\lambda_1\Phi_1+\dots+\lambda_r\Phi_r$ reduces to an endomorphism of $M/\m M$ with at least two distinct eigenvalues.
-(Note that in the graded case the eigenvalues of the reduction of an endomorphism modulo $\m$ are the same as the eigenvalues of endomorphism, so there is no harm in looking only at $M/\m M$.)
+  Let $\phi_1,\dots,\phi_r$ be a basis for $[\End_R(M)]_0$ in the graded case, or minimal generators of $\End_R(M)$ in the local case, and $A_1,\dots,A_r$ their images modulo $\m$, meaning $k$-linear transformations $M/\m M\to M/\m M$ which we view as matrices with entries in $k$.
 
-  It suffices to show that $U$ is a nonempty open subset of $\A^r$. First, we show $U$ is nonempty:
-  Say $M=M_1\oplus M_2$ with $M_1,M_2$  nonzero  summands.
-Choose $\Phi_1$ to be the projection to $M_1$, and $\Phi_2$ the projection to $M_2$.
-Then for any $\lambda_1,\lambda_2\in k$, $\lambda_1\phi_1 + \lambda_2\phi_2$ has eigenvalues $\lambda_1,\lambda_2$; thus in particular there is \emph{an} element of $\End_R(M)$ reducing to an endomorphism of $M/\m M$ with distinct eigenvalues, so $U$ is nonempty.
+  Let $U\subset\AA^r$ be the subset of $r$-tuples $(\lambda_1,\dots,\lambda_r)$ such that $\lambda_1 A_1+\dots+\lambda_r A_r$ has at least two distinct eigenvalues, i.e., such that $\lambda_1\phi_1+\dots+\lambda_r\phi_r$ reduces to an endomorphism of $M/\m M$ with at least two distinct eigenvalues.
+% (Note that in the graded case the eigenvalues of the reduction of an endomorphism modulo $\m$ are the same as the eigenvalues of endomorphism, so there is no harm in looking only at $M/\m M$.)
 
-  Now, we show that $U$ is open.
-This is a purely linear algebraic statement: we claim that given a matrix $\phi$ with at least two distinct eigenvalues, and any $r$ matrices $\phi_1,\dots,\phi_r$, that
-  $$ A_{\lambda_1,\dots,\lambda_r}:=\phi+\lambda_1\phi_1+\dots+\lambda_r\phi_r $$
-  has at least two distinct eigenvalues for $\lambda_1,\dots,\lambda_r$ outside a Zariski-closed subset of $\A^r$.
-  The eigenvalues of $A_{\lambda_1,\dots,\lambda_r}$ are the roots of $\det(A_{\lambda_1,\dots,\lambda_r}-t I)$, which is a polynomial in $t$ with coefficients in $\lambda_1,\dots,\lambda_r$.
-  $A_{\lambda_1,\dots,\lambda_r}$ fails to have at least two distinct eigenvalues exactly when this polynomial factors as a power of a linear term.
+  It suffices to show that $U$ is a nonempty open subset of $\AA^r$. First, we show $U$ is nonempty:
+  say $M = M_1\oplus M_2$ be a nontrivial decomposition and note that we may take $\phi_1,\phi_2$ to be projectors onto each summand.
+Then for any $\lambda_1,\lambda_2\in k$ the combination $\lambda_1 A_1 + \lambda_2 A_2$ has eigenvalues $\lambda_1,\lambda_2$; thus in particular there is \emph{an} element of $\End_R(M)$ reducing to an endomorphism of $M/\m M$ with distinct eigenvalues, so $U$ is nonempty.
 
-  This condition is polynomial in the coefficients of powers of $t$ in $\det(A_{\lambda_1,\dots,\lambda_r}-t I)$ and thus in the $\lambda_i$; to see this, note that
-  if we write
-  $$ f:=\det(A_{\lambda_1,\dots,\lambda_r}-t I)=t^n  +t^{n-1}b_{n-1}+\dots +t b_1 +b_0, $$
-  then $f$
-  has an $n$-fold root exactly when
-  $ f,\d f/\d t,\dots, \d^{n-1} f/\d t^{n-1} $
-  vanish simultaneously; the resultant of these $n$ polynomials in the $n$ variables $b_i$ gives polynomial conditions in the $b_i$ for this to occur.
-  In our setting, the $b_i$ are themselves polynomials in the $\lambda_i$, and thus we have obtained polynomial equations defining the locus where $A_{\lambda_1,\dots,\lambda_r}$ fails to have distinct roots, and thus the complement $U$ is open.
+  Now, we show that $U$ is open. This is a purely linear algebraic statement: we claim that given a matrix $A_0$ with at least two distinct eigenvalues, and any $r$ matrices $A_1,\dots,A_r$, that
+  $$ A_{\lambda_1,\dots,\lambda_r} \coloneqq A_0+\lambda_1 A_1+\dots+\lambda_r A_r $$
+  has at least two distinct eigenvalues for $\lambda_1,\dots,\lambda_r$ outside a Zariski-closed subset of $\AA^r$.
+  The eigenvalues of $A_{\lambda_1,\dots,\lambda_r}$ are the roots of $\chi_{A_{\lambda_1,\dots,\lambda_r}}(\lambda) = \det(A_{\lambda_1,\dots,\lambda_r}-\lambda \id_{M/\m M})$, which is a polynomial in $\lambda$ with coefficients in $\lambda_1,\dots,\lambda_r$.
+Then $A_{\lambda_1,\dots,\lambda_r}$ fails to have at least two distinct eigenvalues exactly when this polynomial factors as a pure power of a linear term.
+
+  This condition is polynomial in the coefficients of powers of $\lambda$ in
+  \[ f \coloneqq \chi_{A_{\lambda_1,\dots,\lambda_r}}(\lambda) = \lambda^n + c_{n-1} \lambda^{n-1} + \dots + c_1 \lambda + c_0, \]
+  which are polynomials in the $\lambda_i$. In particular, $f$ has an $n$-fold root when
+  $ f, \frac{\d f}{\d\lambda}, \dots, \frac{\d^{n-1}f}{\d\lambda^{n-1}}$ 
+  vanish simultaneously. The resultant of these $n$ polynomials in the $n$ variables $c_i$ gives polynomial conditions in the $c_i$ for this to occur.
+  In our setting, the $c_i$ are themselves polynomials in the $\lambda_i$, and thus we have obtained polynomial equations defining the locus where $A_{\lambda_1,\dots,\lambda_r}$ fails to have distinct roots, and thus the complement $U$ is open.
 \end{proof}
 
-\begin{rem}
-  Note that 
-\Cref{alg:graded} and \Cref{alg:local} are 
- quite sensitive to the ground field $k$, because they needs all the eigenvalues of the endomorphism 
-%$A$ of $M/\m M$ 
-to be contained in $k$.
+\begin{rem}\label{rem:extension}
+  Note that \Cref{alg:graded} and \Cref{alg:local} are quite sensitive to the ground field $k$, because they need all the eigenvalues of the endomorphism to be contained in $k$.
 While theoretically the issue can be avoided by working over an algebraically closed ground field $\bar k$, for practical use on a computer algebra system it is better to extend $k$ to some larger finite field.
@@
 
-If the above algorithm fails to produce a nontrivial idempotent, it does not certify that $M$ is indecomposable.
+Note that by nature of probabilistic algorithms, if \Cref{alg:graded} or \ref{alg:local} fail to produce a nontrivial summand, it does not certify that $M$ is indecomposable.
 However, there are a few sufficient conditions to be indecomposable, which in practice often (but not always) produce such a certification.
@@
 
-\begin{lem}
-  Suppose $R$ is graded, and let $M$ be a finitely generated $R$-module and let $[\End_R(M)]_0$ be the $k$-vector space of degree zero endomorphisms. Suppose that either:
+\begin{prop}\label{prop:indecomposability}
+  Let $M$ be a finitely generated $R$-module. Suppose that either:
   \begin{enumerate}
-  \item $[\End_R(M)]_0$ is 1-dimensional and thus spanned by the identity endomorphism, or
-  \item every non-identity element of $[\End_R(M)]_0$, viewed as a matrix, has entries in $\m$;
+  \item\label{item:indec-graded} $R$ is graded and $[\End_R(M)]_0$ is 1-dimensional and thus spanned by $\id_M$, or
+  \item\label{item:indec-local} $R$ is local and the image of $\End_R(M)$ in $\End_k(M/\m M)$ is 1-dimensional;
+  % \item every generator of $\End_R(M)$, viewed as a matrix, is either $\id_M$ or has all entries in $\m$;
   \end{enumerate}
   then $M$ is indecomposable.
-\end{lem}
-
+\end{prop}
 \begin{proof}
-  If $M$ decomposes non-trivially as $M_1\oplus M_2$, then the projections onto each factor are nontrivial degree zero endomorphisms not equal to the identity, and which do not have entries contained in~$\m$.
+  To see \eqref{item:indec-graded}, observe that if $M$ decomposes non-trivially as $M_1\oplus M_2$, then the projections onto each factor are nontrivial degree-zero endomorphisms not equal to the identity, which do not have all entries contained in $\m$.
+The proof for \eqref{item:indec-local} is analogous.
 \end{proof}
 
-The following lemma is the algebraic analogue of the following fact about coherent sheaves:
+%% \Cref{prop:indecomposability} is the algebraic analogue of the following fact about coherent sheaves:
 
-\begin{cor}
-  Let $X$ be a projective variety over a field $k$, and $\cF$ a coherent sheaf on $X$.
-  If $ H^0(\End \cF) = k, $ then $\cF$ is indecomposable.
-\end{cor}
+%% \begin{lem}\label{lem:indecomposablity-sheaf}
+%%   Let $X$ be a projective variety over a field $k$, and $\cF$ a coherent sheaf on $X$.\linebreak
+%%   If $ H^0(\End_{\OO_X}(\cF)) = k, $ then $\cF$ is indecomposable.
+%% \end{lem}
+%% \begin{proof}
+%%   If there is a nontrivial direct sum decomposition $\cF = \cF_1\oplus \cF_2$, then composing the projection and inclusions we obtain $\cF\to \cF_1\inc \cF$, which yields a nonzero endomorphism of $\cF$ that is not a scalar multiple of $\id_\cF$, and thus $H^0(\End_{\OO_X}(\cF)) \neq k$.
+%% \end{proof}
 
-\begin{proof}
-  If there is a nontrivial direct sum decomposition $\cF = \cF_1\oplus \cF_2$, then the composing the projection and inclusions we obtain $\cF\to \cF_1\inc \cF$ yields a nonzero endomorphism of $\cF$ that is a scalar multiple of $\id_\cF$, and thus $H^0(\End \cF) \neq k$.
-\end{proof}
-
-\begin{rem}\label{rem:char}
-  If $R$ is a local ring over a field $k\subset \C$ of characteristic 0, and $M$ a finitely generated $R$-module, it would be useful to have an algorithm to find a direct sum decomposition of $M$.
-  At the moment, we do not have such an algorithm outside the graded case.
-  However, we note that the algorithm above can be used to test indecomposability of $M$ via reduction modulo $p$, as follows:
-  One can choose a finitely generated $\Z$-algebra $A$ and an $A$-algebra $R_A$ such that $R_A\otimes _A k=R$, and likewise an $R_A$-module $M_A$ such that $M_A\otimes _A k = R$ and such that $M_A$ is flat over $A$.
-If $\n$ is a maximal ideal of $A$, one can check that $A/\n \cong \FF_{p^e}$ for some prime $p$ and $e$.
-  The key point is that if $M$ is decomposable, then $M_A$ also can be taken to be decomposable. Thus, the various reductions  $M\otimes A/\n$ will also be decomposable for all $\n$.
-  If our algorithm does not detect an indecomposable summand of $M\otimes A/\n$ for an appropriate choice of $\n$, then, the original module $M$ must have been indecomposable.
-  However, if $M$ is \emph{decomposable}, we do not have a way of patching the decompositions of various $M_A\otimes_A A/\n$ into a decomposition of $M$.
-
-  We also point out that it is possible to ``guess'' an idempotent for $M$, even when there is no algorithm to produce one.
-In practice, if $M$ does have nontrivial idempotents, Macaulay2 often chooses them as some of the generators of $\End_R(M)$.
-By checking if the generators of $\End_R(M)$ are idempotent, and trying to construct idempotents by subtracting eigenvalues off of general endomorphisms and checking idempotency, we can often produce nontrivial direct sum decompositions in characteristic 0.
-\end{rem}
-
-\begin{rem}
+\begin{rem}\label{rem:complexity}
   We omit a computational complexity analysis here because without any assumptions about the module there are too many parameters that in any given application may become prominent.
-We note, however, that $[\End_R(M)]_0$ can be computed using degree-limited syzygy algorithms without computing all of $\End_R(M)$.
-While in many applications this significantly reduces the computational complexity, in general the complexity of the algorithm presented here is at least bounded by the complexity of computing syzygies.
-Compare with \cite{DX22}, which provides a complexity analysis for a decomposition algorithm with the additional assumption that the generators of $R$ as well as generators and relations of $M$ have distinct degrees.
-This assumption is exceedingly rare in commutative algebra and algebraic geometry applications, but it is relevant, though not guaranteed, in the context of multiparameter persistence in topological data analysis.
+We note, however, two specific advantages:
+  \begin{itemize}
+  \item Using degree-limited syzygy algorithms, $[\End_R(M)]_0$ is easier to compute than $\End_R(M)$. In particular, the computation is faster if $M$ can be endowed with a multigrading.
+  \item Over a large enough field \Cref{alg:graded,alg:local} are likely to terminate in one non-recursive iteration, hence we only need to compute a single endomorphism (see \Cref{rem:graded}).
+  \end{itemize}
+   While in many applications these advantages significantly improve the computational complexity, in general the worst case complexity of the algorithms presented here is bounded by the complexity of computing the kernel of the map $f^t\otimes_R M$ where $f^t$ is the transpose of a free presentation $f$ of $M$.
+Compare with \cite{DX22}, which provides a complexity analysis for a decomposition algorithm with the additional assumption that the generators of $R$ as well as generators and relations of $M$ have distinct multidegrees.
+This assumption is exceedingly rare in applications from commutative algebra and algebraic geometry, but it is relevant, though not guaranteed, in the context of multiparameter persistence in topological data analysis.
 \end{rem}
@@
 
-While the preceding section was written in the language of modules, by the standard translation to global (multi)projective varieties, the algorithm works equally well to find indecomposable decompositions of coherent sheaves on (multi)projective varieties.
-In this section, we make a few notes in the single-graded case regarding the relation between the eigenvalues discussed in the preceding section (in this section, called \emph{module-theoretic eigenvalues}) with the notion of eigenvalue of an endomorphism of a vector bundle.
+While the preceding section was written in the language of modules, the standard algebro-geometric dictionary implies that \Cref{alg:graded} can be used to find indecomposable decompositions of coherent sheaves on projective varieties.
+In this section, we make a few notes in the standard graded case regarding the relation between the eigenvalues discussed in the previous section (in this section, called \emph{module-theoretic eigenvalues}) with the notion of eigenvalue of an endomorphism of a vector bundle.
 
 Let $X\subset \P^n$ be a projective variety, with ample line bundle $\OO_X(1)$.
-Throughout, let $E$ be a vector bundle (i.e., a locally free coherent sheaf) on $X$.
+Throughout, let $E$ be a vector bundle (i.e., a locally free coherent sheaf) on $X$, and consider an endomorphism $f\in\End_{\OO_X}(E)$. The following result is well-known.
 
-The following is well-known:
-
-\begin{lem}{{\cite[Exercise~4.1]{Friedman98}}}
-  A morphism $f:E\to E$ is injective if and only if it is an isomorphism.
+\begin{lem}\label{lem:inj-iff-iso}
+  An endomorphism $f\colon E\to E$ is injective if and only if it is an isomorphism.
 \end{lem}
-
 %(It is also true that $f$ is surjective if and only if it's an isomorphism, but this is more well-known.)
 
-%For completeness, we give the standard proof (see, e.g., \cite[Exercise~4.1]{Friedman}).
+For completeness, we give the standard proof (see, e.g., \cite[Exercise~4.1]{Friedman98}).
 
-%\begin{proof}
-%This requires only that $E$ is a coherent sheaf:
-%% since $\End_{\OO_X}(E)$ is finite-dimensional, $f$ satisfies some minimal monic polynomial $\phi(x)$; since $f$ is injective, the constant term of $\phi(x)$ is nonzero.
-%We claim that $f\otimes k(x)$ is injective for any $k(x)$. This then implies that $f\otimes k(x)$ is surjective for every $x\in X$ and thus $f$ is surjective.
-%To see this, note that $f$ satisfies some minimal-degree monic polynomial, since $\End_{\OO_X}(E)$ is finite-dimensional over $k$, and that this monic polynomial has nonzero constant term, since $f$ is injective.
+\begin{proof}
+  This requires only that $E$ is a coherent sheaf.
+We claim that $f\otimes k(x)$ is injective for any $x\in X$, which implies that $f\otimes k(x)$ is also surjective for every $x$ and thus $f$ is surjective.
+
+  To see this, note that $f$ satisfies some minimal-degree monic polynomial, since $\End_{\OO_X}(E)$ is finite-dimensional over $k$.
+This monic polynomial has nonzero constant term since $f$ is injective.
 Now, $f\otimes k(x)$ satisfies this same polynomial, so must be injective.
-%\end{proof}
+\end{proof}
 
 \begin{dfn}
-  For $f\in \End_{\OO_X}(E)$, taking top exterior powers yields the map
-  $$ \bigwedge^{\rank E} f: \det E  \to \det E, $$
-  where $\det E := \bigwedge^{\rank E} E$.
-  Since $\End_{\OO_X}(\det E) = H^0(\OO_X)=k$, $\bigwedge^{\rank E} f$ is multiplication by $\lambda\in k$; we write $\det f$ for this scalar $\lambda$.
+  Let $r\coloneqq\rank E$, then for $f\in \End_{\OO_X}(E)$, taking top exterior powers yields
+  \[\textstyle \bigwedge^r f\colon \bigwedge^r E \to \bigwedge^r E. \]
+  Since $\End_{\OO_X}(\bigwedge^r E) = H^0(\OO_X)=k$, the map $\bigwedge^r f$ is multiplication by some $\lambda\in k$; we write $\det f$ for this scalar $\lambda$.
 \end{dfn}
@@
 \end{lem}
-
 \begin{proof}
-  Since $(\bigwedge^{\rank E} f)\otimes k(x) = \bigwedge^{\rank( E\otimes k(x))}( f\otimes k(x))$ as $k(x)$-linear maps on $E\otimes k(x)$,
-  we have $\det(f\otimes k(x)) =\det(f)\otimes k(x)$, but $\det(f)\in k$ and so is unaffected by going modulo $\m_x$.
+  Note that $\rank( E\otimes k(x)) = \rank E = r$. Thus since $(\bigwedge^r f)\otimes k(x) = \bigwedge^r( f\otimes k(x))$ as $k(x)$-linear maps on $E\otimes k(x)$,
+  we have $\det(f\otimes k(x)) =\det(f)\otimes k(x)$, but $\det(f)\in k$ and so is unaffected by reducing modulo $\m_x$.
 \end{proof}
@@
 \begin{lem}
-  $\det f \neq 0 $ if and only if $f$ is injective (if and only $f$ is an isomorphism).
+  $\det f \neq 0 $ if and only if $f$ is injective, in which case $f$ is an isomorphism.
 \end{lem}
-
 \begin{proof}
-  Let $\ker f \neq 0$. Since localization is exact
-  we have
-  $\ker(f\otimes k(X)) = (\ker f)\otimes k(X)$.
-  Since $\ker f \subset E$ is torsionfree, the localization map $\ker f \to (\ker f)\otimes k(X)$ is injective and thus $(\ker f) \otimes k(X)\neq 0$.
+  Let $\ker f \neq 0$.
+Since localization is exact we have $\ker(f\otimes k(X)) = (\ker f)\otimes k(X)$.
+Since $\ker f \subset E$ is torsion-free, the localization map $\ker f \to (\ker f)\otimes k(X)$ is injective and thus $(\ker f) \otimes k(X)\neq 0$.
 Since the $k(X)$-vector space map $f\otimes k(X)$ is not injective, $\det(f\otimes k(X))= 0$.
 In particular  $\det f = 0$ as well.
-  Conversely,
-  if $\ker f = 0$, then
-  $f$ is an isomorphism, hence an isomorphism on fibers, and hence
-  $\det(f\otimes k(x)) \neq 0$ for any $x$. Thus $\det f\neq 0$.
+
+  Conversely, if $\ker f = 0$, then $f$ is an isomorphism, hence an isomorphism on fibers, and hence $\det(f\otimes k(x)) \neq 0$ for any $x$. Thus $\det f\neq 0$.
 \end{proof}
@@
   $\lambda \in k$ is an \defn{eigenvalue} of $f \in \End_{\OO_X}(E)$ if $\det(f-\lambda \id_E)= 0$.
-In other words, the eigenvalues of $f$ are the zeroes of the univariate polynomial $\det(f-t\id _E)$ with coefficients in $k$.
+In other words, the eigenvalues of $f$ are the zeroes of the univariate polynomial $\det(f - \lambda\id _E)$ with coefficients in $k$.
 \end{dfn}
@@
 
-\begin{dfn}
-  Let $M$ be a homogeneous module over a graded ring $R$ with $R_0=k$.
-  Let $g \in \End_R(M)$. The \defn{module-theoretic eigenvalues} of $g$ are defined to be the eigenvalues of the map of vector spaces $g\otimes R/\m:M/\m M \to M/\m M$.
-\end{dfn}
-
-In other words, the module-theoretic eigenvalues are the eigenvalues modulo the maximal ideal discussed in the previous section.
-
 Now, let $R$ be the homogeneous coordinate ring of $X\subset \P^n$, and let $M$ be a graded $R$-module such that $\wtilde M=E$.
+For $\phi\in\End_R(M)$ we will refer to the eigenvalues of $\phi$ (see \Cref{def:eigenvalues}) as \emph{module-theoretic} eigenvalues of $\phi$.
 
 \begin{lem}
-  Let $f:E\to E$ arise from a map $g:M\to M$ (i.e., $\wtilde M = E$ and $\wtilde g = f$).
-  The eigenvalues of $f$ are a subset of the module-theoretic eigenvalues of $g$.
+  Let $f\colon E\to E$ arise from a map $\phi\colon M\to M$, in the sense that $\wtilde M = E$ and $\wtilde\phi = f$. The eigenvalues of $f$ are a subset of the module-theoretic eigenvalues of $\phi$.
 \end{lem}
@@
 \begin{proof}
-  All that needs to be shown is that if $\det(f-\lambda \id_E) =0$, then $\det((g-\lambda \id_M)\otimes R/\m)=0$.
-  Replacing
-  $f-\lambda \id_E$
-  by
-  $f$
-  and likewise
-  $g-\lambda\id_M$
-  by
-  $g$,
-  we just need to show that $\det(f)=0$ implies $\det(g\otimes R/\m)=0$.
+  All that needs to be shown is that if $\det(f-\lambda \id_E) =0$, then $\det(\phi-\lambda \id_M)=0$. Replacing $f-\lambda \id_E$ by $f$ and likewise $\phi-\lambda\id_M$ by $\phi$,
+  we just need to show that $\det f = 0$ implies $\det\phi = 0$.
 
-  Say $\det(f)=0$.
-  If $\det(g\otimes R/\m)\neq 0$, then $g$ induces a surjection $M/\m M\to M/\m M$, and thus $g$ induces a surjection $M\to M$ by Nakayama's lemma.
-The endomorphism $f$ is thus an isomorphism \cite[Corollary~4.4]{Eisenbud95}, and so $\det(f)\neq 0$.
+  Say $\det f = 0$.
+If $\det\phi = \det(\phi\otimes R/\m)\neq 0$, then $\phi$ induces a surjection $M/\m M\to M/\m M$, and thus $\phi$ induces a surjection $M\to M$ by Nakayama's lemma.
+The endomorphism $f$ is thus an isomorphism \cite[Corollary~4.4]{Eisenbud95}, and so $\det f\neq 0$.
 \end{proof}
@@
 \begin{exa}
-  By adding  irrelevant summands to $M$, one can always add extraneous module-theoretic eigenvalues, so the containment of eigenvalues may be proper.
+  By adding irrelevant summands to $M$ (i.e., summands which sheafify to zero), one can always add extraneous module-theoretic eigenvalues, so the containment of eigenvalues may be proper.
   %However, if $M=\Gamma_*(E)$, set of values will be the same: in this case, if $g\otimes R/m$ is not an isomorphism, then $g$ cannot be surjective.
-  Note that in general the multiplicities will never be equal, even if one takes $M=\Gamma_*(E)$: if $f$ is multiplication by $\lambda$, then $f$ will have $\rank E$ many eigenvalues, while $g$ will have $\mu(M)$ module-theoretic eigenvalues.
+  Note that in general the multiplicities will never be equal, even if one takes $M=\Gamma_*(E)$: if $f$ is multiplication by $\lambda$, then $f$ will have $\rank E$ many eigenvalues, while $\phi$ will have $\mu(M)$ module-theoretic eigenvalues.
 \end{exa}
 
-By combining the discussion above with the results of \Cref{sec:graded-alg}, we thus have:
+Finally, we may combine the results of this section with \Cref{prop:split-surj} to prove the following proposition.
 
 \begin{prop}
-  If $X$ is a variety over a field $k$,
-  $E$ a vector bundle on $X$, and $f:E\to E$ has two distinct eigenvalues $\lambda_1,\lambda_2$, then $E$ is a nontrivial direct sum, with one summand given by $(f-\lambda_1)^{N}$ for some $N$.
+  Suppose $E$ is a vector bundle on a $k$-variety $X$, and $f\in\End_{\OO_X}(E)$ has distinct eigenvalues $\lambda_1,\dots,\lambda_n$.
+Set $g_i = (f - \lambda_i\id_E)^{\mu_i}$, where $\mu_i$ are module-theoretic multiplicities of the corresponding eigenvalues of $\Gamma_*(f)$, and $g = g_1\circ\dots\circ g_n$, then $E$ has a direct sum decomposition
+  \[ E = \ker g_1 \oplus \dots \oplus \ker g_n \oplus \im g. \]
 \end{prop}
-
 \begin{proof}
-  Take $M= \Gamma_*(E)$; then since the eigenvalues $\lambda_1\neq \lambda_2$ are also module-theoretic eigenvalues of $M$, \Cref{alg:graded} produces a nontrivial direct sum decomposition $M_1\oplus M_2$, and since $M$ is torsion-free we know that $\widetilde M_i\neq 0$, and thus $E=\widetilde M_1\oplus \widetilde M_2$ is a nontrivial direct summand of sheaves.
+  Take $M = \Gamma_*(E)$, $\psi_i = \Gamma_*(g_i)$, and $\psi = \Gamma_*(g)$; then since each $\lambda_i$ is also a module-theoretic eigenvalue of $M$, \Cref{prop:split-surj} produces a direct sum decomposition
+  \[ M = \ker\psi_1 \oplus \cdots \oplus \ker\psi_n \oplus \im\psi; \]
+note that sheafification is exact, so the sheaf associated to $\ker \psi_i$ is $\ker \wtilde \psi_i = \ker g_i$ (and this is nonzero), while the sheaf associated to $\im \psi $ is $\im g$.
+Thus, taking the sheafification of the decomposition of $M$,
+we have the desired direct sum decomposition of $E$.
 \end{proof}
@@
 
-  If one extends the ground field to $\FF_{49}$, however, our algorithm calculates the full decomposition
+  After extending the ground field to $\FF_{49}$, our algorithm calculates the full decomposition
   $$ F_* \OO_X=\bigoplus_{p=1}^7 \OO_X(p_i). $$
@@
 
-  The use of our algorithm to compute examples in small $p$ and $e$ suggest the following behavior:
+  The use of our algorithm to compute examples in small $p$ and $e$ suggest the behavior
   $$ F_* \OO_X = \OO_X\oplus M, $$
@@
     0&a-b+c-d&0&0\\
-    0&0&a+i\,b-c-i\,d&0\\
-    0&0&0&a-i\,b-c+i\,d
+    0&0&a+bi-c-di&0\\
+    0&0&0&a-bi-c+di
   \end{pmatrix}.
   \]
+  In particular, the change of basis matrices can be extracted from our algorithm.
 \end{exa}

### 2025-04-28 paper/main.tex
devlin's final edits

@@
 The problems of finding isomorphism classes of indecomposable modules with a given property, or determining the indecomposable summands of a module, are ubiquitous in commutative algebra and representation theory and even in areas such as topological data analysis.
-Within commutative algebra, for instance, the classification of rings $R$ for which there are only finitely many isomorphism classes of indecomposable maximal Cohen--Macaulay $R$-modules (the \emph{finite CM-type} property), or determining whether iterated Frobenius pushforwards of a Noetherian ring in positive characteristic have finitely many isomorphism classes of indecomposable summands (the \emph{finite F-representation type} property) are two well-established research problems (see, for example, \cite{Yoshino90,LW12} for the former and \cite{SVdB97,Hara15,TT08}, among many others, for the latter).
+Within commutative algebra, for instance, the classification of rings $R$ for which there are only finitely many isomorphism classes of indecomposable maximal Cohen--Macaulay $R$-modules (the \emph{finite CM-type} property), or for which iterated Frobenius pushforwards in positive characteristic have finitely many isomorphism classes of indecomposable summands (the \emph{finite F-representation type} property) are two well-established research problems (see, for example, \cite{Yoshino90,LW12} for the former and \cite{SVdB97,Hara15,TT08}, among many others, for the latter).
 For both these problems, and many others, making and testing conjectures depends on computing summands of modules and verifying their indecomposability.
 
-Currently there are no algorithms available for performing these tasks on an arbitrary module over a commutative ring.
+Currently there are no algorithms available for finding direct sum decompositions of an arbitrary module over a commutative ring.
 The need for and absence of such algorithms is noted, for example, in \cite[\S15.10.9]{Eisenbud95} and \cite[\S8.5]{BL23}.
-In contrast, variants of the ``Meat-Axe'' algorithm for determining irreducibility of finite-dimensional modules over a group algebra have wide ranging applications in computational group theory \cite{Parker84,HR94,Holt98,IL00} and are available through symbolic algebra software such as Magma and GAP \cite{MAGMA,GAP}.
+In contrast, variants of the ``Meat-Axe'' algorithm for determining irreducibility of finite-dimensional modules over a group algebra have wide-ranging applications in computational group theory \cite{Parker84,HR94,Holt98,IL00} and are available through symbolic algebra software such as Magma and GAP \cite{MAGMA,GAP}.
+\looseness-1
 
@@
 
-The purpose of this paper is to describe and prove correctness of two practical algorithms for computing indecomposable summands of finitely generated modules over a $k$-algebra $R$: \Cref{alg:graded} if $M$ is a homogeneous module over a (multi)graded ring $R$ with $k$ a field of arbitrary characteristic, and \Cref{alg:local} if $R$ is local and $k$ a field of positive characteristic.
+The purpose of this paper is to describe and prove correctness of two practical algorithms for computing indecomposable summands of finitely generated modules over a $k$-algebra $R$: \Cref{alg:graded} in the case that $M$ is a homogeneous module over a (multi)graded ring $R$ with $k$ a field of arbitrary characteristic, and \Cref{alg:local} when $R$ is local and $k$ a field of positive characteristic contained in $\bar \F_p$.
 Over a large enough field, our algorithms are often capable of completely decomposing a module in a single non-recursive iteration.
@@
 Let $R$ be a $\N$-graded ring with $R_0 = k$ a field.
-Suppose $M = \bigoplus [M]_d$ is a finitely generated, graded $R$-module and consider a general degree-zero endomorphism $\phi \in [\End_R(M)]_0$.
+Suppose $M = \bigoplus [M]_d$ is a finitely generated, graded $R$-module and consider a  degree-zero endomorphism $\phi \in [\End_R(M)]_0$.
 Fix a set of minimal homogeneous generators $m_1,\dots,m_{\mu(M)}$ of $M$ so that $\phi$ may be presented as a $\mu(M)\times\mu(M)$ matrix with entries in $R$.
@@
 \begin{proof}
-  Since $M$ is $\N$-graded, assume its generators are ordered with $\deg m_i$ nonincreasing, and let $d_1,\dots,d_n$ be the decreasing list of unique degrees of generators of $M$.
+  Since $M$ is $\N$-graded, we may order its generators such that $\deg m_i$ is nonincreasing; let $d_1,\dots,d_n$ be the decreasing list of unique degrees of generators of $M$.
 The matrix for $\phi$ is then ``block upper triangular'' with $n$ diagonal blocks with entries in $k$, i.e., of the form
@@
   where the blocks $B_i$ are square matrices \emph{with entries in $k$}, and the remaining entries are in $\m$.
-To see this, note that since $\phi$ has degree zero, $\phi(m_i)$ is a $k$-linear combination of the $m_j$ with $\deg m_j = \deg m_i$ plus an $R$-linear combination of the $m_j$ with $\deg m_j < \deg m_i$.
+To see this, note that since $\phi$ has degree zero, $\phi(m_i)$ is a $k$-linear combination of the $m_j$ with $\deg m_j = \deg m_i$ plus an $\m$-linear combination of the $m_j$ with $\deg m_j < \deg m_i$.
 %In particular, $B_i$ is an $k$-endomorphism on the $k$-vector space spanned by all generators of degree $d_i$.
 
-  Then the characteristic polynomial $\det(\phi - \lambda \id_M) = \prod \det(B_i - \lambda) = \det(A - \lambda \id_{M/\m M})$ is a univariate polynomial in $\lambda$ with coefficients in $k$, thus has solutions in $\bar k$.
+  Thus the characteristic polynomial $\det(\phi - \lambda \id_M) = \prod \det(B_i - \lambda) = \det(A - \lambda \id_{M/\m M})$ is a univariate polynomial in $\lambda$ with coefficients in $k$, thus has solutions in $\bar k$.
 \end{proof}
@@
   The collection of eigenvalues of $B_i$ in the splitting field of $\chi_\phi(\lambda)$ are the \emph{eigenvalues of $\phi$}.
-We will use $\mu_j \coloneqq \mu(\lambda_j)$ for the sum of the (geometric) multiplicities of $\lambda_j$ in each $B_i$.
+We will write $\mu_j \coloneqq \mu(\lambda_j)$ for the sum of the (geometric) multiplicities of $\lambda_j$ in each $B_i$.
 Note that the (algebraic) multiplicity of $\lambda_j$ in $\chi_\phi(\lambda)$ is the sum of the multiplicities of $\lambda_j$ in each $\chi_{B_i}(\lambda)$ and may be larger than $\mu_j$.
@@
 In the proof of \Cref{prop:eigenvalues}, we only need a partial order on the degrees of the generators of $M$ in order to present $\phi$ in a block upper triangular form \eqref{eq:blocks}.
-Suppose $R$ is $\ZZ^r$-graded, then the effective cone $\Eff(R)\subset\Q^n$ generated by the elements $\deg m$ for each monomial $m \in R$ induces a partial order on the degrees of generators of $M$ if it is strongly convex (contains no positive dimensional subspace of $\Q^n$) and $R_0 = k$.
+If $R$ is $\ZZ^r$-graded, then the \emph{effective cone} $\Eff(R)\subset\Q^n$ generated by the elements $\deg m$ for each monomial $m \in R$ induces a partial order on the degrees of generators of $M$ if it is strongly convex (contains no positive dimensional subspace of $\Q^n$) and $R_0 = k$.
 We will call such rings \emph{positively graded}.
@@
 
-Note that as long as each $\psi_i$ in \Cref{lem:decomposition} is nontrivial (not zero or identity endomorphisms), the kernel summands will be nontrivial, however, they may not necessarily be indecomposable.
+Note that as long as each $\psi_i$ in \Cref{lem:decomposition} is nontrivial (not the zero or identity endomorphism), the kernel summands will be nontrivial (although they may not necessarily be indecomposable).
 Also note that $\im\psi$ may be zero.
@@
 \begin{proof}
-  If $\psi_i$ is injective on $\im\psi_i$, then $\im\psi_i$ is a summand: in this case, the composition $\im\psi_i\hookrightarrow M\to \im\psi_i$ is injective, but a degree-zero endomorphism of a finitely generated $R$-module is an isomorphism (it must be injective, hence surjective, on each finite-dimensional $[M]_d$), and thus the inclusion $\im \psi \to M$ is split by the surjection $M\to \im \psi$.
+  If $\psi_i$ is injective on $\im\psi_i$, then $\im\psi_i$ is a summand: in this case, the composition $\im\psi_i\hookrightarrow M\to \im\psi_i$ is injective, but a degree-zero endomorphism of a finitely generated $R$-module is an isomorphism (it must be injective, hence surjective, on each finite-dimensional $[M]_d$).
+Thus, the inclusion $\im \psi \to M$ is split by the surjection $M\to \im \psi$.
 Therefore the short exact sequence $0\to\ker\psi_i\to M\to\im\psi_i\to 0$ is split and $M = \ker\psi_i\oplus\im\psi_i$.
@@
   To see that $\psi_i$ is injective on $\im\psi_i$, first for simplicity replace $\phi$ by $\phi-\lambda_i\id_M$ so $\psi_i = \phi^{\mu_i}$.
-By the Cayley--Hamilton theorem $\psi_i$ divides $\chi_\phi(\phi)$, hence
+By the Cayley--Hamilton theorem (see, e.g., \cite[Theorem~4.3]{Eisenbud95}), $\psi_i$ divides $\chi_\phi(\phi)$, hence
   \begin{align*}
@@
   \( \psi_i = c_0^{-1} \left(\phi^{\mu(M) - \mu_i - 1} - \dots + c_1\right) \circ \phi \circ \psi_i. \)
-  This means that $\phi$ is injective on $\im\psi_i$, therefore $\psi_i=\phi^{\mu_i}$ is injective on $\im\psi_i$.
+  This means that $\phi$ is injective on $\im\psi_i$, and therefore the composition $\psi_i=\phi^{\mu_i}$ is still injective on $\im\psi_i$.
+We thus have that $\psi_i$ is a split surjection, as desired.
 
-  Moreover, using the standard proof for vector spaces, we know $\ker\psi_i\cap\ker\psi_j = 0$ if $i\neq j$: suppose $(\phi - \lambda_i)^m(v) = 0$ for some $v\in M$ and $m$ is minimal so that $w = (\phi-\lambda_i)^{m-1}(v) \neq 0$ satisfies $(\phi-\lambda_i)(w) = 0$, but then since $(\phi - \lambda_j)(\phi - \lambda_i) = (\phi - \lambda_i)(\phi - \lambda_j)$ we get
+  To see that $\ker\psi_i\cap\ker\psi_j = 0$ if $i\neq j$, we mimic the standard proof for vector spaces: suppose $(\phi - \lambda_i)^m(v) = 0$ for some $v\in M$ and $m$ is minimal, so that $w = (\phi-\lambda_i)^{m-1}(v) \neq 0$ satisfies $(\phi-\lambda_i)(w) = 0$.
+If $(\phi - \lambda_j)^{\mu_j}(v) = 0$ as well, then since $(\phi - \lambda_j)(\phi - \lambda_i) = (\phi - \lambda_i)(\phi - \lambda_j)$ (as $\phi$ commutes with itself and the identity) we get
   \[
-  (\phi - \lambda_j)(w)
-  = (\phi - \lambda_j)(\phi - \lambda_i)^{m-1}(v)
-  = (\phi - \lambda_i)^{m-1}(\phi - \lambda_j)(v) = 0,
+  (\phi - \lambda_j)^{\mu_j}(w)
+  = (\phi - \lambda_j)^{\mu_j}(\phi - \lambda_i)^{m-1}(v)
+  = (\phi - \lambda_i)^{m-1}(\phi - \lambda_j)^{\mu_j}(v) = 0.
   \]
-  which implies
+Thus,
   \(
-  0 = (\phi - \lambda_j)(w) = (\phi - \lambda_i + \lambda_i - \lambda_j)(w) = (\lambda_i - \lambda_j)(w),
+  0 = (\phi - \lambda_j)^{\mu_j}(w) = (\phi - \lambda_i + \lambda_i - \lambda_j)^{\mu_j}(w) = (\lambda_i - \lambda_j)^{\mu_j}(w)
   \)
-  which is a contradiction.
+(using that $(\phi - \lambda_i)(w)=0$).
+Since $\lambda_i-\lambda_j \in k$ is nonzero, 
+$(\lambda_i - \lambda_j)^{\mu_j}(w)= 0 $ forces $w=0$, a contradiction.
 
@@
     \State Take a general element $\phi$ in $[\End_R(M)]_0$,
-           the degree zero part of $\End_R(M)$. \label{item:End0}
+           the degree-zero part of $\End_R(M)$. \label{item:End0}
     \ForAll { eigenvalues $\lambda_i$ of $\phi$ }
@@
   Note that \Cref{alg:graded} does not depend on the characteristic of the field.
-Further, since over a large enough field (i.e., when $|k|$ is larger than the number of summands) the eigenvalues of a general endomorphism $\phi$ are distinct (e.g., by a variant of \Cref{lem:distinct}), the summands $\ker\psi_i$ and $\im\psi$ will be indecomposable; thus \Cref{alg:graded} will produce an indecomposable decomposition of $M$ in one non-recursive iteration.
+Further, since over a large enough field (i.e., when $|k|$ is larger than the number of summands) the eigenvalues of a general endomorphism $\phi$ are are as distinct as  possible (by a variant of \Cref{lem:distinct}), the summands $\ker\psi_i$ and $\im\psi$ will be indecomposable; thus \Cref{alg:graded} will produce an indecomposable decomposition of $M$ in one non-recursive iteration.
 When $|k|$ is less than or approximately the same as the number of summands, the algorithm will be recursive.
+\looseness-1
 \end{rem}
@@
   Therefore the inclusion $ 0 \to N \inc M $ is split by the composition
-  \[ M \xra{\psi} N \xra{\alpha\inv} N, \]
+  \( M \xra{\psi} N \xra{\alpha\inv} N, \)
   and thus $M$ decomposes as claimed.
@@
 
-  By the Artin--Rees lemma \cite[Lemma~5.1]{Eisenbud95}, there is a number $\ell>0$ such that for any $i\gg0$ we get
+  By the Artin--Rees lemma \cite[Lemma~5.1]{Eisenbud95}, there is $\ell>0$ such that for $i\gg0$ we get
   \[ \m^i M\cap N = \m^{i-\ell} ( \m^i M\cap N)\subset \m^{i-\ell} N. \]
-  That is, the terms of $n_0-n_1+n_2-\cdots$ are going to 0 in the $\m$-adic topology on $N$. Hence we can write
+  That is, the terms of $n_0-n_1+\cdots$ go to 0 in the $\m$-adic topology on $N$. Thus we can write
   \[ n_0 = \psi(n_0-n_1+n_2-\cdots), \]
@@
 The following lemma allows us to produce idempotents modulo $\m$.
+\vadjust{\goodbreak}
 
 \begin{lem}\label{lem:jordan}
-  Let $k$ be a finite field of characteristic $p$, and let $A$ be an endomorphism of a $k$-vector space such that all eigenvalues of $A$ are contained in $k$.
-If $\lambda$ is an eigenvalue of $A$, then some power of $A-\lambda$ is idempotent.
-  Furthermore, if $\lambda$ is not the only eigenvalue of $A$, then a  nonzero power of $A-\lambda$ is idempotent.
+  Let $k\subset \bar \F_{p}$, and let $A$ be an endomorphism of a $k$-vector space. 
+Choose:
+\begin{itemize}
+\item $e_0$ such that $p^{e_0}$ is larger than the geometric multiplicity of any eigenvalue of $A$ in $\bar k$.
+\item $e$ is such that all eigenvalues of $A$ in $\bar k$ lie in $\F_{p^e}$.
+\end{itemize}
+If $\lambda$ is an eigenvalue of $A$ contained in $k$, then $(A-\lambda)^{p^{e_0} (p^{e}-1)}$ is idempotent.
+  If $\lambda$ is not the only eigenvalue of $A$ over $\bar k$, this is a nontrivial idempotent.
 \end{lem}
 \begin{proof}
-  Since all eigenvalues of $A$ are contained in $k$, we can without loss of generality put $A$ in Jordan canonical form, with each Jordan block being an $r_i\times r_i$ matrix of the form
+To check that this power is idempotent, we may extend to the algebraic closure $\bar k$, because if $A^2=A$ as an endomorphism of $\bar k$-vector spaces the same is true over $k$.
+  Since all eigenvalues of $A$ are contained in $\bar k$, we can without loss of generality put $A$ in Jordan canonical form, with each Jordan block being an $r_i\times r_i$ matrix of the form
   \[ \begin{pmatrix}
@@
     \end{pmatrix}, \]
-  where each $\lambda_i$ is an eigenvalue of $A$.
+  where each $\lambda_i$ is an eigenvalue of $A$ and $r_i$ is bounded above by the geometric multiplicity of $\lambda_i$.
   In this basis, $A-\lambda$ will be block-diagonal with blocks
@@
   \end{pmatrix}. \]
-  If we choose $n > r_i$ and $n$ a power of $p$, say $p^{e_0}$, then all non-diagonal terms will vanish, so all blocks will have the form
+  If we choose $n > r_i$ and a power of $p$, for example, $p^{e_0}$, then all non-diagonal terms will vanish, so all blocks will have the form
   \[ \begin{pmatrix}
@@
   \end{pmatrix}. \]
-  Finally, since $k\subset \FF_{p^e}$ for some $e$, if we choose $n$ to be divisible also by $p^e-1$, we have that $$\nu_i^{n-1} = (\nu_i^{p^{e}-1})^{n/(p^{e}-1)} =1.$$
+  Finally, since $\nu_i =\lambda-\lambda_i \in \FF_{p^e}$ for each $i$, if we choose $n$ to be divisible also by $p^e-1$, we have that $$\nu_i^{n} = (\nu_i^{p^{e}-1})^{n/(p^{e}-1)} =1.$$
   In particular, taking $n=p^{e_0}(p^e-1)$, the matrix $(A-\lambda)^{p^{e_0}(p^e-1)}$ is a diagonal matrix with diagonal entries 1 or 0, hence idempotent.
@@
 \begin{prop}\label{prop:split-idems}
-  Let $\lambda_1,\dots,\lambda_r$ be the eigenvalues of $\phi\otimes R/\m$ defined over $\FF_{p^e}$ and set $\psi_j = (\phi - \lambda_j \id_M)^{p^{e_i} (p^{e} - 1)}$ with $e_i = \lceil \log_p(\mu_i) \rceil$ and $\psi = \psi_1\circ\cdots\circ\psi_r$.
-Then $M$ has a direct sum decomposition
+  Let $\lambda_1,\dots,\lambda_r$ be the eigenvalues of $\phi\otimes R/\m$ defined over $k$; pick $e$ such that all eigenvalues of $\phi \otimes R/\m$ in $\bar k$ are contained in $\F_{p^e}$
+and $e_0 =  \lceil \log_p(\mu(M))\rceil+1$.
+Set $\psi_j = (\phi - \lambda_j \id_M)^{p^{e_0} (p^{e} - 1)}$ 
+and $\psi = \psi_1\circ\cdots\circ\psi_r$. Then $M$ has a direct sum decomposition
   \( M = \ker\psi_1 \oplus \cdots \oplus \ker\psi_r \oplus \im\psi. \)
 \end{prop}
+
+We note that $e$ can be calculated effectively by computing the splitting field of $\chi(\phi\otimes R/\m)$.
+
 \begin{proof}
-  The proof is analogous to \Cref{prop:split-surj}.
-Specifically, using \Cref{lem:jordan} and \Cref{lem:idemp}, each $\psi_i$ is an idempotent, and in particular, a split surjection.
+  The proof is analogous to the proof of \Cref{prop:split-surj}. By \Cref{lem:jordan} and \Cref{lem:idemp},
+since $p^{e_0}> \mu(M)$ is an upper bound on the geometric multiplicity of any eigenvalue in $\bar k$,
+ each $\psi_i$ is idempotent and in particular a split surjection.
 Further, $\ker\psi_i\cap\ker\psi_j = 0$ by the same argument as in the proof of \Cref{prop:split-surj}.
@@
     \State Let $A$ be the induced endomorphism of the $k$-vector space $M/\m M$.
+    \State Let $e_0 = \lceil\log_p(\mu(M))\rceil+1$. %(or, for simplicity, $\lceil \log_p(\mu(M)) \rceil+1$).
+     \State Let $e$ be the smallest integer such that $\chi(\phi\otimes R/\m)$ factors into linear terms in $\F_{p^e}$.
     \ForAll { eigenvalues $\lambda_i$ of $A$ }
       \State Let $\mu_i$ be the (geometric) multiplicity of $\lambda_i$.
-      \State Let $e_i = \lceil \log_p(\mu_i) \rceil$ (or, for simplicity, $\lceil \log_p(\mu(M)) \rceil$).
-      \State Let $e$ be the smallest integer such that $\lambda_i\in \FF_{p^e}$.
+      %\State Let $e_0 = \max_j(\lceil \log_p(\mu_j) \rceil)+1$ (or, for simplicity, $\lceil \log_p(\mu(M)) \rceil+1$).
       \State Let $\psi_i = (\phi - \lambda \id_M)^{p^{e_0} (p^{e} - 1)}$.
@@
 \begin{rem}\label{rem:local}
-  Similar to \Cref{rem:graded}, over a large enough field \Cref{alg:local} will produce indecomposable summands in one non-recursive iteration, and otherwise it will recursively split the summands until it finds an indecomposable decomposition of $M$.
+  Similarly as in \Cref{rem:graded}, over a large enough field \Cref{alg:local} will produce indecomposable summands in one non-recursive iteration, and otherwise it will recursively split the summands until it finds an indecomposable decomposition of $M$.
 \end{rem}
@@
   At the moment, we do not have such an algorithm outside the graded case.
-  However, we note that \Cref{alg:local} can be used to test indecomposability of $M$ via reduction modulo $p$, as follows.
+  However, we note that \Cref{alg:local} can be used to test indecomposability of $M$ via reduction modulo~$p$, as follows.
   One can choose a finitely generated $\Z$-algebra $A$ and an $A$-algebra $R_A$ such that $R_A\otimes _A k=R$, and likewise an $R_A$-module $M_A$ such that $M_A\otimes _A k = R$ and such that $M_A$ is flat over $A$.
-If $\n$ is a maximal ideal of $A$, one can check that $A/\n \cong \FF_{p^e}$ for some prime $p$ and $e$.
+If $\n$ is a maximal ideal of $A$, one can check that $A/\n \cong \FF_{p^e}$ for some prime $p$ and~$e$.
   The key point is that if $M$ is decomposable, then $M_A$ also can be taken to be decomposable. Thus, the various reductions  $M\otimes A/\n$ will also be decomposable for all $\n$.
@@
 \begin{lem}\label{lem:terminates}
-  Assume that either $(R,\m)$ is a graded ring with $R_0=k$, or $(R,\m)$ is a local ring with $k=R/\m$ a field of positive characteristic, and that $M$ is a finitely generated $R$-module (homogeneous in the graded case).
-  If $M$ has an endomorphism with at least two distinct eigenvalues modulo $\m$, then $M$ is decomposable.
+  Assume that $(R,\m)$ is a graded ring with $R_0=k$, or $(R,\m)$ is a local ring with $k=R/\m\subset \bar \F_p$, and $M$ is a finitely generated $R$-module (homogeneous in the graded case).
+  If $M$ has an endomorphism with at least two distinct eigenvalues, then $M$ is decomposable.
 \end{lem}
 
-The converse observation implies that if $M$ is decomposable, then these algorithms will find the direct sum decomposition of $M$:
+The following lemma implies that if $M$ is decomposable, then these algorithms will find the direct sum decomposition of $M$:
 
@@
 % (Note that in the graded case the eigenvalues of the reduction of an endomorphism modulo $\m$ are the same as the eigenvalues of endomorphism, so there is no harm in looking only at $M/\m M$.)
+\vadjust{\goodbreak}
 
@@
 \begin{rem}\label{rem:extension}
-  Note that \Cref{alg:graded} and \Cref{alg:local} are quite sensitive to the ground field $k$, because they need all the eigenvalues of the endomorphism to be contained in $k$.
+  Note that \Cref{alg:graded} and \Cref{alg:local} are quite sensitive to the ground field $k$, because they need eigenvalues of the endomorphism to be contained in $k$.
 While theoretically the issue can be avoided by working over an algebraically closed ground field $\bar k$, for practical use on a computer algebra system it is better to extend $k$ to some larger finite field.
@@
 Note that by nature of probabilistic algorithms, if \Cref{alg:graded} or \ref{alg:local} fail to produce a nontrivial summand, it does not certify that $M$ is indecomposable.
-However, there are a few sufficient conditions to be indecomposable, which in practice often (but not always) produce such a certification.
-The following sufficiency condition is immediate, but can be quite useful in practice for verifying indecomposability:
+However, there are a few sufficient conditions to be indecomposable, which in practice often  yield such a certification.
+The following sufficiency condition is immediate, but can be quite useful in practice:
 
@@
   \end{itemize}
-   While in many applications these advantages significantly improve the computational complexity, in general the worst case complexity of the algorithms presented here is bounded by the complexity of computing the kernel of the map $f^t\otimes_R M$ where $f^t$ is the transpose of a free presentation $f$ of $M$.
+   While in many applications these advantages significantly improve the computational complexity, in general the worst-case complexity of the algorithms presented here is bounded by the complexity of computing the kernel of the map $f^t\otimes_R M$ where $f^t$ is the transpose of a free presentation $f$ of $M$.
 Compare with \cite{DX22}, which provides a complexity analysis for a decomposition algorithm with the additional assumption that the generators of $R$ as well as generators and relations of $M$ have distinct multidegrees.
@@
   %However, if $M=\Gamma_*(E)$, set of values will be the same: in this case, if $g\otimes R/m$ is not an isomorphism, then $g$ cannot be surjective.
-  Note that in general the multiplicities will never be equal, even if one takes $M=\Gamma_*(E)$: if $f$ is multiplication by $\lambda$, then $f$ will have $\rank E$ many eigenvalues, while $\phi$ will have $\mu(M)$ module-theoretic eigenvalues.
+  Note that in general the multiplicities will never be equal, even if one takes $M=\Gamma_*(E)$: if $f$ is multiplication by $\lambda$, then $f$ will have $\rank E$ many eigenvalues with multiplicity, while $\phi$ will have $\mu(M)$ module-theoretic eigenvalues with multiplicity.
 \end{exa}

### 2025-04-28 paper/main.tex
revised submission

@@
 
-The purpose of this paper is to describe and prove correctness of two practical algorithms for computing indecomposable summands of finitely generated modules over a $k$-algebra $R$: \Cref{alg:graded} in the case that $M$ is a homogeneous module over a (multi)graded ring $R$ with $k$ a field of arbitrary characteristic, and \Cref{alg:local} when $R$ is local and $k$ a field of positive characteristic contained in $\bar \F_p$.
+The purpose of this paper is to describe and prove correctness of two practical algorithms for computing indecomposable summands of finitely generated modules over a $k$-algebra $R$: \Cref{alg:graded} in the case that $M$ is a homogeneous module over a (multi)graded ring $R$ with $k$ a field of arbitrary characteristic, and \Cref{alg:local} when $R$ is local and $k$ a field of positive characteristic contained in $\bar \FF_p$.
 Over a large enough field, our algorithms are often capable of completely decomposing a module in a single non-recursive iteration.
@@
 
-  Thus the characteristic polynomial $\det(\phi - \lambda \id_M) = \prod \det(B_i - \lambda) = \det(A - \lambda \id_{M/\m M})$ is a univariate polynomial in $\lambda$ with coefficients in $k$, thus has solutions in $\bar k$.
+  Thus the characteristic polynomial $\det(\phi - \lambda \id_M) = \prod \det(B_i - \lambda) = \det(A - \lambda \id_{M/\m M})$ is a univariate polynomial in $\lambda$ with coefficients in $k$, which takes solutions in $\bar k$.
 \end{proof}
@@
 
-  Since the $e_i$ are commuting idempotents, $e_r\cdots e_1$ is also an idempotent.
+  Since the $e_i$ are commuting idempotents, $e_r\cdots e_1$ is also idempotent.
 It's clear that $\im(e_r\cdots e_1) = \im (\psi_r\circ \dots\circ \psi_1) = \im \psi$.
@@
 \begin{proof}
-  If $\psi_i$ is injective on $\im\psi_i$, then $\im\psi_i$ is a summand: in this case, the composition $\im\psi_i\hookrightarrow M\to \im\psi_i$ is injective, but a degree-zero endomorphism of a finitely generated $R$-module is an isomorphism (it must be injective, hence surjective, on each finite-dimensional $[M]_d$).
+  If $\psi_i$ is injective on $\im\psi_i$, then $\im\psi_i$ is a summand: in this case, the composition $\im\psi_i\hookrightarrow M\to \im\psi_i$ is injective, but such a degree-zero endomorphism of a finitely generated $R$-module is an isomorphism (it must be injective, hence surjective, on each finite-dimensional $[M]_d$).
 Thus, the inclusion $\im \psi \to M$ is split by the surjection $M\to \im \psi$.
@@
   Note that \Cref{alg:graded} does not depend on the characteristic of the field.
-Further, since over a large enough field (i.e., when $|k|$ is larger than the number of summands) the eigenvalues of a general endomorphism $\phi$ are are as distinct as  possible (by a variant of \Cref{lem:distinct}), the summands $\ker\psi_i$ and $\im\psi$ will be indecomposable; thus \Cref{alg:graded} will produce an indecomposable decomposition of $M$ in one non-recursive iteration.
+Further, since over a large enough field (i.e., when $|k|$ is larger than the number of summands) the eigenvalues of a general endomorphism $\phi$ are as distinct as  possible (by a variant of \Cref{lem:distinct}), the summands $\ker\psi_i$ and $\im\psi$ will be indecomposable; thus \Cref{alg:graded} will produce an indecomposable decomposition of $M$ in one non-recursive iteration.
 When $|k|$ is less than or approximately the same as the number of summands, the algorithm will be recursive.
@@
   Therefore the inclusion $ 0 \to N \inc M $ is split by the composition
-  \( M \xra{\psi} N \xra{\alpha\inv} N, \)
+  \[ M \xra{\psi} N \xra{\alpha\inv} N, \]
   and thus $M$ decomposes as claimed.
@@
 \begin{lem}\label{lem:jordan}
-  Let $k\subset \bar \F_{p}$, and let $A$ be an endomorphism of a $k$-vector space. 
+  Let $k\subset \bar \FF_{p}$ be a subfield and $A$ an endomorphism of a $k$-vector space. 
 Choose:
@@
 \item $e_0$ such that $p^{e_0}$ is larger than the geometric multiplicity of any eigenvalue of $A$ in $\bar k$.
-\item $e$ is such that all eigenvalues of $A$ in $\bar k$ lie in $\F_{p^e}$.
+\item $e$ such that all eigenvalues of $A$ in $\bar k$ lie in $\FF_{p^e}$.
 \end{itemize}
@@
     \end{pmatrix}, \]
-  where each $\lambda_i$ is an eigenvalue of $A$ and $r_i$ is bounded above by the geometric multiplicity of $\lambda_i$.
+  where each $\lambda_i$ is an eigenvalue of $A$ and $r_i$ is bounded above by the geometric multiplicity of $\lambda_i$ in $A$.
   In this basis, $A-\lambda$ will be block-diagonal with blocks
@@
 \begin{prop}\label{prop:split-idems}
-  Let $\lambda_1,\dots,\lambda_r$ be the eigenvalues of $\phi\otimes R/\m$ defined over $k$; pick $e$ such that all eigenvalues of $\phi \otimes R/\m$ in $\bar k$ are contained in $\F_{p^e}$
-and $e_0 =  \lceil \log_p(\mu(M))\rceil+1$.
-Set $\psi_j = (\phi - \lambda_j \id_M)^{p^{e_0} (p^{e} - 1)}$ 
+  Let $\lambda_1,\dots,\lambda_r$ be the eigenvalues of $\phi\otimes R/\m$ defined over $k$.
+Pick $e$ such that all eigenvalues of $\phi \otimes R/\m$ in $\bar k$ are contained in $\FF_{p^e}$ and $e_0 =  \lceil \log_p(\mu(M) + 1)\rceil$ and set $\psi_j = (\phi - \lambda_j \id_M)^{p^{e_0} (p^{e} - 1)}$
 and $\psi = \psi_1\circ\cdots\circ\psi_r$. Then $M$ has a direct sum decomposition
@@
 
-We note that $e$ can be calculated effectively by computing the splitting field of $\chi(\phi\otimes R/\m)$.
+We note that $e$ can be calculated effectively by computing the splitting field of $\chi_{\phi\otimes R/\m}(\lambda)$.
 
 \begin{proof}
-  The proof is analogous to the proof of \Cref{prop:split-surj}. By \Cref{lem:jordan} and \Cref{lem:idemp},
+  This is analogous to the proof of \Cref{prop:split-surj}. By \Cref{lem:jordan} and \Cref{lem:idemp},
 since $p^{e_0}> \mu(M)$ is an upper bound on the geometric multiplicity of any eigenvalue in $\bar k$,
@@
     \State Let $A$ be the induced endomorphism of the $k$-vector space $M/\m M$.
-    \State Let $e_0 = \lceil\log_p(\mu(M))\rceil+1$. %(or, for simplicity, $\lceil \log_p(\mu(M)) \rceil+1$).
-     \State Let $e$ be the smallest integer such that $\chi(\phi\otimes R/\m)$ factors into linear terms in $\F_{p^e}$.
+    \State Let $e_0 = \lceil\log_p(\mu(M) + 1)\rceil$, where $\mu(M) \coloneqq \dim_{R/\m}(M/\m M)$.
+    \State Let $e$ be the least integer such that $\chi_{\phi\otimes R/\m}(\lambda)$ factors into linear terms in $\FF_{p^e}$.
     \ForAll { eigenvalues $\lambda_i$ of $A$ }
       \State Let $\mu_i$ be the (geometric) multiplicity of $\lambda_i$.
-      %\State Let $e_0 = \max_j(\lceil \log_p(\mu_j) \rceil)+1$ (or, for simplicity, $\lceil \log_p(\mu(M)) \rceil+1$).
       \State Let $\psi_i = (\phi - \lambda \id_M)^{p^{e_0} (p^{e} - 1)}$.
-      \State By \Cref{prop:split-idems}, $\psi_i$ is an idempotent, hence producing a splitting of $M$.
+      \State By \Cref{prop:split-idems}, $\psi_i$ is idempotent, hence producing a splitting of $M$.
       \State Append \Call{Summands}{$\ker\psi_i$} to $L$.
@@
 \end{algorithm}
+
+%We will see via \Cref{lem:distinct} that this results in an indecomposable decomposition of $M$.
 
@@
 \begin{lem}\label{lem:terminates}
-  Assume that $(R,\m)$ is a graded ring with $R_0=k$, or $(R,\m)$ is a local ring with $k=R/\m\subset \bar \F_p$, and $M$ is a finitely generated $R$-module (homogeneous in the graded case).
+  Assume that $R$ is a graded ring with $R_0=k$, or $(R,\m)$ is a local ring with $k=R/\m\subset \bar \FF_p$, and $M$ is a finitely generated $R$-module (homogeneous in the graded case).\linebreak
   If $M$ has an endomorphism with at least two distinct eigenvalues, then $M$ is decomposable.
@@
 
-The following lemma implies that if $M$ is decomposable, then these algorithms will find the direct sum decomposition of $M$:
+Conversely, the following lemma implies that if $M$ is decomposable, then \Cref{alg:graded,alg:local} will find a decomposition of $M$ as a direct sum of indecomposable modules.
 
 \begin{lem}\label{lem:distinct}
-  Retain the notation of \Cref{lem:terminates}.
+  Retain the notation of \Cref{lem:terminates} and suppose $M$ is a decomposable.
   \begin{itemize}
-  \item If $R$ is graded and $M$ is decomposable, then a general endomorphism of $M$ of degree zero has at least two distinct eigenvalues (in $\bar k$)
-  \item If $R$ is local and $M$ is decomposable, then a general endomorphism of $M$ of degree zero reduces to an endomorphism of $M/\m M$ with at least two distinct eigenvalues.
+  \item If $R$ is graded, let $\phi$ be a general $k$-linear combination of a $k$-basis for $[\End_R(M)]_0$, then $\phi$ has at least two distinct eigenvalues in $\bar k$.
+  \item If $R$ is local, let $\phi$ be a general $\bar k$-linear combination of minimal generators of $\End_R(M)$, then $\phi\otimes R/\m \in \End_k(M/\m M)$ has at least two distinct eigenvalues in $\bar k$.
   \end{itemize}
@@
 
-\begin{rem}\label{rem:general}
-  By ``general'' we mean a general $k$-linear combination of a $k$-basis for $[\End_R(M)]_0$ (in the graded case) or of minimal generators of $\End_R(M)$ (in the local case) over the algebraic closure of $k$, or equivalently over a large enough algebraic extension of $k$.
-\end{rem}
+%% \begin{rem}\label{rem:general}
+%%   By ``general'' we mean a general $k$-linear combination of a $k$-basis for $[\End_R(M)]_0$ (in the graded case) or of minimal generators of $\End_R(M)$ (in the local case) over the algebraic closure of $k$, or equivalently over a large enough algebraic extension of $k$.
+%% \end{rem}
 
+  We will refer to such endomorphisms of $M$ as \emph{general}.
 In particular, if $M$ is decomposable, then \Cref{alg:graded} and \Cref{alg:local} produce a nontrivial direct sum decomposition of $M$, because a general element of $\End_R(M)$ will have distinct eigenvalues and thus the algorithms will produce a nontrivial decomposition.
@@
 \begin{exa}[Local singularities]
-  Let $R=\F_2[x,y,z]_{(x,y,z)}/(x^2y+xy^2+xyz+z^2)$.
+  Let $R = \FF_2[x,y,z]_{(x,y,z)}/(x^2y+xy^2+xyz+z^2)$.
 In the notation of \cite{Artin77}, this is the $D_4^1$ singularity, which is an exceptional version of the usual rational double point/Du Val singularities appearing in characteristic 2.
